# Causal Inference for Observational Health Data {#sec-causal-inference}

```{r}
#| include: false
library(tidyverse)
library(MatchIt)
library(WeightIt)
library(cobalt)
library(survival)
library(broom)
```

This chapter is about a question health researchers ask constantly but rarely state out loud: does the treatment actually *work*, or do the treated patients simply differ from everyone else? With observational data we cannot randomise, so we have to do the next best thing --- carefully adjust for the ways treated and untreated patients differ. The methods here (propensity scores, matching, inverse probability weighting, target trial emulation) are all variations on that single idea. As you read, keep one practical mantra in mind: a causal estimate is only as trustworthy as the assumption that we have measured and adjusted for every important confounder.

## Introduction

Most clinical research relies on observational data --- electronic health records, claims databases, registries, and cohort studies. Yet the questions we care about most are causal: *Does* this drug reduce mortality? *Would* this surgery improve outcomes compared with conservative management? Answering causal questions with observational data requires careful methodology, because the fundamental problem is that patients who receive a treatment are systematically different from those who do not.

This chapter equips you with the conceptual frameworks and practical tools to move from "Drug X is associated with lower mortality" to a defensible causal claim --- or, equally important, to recognise when the data cannot support one.

::: {.callout-important}
## The stakes are real
Confounding by indication --- where sicker patients receive more aggressive treatment --- has led to published studies concluding that treatments *cause harm* when they actually help, and vice versa. The Women's Health Initiative overturned decades of observational evidence on hormone replacement therapy. Causal inference methods do not guarantee correct answers, but they make your assumptions explicit and testable.
:::

## Correlation vs Causation in Medicine

### Why simple regression is not enough

Consider a hospital database showing that patients who receive mechanical ventilation have higher mortality than those who do not. A naive analyst might conclude ventilation is harmful. But the confounding is obvious: ventilation is given to the sickest patients.

This is **confounding by indication**. A *confounder* is something that influences both who gets the treatment and who has the bad outcome, creating a misleading association between the two. Here the very reason a patient is put on a ventilator --- being critically ill --- is also the reason they are more likely to die. Standard regression can adjust for *measured* confounders, but:

1. You must know which variables to include (and which to exclude).
2. You must model the relationships correctly (linearity, interactions).
3. You cannot adjust for *unmeasured* confounders.

Causal inference methods address problems 1 and 2 more systematically, and provide tools to assess sensitivity to problem 3.

### The potential outcomes framework

The modern causal inference framework, formalised by Rubin (1974) and extended by Hernan and Robins, rests on **potential outcomes** --- the two outcomes a patient *could* have, one under each version of care. These are also called **counterfactuals**, because at least one of them is contrary to what actually happened:

- $Y^{a=1}$: the outcome a patient *would have* if treated
- $Y^{a=0}$: the outcome the *same* patient *would have* if not treated

The **individual causal effect** is the difference between these two, $Y^{a=1} - Y^{a=0}$. The catch is that a patient is either treated or not --- we only ever see one of the two outcomes, never both. (Imagine wanting to know whether a specific patient's statin prevented their heart attack: we cannot rewind and rerun their life without the drug.) This is the **fundamental problem of causal inference**.

We cannot get the effect for one person, but we *can* estimate an average. The **average treatment effect (ATE)** is the effect we would see if *everyone* in the population were treated versus if *everyone* went untreated:

$$
\text{ATE} = E[Y^{a=1}] - E[Y^{a=0}]
$$

The **average treatment effect on the treated (ATT)** is narrower: it is the effect among the patients who *actually received* the treatment. This is often the more clinically relevant number when you want to know whether the patients you are already treating are benefiting.

$$
\text{ATT} = E[Y^{a=1} - Y^{a=0} \mid A=1]
$$

The key assumption that allows us to estimate these from observational data is **exchangeability**, also called *no unmeasured confounding*. In plain terms: once we account for the measured patient characteristics $L$, who ended up treated is essentially a coin toss with respect to their potential outcomes --- treated and untreated patients are then "exchangeable" and fair to compare. Formally, conditional on $L$, treatment assignment is independent of the potential outcomes.

$$
Y^a \perp\!\!\!\perp A \mid L
$$

## Directed Acyclic Graphs (DAGs)

### What is a DAG?

A **directed acyclic graph (DAG)** is simply a diagram of what you believe causes what --- a picture of your clinical assumptions. Each node is a variable (age, treatment, outcome, and so on). Each arrow means "this directly affects that." "Acyclic" just means the arrows never loop back on themselves: nothing can end up causing itself.

Why should a clinician bother? Because a DAG forces you to decide, *before touching the data*, which variables you must adjust for and which you must leave alone. Getting that list wrong is one of the commonest reasons observational studies reach the wrong conclusion.

```{mermaid}
%%| echo: false
%%| eval: true
%%| label: fig-dag-basic
%%| fig-cap: "A simple DAG showing confounding, drawn as a top-down triangle. Age (gold) is a common cause: it points into BOTH statin use and cardiovascular death, so the raw statin-death association is confounded. This fork shape (one cause into two) is what makes it confounding rather than a chain."
graph TD
    Age["Age"]
    StatinUse["Statin use"]
    CVDeath["Cardiovascular death"]
    Age --> StatinUse
    Age --> CVDeath
    StatinUse --> CVDeath
    style Age fill:#fbf3e8,color:#111
    style StatinUse fill:#eef3fb,color:#111
    style CVDeath fill:#eef3fb,color:#111
```

### Key structures in DAGs

There are three fundamental structures you must recognise:

**1. Confounders (common causes)**

A variable that causes both treatment and outcome. You MUST adjust for confounders to remove bias.

```
L --> A --> Y
L ---------> Y
```

If `L` is diabetes severity, `A` is insulin therapy, and `Y` is HbA1c, then diabetes severity confounds the insulin-HbA1c relationship.

**2. Mediators (intermediate variables)**

A variable on the causal pathway from treatment to outcome. You should NOT adjust for mediators if you want the total effect.

```
A --> M --> Y
```

If `A` is exercise, `M` is weight loss, and `Y` is blood pressure, then weight loss mediates the exercise-BP relationship. Adjusting for weight loss would block part of the effect you want to estimate.

**3. Colliders (common effects)**

A **collider** is a variable that is *caused* by both treatment and outcome (or by variables on each path) --- the two arrows "collide" into it. Colliders are the counter-intuitive case: you should NOT adjust for them. Doing so *creates* a fake association that was not there to begin with (**collider bias**).

```
A --> C <-- Y
```

If `A` is obesity, `Y` is genetic predisposition to diabetes, and `C` is being in a diabetes clinic, then conditioning on clinic attendance (a collider) creates a spurious negative association between obesity and genetic risk.

::: {.callout-tip}
## The "d-separation" rule of thumb
A path between treatment and outcome is **blocked** if:

- It passes through a variable you condition on (confounder or mediator), OR
- It passes through a collider you do NOT condition on.

A path is **open** if it is not blocked. Bias creeps in when there are open *non-causal* paths between treatment and outcome --- so-called **backdoor paths**, sneaky side-routes (usually running through a confounder) that make treatment and outcome look related even when the treatment does nothing. The goal of adjustment is to close every backdoor path while leaving the genuine causal path alone.
:::

### Drawing your own DAG

Before any causal analysis, draw a DAG. Steps:

1. List all variables relevant to treatment assignment and the outcome.
2. Use clinical knowledge to draw arrows (not data --- DAGs encode assumptions).
3. Identify all backdoor paths from treatment to outcome.
4. Determine the minimal adjustment set that blocks all backdoor paths without opening new ones.

Tools like [DAGitty](https://dagitty.net) automate step 4 once you have drawn the DAG.

## Propensity Score Methods

### The propensity score

The **propensity score** is each patient's probability of receiving the treatment, based on their observed characteristics. Think of it as: "given everything I know about this patient, how likely was a clinician to put them on this drug?"

$$
e(L) = P(A=1 \mid L)
$$

Rosenbaum and Rubin (1983) showed something remarkably useful: if exchangeability holds once we account for all the covariates $L$, then it also holds once we account for just this single number $e(L)$. So instead of trying to balance dozens of patient characteristics one by one, we only need to balance one summary score. That is why propensity scores are so popular in clinical research --- they turn an unwieldy problem into a manageable one.

### Estimating the propensity score

We typically estimate $e(L)$ using logistic regression (or more flexible methods like gradient boosting):

```{r}
#| label: propensity-score-estimation
#| eval: false

# Simulate clinical data
set.seed(42)
n <- 2000

dat <- tibble(
  age = rnorm(n, 65, 10),
  diabetes = rbinom(n, 1, 0.3),
  egfr = rnorm(n, 60, 15),
  smoking = rbinom(n, 1, 0.25),
  # Treatment assignment depends on covariates (confounding)
  ps_true = plogis(-2 + 0.03 * age + 0.5 * diabetes - 0.02 * egfr + 0.4 * smoking),
  treatment = rbinom(n, 1, ps_true),
  # Outcome depends on treatment AND covariates
  mortality = rbinom(n, 1, plogis(-3 + 0.04 * age + 0.6 * diabetes -
                                    0.03 * egfr + 0.3 * smoking -
                                    0.5 * treatment))
)

# Estimate propensity score via logistic regression
ps_model <- glm(treatment ~ age + diabetes + egfr + smoking,
                data = dat, family = binomial)
dat$ps <- predict(ps_model, type = "response")

# Visualise overlap
ggplot(dat, aes(x = ps, fill = factor(treatment))) +
  geom_density(alpha = 0.5) +
  labs(x = "Propensity Score", fill = "Treatment",
       title = "Propensity Score Distribution by Treatment Group") +
  theme_minimal()
```

**What the code shows.** We first simulate a clinical cohort where treatment is *deliberately* confounded --- older, diabetic, smoking patients with worse kidney function are more likely to be treated, and those same characteristics also raise mortality. This mimics confounding by indication. We then fit a logistic regression of treatment on the covariates; its fitted probabilities are the propensity scores, stored in `dat$ps`. The final density plot shows the propensity-score distribution separately for treated and untreated patients. What you want to see is substantial *overlap* between the two curves: that means for most patients there exist comparable patients in the other group. If the two curves barely overlap (treated patients piled up near 1, controls near 0), the groups are too different to compare fairly --- the warning below.

::: {.callout-warning}
## Positivity assumption
**Positivity** means every type of patient must have at least *some* chance of receiving *either* treatment --- nobody is guaranteed to be treated or guaranteed to be untreated. If a group of patients always gets the drug (propensity score near 1) or never gets it (near 0), there are no comparable patients in the other arm, so the comparison breaks down and estimates become unstable. Always check the propensity score distributions visually.
:::

### Propensity score matching

**Matching** pairs each treated patient with one (or more) untreated patient who had a similar propensity score --- in effect, finding each treated patient a "statistical twin" who looked equally likely to be treated but happened not to be. Comparing these matched pairs mimics what a randomised trial does naturally: it puts like next to like.

```{r}
#| label: matching-example
#| eval: false

# Using MatchIt for nearest-neighbour matching with caliper
m_out <- matchit(treatment ~ age + diabetes + egfr + smoking,
                 data = dat,
                 method = "nearest",
                 distance = "glm",        # logistic regression PS
                 caliper = 0.2,           # 0.2 SD of logit PS
                 ratio = 1)               # 1:1 matching

summary(m_out)

# Check covariate balance using cobalt
love.plot(m_out,
          thresholds = c(m = 0.1),    # SMD threshold of 0.1
          binary = "std",
          title = "Covariate Balance: Before and After Matching")

# Extract matched data and estimate treatment effect
m_data <- match.data(m_out)

# Outcome model in matched sample
outcome_model <- glm(mortality ~ treatment,
                     data = m_data,
                     family = binomial,
                     weights = weights)

tidy(outcome_model, conf.int = TRUE, exponentiate = TRUE)
```

**What the code shows.** `matchit` pairs each treated patient with the most similar untreated patient by propensity score, refusing any match further apart than the caliper (0.2 standard deviations of the logit score). `summary(m_out)` reports how many patients were matched and how well-balanced the groups now are. `love.plot` then draws a "Love plot": one dot per covariate showing the standardised mean difference (SMD) before matching and after. The SMD is just the difference in a covariate's mean between treated and control groups, divided by its standard deviation --- a unit-free measure of how far apart the groups are. The whole point of matching is to pull every "after" dot inside the 0.1 reference line, signalling the matched groups now look alike on the measured confounders. Finally we estimate the treatment effect on the matched sample; with `exponentiate = TRUE` the `treatment` row is an odds ratio for mortality (below 1 means the treatment is protective), with its 95% confidence interval.

**Checking balance** is the most important step after matching. A standardised mean difference (SMD) below 0.1 for all covariates is the conventional threshold for acceptable balance. Balance, here, simply means the treated and control groups have a similar mix of patients --- the situation a randomised trial would create automatically, and what we are trying to recreate by hand.

### Inverse probability of treatment weighting (IPTW)

Matching throws away patients who cannot be paired. **Inverse probability of treatment weighting (IPTW)** avoids that waste. Instead of discarding anyone, it keeps the whole sample but gives each patient a *weight* so that, after weighting, the treated and untreated groups have the same mix of characteristics. The intuition is simple: a treated patient who, by their covariates, looked very *unlikely* to be treated is rare and therefore informative, so we give them more weight; a treated patient who looked almost certain to be treated is over-represented, so we down-weight them. After weighting, the treated and control "pseudo-populations" have the same covariate mix, just as randomisation would have produced. The weight for each patient is:

$$
w_i = \frac{A_i}{e(L_i)} + \frac{1 - A_i}{1 - e(L_i)}
$$

This estimates the ATE. For the ATT, the weights are:

$$
w_i = A_i + \frac{(1 - A_i) \cdot e(L_i)}{1 - e(L_i)}
$$

```{r}
#| label: iptw-example
#| eval: false

# Using WeightIt
w_out <- weightit(treatment ~ age + diabetes + egfr + smoking,
                  data = dat,
                  method = "ps",
                  estimand = "ATE")

# Check balance
bal.tab(w_out, thresholds = c(m = 0.1))

# Stabilised weights (recommended to reduce variance)
summary(w_out)

# Estimate treatment effect with IPTW
library(survey)
d_weighted <- svydesign(ids = ~1, weights = w_out$weights, data = dat)
svyglm(mortality ~ treatment, design = d_weighted, family = quasibinomial) |>
  tidy(conf.int = TRUE, exponentiate = TRUE)
```

**What the code shows.** `weightit` computes a propensity score and turns it into the weights from the formula above (here for the ATE). `bal.tab` is the weighting equivalent of the Love plot: it prints a table of standardised mean differences for each covariate after weighting --- again, you want them comfortably under 0.1. `summary(w_out)` reports the range of the weights so you can spot any extreme values (see the note below). We then feed those weights into a survey-weighted regression (`svydesign` plus `svyglm`); the survey machinery is needed to get correct, slightly wider confidence intervals that account for the weighting. The exponentiated `treatment` coefficient is the estimated odds ratio in the balanced pseudo-population --- our IPTW estimate of the treatment effect on mortality.

::: {.callout-note}
## Stabilised weights
Extreme weights can inflate variance. **Stabilised weights** replace the numerator 1 with $P(A=1)$ for treated and $P(A=0)$ for untreated. The `WeightIt` package does this by default. Always check that weights are not excessively large (e.g., max weight > 20 warrants investigation).
:::

### Doubly robust estimation

**Doubly robust** methods build in a safety net. They fit two models at once --- a propensity score model (who got treated) and an outcome model (who had the event) --- and give the right answer as long as *at least one* of the two is correct. You get two chances to be right instead of one, which is reassuring when you are never quite sure your model is perfect.

The augmented inverse probability weighted (AIPW) estimator is:

$$
\hat{\tau}_{DR} = \frac{1}{n} \sum_{i=1}^n \left[ \frac{A_i Y_i}{e(L_i)} - \frac{A_i - e(L_i)}{e(L_i)} \hat{\mu}_1(L_i) \right] - \frac{1}{n} \sum_{i=1}^n \left[ \frac{(1-A_i) Y_i}{1-e(L_i)} + \frac{A_i - e(L_i)}{1-e(L_i)} \hat{\mu}_0(L_i) \right]
$$

where $\hat{\mu}_a(L)$ is the predicted outcome under treatment $a$.

Do not be put off by the formula. The practical message is the one above: with ordinary propensity-score methods you have to get the treatment model right, and with a plain outcome regression you have to get the outcome model right, whereas doubly robust estimation stays valid if *either* is correct. Since we are rarely certain which model is the "true" one, this makes it a sensible default for a cautious health researcher.

In practice, packages like `AIPW` in R or `DoWhy` in Python handle this computation.

## Target Trial Emulation

### The framework

**Target trial emulation**, developed by Hernan and Robins, is a disciplined way to use observational data. You first imagine the ideal randomised trial you *wish* you could run --- the "target trial" --- writing down its protocol in full. You then build your analysis from the existing data to copy that protocol as closely as possible. The clinical payoff is that many classic observational blunders simply cannot happen if you specify the trial properly first.

```{mermaid}
%%| echo: false
%%| eval: true
%%| label: fig-target-trial
%%| fig-cap: "Target trial emulation: specify the ideal trial you wish you could run, then mirror each component using the observational data. Based on the target trial emulation framework of @hernan2024causal."
graph LR
    A["Imagine the ideal<br/>randomised trial"] --> B["Write its protocol<br/>(eligibility, strategies,<br/>time zero, outcome)"]
    B --> C["Mirror each component<br/>in the observational data"]
    C --> D["Adjust for confounding<br/>(e.g. IPTW)"]
    D --> E["Estimate the<br/>causal effect"]
```

The key components to specify are:

| Protocol component | Target trial | Observational emulation |
|---|---|---|
| Eligibility criteria | Defined inclusion/exclusion | Same criteria applied to database |
| Treatment strategies | e.g., start drug A vs drug B | Same, defined at "time zero" |
| Treatment assignment | Random | Adjusted via PS methods or g-estimation |
| Start of follow-up | Randomisation date | Date eligibility + treatment criteria met |
| Outcome | e.g., 5-year mortality | Same |
| Causal contrast | Intention-to-treat or per-protocol | Same (with appropriate methods) |
| Analysis plan | Intention-to-treat analysis | Clone-censor-weight or similar |

### Why this reduces bias

Target trial emulation directly addresses several common biases:

**Immortal time bias**: In a naive observational study comparing drug users to non-users, the time between cohort entry and first prescription is "immortal" --- the patient must survive to receive the drug. This artificially inflates survival in the treated group. Target trial emulation avoids this by aligning "time zero" (start of follow-up) with treatment assignment.

**Selection bias from prevalent user designs**: If you study patients who are *already* established on a drug ("prevalent users"), you have quietly excluded everyone who started it and did badly early on --- they dropped out before your study began. Survivors look healthier than they really are. Target trial emulation instead uses a **new-user (incident user) design**: it counts everyone from the moment they *start* the drug, just as a trial enrols patients at randomisation.

**Vague eligibility and time zero**: When eligibility criteria and the start of follow-up are not clearly defined, analyses can be plagued by look-ahead bias. The target trial framework forces explicit specification.

::: {.callout-tip}
## The TARGET reporting guideline (2025)
The TARGET (TrAnsparent ReportinG of observational studies Emulating a Target trial) guideline provides a structured checklist for reporting target trial emulation studies. It was published in the BMJ in 2025 and should be followed for any observational study making causal claims.
:::

### Example: Statin initiation and cardiovascular events

Suppose we want to estimate the effect of initiating statins within 6 months of a type 2 diabetes diagnosis on 5-year cardiovascular events.

**Target trial specification:**

- **Eligibility**: Adults aged 40--75 with new T2DM diagnosis, no prior CVD, no prior statin use
- **Treatment strategies**: (1) Initiate statin within 6 months, (2) Do not initiate statin within 6 months
- **Time zero**: Date of T2DM diagnosis
- **Outcome**: First major adverse cardiovascular event (MACE) within 5 years
- **Analysis**: Intention-to-treat using IPTW for confounding, clone-censor-weight for treatment strategy adherence

## Regression Discontinuity Design

**Regression discontinuity (RD)** takes advantage of the arbitrary cut-offs that pervade clinical guidelines. When a treatment decision hinges on whether some measurement crosses a threshold, patients who fall *just* above and *just* below that line are, for all practical purposes, the same kind of patient --- yet one group gets treated and the other does not. That near-random split around the cut-off is what we exploit.

**Clinical example**: Many guidelines recommend antihypertensive therapy when systolic blood pressure exceeds 140 mmHg. Patients with SBP of 141 are very similar to patients with SBP of 139, but the former are much more likely to receive treatment. Comparing outcomes near the cutoff provides a quasi-experimental estimate.

Key requirements for a valid RD design:

1. **Treatment probability changes sharply at the cutoff** (first stage)
2. **Other covariates do not jump at the cutoff** (continuity assumption)
3. **Patients cannot precisely manipulate their score** (no bunching)

RD designs are less common in clinical research than propensity score methods, but when applicable, they provide strong evidence because the assumptions are more testable.

## Implementation in Python

Python's causal inference ecosystem is maturing. Here we demonstrate propensity score matching and IPTW using `dowhy` and `causalinference`.

```{python}
#| label: python-causal
#| eval: false

import numpy as np
import pandas as pd
from sklearn.linear_model import LogisticRegression
from scipy import stats
import matplotlib.pyplot as plt

# Simulate clinical data
np.random.seed(42)
n = 2000

age = np.random.normal(65, 10, n)
diabetes = np.random.binomial(1, 0.3, n)
egfr = np.random.normal(60, 15, n)
smoking = np.random.binomial(1, 0.25, n)

# Confounded treatment assignment
ps_true = 1 / (1 + np.exp(-(-2 + 0.03*age + 0.5*diabetes - 0.02*egfr + 0.4*smoking)))
treatment = np.random.binomial(1, ps_true)

# Outcome
mortality_prob = 1 / (1 + np.exp(-(-3 + 0.04*age + 0.6*diabetes -
                                     0.03*egfr + 0.3*smoking -
                                     0.5*treatment)))
mortality = np.random.binomial(1, mortality_prob)

df = pd.DataFrame({
    'age': age, 'diabetes': diabetes, 'egfr': egfr,
    'smoking': smoking, 'treatment': treatment, 'mortality': mortality
})

# --- Propensity Score Estimation ---
covariates = ['age', 'diabetes', 'egfr', 'smoking']
ps_model = LogisticRegression(max_iter=1000)
ps_model.fit(df[covariates], df['treatment'])
df['ps'] = ps_model.predict_proba(df[covariates])[:, 1]

# --- IPTW ---
# Compute ATE weights
df['iptw'] = np.where(
    df['treatment'] == 1,
    1 / df['ps'],
    1 / (1 - df['ps'])
)

# Stabilised weights
p_treat = df['treatment'].mean()
df['sw'] = np.where(
    df['treatment'] == 1,
    p_treat / df['ps'],
    (1 - p_treat) / (1 - df['ps'])
)

# Check balance: weighted standardised mean differences
def weighted_smd(data, var, treatment_col, weight_col):
    treated = data[data[treatment_col] == 1]
    control = data[data[treatment_col] == 0]
    mean_t = np.average(treated[var], weights=treated[weight_col])
    mean_c = np.average(control[var], weights=control[weight_col])
    sd_pooled = np.sqrt((treated[var].var() + control[var].var()) / 2)
    return (mean_t - mean_c) / sd_pooled

print("Covariate balance (weighted SMD):")
for var in covariates:
    smd = weighted_smd(df, var, 'treatment', 'sw')
    print(f"  {var}: {smd:.4f}")

# --- Weighted outcome estimation ---
from statsmodels.api import WLS, add_constant

X = add_constant(df['treatment'])
wls_model = WLS(df['mortality'], X, weights=df['sw']).fit()
print("\nIPTW Treatment Effect Estimate:")
print(f"  Coefficient: {wls_model.params[1]:.4f}")
print(f"  95% CI: ({wls_model.conf_int()[1][0]:.4f}, {wls_model.conf_int()[1][1]:.4f})")
```

**What the code shows.** This is the Python counterpart to the R workflow above, written from scratch so you can see every step rather than relying on a package. We simulate the same confounded cohort, fit a logistic regression for the propensity score, then build the inverse-probability weights (both raw and stabilised). The `weighted_smd` helper prints the standardised mean difference for each covariate *after* weighting --- the printed numbers should all be close to zero (well under 0.1) if weighting has balanced the groups. The final weighted least-squares model regresses mortality on treatment using the stabilised weights; its `treatment` coefficient is the estimated effect on the probability of death (a risk difference, since the outcome is on the probability scale), reported with a 95% confidence interval.

### Using DoWhy for causal analysis

```{python}
#| label: dowhy-example
#| eval: false

# DoWhy provides a structured causal inference workflow
# Install: pip install dowhy

import dowhy
from dowhy import CausalModel

# Step 1: Model - encode causal assumptions
model = CausalModel(
    data=df,
    treatment='treatment',
    outcome='mortality',
    common_causes=['age', 'diabetes', 'egfr', 'smoking']
)

# Step 2: Identify - find valid adjustment strategy
identified_estimand = model.identify_effect()
print(identified_estimand)

# Step 3: Estimate - using propensity score weighting
estimate = model.estimate_effect(
    identified_estimand,
    method_name="backdoor.propensity_score_weighting",
    method_params={"weighting_scheme": "ips_stabilized_weight"}
)
print(f"Causal Estimate: {estimate.value:.4f}")

# Step 4: Refute - sensitivity analysis
refutation = model.refute_estimate(
    identified_estimand, estimate,
    method_name="random_common_cause"
)
print(refutation)
```

**What the code shows.** `DoWhy` makes the four logical steps of a causal analysis explicit rather than burying them in modelling code. **Model** records your assumptions (treatment, outcome, and the confounders to adjust for). **Identify** uses those assumptions to work out, on paper, which quantity can actually be estimated --- printing the adjustment formula. **Estimate** then computes it; here using stabilised propensity-score weighting, so the printed `estimate.value` is comparable to the IPTW number from the manual code above. **Refute** is the part most analysts skip: it stress-tests the estimate by, for example, adding a randomly generated extra "common cause." A trustworthy estimate should barely move when a *random* (and therefore non-confounding) variable is added; a large change is a red flag that the analysis is fragile.

## Sensitivity Analysis for Unmeasured Confounding

No observational study can ever prove there is no hidden confounder lurking in the background. The honest response is not to ignore that worry but to quantify it. **Sensitivity analysis** asks: how strong would an unmeasured confounder have to be before it could wipe out the effect we observed? If the answer is "implausibly strong," your finding is robust; if "a mild, common confounder would do it," be cautious.

The **E-value** (VanderWeele and Ding, 2017) puts a single number on this. It is the minimum strength of association (on the risk ratio scale) that an unmeasured confounder would need to have with *both* the treatment and the outcome to fully explain away the result you observed.

$$
\text{E-value} = RR + \sqrt{RR \times (RR - 1)}
$$

where $RR$ is the observed risk ratio. The E-value for the confidence interval bound closest to the null is also reported.

```{r}
#| label: evalue
#| eval: false

# E-value calculation
library(EValue)

# Suppose our IPTW analysis found RR = 0.65 (95% CI: 0.50, 0.85)
# Convert protective RR to above-null: 1/0.65 = 1.538
evalues.RR(est = 1/0.65, lo = 1/0.85, hi = 1/0.50)
```

**What the code shows.** `evalues.RR` takes our observed risk ratio (and its confidence limits) and returns the E-value. Because a protective effect (RR below 1) and a harmful one of the same magnitude are equivalent for this calculation, we flip the protective RR of 0.65 to `1/0.65 = 1.54` first. The function prints two numbers: the E-value for the point estimate and the E-value for the confidence-interval bound nearest the null. Read them like this: "an unmeasured confounder would need to be associated with both treatment and mortality by a risk ratio of *at least this value*, over and above the confounders we already adjusted for, to fully explain away our result." A large E-value (say 3 or more) means only a very strong hidden confounder could overturn the finding, so the result is fairly robust; an E-value close to 1 means a weak, plausible confounder could erase it.

## Practical Guidance: Choosing Your Method

| Scenario | Recommended approach |
|---|---|
| Binary treatment, many controls available | Propensity score matching |
| Binary treatment, need full sample | IPTW |
| Concern about model misspecification | Doubly robust estimation |
| Time-varying treatment | Marginal structural models (IPTW over time) |
| Natural threshold for treatment | Regression discontinuity |
| Emulating a specific trial design | Target trial emulation framework |

## Exercises

::: {.callout-tip title="Exercise 1: DAG construction (Conceptual)"}

You are studying the relationship between **ACE inhibitor use** and **acute kidney injury (AKI)** in hospitalised patients.

a) List at least five variables that might be relevant to this relationship.
b) Draw a DAG (on paper or using DAGitty) encoding your causal assumptions.
c) Identify the minimal sufficient adjustment set for estimating the effect of ACE inhibitors on AKI.
d) Identify at least one potential collider. What would happen if you adjusted for it?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch17_ex1.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch17_ex1.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 2: Propensity score matching in R"}

Using the simulated dataset below, perform propensity score matching and estimate the treatment effect.

```{r}
#| label: exercise-2-data
#| eval: false

library(tidyverse)  # tibble()

set.seed(123)
n <- 1500
exercise_dat <- tibble(
  age = rnorm(n, 70, 8),
  creatinine = rnorm(n, 1.2, 0.4),
  heart_failure = rbinom(n, 1, 0.35),
  prior_mi = rbinom(n, 1, 0.20),
  # Confounded treatment (beta-blocker use)
  treatment = rbinom(n, 1, plogis(-1 + 0.02*age + 0.3*heart_failure +
                                     0.5*prior_mi - 0.8*creatinine)),
  # Outcome: 1-year mortality
  death_1yr = rbinom(n, 1, plogis(-2 + 0.05*age + 0.4*heart_failure +
                                     0.6*prior_mi + 1.0*creatinine -
                                     0.7*treatment))
)
```

a) Estimate propensity scores using logistic regression.
b) Perform 1:1 nearest-neighbour matching with a caliper of 0.2 SD.
c) Create a Love plot to assess balance before and after matching.
d) Estimate the ATT for beta-blocker use on 1-year mortality.
e) Calculate the E-value for your estimate. Interpret it.

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch17_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch17_ex2.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 3: IPTW in Python"}

Using the same data-generating process as Exercise 2 (replicate it in Python):

```{python}
#| label: exercise-3-data
#| eval: false

import numpy as np
import pandas as pd

np.random.seed(123)
n = 1500
age = np.random.normal(70, 8, n)
creatinine = np.random.normal(1.2, 0.4, n)
heart_failure = np.random.binomial(1, 0.35, n)
prior_mi = np.random.binomial(1, 0.20, n)

ps_true = 1 / (1 + np.exp(-(-1 + 0.02*age + 0.3*heart_failure +
                               0.5*prior_mi - 0.8*creatinine)))
treatment = np.random.binomial(1, ps_true)

mort_prob = 1 / (1 + np.exp(-(-2 + 0.05*age + 0.4*heart_failure +
                                 0.6*prior_mi + 1.0*creatinine -
                                 0.7*treatment)))
death_1yr = np.random.binomial(1, mort_prob)

df = pd.DataFrame({
    'age': age, 'creatinine': creatinine, 'heart_failure': heart_failure,
    'prior_mi': prior_mi, 'treatment': treatment, 'death_1yr': death_1yr
})
```

a) Fit a propensity score model and compute stabilised IPTW weights.
b) Assess covariate balance using weighted standardised mean differences.
c) Estimate the ATE using weighted regression.
d) Perform a sensitivity analysis: add a simulated unmeasured confounder and show how the estimate changes.

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch17_ex3.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch17_ex3.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 4: Target trial emulation (Conceptual)"}

You have access to a large electronic health records database and want to estimate the effect of **early metformin initiation** (within 3 months of type 2 diabetes diagnosis) vs **delayed initiation** (3--12 months after diagnosis) on 5-year cardiovascular events.

a) Write a complete target trial protocol table (eligibility, treatment strategies, assignment, start of follow-up, outcome, causal contrast, analysis plan).
b) Identify potential sources of immortal time bias in a naive analysis.
c) Describe how the new-user, active comparator design addresses confounding by indication.
d) What unmeasured confounders might remain? How would you conduct a sensitivity analysis?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch17_ex4.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch17_ex4.py
```
:::
:::

:::

## Summary

| Concept | Key point |
|---|---|
| Confounding by indication | Sicker patients get more treatment --- naive comparisons are biased |
| DAGs | Encode causal assumptions visually; identify what to adjust for |
| Propensity score | $P(\text{treatment} \mid \text{covariates})$; reduces dimensionality |
| Matching | Creates balanced pseudo-randomised groups; check balance |
| IPTW | Reweights full sample; preserves sample size |
| Doubly robust | Combines PS + outcome model; consistent if either is correct |
| Target trial emulation | Design observational analysis to mimic an ideal RCT |
| E-value | Quantifies robustness to unmeasured confounding |

## References and Further Reading

- For the foundations of causal inference, see @hernan2024causal (the definitive modern textbook, freely available online) and @rosenbaum2002observational.
- For propensity scores, see @rosenbaum1983propensity (the original paper) and @ho2011matchit for the `MatchIt` implementation.
- For drawing and interrogating DAGs, see @textor2016dagitty.
- For sensitivity analysis to unmeasured confounding, see @vanderweele2017evalue.
- For target trial emulation, see @cashin2025target for the reporting checklist and @lodi2019effect for what happens when observational analyses are made to mirror trials properly.
- For software, see @greifer2024weightit in R and @sharma2020dowhy in Python.

::: {.sectionrefs}
:::
