# Causal Inference for Observational Health Data {#sec-causal-inference}

```{r}
#| include: false
library(tidyverse)
library(MatchIt)
library(WeightIt)
library(cobalt)
library(survival)
library(broom)
```

This chapter is about a question health researchers ask constantly but rarely state out loud: does the treatment actually *work*, or do the treated patients simply differ from everyone else? With observational data we cannot randomise, so we have to do the next best thing --- carefully adjust for the ways treated and untreated patients differ. The methods here (propensity scores, matching, inverse probability weighting, target trial emulation) are all variations on that single idea. As you read, keep one practical mantra in mind: a causal estimate is only as trustworthy as the assumption that we have measured and adjusted for every important confounder. The code chunks in this chapter are set to `eval: false`, so you will see the commands but not live output --- the surrounding text describes what each command produces and how to read it.

## Introduction

Most clinical research relies on observational data --- electronic health records, claims databases, registries, and cohort studies. Yet the questions we care about most are causal: *Does* this drug reduce mortality? *Would* this surgery improve outcomes compared with conservative management? Answering causal questions with observational data requires careful methodology, because the fundamental problem is that patients who receive a treatment are systematically different from those who do not.

This chapter equips you with the conceptual frameworks and practical tools to move from "Drug X is associated with lower mortality" to a defensible causal claim --- or, equally important, to recognise when the data cannot support one.

::: {.callout-important}
## The stakes are real
Confounding by indication --- where sicker patients receive more aggressive treatment --- has led to published studies concluding that treatments *cause harm* when they actually help, and vice versa. The Women's Health Initiative overturned decades of observational evidence on hormone replacement therapy. Causal inference methods do not guarantee correct answers, but they make your assumptions explicit and testable.
:::

## Correlation vs Causation in Medicine

### Why simple regression is not enough

Consider a hospital database showing that patients who receive mechanical ventilation have higher mortality than those who do not. A naive analyst might conclude ventilation is harmful. But the confounding is obvious: ventilation is given to the sickest patients.

This is **confounding by indication** --- the indication for treatment is itself a risk factor for the outcome. Standard regression can adjust for *measured* confounders, but:

1. You must know which variables to include (and which to exclude).
2. You must model the relationships correctly (linearity, interactions).
3. You cannot adjust for *unmeasured* confounders.

Causal inference methods address problems 1 and 2 more systematically, and provide tools to assess sensitivity to problem 3.

### The potential outcomes framework

The modern causal inference framework, formalised by Rubin (1974) and extended by Hernan and Robins, rests on **potential outcomes**:

- $Y^{a=1}$: the outcome a patient *would have* under treatment
- $Y^{a=0}$: the outcome a patient *would have* under control

The **individual causal effect** is $Y^{a=1} - Y^{a=0}$, but we can never observe both for the same person. This is the **fundamental problem of causal inference**.

We can, however, estimate the **average treatment effect (ATE)**:

$$
\text{ATE} = E[Y^{a=1}] - E[Y^{a=0}]
$$

or the **average treatment effect on the treated (ATT)**:

$$
\text{ATT} = E[Y^{a=1} - Y^{a=0} \mid A=1]
$$

The key assumption that allows us to estimate these from observational data is **exchangeability** (no unmeasured confounding): conditional on measured covariates $L$, treatment assignment is independent of potential outcomes.

$$
Y^a \perp\!\!\!\perp A \mid L
$$

## Directed Acyclic Graphs (DAGs)

### What is a DAG?

A **directed acyclic graph** is a visual representation of your causal assumptions. Each node represents a variable. Each arrow represents a direct causal effect. "Acyclic" means no variable can cause itself through a chain of arrows.

```{mermaid}
%%| eval: true
%%| label: fig-dag-basic
%%| fig-cap: "A simple DAG showing confounding. Age affects both statin use and cardiovascular death."
graph LR
    Age --> StatinUse
    Age --> CVDeath
    StatinUse --> CVDeath
```

### Key structures in DAGs

There are three fundamental structures you must recognise:

**1. Confounders (common causes)**

A variable that causes both treatment and outcome. You MUST adjust for confounders to remove bias.

```
L --> A --> Y
L ---------> Y
```

If `L` is diabetes severity, `A` is insulin therapy, and `Y` is HbA1c, then diabetes severity confounds the insulin-HbA1c relationship.

**2. Mediators (intermediate variables)**

A variable on the causal pathway from treatment to outcome. You should NOT adjust for mediators if you want the total effect.

```
A --> M --> Y
```

If `A` is exercise, `M` is weight loss, and `Y` is blood pressure, then weight loss mediates the exercise-BP relationship. Adjusting for weight loss would block part of the effect you want to estimate.

**3. Colliders (common effects)**

A variable caused by both treatment and outcome (or by variables on each path). You should NOT adjust for colliders --- doing so *creates* a spurious association (collider bias).

```
A --> C <-- Y
```

If `A` is obesity, `Y` is genetic predisposition to diabetes, and `C` is being in a diabetes clinic, then conditioning on clinic attendance (a collider) creates a spurious negative association between obesity and genetic risk.

::: {.callout-tip}
## The "d-separation" rule of thumb
A path between treatment and outcome is **blocked** if:

- It passes through a variable you condition on (confounder or mediator), OR
- It passes through a collider you do NOT condition on.

A path is **open** if it is not blocked. Bias exists when there are open non-causal paths (backdoor paths) between treatment and outcome.
:::

### Drawing your own DAG

Before any causal analysis, draw a DAG. Steps:

1. List all variables relevant to treatment assignment and the outcome.
2. Use clinical knowledge to draw arrows (not data --- DAGs encode assumptions).
3. Identify all backdoor paths from treatment to outcome.
4. Determine the minimal adjustment set that blocks all backdoor paths without opening new ones.

Tools like [DAGitty](https://dagitty.net) automate step 4 once you have drawn the DAG.

## Propensity Score Methods

### The propensity score

The **propensity score** is the probability of receiving treatment given observed covariates:

$$
e(L) = P(A=1 \mid L)
$$

Rosenbaum and Rubin (1983) showed that, if exchangeability holds conditional on $L$, it also holds conditional on the scalar $e(L)$. This reduces a high-dimensional adjustment problem to a single dimension.

### Estimating the propensity score

We typically estimate $e(L)$ using logistic regression (or more flexible methods like gradient boosting):

```{r}
#| label: propensity-score-estimation
#| eval: false

# Simulate clinical data
set.seed(42)
n <- 2000

dat <- tibble(
  age = rnorm(n, 65, 10),
  diabetes = rbinom(n, 1, 0.3),
  egfr = rnorm(n, 60, 15),
  smoking = rbinom(n, 1, 0.25),
  # Treatment assignment depends on covariates (confounding)
  ps_true = plogis(-2 + 0.03 * age + 0.5 * diabetes - 0.02 * egfr + 0.4 * smoking),
  treatment = rbinom(n, 1, ps_true),
  # Outcome depends on treatment AND covariates
  mortality = rbinom(n, 1, plogis(-3 + 0.04 * age + 0.6 * diabetes -
                                    0.03 * egfr + 0.3 * smoking -
                                    0.5 * treatment))
)

# Estimate propensity score via logistic regression
ps_model <- glm(treatment ~ age + diabetes + egfr + smoking,
                data = dat, family = binomial)
dat$ps <- predict(ps_model, type = "response")

# Visualise overlap
ggplot(dat, aes(x = ps, fill = factor(treatment))) +
  geom_density(alpha = 0.5) +
  labs(x = "Propensity Score", fill = "Treatment",
       title = "Propensity Score Distribution by Treatment Group") +
  theme_minimal()
```

**What the code shows.** We first simulate a clinical cohort where treatment is *deliberately* confounded --- older, diabetic, smoking patients with worse kidney function are more likely to be treated, and those same characteristics also raise mortality. This mimics confounding by indication. We then fit a logistic regression of treatment on the covariates; its fitted probabilities are the propensity scores, stored in `dat$ps`. The final density plot shows the propensity-score distribution separately for treated and untreated patients. What you want to see is substantial *overlap* between the two curves: that means for most patients there exist comparable patients in the other group. If the two curves barely overlap (treated patients piled up near 1, controls near 0), the groups are too different to compare fairly --- the warning below.

::: {.callout-warning}
## Positivity assumption
Every patient must have a non-zero probability of receiving *either* treatment. If some patients have propensity scores near 0 or 1, the overlap assumption is violated and estimates become unstable. Always check the propensity score distributions visually.
:::

### Propensity score matching

**Matching** creates a pseudo-randomised subset by pairing treated and control patients with similar propensity scores.

```{r}
#| label: matching-example
#| eval: false

# Using MatchIt for nearest-neighbour matching with caliper
m_out <- matchit(treatment ~ age + diabetes + egfr + smoking,
                 data = dat,
                 method = "nearest",
                 distance = "glm",        # logistic regression PS
                 caliper = 0.2,           # 0.2 SD of logit PS
                 ratio = 1)               # 1:1 matching

summary(m_out)

# Check covariate balance using cobalt
love.plot(m_out,
          thresholds = c(m = 0.1),    # SMD threshold of 0.1
          binary = "std",
          title = "Covariate Balance: Before and After Matching")

# Extract matched data and estimate treatment effect
m_data <- match.data(m_out)

# Outcome model in matched sample
outcome_model <- glm(mortality ~ treatment,
                     data = m_data,
                     family = binomial,
                     weights = weights)

tidy(outcome_model, conf.int = TRUE, exponentiate = TRUE)
```

**What the code shows.** `matchit` pairs each treated patient with the most similar untreated patient by propensity score, refusing any match further apart than the caliper (0.2 standard deviations of the logit score). `summary(m_out)` reports how many patients were matched and how well-balanced the groups now are. `love.plot` then draws a "Love plot": one dot per covariate showing the standardised mean difference (SMD) before matching and after. The SMD is just the difference in a covariate's mean between treated and control groups, divided by its standard deviation --- a unit-free measure of how far apart the groups are. The whole point of matching is to pull every "after" dot inside the 0.1 reference line, signalling the matched groups now look alike on the measured confounders. Finally we estimate the treatment effect on the matched sample; with `exponentiate = TRUE` the `treatment` row is an odds ratio for mortality (below 1 means the treatment is protective), with its 95% confidence interval.

**Checking balance** is the most important step after matching. A standardised mean difference (SMD) below 0.1 for all covariates is the conventional threshold for acceptable balance. Balance, here, simply means the treated and control groups have a similar mix of patients --- the situation a randomised trial would create automatically, and what we are trying to recreate by hand.

### Inverse probability of treatment weighting (IPTW)

Instead of discarding unmatched patients, **IPTW** reweights the entire sample so that the distribution of covariates is balanced between groups. The intuition is simple: a treated patient who, by their covariates, looked very *unlikely* to be treated is rare and therefore informative, so we give them more weight; a treated patient who looked almost certain to be treated is over-represented, so we down-weight them. After weighting, the treated and control "pseudo-populations" have the same covariate mix, just as randomisation would have produced. The weight for each patient is:

$$
w_i = \frac{A_i}{e(L_i)} + \frac{1 - A_i}{1 - e(L_i)}
$$

This estimates the ATE. For the ATT, the weights are:

$$
w_i = A_i + \frac{(1 - A_i) \cdot e(L_i)}{1 - e(L_i)}
$$

```{r}
#| label: iptw-example
#| eval: false

# Using WeightIt
w_out <- weightit(treatment ~ age + diabetes + egfr + smoking,
                  data = dat,
                  method = "ps",
                  estimand = "ATE")

# Check balance
bal.tab(w_out, thresholds = c(m = 0.1))

# Stabilised weights (recommended to reduce variance)
summary(w_out)

# Estimate treatment effect with IPTW
library(survey)
d_weighted <- svydesign(ids = ~1, weights = w_out$weights, data = dat)
svyglm(mortality ~ treatment, design = d_weighted, family = quasibinomial) |>
  tidy(conf.int = TRUE, exponentiate = TRUE)
```

**What the code shows.** `weightit` computes a propensity score and turns it into the weights from the formula above (here for the ATE). `bal.tab` is the weighting equivalent of the Love plot: it prints a table of standardised mean differences for each covariate after weighting --- again, you want them comfortably under 0.1. `summary(w_out)` reports the range of the weights so you can spot any extreme values (see the note below). We then feed those weights into a survey-weighted regression (`svydesign` plus `svyglm`); the survey machinery is needed to get correct, slightly wider confidence intervals that account for the weighting. The exponentiated `treatment` coefficient is the estimated odds ratio in the balanced pseudo-population --- our IPTW estimate of the treatment effect on mortality.

::: {.callout-note}
## Stabilised weights
Extreme weights can inflate variance. **Stabilised weights** replace the numerator 1 with $P(A=1)$ for treated and $P(A=0)$ for untreated. The `WeightIt` package does this by default. Always check that weights are not excessively large (e.g., max weight > 20 warrants investigation).
:::

### Doubly robust estimation

**Doubly robust** methods combine a propensity score model and an outcome model. The estimate is consistent if *either* model is correctly specified (but not necessarily both). This provides an extra layer of protection against model misspecification.

The augmented inverse probability weighted (AIPW) estimator is:

$$
\hat{\tau}_{DR} = \frac{1}{n} \sum_{i=1}^n \left[ \frac{A_i Y_i}{e(L_i)} - \frac{A_i - e(L_i)}{e(L_i)} \hat{\mu}_1(L_i) \right] - \frac{1}{n} \sum_{i=1}^n \left[ \frac{(1-A_i) Y_i}{1-e(L_i)} + \frac{A_i - e(L_i)}{1-e(L_i)} \hat{\mu}_0(L_i) \right]
$$

where $\hat{\mu}_a(L)$ is the predicted outcome under treatment $a$.

Do not be put off by the formula --- the practical message is reassuring. With ordinary propensity-score methods you have to get the treatment model right; with an outcome regression you have to get the outcome model right. Doubly robust estimation gives you *two chances*: it stays valid if you correctly specify **either** the propensity model **or** the outcome model. Since we are rarely certain which model is "correct," this is a sensible default for a cautious health researcher.

In practice, packages like `AIPW` in R or `DoWhy` in Python handle this computation.

## Target Trial Emulation

### The framework

**Target trial emulation**, developed by Hernan and Robins, asks: "What randomised trial would we *like* to conduct?" and then designs the observational analysis to mimic that trial as closely as possible.

The key components to specify are:

| Protocol component | Target trial | Observational emulation |
|---|---|---|
| Eligibility criteria | Defined inclusion/exclusion | Same criteria applied to database |
| Treatment strategies | e.g., start drug A vs drug B | Same, defined at "time zero" |
| Treatment assignment | Random | Adjusted via PS methods or g-estimation |
| Start of follow-up | Randomisation date | Date eligibility + treatment criteria met |
| Outcome | e.g., 5-year mortality | Same |
| Causal contrast | Intention-to-treat or per-protocol | Same (with appropriate methods) |
| Analysis plan | Intention-to-treat analysis | Clone-censor-weight or similar |

### Why this reduces bias

Target trial emulation directly addresses several common biases:

**Immortal time bias**: In a naive observational study comparing drug users to non-users, the time between cohort entry and first prescription is "immortal" --- the patient must survive to receive the drug. This artificially inflates survival in the treated group. Target trial emulation avoids this by aligning "time zero" (start of follow-up) with treatment assignment.

**Selection bias from prevalent user designs**: Studying patients already on a drug (prevalent users) conflates treatment initiation effects with survival conditioning. Target trial emulation uses a **new-user (incident user) design**.

**Vague eligibility and time zero**: When eligibility criteria and the start of follow-up are not clearly defined, analyses can be plagued by look-ahead bias. The target trial framework forces explicit specification.

::: {.callout-tip}
## The TARGET reporting guideline (2025)
The TARGET (TrAnsparent ReportinG of observational studies Emulating a Target trial) guideline provides a structured checklist for reporting target trial emulation studies. It was published in the BMJ in 2025 and should be followed for any observational study making causal claims.
:::

### Example: Statin initiation and cardiovascular events

Suppose we want to estimate the effect of initiating statins within 6 months of a type 2 diabetes diagnosis on 5-year cardiovascular events.

**Target trial specification:**

- **Eligibility**: Adults aged 40--75 with new T2DM diagnosis, no prior CVD, no prior statin use
- **Treatment strategies**: (1) Initiate statin within 6 months, (2) Do not initiate statin within 6 months
- **Time zero**: Date of T2DM diagnosis
- **Outcome**: First major adverse cardiovascular event (MACE) within 5 years
- **Analysis**: Intention-to-treat using IPTW for confounding, clone-censor-weight for treatment strategy adherence

## Regression Discontinuity Design

**Regression discontinuity (RD)** exploits arbitrary thresholds in clinical decision-making. When treatment is assigned based on whether a continuous variable exceeds a cutoff, patients just above and just below the cutoff are essentially randomly assigned.

**Clinical example**: Many guidelines recommend antihypertensive therapy when systolic blood pressure exceeds 140 mmHg. Patients with SBP of 141 are very similar to patients with SBP of 139, but the former are much more likely to receive treatment. Comparing outcomes near the cutoff provides a quasi-experimental estimate.

Key requirements for a valid RD design:

1. **Treatment probability changes sharply at the cutoff** (first stage)
2. **Other covariates do not jump at the cutoff** (continuity assumption)
3. **Patients cannot precisely manipulate their score** (no bunching)

RD designs are less common in clinical research than propensity score methods, but when applicable, they provide strong evidence because the assumptions are more testable.

## Implementation in Python

Python's causal inference ecosystem is maturing. Here we demonstrate propensity score matching and IPTW using `dowhy` and `causalinference`.

```{python}
#| label: python-causal
#| eval: false

import numpy as np
import pandas as pd
from sklearn.linear_model import LogisticRegression
from scipy import stats
import matplotlib.pyplot as plt

# Simulate clinical data
np.random.seed(42)
n = 2000

age = np.random.normal(65, 10, n)
diabetes = np.random.binomial(1, 0.3, n)
egfr = np.random.normal(60, 15, n)
smoking = np.random.binomial(1, 0.25, n)

# Confounded treatment assignment
ps_true = 1 / (1 + np.exp(-(-2 + 0.03*age + 0.5*diabetes - 0.02*egfr + 0.4*smoking)))
treatment = np.random.binomial(1, ps_true)

# Outcome
mortality_prob = 1 / (1 + np.exp(-(-3 + 0.04*age + 0.6*diabetes -
                                     0.03*egfr + 0.3*smoking -
                                     0.5*treatment)))
mortality = np.random.binomial(1, mortality_prob)

df = pd.DataFrame({
    'age': age, 'diabetes': diabetes, 'egfr': egfr,
    'smoking': smoking, 'treatment': treatment, 'mortality': mortality
})

# --- Propensity Score Estimation ---
covariates = ['age', 'diabetes', 'egfr', 'smoking']
ps_model = LogisticRegression(max_iter=1000)
ps_model.fit(df[covariates], df['treatment'])
df['ps'] = ps_model.predict_proba(df[covariates])[:, 1]

# --- IPTW ---
# Compute ATE weights
df['iptw'] = np.where(
    df['treatment'] == 1,
    1 / df['ps'],
    1 / (1 - df['ps'])
)

# Stabilised weights
p_treat = df['treatment'].mean()
df['sw'] = np.where(
    df['treatment'] == 1,
    p_treat / df['ps'],
    (1 - p_treat) / (1 - df['ps'])
)

# Check balance: weighted standardised mean differences
def weighted_smd(data, var, treatment_col, weight_col):
    treated = data[data[treatment_col] == 1]
    control = data[data[treatment_col] == 0]
    mean_t = np.average(treated[var], weights=treated[weight_col])
    mean_c = np.average(control[var], weights=control[weight_col])
    sd_pooled = np.sqrt((treated[var].var() + control[var].var()) / 2)
    return (mean_t - mean_c) / sd_pooled

print("Covariate balance (weighted SMD):")
for var in covariates:
    smd = weighted_smd(df, var, 'treatment', 'sw')
    print(f"  {var}: {smd:.4f}")

# --- Weighted outcome estimation ---
from statsmodels.api import WLS, add_constant

X = add_constant(df['treatment'])
wls_model = WLS(df['mortality'], X, weights=df['sw']).fit()
print("\nIPTW Treatment Effect Estimate:")
print(f"  Coefficient: {wls_model.params[1]:.4f}")
print(f"  95% CI: ({wls_model.conf_int()[1][0]:.4f}, {wls_model.conf_int()[1][1]:.4f})")
```

**What the code shows.** This is the Python counterpart to the R workflow above, written from scratch so you can see every step rather than relying on a package. We simulate the same confounded cohort, fit a logistic regression for the propensity score, then build the inverse-probability weights (both raw and stabilised). The `weighted_smd` helper prints the standardised mean difference for each covariate *after* weighting --- the printed numbers should all be close to zero (well under 0.1) if weighting has balanced the groups. The final weighted least-squares model regresses mortality on treatment using the stabilised weights; its `treatment` coefficient is the estimated effect on the probability of death (a risk difference, since the outcome is on the probability scale), reported with a 95% confidence interval.

### Using DoWhy for causal analysis

```{python}
#| label: dowhy-example
#| eval: false

# DoWhy provides a structured causal inference workflow
# Install: pip install dowhy

import dowhy
from dowhy import CausalModel

# Step 1: Model - encode causal assumptions
model = CausalModel(
    data=df,
    treatment='treatment',
    outcome='mortality',
    common_causes=['age', 'diabetes', 'egfr', 'smoking']
)

# Step 2: Identify - find valid adjustment strategy
identified_estimand = model.identify_effect()
print(identified_estimand)

# Step 3: Estimate - using propensity score weighting
estimate = model.estimate_effect(
    identified_estimand,
    method_name="backdoor.propensity_score_weighting",
    method_params={"weighting_scheme": "ips_stabilized_weight"}
)
print(f"Causal Estimate: {estimate.value:.4f}")

# Step 4: Refute - sensitivity analysis
refutation = model.refute_estimate(
    identified_estimand, estimate,
    method_name="random_common_cause"
)
print(refutation)
```

**What the code shows.** `DoWhy` makes the four logical steps of a causal analysis explicit rather than burying them in modelling code. **Model** records your assumptions (treatment, outcome, and the confounders to adjust for). **Identify** uses those assumptions to work out, on paper, which quantity can actually be estimated --- printing the adjustment formula. **Estimate** then computes it; here using stabilised propensity-score weighting, so the printed `estimate.value` is comparable to the IPTW number from the manual code above. **Refute** is the part most analysts skip: it stress-tests the estimate by, for example, adding a randomly generated extra "common cause." A trustworthy estimate should barely move when a *random* (and therefore non-confounding) variable is added; a large change is a red flag that the analysis is fragile.

## Sensitivity Analysis for Unmeasured Confounding

No observational study can prove the absence of unmeasured confounding. **Sensitivity analysis** quantifies how strong unmeasured confounding would need to be to explain away the observed effect.

The **E-value** (VanderWeele and Ding, 2017) provides a simple summary: it is the minimum strength of association (on the risk ratio scale) that an unmeasured confounder would need to have with both the treatment and the outcome to fully explain away the observed treatment-outcome association.

$$
\text{E-value} = RR + \sqrt{RR \times (RR - 1)}
$$

where $RR$ is the observed risk ratio. The E-value for the confidence interval bound closest to the null is also reported.

```{r}
#| label: evalue
#| eval: false

# E-value calculation
library(EValue)

# Suppose our IPTW analysis found RR = 0.65 (95% CI: 0.50, 0.85)
# Convert protective RR to above-null: 1/0.65 = 1.538
evalues.RR(est = 1/0.65, lo = 1/0.85, hi = 1/0.50)
```

**What the code shows.** `evalues.RR` takes our observed risk ratio (and its confidence limits) and returns the E-value. Because a protective effect (RR below 1) and a harmful one of the same magnitude are equivalent for this calculation, we flip the protective RR of 0.65 to `1/0.65 = 1.54` first. The function prints two numbers: the E-value for the point estimate and the E-value for the confidence-interval bound nearest the null. Read them like this: "an unmeasured confounder would need to be associated with both treatment and mortality by a risk ratio of *at least this value*, over and above the confounders we already adjusted for, to fully explain away our result." A large E-value (say 3 or more) means only a very strong hidden confounder could overturn the finding, so the result is fairly robust; an E-value close to 1 means a weak, plausible confounder could erase it.

## Practical Guidance: Choosing Your Method

| Scenario | Recommended approach |
|---|---|
| Binary treatment, many controls available | Propensity score matching |
| Binary treatment, need full sample | IPTW |
| Concern about model misspecification | Doubly robust estimation |
| Time-varying treatment | Marginal structural models (IPTW over time) |
| Natural threshold for treatment | Regression discontinuity |
| Emulating a specific trial design | Target trial emulation framework |

## Exercises

### Exercise 1: DAG construction (Conceptual)

You are studying the relationship between **ACE inhibitor use** and **acute kidney injury (AKI)** in hospitalised patients.

a) List at least five variables that might be relevant to this relationship.
b) Draw a DAG (on paper or using DAGitty) encoding your causal assumptions.
c) Identify the minimal sufficient adjustment set for estimating the effect of ACE inhibitors on AKI.
d) Identify at least one potential collider. What would happen if you adjusted for it?

### Exercise 2: Propensity score matching in R

Using the simulated dataset below, perform propensity score matching and estimate the treatment effect.

```{r}
#| label: exercise-2-data
#| eval: false

library(tidyverse)  # tibble()

set.seed(123)
n <- 1500
exercise_dat <- tibble(
  age = rnorm(n, 70, 8),
  creatinine = rnorm(n, 1.2, 0.4),
  heart_failure = rbinom(n, 1, 0.35),
  prior_mi = rbinom(n, 1, 0.20),
  # Confounded treatment (beta-blocker use)
  treatment = rbinom(n, 1, plogis(-1 + 0.02*age + 0.3*heart_failure +
                                     0.5*prior_mi - 0.8*creatinine)),
  # Outcome: 1-year mortality
  death_1yr = rbinom(n, 1, plogis(-2 + 0.05*age + 0.4*heart_failure +
                                     0.6*prior_mi + 1.0*creatinine -
                                     0.7*treatment))
)
```

a) Estimate propensity scores using logistic regression.
b) Perform 1:1 nearest-neighbour matching with a caliper of 0.2 SD.
c) Create a Love plot to assess balance before and after matching.
d) Estimate the ATT for beta-blocker use on 1-year mortality.
e) Calculate the E-value for your estimate. Interpret it.

### Exercise 3: IPTW in Python

Using the same data-generating process as Exercise 2 (replicate it in Python):

```{python}
#| label: exercise-3-data
#| eval: false

import numpy as np
import pandas as pd

np.random.seed(123)
n = 1500
age = np.random.normal(70, 8, n)
creatinine = np.random.normal(1.2, 0.4, n)
heart_failure = np.random.binomial(1, 0.35, n)
prior_mi = np.random.binomial(1, 0.20, n)

ps_true = 1 / (1 + np.exp(-(-1 + 0.02*age + 0.3*heart_failure +
                               0.5*prior_mi - 0.8*creatinine)))
treatment = np.random.binomial(1, ps_true)

mort_prob = 1 / (1 + np.exp(-(-2 + 0.05*age + 0.4*heart_failure +
                                 0.6*prior_mi + 1.0*creatinine -
                                 0.7*treatment)))
death_1yr = np.random.binomial(1, mort_prob)

df = pd.DataFrame({
    'age': age, 'creatinine': creatinine, 'heart_failure': heart_failure,
    'prior_mi': prior_mi, 'treatment': treatment, 'death_1yr': death_1yr
})
```

a) Fit a propensity score model and compute stabilised IPTW weights.
b) Assess covariate balance using weighted standardised mean differences.
c) Estimate the ATE using weighted regression.
d) Perform a sensitivity analysis: add a simulated unmeasured confounder and show how the estimate changes.

### Exercise 4: Target trial emulation (Conceptual)

You have access to a large electronic health records database and want to estimate the effect of **early metformin initiation** (within 3 months of type 2 diabetes diagnosis) vs **delayed initiation** (3--12 months after diagnosis) on 5-year cardiovascular events.

a) Write a complete target trial protocol table (eligibility, treatment strategies, assignment, start of follow-up, outcome, causal contrast, analysis plan).
b) Identify potential sources of immortal time bias in a naive analysis.
c) Describe how the new-user, active comparator design addresses confounding by indication.
d) What unmeasured confounders might remain? How would you conduct a sensitivity analysis?

## Summary

| Concept | Key point |
|---|---|
| Confounding by indication | Sicker patients get more treatment --- naive comparisons are biased |
| DAGs | Encode causal assumptions visually; identify what to adjust for |
| Propensity score | $P(\text{treatment} \mid \text{covariates})$; reduces dimensionality |
| Matching | Creates balanced pseudo-randomised groups; check balance |
| IPTW | Reweights full sample; preserves sample size |
| Doubly robust | Combines PS + outcome model; consistent if either is correct |
| Target trial emulation | Design observational analysis to mimic an ideal RCT |
| E-value | Quantifies robustness to unmeasured confounding |

## References and Further Reading

1. **Hernan MA, Robins JM.** *Causal Inference: What If.* Chapman & Hall/CRC, 2024. Freely available at [https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/). The definitive modern textbook on causal inference.

2. **Rosenbaum PR.** *Observational Studies.* 2nd ed. Springer, 2002. The classic text on propensity score methods and sensitivity analysis.

3. **Rosenbaum PR, Rubin DB.** The central role of the propensity score in observational studies for causal effects. *Biometrika.* 1983;70(1):41--55.

4. **TARGET Guideline.** TrAnsparent ReportinG of observational studies Emulating a Target trial. *BMJ.* 2025. Reporting checklist for target trial emulation studies.

5. **VanderWeele TJ, Ding P.** Sensitivity analysis in observational research: introducing the E-value. *Annals of Internal Medicine.* 2017;167(4):268--274.

6. **Textor J, van der Zander B, Gilthorpe MS, et al.** Robust causal inference using directed acyclic graphs: the R package 'dagitty'. *International Journal of Epidemiology.* 2016;45(6):1887--1894.

7. **Ho DE, Imai K, King G, Stuart EA.** MatchIt: nonparametric preprocessing for parametric causal inference. *Journal of Statistical Software.* 2011;42(8):1--28.

8. **Greifer N.** WeightIt: weighting for covariate balance in observational studies. R package. [https://ngreifer.github.io/WeightIt/](https://ngreifer.github.io/WeightIt/).

9. **Sharma A, Kiciman E.** DoWhy: An end-to-end library for causal inference. arXiv:2011.04216, 2020.

10. **Lodi S, Phillips A, Lundgren J, et al.** Effect estimates in randomized trials and observational studies: comparing apples with apples. *American Journal of Epidemiology.* 2019;188(8):1569--1577.
