# Estimating Treatment Effects: Inverse Probability Weighting and G-Computation {#sec-g-methods}

Imagine you are handed a hospital database and a single, deceptively simple question: does this blood-pressure drug actually *prevent* deaths, or does it only *look* protective because the doctors who prescribe it tend to pick their healthier patients? In the previous chapter we met this problem --- confounding by indication --- and we learned to draw it out with a directed acyclic graph (DAG), to estimate a propensity score, and to balance groups by matching. This chapter takes the next step: once the groups are balanced, how do we actually *compute* the treatment effect, with an honest confidence interval, in a way that recovers the truth? Two methods do almost all the heavy lifting in modern causal analysis, and both come from the work of the epidemiologist James Robins. They are called the **g-methods** ("g" for *generalised*). This chapter is a slow, hands-on tour of the two foundational ones: **inverse probability weighting** and **g-computation**.

## Introduction

In the last chapter we built the conceptual scaffolding. We said that every patient has two *potential outcomes* --- what would happen if treated, and what would happen if untreated --- and that the causal effect is the contrast between them. We also said that the central obstacle is confounding: the treated and untreated patients differ systematically, so a raw comparison is biased. We introduced the propensity score (a patient's probability of being treated given their characteristics) and used it to match patients and, briefly, to weight them.

This chapter answers the practical question that follows: **given a balanced or balanceable dataset, how do I produce a single, defensible number for the average treatment effect?** There are two foundational answers, and it is worth knowing both because they make their assumptions in different places:

- **Inverse probability weighting (IPW)** models *who gets treated*. It builds a re-weighted "pseudo-population" in which treatment is no longer tied to the confounders, then compares outcomes in that pseudo-population.
- **G-computation** (also called the *parametric g-formula* or *standardisation*) models *what the outcome would be*. It fits an outcome model, then uses it to predict every patient's outcome twice --- once as if treated, once as if untreated --- and contrasts the averages.

Both target the same quantity, the average treatment effect (ATE) we defined last chapter, and under the same core assumption (no unmeasured confounding) both recover it. We will build each from first principles, run them on simulated data where we *know* the right answer, and show that a naive regression gets it wrong while the g-methods get it right.

::: {.callout-important}
## Why two methods instead of one?
A naive analyst fits one regression and reads off a coefficient. The trouble is that a regression coefficient for treatment is only the causal effect under restrictive conditions (no interactions, correct functional form, a collapsible effect measure). The g-methods sidestep these traps by computing the effect *the way the definition demands* --- as a contrast of average potential outcomes --- rather than trusting a single coefficient to mean what we hope it means. And because IPW and g-computation lean on different models, agreement between them is itself reassuring evidence.
:::

## The Running Clinical Example

Throughout this chapter we use one scenario. We have observational data on patients started (or not) on a hypothetical drug, and we want its effect on one-year mortality. The catch, as always, is **confounding by indication**: sicker patients --- older, with worse kidney function, more comorbidity --- are *both* more likely to be treated *and* more likely to die regardless of treatment. A raw comparison will therefore make the drug look useless or even harmful, even if it genuinely saves lives. Our job is to undo that distortion.

```{mermaid}
%%| eval: true
%%| echo: false
%%| label: fig-confounding-gmethods
%%| fig-cap: "Confounding by indication, drawn as the classic confounding triangle. Illness severity (the gold node) is a *common cause*: it points to BOTH the drug and mortality, so the raw drug-mortality association is contaminated. The g-methods aim to isolate the genuine drug -> mortality arrow. (This is confounding, not mediation: a confounder points *into* both treatment and outcome, whereas a mediator would sit *on* the path, treatment -> mediator -> outcome.)"
graph TD
    L["Illness severity<br/>(age, kidney function, comorbidity)"]
    A["Drug (treatment)"]
    Y["One-year mortality"]
    L --> A
    L --> Y
    A --> Y
    style L fill:#fbf3e8,color:#111
    style A fill:#eef3fb,color:#111
    style Y fill:#eef3fb,color:#111
```

The single arrow from drug to mortality is the *causal effect* we want. The two arrows out of illness severity are the *backdoor path* that contaminates the raw comparison. Everything below is machinery for shutting that backdoor.

## Inverse Probability Weighting

### The idea in plain language

Recall the propensity score from last chapter: each patient's probability of receiving the treatment they were eligible for, given their measured characteristics [@haukoos2015propensity]. Inverse probability of treatment weighting (often shortened to IPW or IPTW) uses that score to **re-weight** the sample.

Here is the intuition, set out accessibly for clinical readers by @mansournia2016ipw and @chesnaye2022ipw. Suppose a very sick patient --- the kind clinicians almost always treat --- nevertheless went *untreated*. That patient is rare and precious: they are one of the few untreated people who *look like* the treated group, so they carry a lot of information about what happens to treated-type patients when they go without the drug. IPW gives such a patient a large weight. Conversely, an untreated patient who looked very unlikely to be treated anyway is "over-represented" among controls, so IPW shrinks their weight. Do this for everyone and you build a **pseudo-population** in which being treated is statistically unrelated to the confounders --- exactly the situation a randomised trial creates by design.

The weight for each patient is the inverse of the probability of the treatment they actually received:

$$
w_i = \frac{A_i}{e(L_i)} + \frac{1 - A_i}{1 - e(L_i)}
$$

where $A_i$ is 1 if treated and 0 if not, and $e(L_i)$ is the propensity score. A treated patient with propensity score 0.2 gets weight $1/0.2 = 5$; an untreated patient with propensity score 0.9 (so probability of being untreated $= 0.1$) gets weight $1/0.1 = 10$. These weights are larger precisely for the informative, "surprising" patients.

### Stabilised weights

The raw weights above can occasionally become enormous --- if a treated patient had a propensity score of 0.01, their weight is 100, and a single patient can then dominate the analysis and inflate the variance. **Stabilised weights** tame this by replacing the "1" in the numerator with the overall probability of receiving that treatment:

$$
sw_i = \frac{A_i \cdot P(A=1)}{e(L_i)} + \frac{(1 - A_i) \cdot P(A=0)}{1 - e(L_i)}
$$

The estimate is unchanged on average, but the weights cluster sensibly around 1 and the confidence intervals are narrower and more honest. Stabilised weights are the default recommendation and what modern packages produce automatically.

::: {.callout-note}
## Marginal structural models in one line
When you fit an outcome model (say, a regression of mortality on treatment) *using these weights*, the model you have fitted is called a **marginal structural model (MSM)**. "Marginal" because it describes the whole population (not conditional on confounders), and "structural" because its coefficients have a causal interpretation. MSMs are the natural home of IPW and the standard tool when a treatment changes over time --- a setting where ordinary regression genuinely cannot give the right answer.
:::

### Checking the weights worked: balance and positivity

Weighting is only trustworthy if it actually balanced the confounders, and only valid if every kind of patient could in principle have received either treatment. So two checks are mandatory:

1. **Covariate balance.** After weighting, the treated and control groups should have a near-identical mix of confounders. We measure this with the *standardised mean difference* (SMD) --- the difference in a covariate's mean between groups, divided by its standard deviation --- and want every value comfortably below 0.1.
2. **Positivity.** Recall from last chapter: every patient must have a non-zero chance of either treatment. In weighting terms, a propensity score near 0 or 1 produces an explosive weight. So we inspect the largest weights; a maximum weight above roughly 10--20 is a warning sign that positivity is shaky and the estimate may be unstable.

### IPW in practice

::: {.panel-tabset}

### R

```{r}
#| label: ipw-r
#| eval: false

library(WeightIt)        # estimate balancing weights
library(cobalt)          # check covariate balance
library(marginaleffects) # turn the weighted model into an effect

# --- Simulate a confounded cohort with a KNOWN true effect ---
set.seed(2024)
n <- 3000
age       <- rnorm(n, 65, 10)
egfr      <- rnorm(n, 60, 15)               # kidney function
comorbid  <- rbinom(n, 1, 0.4)              # any major comorbidity

# Sicker patients (older, worse kidneys, comorbid) are MORE likely treated
lp_treat  <- -1 + 0.03 * (age - 65) - 0.03 * (egfr - 60) + 0.5 * comorbid
ps_true   <- plogis(lp_treat)
treat     <- rbinom(n, 1, ps_true)

# TRUE causal effect of treatment on mortality: log-odds -0.7 (protective)
lp_death  <- -1 + 0.04 * (age - 65) - 0.03 * (egfr - 60) + 0.6 * comorbid - 0.7 * treat
death     <- rbinom(n, 1, plogis(lp_death))

dat <- data.frame(age, egfr, comorbid, treat, death)

# --- Step 1: estimate stabilised weights for the ATE ---
W <- weightit(treat ~ age + egfr + comorbid,
              data = dat, method = "glm", estimand = "ATE",
              stabilize = TRUE)

# --- Step 2: check balance and the size of the weights ---
bal.tab(W, thresholds = c(m = 0.1))   # SMDs after weighting; want < 0.1
summary(W)                            # inspect the largest weights (positivity)

# --- Step 3: fit the weighted outcome model (a marginal structural model) ---
msm <- glm_weightit(death ~ treat, data = dat, weightit = W,
                    family = binomial)

# --- Step 4: read off the causal effect on the risk-difference scale ---
avg_comparisons(msm, variables = "treat")
```

**What the code shows.** We first manufacture a cohort where treatment is deliberately confounded and the *true* protective effect is built in (the `- 0.7 * treat` term in the mortality equation). `weightit()` estimates the propensity score and converts it into stabilised ATE weights in one call. `bal.tab()` then prints a table of standardised mean differences for `age`, `egfr`, and `comorbid` *after* weighting --- you want every number under 0.1, confirming the pseudo-population is balanced. `summary(W)` reports the range of weights; a sane maximum (well under 20 here) means positivity is not violated. Finally `glm_weightit()` fits the weighted outcome model --- this is the marginal structural model --- and `avg_comparisons()` converts it into a clean *risk difference*: the average change in the probability of death caused by the drug, with a confidence interval that correctly accounts for the weighting. Because we hard-wired a protective effect, you should see a negative risk difference (fewer deaths under treatment), unlike the naive analysis we run later.

### Python

```{python}
#| label: ipw-py
#| eval: false

import numpy as np
import pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf

# --- Simulate a confounded cohort with a KNOWN true effect ---
rng = np.random.default_rng(2024)
n = 3000
age      = rng.normal(65, 10, n)
egfr     = rng.normal(60, 15, n)
comorbid = rng.binomial(1, 0.4, n)

lp_treat = -1 + 0.03*(age-65) - 0.03*(egfr-60) + 0.5*comorbid
ps_true  = 1/(1+np.exp(-lp_treat))
treat    = rng.binomial(1, ps_true)

# TRUE causal effect: log-odds -0.7 (protective)
lp_death = -1 + 0.04*(age-65) - 0.03*(egfr-60) + 0.6*comorbid - 0.7*treat
death    = rng.binomial(1, 1/(1+np.exp(-lp_death)))

df = pd.DataFrame(dict(age=age, egfr=egfr, comorbid=comorbid,
                       treat=treat, death=death))

# --- Step 1: estimate the propensity score ---
ps_model = smf.logit("treat ~ age + egfr + comorbid", data=df).fit(disp=0)
df["ps"] = ps_model.predict(df)

# --- Step 2: stabilised weights ---
p_treat = df["treat"].mean()
df["sw"] = np.where(df["treat"] == 1,
                    p_treat / df["ps"],
                    (1 - p_treat) / (1 - df["ps"]))

# --- Step 3: check balance (weighted standardised mean differences) ---
def wsmd(v):
    t, c = df[df.treat == 1], df[df.treat == 0]
    mt = np.average(t[v], weights=t.sw)
    mc = np.average(c[v], weights=c.sw)
    sd = np.sqrt((t[v].var() + c[v].var()) / 2)
    return (mt - mc) / sd

print("Weighted SMDs (want |value| < 0.1):")
for v in ["age", "egfr", "comorbid"]:
    print(f"  {v}: {wsmd(v):+.3f}")
print("Max weight (positivity check):", df["sw"].max().round(2))

# --- Step 4: weighted outcome model on the risk-difference scale ---
msm = smf.glm("death ~ treat", data=df, family=sm.families.Binomial(),
              var_weights=df["sw"]).fit()
# Convert to a marginal risk difference by predicting under treat=0/1
d1 = df.assign(treat=1); d0 = df.assign(treat=0)
rd = (msm.predict(d1) - msm.predict(d0)).mean()
print(f"\nIPW risk difference (treat - control): {rd:+.4f}")
```

**What the code shows.** This is the same workflow written from scratch so every step is visible. We simulate the identical confounded cohort, fit a logistic propensity-score model with `statsmodels`, and build stabilised weights by hand. The `wsmd` helper prints the weighted standardised mean difference for each confounder --- all should be near zero (under 0.1) if weighting balanced the groups --- and we print the maximum weight as a positivity check. We then fit the weighted outcome model and, rather than reading a single coefficient, we predict each patient's death probability twice (everyone treated, then everyone untreated) and average the difference. That average is the *risk difference*, directly comparable to the R output, and should be negative because we built in a protective effect.

:::

## G-Computation

### The idea in plain language

IPW models the *treatment*. **G-computation** instead models the *outcome*, and it follows the definition of the average treatment effect almost literally.

Recall the definition: the ATE is the average outcome if *everyone* were treated minus the average outcome if *everyone* were untreated. We never observe both worlds for any patient --- but we can *predict* them. The recipe has three steps:

1. **Fit one outcome model** that includes treatment *and* all the confounders (and, ideally, their interactions). This model learns how outcome depends jointly on treatment and patient characteristics.
2. **Predict twice for every patient.** First set everyone's treatment to "treated" (leaving their real confounders untouched) and predict each patient's outcome. Then reset everyone to "untreated" and predict again. Each patient now has two predicted outcomes --- their estimated potential outcomes.
3. **Average and contrast.** Average the "all treated" predictions, average the "all untreated" predictions, and take the difference. That difference *is* the average treatment effect.

This procedure is also called **standardisation**, because it standardises both arms to the *same* confounder distribution (the whole sample's). It is the oldest idea in the chapter --- Robins introduced the g-formula in 1986 --- and it remains one of the most reliable.

```{mermaid}
%%| eval: true
%%| echo: false
%%| label: fig-gcomp-steps
%%| fig-cap: "G-computation in three steps: fit one outcome model, predict every patient's outcome under each treatment, then average and contrast to obtain the average treatment effect."
graph LR
    A["Fit outcome model:<br/>outcome ~ treatment<br/>+ confounders"] --> B["Predict each patient<br/>as if TREATED"]
    A --> C["Predict each patient<br/>as if UNTREATED"]
    B --> D["Average each set,<br/>then subtract"]
    C --> D
    D --> E["Average treatment<br/>effect"]
    style A fill:#eef3fb,color:#111
    style B fill:#eef3fb,color:#111
    style C fill:#eef3fb,color:#111
    style D fill:#eef3fb,color:#111
    style E fill:#eef3fb,color:#111
```

### Confidence intervals: use the bootstrap

Because g-computation predicts, averages, and contrasts, the uncertainty in the final number does not come out of the outcome model directly. The standard solution is the **bootstrap**: resample the patients with replacement many times, rerun the whole predict-average-contrast procedure on each resample, and use the spread of the resulting estimates as the confidence interval. Modern tooling automates this --- in R, `marginaleffects::inferences()` wraps the bootstrap around the whole calculation.

### G-computation in practice

::: {.panel-tabset}

### R

```{r}
#| label: gcomp-r
#| eval: false

library(marginaleffects)  # avg_comparisons() + inferences() bootstrap

# --- Simulate the SAME confounded cohort with a known true effect ---
set.seed(2024)
n <- 3000
age       <- rnorm(n, 65, 10)
egfr      <- rnorm(n, 60, 15)
comorbid  <- rbinom(n, 1, 0.4)
ps_true   <- plogis(-1 + 0.03*(age-65) - 0.03*(egfr-60) + 0.5*comorbid)
treat     <- rbinom(n, 1, ps_true)
death     <- rbinom(n, 1, plogis(-1 + 0.04*(age-65) - 0.03*(egfr-60) +
                                   0.6*comorbid - 0.7*treat))
dat <- data.frame(age, egfr, comorbid, treat, death)

# --- Step 1: fit ONE outcome model with treatment + confounders ---
#     Include treatment-confounder interactions so effects can vary by patient
out_model <- glm(death ~ treat * (age + egfr + comorbid),
                 data = dat, family = binomial)

# --- Steps 2-4: predict under treat=1 and treat=0, average, contrast ---
#     avg_comparisons does the standardisation; inferences() bootstraps the CI
avg_comparisons(out_model, variables = "treat") |>
  inferences(method = "boot", R = 500)
```

**What the code shows.** We rebuild the identical confounded cohort (this chunk is fully self-contained). We then fit a *single* logistic outcome model, deliberately including `treat * (age + egfr + comorbid)` so the drug's effect is allowed to differ across patient types --- a flexibility plain regression coefficients cannot capture. `avg_comparisons(..., variables = "treat")` carries out the g-computation automatically: it predicts every patient's death probability with `treat` set to 1, again with it set to 0, averages each, and reports the difference as a risk difference. Piping into `inferences(method = "boot", R = 500)` wraps the whole procedure in a 500-replicate bootstrap so the confidence interval is honest. The estimate should be negative and close to the IPW estimate above --- two different models, agreeing on the truth.

### Python

```{python}
#| label: gcomp-py
#| eval: false

import numpy as np
import pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf

# --- Simulate the SAME confounded cohort ---
rng = np.random.default_rng(2024)
n = 3000
age      = rng.normal(65, 10, n)
egfr     = rng.normal(60, 15, n)
comorbid = rng.binomial(1, 0.4, n)
ps_true  = 1/(1+np.exp(-(-1 + 0.03*(age-65) - 0.03*(egfr-60) + 0.5*comorbid)))
treat    = rng.binomial(1, ps_true)
death    = rng.binomial(1, 1/(1+np.exp(-(-1 + 0.04*(age-65) - 0.03*(egfr-60)
                                          + 0.6*comorbid - 0.7*treat))))
df = pd.DataFrame(dict(age=age, egfr=egfr, comorbid=comorbid,
                       treat=treat, death=death))

# --- The g-computation procedure as a reusable function ---
def gcomp(data):
    m = smf.glm("death ~ treat * (age + egfr + comorbid)",
                data=data, family=sm.families.Binomial()).fit(disp=0)
    d1 = data.assign(treat=1)   # everyone treated
    d0 = data.assign(treat=0)   # everyone untreated
    return m.predict(d1).mean() - m.predict(d0).mean()

point = gcomp(df)

# --- Bootstrap the confidence interval ---
rng_boot = np.random.default_rng(1)
boot = np.array([
    gcomp(df.sample(len(df), replace=True, random_state=int(rng_boot.integers(1e9))))
    for _ in range(500)
])
lo, hi = np.percentile(boot, [2.5, 97.5])
print(f"G-computation risk difference: {point:+.4f}  (95% CI {lo:+.4f}, {hi:+.4f})")
```

**What the code shows.** We wrap the predict-average-contrast recipe in a function, `gcomp`, that fits the interaction outcome model, predicts every patient's death probability under "everyone treated" and "everyone untreated", and returns the average difference. Calling it once on the real data gives the point estimate. We then call it 500 more times on bootstrap resamples (sampling patients with replacement) and take the 2.5th and 97.5th percentiles of those estimates as a 95 percent confidence interval. The printed risk difference should be negative and close to the R g-computation result and the IPW result --- demonstrating that, when the assumptions hold, the route you take to the answer matters less than getting the confounding adjustment right.

:::

::: {.callout-tip}
## A ready-made alternative in R
You do not have to hand-roll the outcome model. The `stdReg2` package implements standardisation (g-computation) directly with built-in confidence intervals, and for *time-varying* treatments --- where a patient's treatment and confounders evolve over follow-up, and ordinary methods break down --- the `gfoRmula` package implements the full longitudinal parametric g-formula. We use the `glm` + `marginaleffects` route here because it makes every step transparent, but `stdReg2` is an excellent production tool.
:::

## Putting It Together: Naive vs IPW vs G-Computation

The point of the chapter is best seen in one side-by-side comparison. The chunk below estimates the drug's effect three ways on identical data: a *naive* regression that ignores confounding, IPW, and g-computation.

::: {.panel-tabset}

### R

```{r}
#| label: compare-r
#| eval: false

library(WeightIt)
library(marginaleffects)

# --- One confounded cohort, known true effect (protective) ---
set.seed(2024)
n <- 3000
age      <- rnorm(n, 65, 10)
egfr     <- rnorm(n, 60, 15)
comorbid <- rbinom(n, 1, 0.4)
treat    <- rbinom(n, 1, plogis(-1 + 0.03*(age-65) - 0.03*(egfr-60) + 0.5*comorbid))
death    <- rbinom(n, 1, plogis(-1 + 0.04*(age-65) - 0.03*(egfr-60) +
                                  0.6*comorbid - 0.7*treat))
dat <- data.frame(age, egfr, comorbid, treat, death)

# (a) NAIVE: ignore confounders entirely, compare raw mortality
naive <- glm(death ~ treat, data = dat, family = binomial)
naive_rd <- avg_comparisons(naive, variables = "treat")

# (b) IPW: stabilised weights + weighted outcome model
W   <- weightit(treat ~ age + egfr + comorbid, data = dat,
                method = "glm", estimand = "ATE", stabilize = TRUE)
ipw_rd <- avg_comparisons(glm_weightit(death ~ treat, data = dat, weightit = W,
                                       family = binomial), variables = "treat")

# (c) G-COMPUTATION: outcome model + standardisation
gcomp_rd <- avg_comparisons(glm(death ~ treat * (age + egfr + comorbid),
                                data = dat, family = binomial),
                            variables = "treat")

naive_rd; ipw_rd; gcomp_rd
```

**What the code shows.** Three estimates of the same risk difference on the same data. The **naive** model regresses death on treatment alone, so it absorbs all the confounding: because sicker patients were preferentially treated, it will report a risk difference near zero or even positive --- making a genuinely protective drug look useless or harmful. The **IPW** and **g-computation** estimates each adjust for `age`, `egfr`, and `comorbid` (one through weighting, one through outcome modelling) and should both land close to the true protective effect, i.e. a clearly negative risk difference. Seeing the naive estimate disagree with two methods that agree with each other is the whole lesson: confounding is not a rounding error, and the route you use to adjust matters far less than adjusting at all.

### Python

```{python}
#| label: compare-py
#| eval: false

import numpy as np, pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf

# --- One confounded cohort, known true effect (protective) ---
rng = np.random.default_rng(2024)
n = 3000
age      = rng.normal(65, 10, n)
egfr     = rng.normal(60, 15, n)
comorbid = rng.binomial(1, 0.4, n)
treat    = rng.binomial(1, 1/(1+np.exp(-(-1 + 0.03*(age-65) - 0.03*(egfr-60) + 0.5*comorbid))))
death    = rng.binomial(1, 1/(1+np.exp(-(-1 + 0.04*(age-65) - 0.03*(egfr-60)
                                          + 0.6*comorbid - 0.7*treat))))
df = pd.DataFrame(dict(age=age, egfr=egfr, comorbid=comorbid, treat=treat, death=death))

# (a) NAIVE: raw difference in mortality
naive = smf.glm("death ~ treat", data=df, family=sm.families.Binomial()).fit(disp=0)
naive_rd = naive.predict(df.assign(treat=1)).mean() - naive.predict(df.assign(treat=0)).mean()

# (b) IPW: stabilised weights + weighted outcome model
ps = smf.logit("treat ~ age + egfr + comorbid", data=df).fit(disp=0).predict(df)
p = df.treat.mean()
sw = np.where(df.treat == 1, p/ps, (1-p)/(1-ps))
mw = smf.glm("death ~ treat", data=df, family=sm.families.Binomial(),
             var_weights=sw).fit()
ipw_rd = mw.predict(df.assign(treat=1)).mean() - mw.predict(df.assign(treat=0)).mean()

# (c) G-COMPUTATION
gm = smf.glm("death ~ treat * (age + egfr + comorbid)", data=df,
             family=sm.families.Binomial()).fit(disp=0)
gcomp_rd = gm.predict(df.assign(treat=1)).mean() - gm.predict(df.assign(treat=0)).mean()

print(f"Naive risk difference:        {naive_rd:+.4f}")
print(f"IPW risk difference:          {ipw_rd:+.4f}")
print(f"G-computation risk difference:{gcomp_rd:+.4f}")
```

**What the code shows.** The Python version of the same three-way comparison, written out so each estimator is fully visible. The naive risk difference ignores confounding and will be biased toward (or past) zero; the IPW and g-computation risk differences adjust for the three confounders and should both be clearly negative, recovering the protective effect we built in. The take-home is identical to the R version: naive disagrees, the two principled methods agree.

:::

## IPW vs G-Computation, and a Word on "Best of Both"

The two methods are complementary, and choosing between them comes down to which model you trust more.

| | Inverse probability weighting | G-computation |
|---|---|---|
| What you model | The **treatment** (propensity score) | The **outcome** |
| Main risk | A wrong treatment model, or extreme weights (positivity) | A wrong outcome model (functional form, interactions) |
| Strengths | Natural for time-varying treatment (marginal structural models); transparent pseudo-population | Uses all data efficiently; no weight explosions; easy to add interactions |
| Confidence interval | Robust/sandwich or bootstrap | Bootstrap |

You do not have to pick. **Doubly robust** estimators --- augmented IPW (AIPW) and targeted maximum likelihood estimation (TMLE), which we previewed last chapter --- combine *both* a treatment model and an outcome model and stay valid as long as *at least one* of the two is correct. They are the modern "best of both worlds" default, and once you understand IPW and g-computation you already understand their two ingredients. In R they live in packages such as `AIPW` and `tmle`; the simplest mental model is "g-computation, then nudge the prediction using the weights to correct any outcome-model mistakes."

::: {.callout-warning}
## The assumption that no method can rescue
Every estimator in this chapter --- naive, IPW, g-computation, even doubly robust --- relies on **no unmeasured confounding** (the exchangeability assumption from last chapter). If illness severity has a component you did not measure, all of them are biased in the same direction. The g-methods give you a correct answer *to the question you asked of the data you have*; they cannot conjure information you never collected. This is why the E-value sensitivity analysis from last chapter belongs in every causal report.
:::

## Exercises

::: {.callout-tip title="Exercise 1: Recover a known effect (R or Python)"}

Simulate a confounded cohort of 2000 patients in which a single binary confounder `frail` raises both the chance of treatment and the risk of a binary bad outcome, and the treatment has a true protective log-odds of $-0.8$.

a) Estimate the effect by naive regression. By how much is it biased?
b) Estimate it by IPW (stabilised weights) and by g-computation.
c) Confirm both adjusted estimates are close to the truth and to each other.


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch17b_ex1.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch17b_ex1.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 2: When weights explode (Conceptual + code)"}

Modify your Exercise 1 simulation so that frail patients are treated with probability 0.98 and non-frail patients with probability 0.03.

a) Inspect the maximum stabilised weight. What does it tell you about positivity?
b) Which method --- IPW or g-computation --- is more affected, and why?
c) What would you report to a clinical collaborator about the trustworthiness of the IPW estimate here?


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch17b_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch17b_ex2.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 3: Interactions and standardisation (Conceptual)"}

In the g-computation example we wrote `death ~ treat * (age + egfr + comorbid)`.

a) Explain in plain words what the interaction terms allow the drug's effect to do.
b) If you omitted the interactions and the true effect really did differ by age, would a single regression coefficient still equal the average treatment effect? Why might standardisation still recover the right average?


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch17b_ex3.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch17b_ex3.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 4: Choosing a method (Conceptual)"}

For each scenario, say whether you would lead with IPW, g-computation, or a doubly robust estimator, and why: (a) a treatment whose dose changes month to month over two years; (b) a setting with a rich, well-understood outcome but a poorly understood treatment-assignment process; (c) a high-stakes regulatory analysis where you want a safety net against model misspecification.


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch17b_ex4.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch17b_ex4.py
```
:::
:::
:::

## Summary

| Concept | Key point |
|---|---|
| G-methods | Robins' family of methods for estimating causal effects under confounding |
| Inverse probability weighting | Re-weight by 1/P(received treatment) to build an unconfounded pseudo-population |
| Stabilised weights | Multiply by the marginal treatment probability to tame extreme weights |
| Marginal structural model | An outcome model fitted with IP weights; coefficients are causal |
| Positivity check | Inspect the largest weights; a maximum above ~20 is a red flag |
| G-computation | Fit one outcome model, predict everyone treated then untreated, average and contrast |
| Standardisation | Another name for g-computation; both arms standardised to the same confounder mix |
| Bootstrap | The standard way to get confidence intervals for g-computation |
| Doubly robust (AIPW/TMLE) | Combine treatment and outcome models; valid if either is correct |

::: {.callout-tip}
## Key Takeaways
- A treatment-coefficient from a single naive regression is **not** generally a causal effect. The g-methods compute the effect the way its definition demands.
- **IPW** models *who gets treated* and re-weights the sample; **g-computation** models *the outcome* and standardises. Both target the average treatment effect and both recover it when confounding is fully measured.
- Always **check balance and positivity** for IPW, and always **bootstrap** g-computation's confidence interval.
- When the two methods agree, that agreement is itself reassuring; when they disagree, suspect a misspecified model or shaky positivity.
- **Doubly robust** estimators (AIPW, TMLE) combine both ingredients and forgive a single modelling mistake --- a sensible default once you understand the two foundations.
- No method overcomes **unmeasured confounding**. Pair every causal estimate with a sensitivity analysis such as the E-value.
:::

## References and Further Reading

- For the foundations, see @hernan2024causal --- Part II covers IP weighting and the g-formula in depth --- and @robins1986gformula for the original g-formula.
- For marginal structural models and time-varying confounding, see @robins2000msm.
- For clinician-friendly introductions to weighting, see @mansournia2016ipw, @chesnaye2022ipw, and @haukoos2015propensity.
- For a code-first tutorial covering exactly these methods, see @chatton2024cookbook.
- For software, see @greifer2024weightit and @arelbundock2024marginaleffects in R, and @zivich2024zepid in Python.

::: {.sectionrefs}
:::
