# Producing Journal-Ready Statistical Analyses {#sec-journal-ready}

```{r}
#| include: false
library(tidyverse)
library(gtsummary)
library(gt)
library(survival)
library(broom)
library(patchwork)
```

## Introduction

You have learned how to fit models, validate predictions, and draw causal inferences. The final --- and often most undervalued --- skill is **communicating your results** in a form that meets the standards of peer-reviewed medical journals. Reviewers and editors routinely reject papers not because the statistics are wrong, but because the presentation is unclear, tables are poorly formatted, figures are low resolution, or the statistical methods section is incomplete.

This chapter walks you through the complete workflow: from creating publication-quality tables and figures to writing a statistical methods section, meeting reporting guideline requirements, and building a reproducible research pipeline. We conclude with three capstone project descriptions that bring together everything in the course.

For a clinician or epidemiologist, this is where months of analysis either land or fall flat. A clear "Table 1", a well-formatted regression table, and a figure that a reader can interpret at a glance are what convince reviewers --- and ultimately fellow clinicians --- that your findings are trustworthy. The code chunks below are set to `eval: false`, so the rendered page shows the code but not its output; the surrounding text explains what each table or figure *would* look like and how to read it, so you know exactly what to expect when you run the code on your own data.

## What Do Top Journals Expect?

### General standards

Major medical journals (The Lancet, BMJ, JAMA, NEJM, Lancet Digital Health) share common expectations:

1. **Transparent reporting**: Follow the relevant reporting guideline (CONSORT for RCTs, STROBE for observational studies, TRIPOD+AI for prediction models, PRISMA for systematic reviews).
2. **Pre-specified analysis plan**: Distinguish confirmatory from exploratory analyses.
3. **Effect estimates with uncertainty**: Report point estimates, confidence intervals, and (where appropriate) p-values. Never report p-values alone.
4. **Reproducibility**: Many journals now require data availability statements and code sharing.
5. **Figure quality**: Minimum 300 DPI, specific file formats (TIFF, EPS, or high-resolution PDF), accessible colour palettes.

### Journal-specific requirements

| Journal | Key statistical requirements |
|---|---|
| **Lancet / Lancet Digital Health** | TRIPOD+AI for prediction models; structured statistical methods; data sharing encouraged |
| **BMJ** | EQUATOR reporting guidelines mandatory; patient involvement statement; open access to data |
| **JAMA** | Statistical review by dedicated biostatistician; eTable/eFigure supplements expected |
| **NEJM** | Extreme rigour; independent statistical review; full protocol as supplement |

::: {.callout-important}
## The statistical review
All top medical journals employ dedicated statistical reviewers. Common reasons for rejection on statistical grounds include: inappropriate handling of missing data, failure to account for multiple comparisons, presenting associations as causal claims without proper methodology, and inadequate model validation.
:::

## Table 1: Baseline Characteristics

"Table 1" is the standard first table in any clinical paper, describing the study population by group (treatment arms, exposure groups, or clusters).

### What belongs in Table 1

- Demographics: age, sex, race/ethnicity
- Clinical characteristics: comorbidities, disease severity, vital signs
- Laboratory values: relevant biomarkers
- Medications and prior treatments
- Summary statistics: mean (SD) or median (IQR) for continuous variables; n (%) for categorical variables

### Creating Table 1 in R with gtsummary

```{r}
#| label: table1-r
#| eval: false

library(gtsummary)

# Using built-in trial dataset (modify for your data)
trial |>
  select(trt, age, grade, stage, marker, response) |>
  tbl_summary(
    by = trt,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 1,
      all_categorical() ~ c(0, 1)
    ),
    label = list(
      age ~ "Age, years",
      grade ~ "Tumour grade",
      stage ~ "Cancer stage",
      marker ~ "Biomarker level, ng/mL",
      response ~ "Tumour response"
    ),
    missing_text = "Missing"
  ) |>
  add_overall() |>
  add_p(
    test = list(
      all_continuous() ~ "t.test",
      all_categorical() ~ "chisq.test"
    )
  ) |>
  add_n() |>
  modify_header(label = "**Characteristic**") |>
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Treatment Group**") |>
  bold_labels() |>
  as_gt() |>
  gt::tab_header(
    title = "Table 1. Baseline Characteristics of the Study Population"
  )
```

**What the code shows.** This builds a complete, publication-ready "Table 1" from the `trial` dataset using `gtsummary`, the most widely used R package for clinical tables. The pipeline reads top to bottom: `tbl_summary(by = trt)` produces one column per treatment group, summarising continuous variables as mean (SD) and categorical ones as n (%); the later steps add an overall column (`add_overall`), a column of p-values comparing the groups (`add_p`), and the sample size per variable (`add_n`), then relabel and bold the rows and hand off to `gt` to add a title. The rendered result is the familiar table you see at the start of almost every clinical paper: rows are patient characteristics (age, tumour grade, stage, and so on), columns are the treatment arms, and each cell tells you the typical value in that group. Reading across a row lets you check whether the arms were balanced at baseline; the `label` argument is what turns raw variable names like `ph.ecog` into human-readable text for the reader.

::: {.callout-tip}
## Should Table 1 include p-values?
In a randomised trial, baseline differences are due to chance, so p-values for baseline comparisons are widely considered inappropriate (they test a hypothesis you know to be true). In observational studies, standardised mean differences (SMDs) are preferred over p-values because SMDs are not influenced by sample size.
:::

### Creating Table 1 in Python with tableone

```{python}
#| label: table1-python
#| eval: false

import pandas as pd
from tableone import TableOne
import numpy as np

# Simulate clinical trial data
np.random.seed(42)
n = 500

data = pd.DataFrame({
    'Treatment': np.random.choice(['Drug A', 'Drug B'], n),
    'Age': np.random.normal(65, 10, n).round(1),
    'Sex': np.random.choice(['Male', 'Female'], n, p=[0.55, 0.45]),
    'BMI': np.random.normal(28, 5, n).round(1),
    'Diabetes': np.random.choice(['Yes', 'No'], n, p=[0.3, 0.7]),
    'eGFR': np.random.normal(62, 18, n).round(0),
    'Systolic_BP': np.random.normal(140, 20, n).round(0),
    'Smoking': np.random.choice(['Never', 'Former', 'Current'], n, p=[0.4, 0.35, 0.25])
})

# Define variable types
columns = ['Age', 'Sex', 'BMI', 'Diabetes', 'eGFR', 'Systolic_BP', 'Smoking']
categorical = ['Sex', 'Diabetes', 'Smoking']
nonnormal = ['eGFR']  # Report as median [IQR]

# Create Table 1
table1 = TableOne(
    data,
    columns=columns,
    categorical=categorical,
    groupby='Treatment',
    nonnormal=nonnormal,
    pval=True,
    smd=True  # Standardised mean differences
)

print(table1.tabulate(tablefmt="github"))

# Export to various formats
# table1.to_excel('table1.xlsx')
# table1.to_latex('table1.tex')
# table1.to_html('table1.html')
```

**What the code shows.** This is the Python equivalent using the `tableone` package. After simulating a clinical dataset, we tell `TableOne` which variables are categorical (so they are reported as n (%)) and which non-normal continuous variables should be shown as median [IQR] rather than mean (SD). Setting `groupby='Treatment'` splits the table by treatment arm, and `smd=True` adds standardised mean differences --- the preferred way to judge baseline balance in observational studies, because (unlike p-values) an SMD does not shrink just because the sample is large. The printed table mirrors the R output: characteristics down the rows, treatment groups across the columns, with the final SMD column flagging any variable that is meaningfully imbalanced (a common rule of thumb is that an SMD above 0.1 is worth noting). The commented `to_excel`/`to_latex`/`to_html` lines show how to export it straight into a manuscript or supplement.

## Regression Tables

### Presenting regression results

A well-formatted regression table should include:

- Variable names (not raw column names)
- Effect estimates (OR, HR, RR, or coefficients) with 95% CIs
- P-values (to appropriate decimal places)
- Sample size and number of events
- Model fit statistics (R-squared, C-statistic, AIC)

### Regression tables in R

```{r}
#| label: regression-table-r
#| eval: false

library(gtsummary)
library(survival)

# Fit a Cox model
cox_model <- coxph(Surv(time, status) ~ age + sex + ph.ecog + ph.karno + wt.loss,
                   data = lung)

# Publication-quality table
tbl_regression(
  cox_model,
  exponentiate = TRUE,
  label = list(
    age ~ "Age, per year",
    sex ~ "Sex (female vs male)",
    ph.ecog ~ "ECOG performance status",
    ph.karno ~ "Karnofsky performance score",
    wt.loss ~ "Weight loss, kg"
  ),
  pvalue_fun = ~ style_pvalue(.x, digits = 3)
) |>
  add_global_p() |>
  bold_p(t = 0.05) |>
  modify_header(label = "**Variable**") |>
  modify_footnote(
    estimate ~ "HR = hazard ratio; CI = confidence interval",
  ) |>
  as_gt() |>
  gt::tab_header(
    title = "Table 2. Multivariable Cox Regression for Overall Survival"
  )
```

**What the code shows.** This turns a fitted Cox proportional-hazards model into a polished "Table 2". We first fit the model with `coxph`, then `tbl_regression(exponentiate = TRUE)` exponentiates the coefficients so they appear as hazard ratios (HRs) rather than raw log-hazards --- the scale clinicians actually interpret. Each row is a predictor with its HR, 95% confidence interval, and p-value; `add_global_p` adds an overall p-value for multi-level variables, and `bold_p` highlights the statistically significant ones. To read it: an HR above 1 means higher hazard (worse survival) per unit increase in that predictor, an HR below 1 means lower hazard, and an HR of 1 means no association. For example, an HR of 1.02 for "Age, per year" would mean each additional year of age is associated with a 2% higher hazard of death, holding the other covariates fixed. The footnote spelling out the abbreviations is the kind of detail reviewers expect.

### Regression tables in Python

```{python}
#| label: regression-table-python
#| eval: false

import pandas as pd
import numpy as np
from lifelines import CoxPHFitter
from lifelines.datasets import load_lung

# Load and prepare data
lung = load_lung()
lung = lung[['T', 'status', 'age', 'sex', 'ph.ecog', 'ph.karno', 'wt.loss']].dropna()
lung['status'] = lung['status'] - 1  # Convert to 0/1

# Fit Cox model
cph = CoxPHFitter()
cph.fit(lung, duration_col='T', event_col='status')

# Extract results as a formatted table
summary_df = cph.summary[['exp(coef)', 'exp(coef) lower 95%',
                           'exp(coef) upper 95%', 'p']].copy()
summary_df.columns = ['HR', 'Lower 95% CI', 'Upper 95% CI', 'p-value']
summary_df = summary_df.round({'HR': 2, 'Lower 95% CI': 2,
                                'Upper 95% CI': 2, 'p-value': 4})

# Create formatted CI column
summary_df['HR (95% CI)'] = summary_df.apply(
    lambda r: f"{r['HR']:.2f} ({r['Lower 95% CI']:.2f}-{r['Upper 95% CI']:.2f})",
    axis=1
)

# Rename index for publication
name_map = {
    'age': 'Age, per year',
    'sex': 'Sex (female vs male)',
    'ph.ecog': 'ECOG performance status',
    'ph.karno': 'Karnofsky score, per 10 points',
    'wt.loss': 'Weight loss, per kg'
}
summary_df.index = summary_df.index.map(lambda x: name_map.get(x, x))

print(summary_df[['HR (95% CI)', 'p-value']].to_string())
```

**What the code shows.** This produces the same hazard-ratio table in Python using `lifelines`. After fitting the Cox model with `CoxPHFitter`, we pull the relevant columns out of `cph.summary` --- the exponentiated coefficient (the HR) and its lower and upper 95% confidence limits --- round them, and stitch them into a single tidy "HR (95% CI)" string per predictor. The `name_map` step renames the raw variable names to clinician-friendly labels, exactly as `gtsummary` does in R. The printed result is a compact two-column table (HR with CI, and p-value) ready to drop into a manuscript, and it is interpreted identically: HRs above 1 indicate higher hazard, below 1 lower hazard.

## Publication-Quality Figures

### Journal specifications

Most journals require:

- **Resolution**: minimum 300 DPI (600 DPI for line art)
- **Format**: TIFF, EPS, or high-resolution PDF
- **Size**: single column (89 mm) or double column (183 mm)
- **Fonts**: Arial, Helvetica, or Times New Roman, minimum 8pt
- **Colours**: Accessible to colour-blind readers (avoid red-green combinations)

### Creating journal-quality figures in R

```{r}
#| label: pub-figures-r
#| eval: false

library(ggplot2)
library(survival)
library(survminer)
library(RColorBrewer)

# Define a publication theme
theme_publication <- function(base_size = 11) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      text = element_text(family = "Arial"),
      plot.title = element_text(size = base_size + 2, face = "bold",
                                 hjust = 0, margin = margin(b = 10)),
      axis.title = element_text(size = base_size, face = "bold"),
      axis.text = element_text(size = base_size - 1, colour = "black"),
      legend.title = element_text(size = base_size, face = "bold"),
      legend.text = element_text(size = base_size - 1),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey90"),
      strip.text = element_text(size = base_size, face = "bold"),
      plot.margin = margin(10, 10, 10, 10)
    )
}

# Colour-blind friendly palette
cb_palette <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7",
                "#F0E442", "#56B4E9", "#E69F00", "#000000")

# --- Kaplan-Meier survival curve ---
fit <- survfit(Surv(time, status) ~ sex, data = lung)

p_surv <- ggsurvplot(
  fit,
  data = lung,
  palette = cb_palette[1:2],
  risk.table = TRUE,
  risk.table.col = "strata",
  pval = TRUE,
  conf.int = TRUE,
  xlab = "Time (days)",
  ylab = "Survival probability",
  legend.labs = c("Male", "Female"),
  legend.title = "Sex",
  font.main = c(14, "bold"),
  font.x = c(12, "bold"),
  font.y = c(12, "bold"),
  font.tickslab = 11,
  risk.table.fontsize = 3.5,
  ggtheme = theme_publication()
)

print(p_surv)

# --- Saving at journal quality ---
# For TIFF (most journals)
ggsave("figure1_survival.tiff",
       plot = p_surv$plot,
       width = 183, height = 120,
       units = "mm", dpi = 300,
       compression = "lzw")

# For PDF (vector format, scalable)
ggsave("figure1_survival.pdf",
       plot = p_surv$plot,
       width = 183, height = 120,
       units = "mm", device = cairo_pdf)
```

**What the code shows.** This produces a journal-ready Kaplan-Meier survival figure and saves it at the resolution and dimensions a journal expects. The first part defines a reusable `theme_publication()` so every figure in the paper shares consistent fonts, sizes, and gridlines, and sets up a colour-blind-friendly palette (avoiding red-green combinations that some readers cannot distinguish). `ggsurvplot` then draws the survival curves by sex, with a risk table beneath showing how many patients remain at risk over time, a log-rank p-value, and shaded confidence bands. The two `ggsave` calls export the same figure twice: a TIFF at 300 DPI (the raster format most journals require) and a vector PDF (which stays sharp at any size). The point is that the *figure content* and the *export settings* are equally important --- a beautiful plot saved at screen resolution will be rejected at the production stage.

### Publication figures in Python

```{python}
#| label: pub-figures-python
#| eval: false

import matplotlib.pyplot as plt
import matplotlib as mpl
import numpy as np
from lifelines import KaplanMeierFitter
from lifelines.datasets import load_lung

# Set publication defaults
plt.rcParams.update({
    'font.family': 'Arial',
    'font.size': 11,
    'axes.labelsize': 12,
    'axes.titlesize': 13,
    'axes.labelweight': 'bold',
    'figure.dpi': 300,
    'savefig.dpi': 300,
    'figure.figsize': (7.2, 4.7),  # ~183mm x 120mm
    'axes.spines.top': False,
    'axes.spines.right': False,
})

# Colour-blind friendly palette
cb_palette = ['#0072B2', '#D55E00', '#009E73', '#CC79A7']

# Prepare data
lung = load_lung()
lung['status'] = lung['status'] - 1

fig, ax = plt.subplots()

kmf = KaplanMeierFitter()

for i, (sex_val, label) in enumerate([(1, 'Male'), (2, 'Female')]):
    mask = lung['sex'] == sex_val
    kmf.fit(lung.loc[mask, 'T'], lung.loc[mask, 'status'], label=label)
    kmf.plot_survival_function(ax=ax, color=cb_palette[i], ci_show=True)

ax.set_xlabel('Time (days)')
ax.set_ylabel('Survival Probability')
ax.set_title('Kaplan-Meier Survival by Sex')
ax.legend(loc='lower left', frameon=False)
ax.set_ylim(0, 1.05)

plt.tight_layout()
plt.savefig('figure1_survival.tiff', dpi=300, bbox_inches='tight')
plt.savefig('figure1_survival.pdf', bbox_inches='tight')
plt.show()
```

**What the code shows.** This is the Python counterpart, producing the same Kaplan-Meier figure with `matplotlib` and `lifelines`. The `plt.rcParams.update(...)` block sets publication defaults once --- Arial font, bold axis labels, 300 DPI, the correct double-column width (~183 mm), and clean spines --- so they apply to the whole figure without repeating yourself. We then loop over the two sexes, fitting and plotting a survival curve for each in a colour-blind-safe colour with confidence bands. The result is a survival plot with time on the x-axis and survival probability (0 to 1) on the y-axis: curves that drop more steeply indicate worse survival, and the gap between the two curves shows the survival difference between groups. As in R, it is saved both as a 300-DPI TIFF and as a scalable PDF.

### Multi-panel figures

Journals often prefer combined multi-panel figures (Figure 1A, 1B, etc.) rather than separate images.

```{r}
#| label: multipanel-r
#| eval: false

library(patchwork)

p1 <- ggplot(lung, aes(x = factor(sex), y = age)) +
  geom_boxplot(fill = cb_palette[1:2], alpha = 0.7) +
  labs(x = "Sex", y = "Age (years)", title = "A") +
  scale_x_discrete(labels = c("Male", "Female")) +
  theme_publication()

p2 <- ggplot(lung, aes(x = age, y = ph.karno)) +
  geom_point(alpha = 0.5, colour = cb_palette[1]) +
  geom_smooth(method = "lm", colour = cb_palette[2], se = TRUE) +
  labs(x = "Age (years)", y = "Karnofsky Score", title = "B") +
  theme_publication()

combined <- p1 + p2 +
  plot_layout(ncol = 2, widths = c(1, 1))

ggsave("figure2_panels.tiff",
       plot = combined,
       width = 183, height = 90,
       units = "mm", dpi = 300)
```

**What the code shows.** Journals usually want related plots combined into one labelled figure (Figure 1A, 1B, ...) rather than several separate files, and the `patchwork` package makes this trivial in R. Here we build two `ggplot` panels --- a boxplot of age by sex (panel A) and a scatter plot of Karnofsky score against age with a fitted regression line (panel B) --- giving each a `title` of "A" or "B". The expression `p1 + p2` simply places them side by side, and `plot_layout` controls the arrangement. The single exported TIFF then contains both panels at journal width, ready to reference as "Figure 2A" and "Figure 2B" in the text. The lesson is mechanical but important: assembling panels in code (rather than pasting images together in other software) keeps the figure reproducible and correctly sized.

## Writing a Statistical Methods Section

### Structure

A complete statistical methods section should address:

1. **Study design and population**: Brief recap (detailed in Methods)
2. **Primary and secondary outcomes**: How each was defined and measured
3. **Sample size / power calculation**: How you determined the required sample size
4. **Descriptive statistics**: How continuous and categorical variables are summarised
5. **Primary analysis**: The statistical model, its assumptions, and how they were checked
6. **Missing data**: How much was missing, the assumed mechanism (MCAR/MAR/MNAR), and the handling method
7. **Sensitivity analyses**: Pre-specified alternative approaches
8. **Multiple comparisons**: Whether and how adjusted (Bonferroni, FDR, etc.)
9. **Software**: Packages, versions, and random seeds

### Example statistical methods paragraph

> Baseline characteristics were summarised as means (SD) for normally distributed continuous variables, medians (IQR) for skewed continuous variables, and frequencies (percentages) for categorical variables. The primary outcome (time to first major adverse cardiovascular event) was analysed using multivariable Cox proportional hazards regression, adjusting for age, sex, diabetes, eGFR, systolic blood pressure, and smoking status. The proportional hazards assumption was tested using Schoenfeld residuals. Missing covariate data (3.2% overall, assumed missing at random) were handled using multiple imputation by chained equations (50 imputed datasets, 20 iterations). Rubin's rules were used to pool effect estimates. Model discrimination was assessed using Harrell's C-statistic with 95% confidence intervals estimated via 500 bootstrap resamples. Calibration was assessed graphically and using the Greenwood-Nam-D'Agostino test. Internal-external cross-validation was performed across the three recruitment centres. All analyses were pre-specified and conducted using R version 4.4.1 (R Foundation for Statistical Computing) with the survival (v3.7), mice (v3.16), and rms (v6.8) packages. Two-sided p-values < 0.05 were considered statistically significant.

::: {.callout-warning}
## Common mistakes in statistical methods sections
1. Not specifying the handling of missing data
2. Claiming "data were normally distributed" without stating how this was assessed
3. Using stepwise variable selection without acknowledging its limitations
4. Reporting only p-values without effect estimates and confidence intervals
5. Not specifying the software and package versions
6. Not describing how the proportional hazards or linearity assumptions were checked
7. Claiming causation from an observational study without using causal inference methods
:::

## TRIPOD+AI Compliance

The **TRIPOD+AI** statement (Collins et al., 2024) extends the original TRIPOD guideline for prediction models to include AI-specific items. If your study develops, validates, or updates a clinical prediction model --- whether based on logistic regression or deep learning --- you should follow TRIPOD+AI.

### Key TRIPOD+AI items

| Item | Requirement | How to address |
|---|---|---|
| Title | State it is a prediction model study and the type (development, validation, both) | "Development and external validation of a machine learning model for..." |
| Source of data | Describe the data source, setting, dates | EHR data from Hospital X, 2015-2023 |
| Participants | Eligibility criteria, sampling, flow diagram | Include a CONSORT-style flow diagram |
| Outcome | Clear definition, timing, blinding | "30-day all-cause mortality ascertained from hospital records" |
| Sample size | Justify based on Riley et al. criteria | Use `pmsampsize` package |
| Missing data | Report extent and handling | Multiple imputation; report complete case sensitivity analysis |
| Model development | Full specification including hyperparameters | Report all hyperparameters; include code supplement |
| Model performance | Discrimination (C-statistic, AUROC) AND calibration (calibration plot, slope, intercept) | Never report AUROC alone --- calibration is equally important |
| Model fairness | Assess performance across demographic subgroups | Report C-statistic and calibration by sex, age group, ethnicity |
| Code and data | Availability statement | Share code via GitHub; data via controlled access repository |

### Step-by-step TRIPOD+AI walkthrough

```{r}
#| label: tripod-checklist
#| eval: false

# Step 1: Sample size calculation for a prediction model
library(pmsampsize)

# Binary outcome model
# Assume: 15 candidate predictors, outcome prevalence 10%,
# target C-statistic 0.80, max shrinkage 10%
ss <- pmsampsize(type = "b",
                 rsquared = 0.15,  # Cox-Snell R-squared
                 parameters = 15,
                 prevalence = 0.10)
ss

# Step 2: Develop model with full pre-specification
library(rms)

# Set up data distribution summaries
dd <- datadist(your_data)
options(datadist = "dd")

# Fit model
full_model <- lrm(outcome ~ age + sex + rcs(sbp, 4) + diabetes + egfr +
                     smoking + rcs(bmi, 3),
                   data = your_data, x = TRUE, y = TRUE)

# Step 3: Internal validation with bootstrap
val <- validate(full_model, B = 200)
print(val)

# Optimism-corrected C-statistic
c_stat_apparent <- full_model$stats["C"]
c_stat_corrected <- c_stat_apparent - (val["Dxy", "index.orig"] -
                                        val["Dxy", "index.corrected"]) / 2

# Step 4: Calibration plot
cal <- calibrate(full_model, B = 200)
plot(cal, xlab = "Predicted probability", ylab = "Observed proportion",
     main = "Calibration Plot (Bootstrap-Corrected)")

# Step 5: Fairness assessment
# Evaluate performance in subgroups
for (subgroup in c("Male", "Female")) {
  sub_data <- your_data |> filter(sex == subgroup)
  pred <- predict(full_model, newdata = sub_data, type = "fitted")
  auc <- pROC::auc(sub_data$outcome, pred)
  cat(subgroup, ": C-statistic =", round(auc, 3), "\n")
}
```

**What the code shows.** This walks through the core statistical steps a TRIPOD+AI-compliant prediction-model paper must report, in order. Step 1 uses `pmsampsize` to calculate how many patients (and events) you need before you start --- guarding against the common mistake of developing a model on too little data. Step 2 fits the model with `rms::lrm`, using restricted cubic splines (`rcs`) so continuous predictors like blood pressure are not forced to be perfectly linear. Step 3's `validate(..., B = 200)` runs bootstrap internal validation, which estimates *optimism* --- how much the model's apparent performance is inflated by being tested on the same data it was trained on --- and the code then subtracts that optimism to give an honest, corrected C-statistic (discrimination). Step 4's `calibrate` produces a calibration plot, checking whether predicted risks match observed risks (a model can discriminate well yet be poorly calibrated). Step 5 loops over sex subgroups to report performance separately --- the fairness check TRIPOD+AI now requires. The key message reinforced throughout: never report discrimination (the C-statistic / AUROC) without also reporting calibration.

## Reproducible Research Workflow

### Project structure

A reproducible research project should follow a consistent directory structure:

```
project/
  |-- README.md               # Project overview
  |-- renv.lock                # R package versions (or environment.yml for conda)
  |-- _targets.R               # Pipeline definition (or Makefile)
  |-- data/
  |     |-- raw/               # Never modify raw data
  |     |-- processed/         # Cleaned datasets
  |     |-- metadata/          # Data dictionaries
  |-- R/                       # R functions
  |-- python/                  # Python scripts
  |-- analysis/
  |     |-- 01_data_cleaning.qmd
  |     |-- 02_descriptive.qmd
  |     |-- 03_primary_analysis.qmd
  |     |-- 04_sensitivity.qmd
  |-- output/
  |     |-- figures/
  |     |-- tables/
  |-- manuscript/
  |     |-- manuscript.qmd     # The paper itself
  |     |-- references.bib
  |-- Dockerfile               # For full computational reproducibility
```

### Package management

**R (renv)**:

```{r}
#| label: renv-setup
#| eval: false

# Initialise renv in your project
renv::init()

# After installing packages, take a snapshot
renv::snapshot()

# Collaborator restores exact versions
renv::restore()
```

**Python (conda/pip)**:

```{python}
#| label: conda-setup
#| eval: false

# Create environment
# conda create -n my_project python=3.11
# conda activate my_project
# pip install pandas numpy scikit-learn lifelines tableone

# Export environment
# pip freeze > requirements.txt
# OR
# conda env export > environment.yml

# Collaborator recreates environment
# pip install -r requirements.txt
# OR
# conda env create -f environment.yml
```

### Version control with Git

Essential git practices for research:

1. **Commit analysis scripts, not data** (unless data are small and non-sensitive)
2. **Use `.gitignore`** to exclude data files, outputs, and sensitive information
3. **Write meaningful commit messages**: "Add multiple imputation for missing eGFR values" not "update script"
4. **Tag milestones**: `git tag -a v1.0-submission -m "Initial journal submission"`
5. **Use branches** for exploratory analyses that may not make it into the paper

### Docker for full reproducibility (brief)

A Dockerfile captures the complete computational environment:

```dockerfile
FROM rocker/verse:4.4.1

# Install system dependencies
RUN apt-get update && apt-get install -y libglpk-dev

# Install R packages
RUN R -e "install.packages(c('survival', 'rms', 'mice', 'gtsummary', 'pmsampsize'))"

# Copy project files
COPY . /home/rstudio/project

WORKDIR /home/rstudio/project
```

This guarantees that anyone, anywhere, at any time can reproduce your exact results.

## Case Study: Replicating a Lancet Digital Health Analytical Approach

To illustrate the complete workflow, we replicate the analytical structure of a 2025 Lancet Digital Health prediction model study.

**Study design**: Development and external validation of a machine learning model to predict 30-day hospital readmission in heart failure patients.

```{r}
#| label: case-study
#| eval: false

# --- STEP 1: Data preparation ---
library(tidyverse)
library(mice)

# Load data (using a public dataset as proxy)
# In practice, this would be your EHR extract
data("heart", package = "survival")  # Stanford heart transplant data

# Document missingness
missing_summary <- data.frame(
  variable = names(heart),
  n_missing = sapply(heart, function(x) sum(is.na(x))),
  pct_missing = sapply(heart, function(x) mean(is.na(x)) * 100)
)
print(missing_summary)

# Multiple imputation
imp <- mice(heart, m = 50, maxit = 20, method = "pmm", seed = 42,
            printFlag = FALSE)

# --- STEP 2: Table 1 ---
library(gtsummary)

heart |>
  tbl_summary(
    by = transplant,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    )
  ) |>
  add_overall() |>
  add_p()

# --- STEP 3: Model development (in imputed data) ---
library(rms)
library(pROC)

# Pool results across imputations using Rubin's rules
# (Simplified example with complete cases for illustration)
dd <- datadist(heart)
options(datadist = "dd")

# Pre-specify model based on clinical knowledge
model <- lrm(event ~ age + year + surgery + rcs(futime, 4),
             data = heart, x = TRUE, y = TRUE)

print(model)

# --- STEP 4: Internal validation ---
set.seed(42)
val <- validate(model, B = 500)
print(val)

# Optimism-corrected C-statistic
cat("Apparent C-statistic:", model$stats["C"], "\n")
cat("Optimism:", (val["Dxy", "index.orig"] - val["Dxy", "index.corrected"])/2, "\n")

# --- STEP 5: Calibration plot ---
cal <- calibrate(model, B = 200)
plot(cal,
     xlab = "Predicted Probability",
     ylab = "Observed Proportion",
     main = "Calibration Plot (Internal Bootstrap Validation)")
abline(0, 1, lty = 2, col = "grey50")

# --- STEP 6: Decision curve analysis ---
library(dcurves)

dca_result <- dca(event ~ predicted_prob,
                  data = heart_with_preds,
                  thresholds = seq(0, 0.5, by = 0.01))
plot(dca_result)

# --- STEP 7: Export all results ---
# Tables to Word/HTML
tbl_summary_result |> as_gt() |> gt::gtsave("output/tables/table1.docx")

# Figures at publication quality
ggsave("output/figures/figure1_calibration.tiff",
       width = 89, height = 89, units = "mm", dpi = 300)
```

**What the code shows.** This is the chapter's capstone example, stitching every earlier skill into one end-to-end pipeline that mirrors how a real prediction-model paper is built. It moves through seven numbered steps: (1) load the data and document missingness, then handle it with multiple imputation (`mice`) rather than silently dropping incomplete records; (2) generate the Table 1 of baseline characteristics by group; (3) develop a pre-specified logistic model with `rms::lrm`, using splines for non-linear effects; (4) run bootstrap internal validation to obtain an optimism-corrected C-statistic; (5) draw a calibration plot with the diagonal reference line, where points hugging the line mean predicted and observed risks agree; (6) perform a decision curve analysis with `dcurves`, which translates statistical performance into net clinical benefit across the range of risk thresholds a clinician might act on; and (7) export the finished tables and figures at journal quality. Read as a whole, this is the template you would adapt for your own study --- the deliberate sequence (impute, describe, develop, validate, calibrate, assess clinical usefulness, export) is itself the lesson.

## Capstone Projects

Choose one of the following three capstone projects. Each requires you to produce a complete, journal-ready analysis with Table 1, regression/model results, publication-quality figures, and a statistical methods section.

### Capstone 1: Cardiovascular Risk Prediction Model

**Objective**: Develop a cardiovascular risk prediction model using Framingham Heart Study methodology and validate it in NHANES data.

**Dataset**: Framingham Heart Study (public teaching dataset from `riskCommunicator` R package) for development; NHANES 2017-2020 (via `nhanesA` package) for external validation.

**Requirements**:

1. **Table 1**: Baseline characteristics of the development cohort, stratified by 10-year CVD event status.
2. **Model development**: Logistic regression predicting 10-year CVD risk using age, sex, total cholesterol, HDL cholesterol, systolic blood pressure, blood pressure treatment, smoking, and diabetes.
3. **Internal validation**: Bootstrap-corrected C-statistic and calibration slope.
4. **External validation in NHANES**: Apply the model, report discrimination and calibration in the external population.
5. **Fairness assessment**: Report model performance stratified by sex and race/ethnicity.
6. **Figures**: (A) Calibration plot (development, bootstrap-corrected), (B) Calibration plot (NHANES external validation), (C) Decision curve analysis.
7. **Statistical methods section**: Following TRIPOD+AI structure.
8. **Code**: Fully reproducible R or Python scripts with renv/conda environment.

### Capstone 2: Metabolic Phenotype Clustering

**Objective**: Identify metabolic phenotypes in the US adult population using unsupervised learning and assess their association with mortality.

**Dataset**: NHANES 2017-2020 (demographics, laboratory, examination, and mortality linkage files).

**Requirements**:

1. **Data preparation**: Select metabolic variables (glucose, HbA1c, insulin, triglycerides, HDL, LDL, waist circumference, BMI, blood pressure, CRP). Handle missing data with multiple imputation.
2. **Dimensionality reduction**: PCA to identify major axes of metabolic variation. Present scree plot and loadings table.
3. **Clustering**: Apply k-means and Gaussian mixture models to PC scores. Use the gap statistic, silhouette width, and BIC to select the number of clusters. Visualise clusters in PC space.
4. **Table 1**: Baseline characteristics stratified by cluster assignment.
5. **Cluster characterisation**: Describe the clinical profile of each cluster (the "metabolic phenotypes").
6. **Survival analysis**: Kaplan-Meier curves and Cox regression for all-cause mortality by cluster, adjusting for age, sex, race/ethnicity, smoking, and education.
7. **Figures**: (A) Scree plot, (B) Cluster visualisation in PC1-PC2 space, (C) Kaplan-Meier curves by cluster, (D) Forest plot of hazard ratios.
8. **Statistical methods section**: Including justification for unsupervised approach and cluster validation.

### Capstone 3: Bayesian Survival Analysis

**Objective**: Fit a Bayesian hierarchical survival model to the primary biliary cholangitis (PBC) dataset, comparing Bayesian and frequentist approaches.

**Dataset**: `pbc` dataset from the `survival` R package (Mayo Clinic PBC trial, 418 patients).

**Requirements**:

1. **Table 1**: Baseline characteristics stratified by treatment arm (D-penicillamine vs placebo).
2. **Frequentist Cox model**: Standard Cox PH regression with selected predictors. Check proportional hazards assumption.
3. **Bayesian Cox model**: Fit using `brms` (R) or `PyMC` (Python) with:
   - Weakly informative priors on log-hazard ratios: $\beta \sim N(0, 1)$
   - Compare with informative priors based on prior literature
   - Report posterior medians, 95% credible intervals, and posterior probability of benefit
4. **Hierarchical extension**: If the data include centre/site information, fit a model with random intercepts. Otherwise, use histological stage as a grouping variable.
5. **Model comparison**: Compare DIC/WAIC/LOO-CV between Bayesian models.
6. **Posterior predictive checks**: Simulate survival curves from the posterior and overlay on observed Kaplan-Meier.
7. **Figures**: (A) Prior vs posterior distributions for key parameters, (B) Posterior predictive survival curves, (C) Forest plot comparing Bayesian vs frequentist estimates.
8. **Statistical methods section**: Justify the Bayesian approach, describe priors and sensitivity to prior choice.

::: {.callout-tip}
## Capstone assessment criteria
Each capstone will be assessed on:

- **Statistical correctness** (40%): Appropriate methods, valid assumptions, correct interpretation
- **Reproducibility** (20%): Code runs from start to finish; environment documented
- **Presentation quality** (20%): Journal-ready tables, figures, and methods section
- **Critical thinking** (20%): Discussion of limitations, sensitivity analyses, clinical implications
:::

## Exercises

### Exercise 1: Table 1 in both R and Python

Using the `lung` dataset (R) or an equivalent simulated dataset (Python):

a) Create a Table 1 stratified by sex, with appropriate summary statistics for age, ECOG score, Karnofsky score, meal calories, and weight loss.
b) Include standardised mean differences instead of p-values.
c) Export the table to a format suitable for journal submission (Word or LaTeX).

**R**:
```{r}
#| label: ex1-r
#| eval: false

data(lung, package = "survival")
# Your code here using gtsummary
```

**Python**:
```{python}
#| label: ex1-python
#| eval: false

from lifelines.datasets import load_lung
lung = load_lung()
# Your code here using tableone
```

### Exercise 2: Publication-quality figure

Create a multi-panel figure (at least 3 panels) summarising a survival analysis of the `lung` dataset:

- Panel A: Kaplan-Meier curve by sex
- Panel B: Forest plot of hazard ratios from a multivariable Cox model
- Panel C: Calibration plot for a 1-year survival prediction

Requirements: colour-blind accessible palette, Arial font, 300 DPI, single figure file suitable for a double-column journal layout (183 mm wide).

### Exercise 3: Write a statistical methods section

Write a complete statistical methods section (250-400 words) for the following study:

> A retrospective cohort study of 5,000 patients with type 2 diabetes from a hospital EHR database (2015-2023). The primary outcome is time to first major adverse cardiovascular event (MACE). The exposure is SGLT2 inhibitor use (vs DPP-4 inhibitor use). Covariates include age, sex, BMI, HbA1c, eGFR, prior CVD, hypertension, and smoking. Missing data range from 2% (age) to 15% (BMI). The analysis uses propensity score matching followed by Cox regression in the matched cohort.

Your methods section should cover all items listed in the "Writing a statistical methods section" subsection above.

### Exercise 4: TRIPOD+AI checklist

Download the TRIPOD+AI checklist (from the EQUATOR Network website). For one of the three capstone projects above, complete the checklist, indicating:

- The page/section number where each item is addressed
- Items that are not applicable and why
- Items that you cannot fully address and what you would need

## Summary

| Component | Tool/Package | Key considerations |
|---|---|---|
| Table 1 | gtsummary (R), tableone (Python) | SMDs over p-values in observational studies |
| Regression tables | gtsummary, broom, gt (R) | Always include CIs; exponentiate where appropriate |
| Figures | ggplot2 + patchwork (R), matplotlib (Python) | 300+ DPI, colour-blind safe, journal size specs |
| Methods section | - | Missing data, assumptions, software versions |
| Reporting guideline | TRIPOD+AI, STROBE, CONSORT | Choose based on study design |
| Reproducibility | renv (R), conda (Python), Git, Docker | Capture full computational environment |
| Calibration | rms, probably (R), scikit-learn (Python) | Never report discrimination without calibration |

## References and Further Reading

1. **Collins GS, Moons KGM, Dhiman P, et al.** TRIPOD+AI statement: updated guidance for reporting clinical prediction models that use regression or machine learning methods. *BMJ.* 2024;385:e078378.

2. **von Elm E, Altman DG, Egger M, et al.** The Strengthening the Reporting of Observational Studies in Epidemiology (STROBE) statement. *Lancet.* 2007;370(9596):1453--1457.

3. **Sjoberg DD, Whiting K, Curry M, Lavery JA, Larmarange J.** Reproducible summary tables with the gtsummary package. *The R Journal.* 2021;13(1):570--580.

4. **Riley RD, Ensor J, Snell KIE, et al.** Calculating the sample size required for developing a clinical prediction model. *BMJ.* 2020;368:m441.

5. **Steyerberg EW.** *Clinical Prediction Models: A Practical Approach to Development, Validation, and Updating.* 2nd ed. Springer, 2019.

6. **Van Calster B, McLernon DJ, van Smeden M, et al.** Calibration: the Achilles heel of predictive analytics. *BMC Medicine.* 2019;17:230.

7. **Vickers AJ, Elkin EB.** Decision curve analysis: a novel method for evaluating prediction models. *Medical Decision Making.* 2006;26(6):565--574.

8. **Hernan MA, Robins JM.** *Causal Inference: What If.* Chapman & Hall/CRC, 2024.

9. **Harrell FE.** *Regression Modeling Strategies.* 2nd ed. Springer, 2015.

10. **The Lancet Digital Health.** Author guidelines. Available at [https://www.thelancet.com/digital-health](https://www.thelancet.com/digital-health).

11. **EQUATOR Network.** Reporting guidelines for health research. Available at [https://www.equator-network.org](https://www.equator-network.org).

12. **Wickham H.** *ggplot2: Elegant Graphics for Data Analysis.* 3rd ed. Springer, 2024.

13. **Wilson G, Bryan J, Cranston K, et al.** Good enough practices in scientific computing. *PLoS Computational Biology.* 2017;13(6):e1005510.
