# Decision Trees, Random Forests, and Gradient Boosting {#sec-trees-ensembles}

```{r}
#| label: setup-trees
#| include: false
library(tidyverse)
library(tidymodels)
library(rpart)
library(rpart.plot)
library(ranger)
library(xgboost)
library(vip)
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Of all the machine learning methods in this book, the tree-based family is the one clinicians tend to adopt fastest --- because it mirrors how clinical reasoning already works. A doctor triaging chest pain mentally runs through a series of questions ("Is the troponin elevated? Are there ECG changes? Is the patient over 65?"), and each answer narrows down the likely diagnosis. That branching logic *is* a decision tree. This chapter starts from that intuition and builds up to the methods that win most clinical prediction competitions on tabular data: random forests and gradient boosting, which combine hundreds of trees into a single, far more accurate model.

Why invest in these methods rather than stopping at logistic regression? Because they automatically capture two things that are pervasive in medicine and awkward for standard regression: **non-linear effects** (risk that rises sharply only past a threshold) and **interactions** (a lab value that matters only in diabetic patients). You do not have to specify these by hand --- the trees discover them. The price is interpretability, which is why the chapter spends time on how to read a tree and how to inspect variable importance.

## Decision Trees (CART)

A **decision tree** is perhaps the most intuitive machine learning model. It makes predictions by asking a series of yes/no questions about the input features, splitting the data at each step into increasingly homogeneous groups. The result looks like an inverted tree — hence the name.

**Why start with CART?** CART (Classification And Regression Trees) is the foundation for everything else in this chapter --- random forests and gradient boosting are both just clever ways of combining many CART trees. But a single tree is also valuable in its own right for clinical work: it produces a rule you can literally read off and apply at the bedside without a computer, and it handles a mix of continuous and categorical predictors, missing values, and interactions with no special preprocessing. When you need a model that a clinical team will actually trust and use, a small decision tree is often the most practical choice. @fig-cart-example shows the kind of transparent rule a tree produces.

```{mermaid}
%%| eval: true
%%| label: fig-cart-example
%%| fig-cap: "A simple, readable decision tree for 30-day readmission risk. To use it, start at the top and follow the answers to your patient's questions down to a leaf, which gives the predicted risk group. This is the kind of transparent rule that makes trees appealing at the bedside."
flowchart TD
    A["Prior admissions ≥ 2?"] -->|Yes| B["Has heart failure?"]
    A -->|No| C["Age ≥ 75?"]
    B -->|Yes| D["High risk<br/>~62% readmit"]
    B -->|No| E["Moderate risk<br/>~34% readmit"]
    C -->|Yes| F["Moderate risk<br/>~28% readmit"]
    C -->|No| G["Low risk<br/>~9% readmit"]
    style D fill:#e8736a,color:#fff
    style E fill:#f4c06a
    style F fill:#f4c06a
    style G fill:#7fc08a,color:#fff
```

### How Trees Split

The algorithm works top-down, **greedily** selecting the best split at each node. "Greedy" means it takes the single best step available right now without looking ahead --- like a clinician ordering the one test that best separates the likely diagnoses at this moment, then reassessing. Concretely:

1. **Consider every feature and every possible split point.** For a continuous variable like age, the tree tries every threshold ("age < 60?", "age < 61?", …); for a binary variable like *has diabetes*, there is only one possible split. So at the very first node the algorithm might evaluate thousands of candidate questions.
2. **Score each candidate split by how much it improves the "purity" of the two resulting groups.** A split is good if it sends mostly-readmitted patients down one branch and mostly-not-readmitted patients down the other. "Purity" is just a number measuring how one-sided a group is (defined precisely in the next section).
3. **Choose the single split that improves purity the most**, and use it to divide the patients into two child nodes.
4. **Repeat recursively** on each child node --- each becomes a new little problem with its own best question --- until a **stopping criterion** is met: the node is too small to split further (`minsplit`), the tree has reached its maximum allowed depth (`maxdepth`), or no remaining split improves purity enough to be worthwhile.

A worked picture: at the root, the algorithm might find that "prior admissions ≥ 2?" best separates readmitted from non-readmitted patients. Among the patients who answered "yes," it then searches again and might pick "has heart failure?"; among those who answered "no," it might instead pick "age ≥ 75?". Different branches can ask about different variables --- this is precisely how trees capture interactions automatically.

### Splitting Criteria

"Purity" needs to be turned into a number the computer can compare. **Impurity** is simply a score for *how mixed* a group of patients is: it is **0 when everyone in the group has the same outcome** (all readmitted, or all not) and **largest when the group is a 50/50 coin-flip**. A split is judged good when it produces child groups that are *purer* (lower impurity) than the parent. Two formulas are used to measure this; do not be put off by the notation --- each one is just a one-line recipe.

In the formulas below, imagine a group of patients in which a fraction $p_1$ have outcome 1 (readmitted), $p_2$ have outcome 2, and so on, across $K$ possible outcomes. For the binary readmission example there are just two: $p_{\text{yes}}$ and $p_{\text{no}} = 1 - p_{\text{yes}}$.

**Gini impurity** --- the chance two randomly drawn patients from the group have *different* outcomes:

$$G = \sum_{k=1}^{K} p_k (1 - p_k) = 1 - \sum_{k=1}^{K} p_k^2$$

Read it as: take each outcome's proportion, and you are summing how "spread out" the group is across outcomes. Gini is 0 when a node is pure (all one class) and maximised when classes are equally mixed. For a binary problem the maximum is 0.5, reached at $p = 0.5$.

*Worked example.* A node of 100 patients with 30 readmitted and 70 not has $p_{\text{yes}} = 0.30$, $p_{\text{no}} = 0.70$, so $G = 1 - (0.30^2 + 0.70^2) = 1 - (0.09 + 0.49) = 0.42$ --- fairly impure. If a split sends 40 patients to a child node that is 35 readmitted / 5 not ($p = 0.875$), that child has $G = 1 - (0.875^2 + 0.125^2) = 0.22$ --- much purer. The tree favours splits that drive these child Gini values down.

**Entropy (information gain)** --- an alternative impurity measure borrowed from information theory:

$$H = -\sum_{k=1}^{K} p_k \log_2(p_k)$$

It behaves almost identically to Gini: 0 when the node is pure, and maximised (at $\log_2 K$) when outcomes are evenly split. The reduction in entropy achieved by a split is called the **information gain**.

In practice, Gini and entropy produce very similar trees, so the choice rarely matters. Gini is the default in most implementations because it is slightly faster to compute (no logarithm). **You do not need to compute these by hand** --- the software does it for every candidate split. The reason to understand them is to know what the tree is optimising: at each step it is hunting for the question that most cleanly separates the outcomes.

For **regression trees** (continuous outcomes, e.g. predicting length of stay), there are no classes to count, so impurity is instead the **variance** within the node. A split is good when it produces child nodes whose values are tightly clustered --- equivalently, when it gives the largest **reduction in residual sum of squares**.

### Clinical Example: Hospital Readmission

Let us build a decision tree to predict 30-day hospital readmission — a common quality metric tied to reimbursement penalties in the United States.

::: {.panel-tabset}

#### R

```{r}
#| label: tree-data-r
# Simulate hospital readmission data
set.seed(42)
n <- 1000

readmit_data <- tibble(
  age = rnorm(n, 68, 12),
  length_of_stay = rpois(n, 5) + 1,
  num_comorbidities = rpois(n, 3),
  prior_admissions = rpois(n, 1),
  discharge_hemoglobin = rnorm(n, 11, 2),
  discharge_creatinine = rlnorm(n, 0.2, 0.5),
  has_diabetes = rbinom(n, 1, 0.35),
  has_chf = rbinom(n, 1, 0.25),
  readmitted = factor(
    rbinom(n, 1, plogis(-3 + 0.02 * (rnorm(n, 68, 12) - 68) +
                          0.15 * rpois(n, 1) +
                          0.1 * rpois(n, 3) +
                          0.3 * rbinom(n, 1, 0.25) -
                          0.1 * rnorm(n, 11, 2))),
    labels = c("No", "Yes")
  )
)

cat("Readmission rate:", mean(readmit_data$readmitted == "Yes"), "\n")
cat("N =", nrow(readmit_data), "\n")
```

```{r}
#| label: tree-fit-r
#| fig-cap: "A classification tree for 30-day readmission. Each node shows the predicted class, probability, and percentage of observations."
# Fit a decision tree
tree_model <- rpart(
  readmitted ~ age + length_of_stay + num_comorbidities +
    prior_admissions + discharge_hemoglobin +
    discharge_creatinine + has_diabetes + has_chf,
  data = readmit_data,
  method = "class",
  control = rpart.control(cp = 0.01, maxdepth = 5, minsplit = 20)
)

# Visualize
rpart.plot(tree_model, type = 4, extra = 106, under = TRUE,
           box.palette = "RdYlGn", roundint = FALSE,
           main = "Decision Tree: 30-Day Readmission")
```

#### Python

```{python}
#| label: tree-fit-python
import numpy as np
import pandas as pd
from sklearn.tree import DecisionTreeClassifier, plot_tree
from sklearn.model_selection import cross_val_score, StratifiedKFold
import matplotlib.pyplot as plt

np.random.seed(42)
n = 1000

df = pd.DataFrame({
    'age': np.random.normal(68, 12, n),
    'length_of_stay': np.random.poisson(5, n) + 1,
    'num_comorbidities': np.random.poisson(3, n),
    'prior_admissions': np.random.poisson(1, n),
    'discharge_hemoglobin': np.random.normal(11, 2, n),
    'discharge_creatinine': np.random.lognormal(0.2, 0.5, n),
    'has_diabetes': np.random.binomial(1, 0.35, n),
    'has_chf': np.random.binomial(1, 0.25, n)
})
y = np.random.binomial(1, 0.18, n)

feature_names = list(df.columns)

# Fit decision tree
tree = DecisionTreeClassifier(max_depth=5, min_samples_split=20,
                               min_samples_leaf=10, random_state=42)
tree.fit(df, y)

fig, ax = plt.subplots(figsize=(16, 8))
plot_tree(tree, feature_names=feature_names, class_names=['No', 'Yes'],
          filled=True, rounded=True, ax=ax, fontsize=8, max_depth=3)
ax.set_title("Decision Tree: 30-Day Readmission")
plt.tight_layout()
plt.show()
```

:::

### Reading a Decision Tree

Each node in the tree displays:

- The **splitting rule** (e.g., "age >= 72")
- The **predicted class** (the majority class in that node)
- The **class probabilities** (e.g. 0.34 readmitted / 0.66 not)
- The **percentage of observations** that reach this node

To classify a new patient, start at the root and follow the branches based on the patient's feature values until you reach a terminal node (leaf). The prediction is the majority class of that leaf, and the leaf's class probability is the patient's estimated risk.

*Worked example.* Suppose a 78-year-old patient has had no prior admissions and no heart failure. Starting at the root of @fig-cart-example, "prior admissions ≥ 2?" is **No**, so we go right to "age ≥ 75?", which is **Yes**, landing in the *moderate risk* leaf (~28% readmission risk). A different patient with three prior admissions and heart failure follows the left branches into the *high risk* leaf (~62%). Notice two clinically useful properties: the prediction is reached by at most a handful of questions, and you can explain *exactly why* any patient received their risk estimate --- a transparency that "black box" models cannot offer. The flip side is that a single tree's boundaries are coarse (everyone in a leaf gets the same risk) and can shift noticeably if the data change slightly, which is the motivation for the ensemble methods later in this chapter.

### Pruning: Controlling Tree Complexity

**Why pruning matters.** A tree left to grow without limits will keep asking questions until each leaf contains just one or two patients --- effectively building a private rule for every individual in the training set. It will look *perfect* on the data it was built from and then fail badly on new patients, because it has memorised noise (one patient's quirky lab value) instead of learning generalisable patterns. This is **overfitting**, and for a clinical tool it is dangerous: a model that boasts 99% accuracy in development but collapses in the next clinic is worse than no model, because people trusted it.

**Pruning** is the cure: it deliberately cuts the tree back to a smaller, simpler set of rules that capture the real signal and ignore the noise. A pruned tree usually performs slightly *worse* on the training data but *better* on new patients --- which is the only performance that matters. Think of it as the bias--variance tradeoff (from the previous chapter) made tangible: a few extra mistakes in development buy a large gain in reliability. The goal is the smallest tree whose performance is statistically indistinguishable from the bigger one.

Two strategies:

1. **Pre-pruning** (early stopping): Set constraints *before* growing the tree — maximum depth, minimum samples per leaf, minimum information gain to split. Simple but can stop too early.

2. **Post-pruning** (cost-complexity pruning): Grow a full tree, then prune branches that provide minimal improvement. The **complexity parameter (cp)** in R's `rpart` controls this: higher cp = more pruning. Select cp via cross-validation.

::: {.panel-tabset}

#### R

```{r}
#| label: prune-r
#| fig-cap: "Cross-validation error as a function of the complexity parameter. The optimal cp minimizes the cross-validated error."
# View the CP table
printcp(tree_model)

# Plot CV error vs cp
plotcp(tree_model)

# Prune to optimal cp
optimal_cp <- tree_model$cptable[which.min(tree_model$cptable[, "xerror"]), "CP"]
pruned_tree <- prune(tree_model, cp = optimal_cp)

cat("Optimal cp:", optimal_cp, "\n")
cat("Number of terminal nodes (full):", sum(tree_model$frame$var == "<leaf>"), "\n")
cat("Number of terminal nodes (pruned):", sum(pruned_tree$frame$var == "<leaf>"), "\n")
```

#### Python

```{python}
#| label: prune-python
from sklearn.tree import DecisionTreeClassifier
from sklearn.model_selection import cross_val_score
import numpy as np

# Compare trees of different depths
depths = [2, 3, 5, 7, 10, 15, None]
cv = StratifiedKFold(n_splits=5, shuffle=True, random_state=42)

results = []
for d in depths:
    t = DecisionTreeClassifier(max_depth=d, random_state=42)
    scores = cross_val_score(t, df, y, cv=cv, scoring='roc_auc')
    depth_label = str(d) if d is not None else "None (full)"
    results.append({'max_depth': depth_label,
                    'mean_auc': scores.mean(),
                    'std_auc': scores.std()})
    print(f"max_depth={depth_label:>6s}: AUC = {scores.mean():.3f} (+/- {scores.std():.3f})")

best = max(results, key=lambda x: x['mean_auc'])
print(f"\nBest: max_depth={best['max_depth']} with AUC={best['mean_auc']:.3f}")
```

:::

**What the code is showing.** Both versions answer the same question --- *how big should the tree be?* --- by measuring performance on held-out data rather than the training data:

- The **R** version uses cost-complexity pruning. `printcp()` prints a table where each row is a candidate tree size, and the key column is `xerror` (the cross-validated error). As the tree grows, `xerror` first falls (the tree is learning real structure) and then rises again (it has started memorising noise). `plotcp()` shows this as a U-shape. The code picks the complexity parameter `cp` at the bottom of the U and prunes to it; the final two lines print how many leaves the full tree had versus the smaller pruned tree --- you should see the pruned tree is considerably smaller.
- The **Python** version makes the same point more directly by sweeping `max_depth` from 2 up to "no limit" and printing the cross-validated AUC for each. Reading down the list, AUC climbs, peaks at a modest depth, then stops improving (or worsens) as the tree gets deeper --- and the script reports the depth that won. The deepest tree is *not* the best.

The shared lesson: deeper is not better. The right-sized tree is the one that does best on patients it has never seen, and cross-validation is how you find it.

## Why Single Trees Are Unstable

Decision trees have a critical weakness: **high variance**. Small changes in the training data can produce a completely different tree. This instability means that:

- Predictions are unreliable for individual trees.
- Results may not replicate across datasets.
- Performance on new data is often poor.

The solution is to combine many trees into an **ensemble**. Two main strategies exist: **bagging** (reduce variance by averaging) and **boosting** (reduce bias by learning sequentially).

## Bagging: Bootstrap Aggregating

**Bagging** (Bootstrap AGGregatING), introduced by Leo Breiman in 1996, reduces variance by:

1. Drawing $B$ bootstrap samples (random samples with replacement) from the training data.
2. Fitting a separate decision tree to each bootstrap sample.
3. Averaging predictions (regression) or taking a majority vote (classification) across all $B$ trees.

By averaging many high-variance, low-bias trees, bagging dramatically reduces the overall variance without increasing bias. The key insight: the variance of an average of $B$ identically distributed random variables decreases with $B$ (provided they are not perfectly correlated).

## Random Forests

**Random forests** extend bagging with one additional trick: at each split, instead of considering all features, the algorithm randomly selects a **subset of features** (typically $\sqrt{p}$ for classification or $p/3$ for regression, where $p$ is the total number of features).

This feature randomization **decorrelates the trees**. In bagging without feature randomization, if one feature is very strong, most trees will split on it first, making the trees correlated. Random forests force trees to consider different features, producing more diverse trees whose average is more stable.

### Key Hyperparameters

| Parameter | What It Controls | Typical Default |
|-----------|-----------------|----------------|
| `num.trees` / `n_estimators` | Number of trees in the forest | 500–2000 |
| `mtry` / `max_features` | Features considered at each split | $\sqrt{p}$ (classification), $p/3$ (regression) |
| `min.node.size` / `min_samples_leaf` | Minimum observations in a leaf | 1 (classification), 5 (regression) |
| `max.depth` | Maximum tree depth | Unlimited (let trees grow fully) |

### Out-of-Bag (OOB) Error

Each bootstrap sample includes about 63% of the original observations (due to sampling with replacement). The remaining 37% — the **out-of-bag** observations — were not used to build that particular tree. Random forests use these OOB observations as a built-in validation set: each observation is predicted by the trees that did *not* include it in their bootstrap sample.

The OOB error is a nearly unbiased estimate of the test error — no separate validation set required. This is one of the most elegant features of random forests.

### Fitting a Random Forest

::: {.panel-tabset}

#### R

```{r}
#| label: rf-r
library(ranger)

# Fit random forest
set.seed(42)
rf_model <- ranger(
  readmitted ~ age + length_of_stay + num_comorbidities +
    prior_admissions + discharge_hemoglobin +
    discharge_creatinine + has_diabetes + has_chf,
  data = readmit_data,
  num.trees = 500,
  mtry = 3,
  min.node.size = 10,
  importance = "impurity",
  probability = TRUE,   # predict probabilities
  seed = 42
)

cat("OOB prediction error:", rf_model$prediction.error, "\n")
```

```{r}
#| label: rf-importance-r
#| fig-cap: "Variable importance from the random forest model, measured by Gini impurity reduction."
# Variable importance
importance_df <- tibble(
  Variable = names(rf_model$variable.importance),
  Importance = rf_model$variable.importance
) %>%
  arrange(desc(Importance))

ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "#2E86AB") +
  coord_flip() +
  labs(x = NULL, y = "Importance (Gini impurity reduction)",
       title = "Random Forest Variable Importance") +
  theme_minimal(base_size = 14)
```

#### Python

```{python}
#| label: rf-python
from sklearn.ensemble import RandomForestClassifier
from sklearn.model_selection import cross_val_score, StratifiedKFold
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

np.random.seed(42)

rf = RandomForestClassifier(
    n_estimators=500,
    max_features='sqrt',
    min_samples_leaf=10,
    oob_score=True,
    random_state=42,
    n_jobs=-1
)
rf.fit(df, y)

print(f"OOB accuracy: {rf.oob_score_:.3f}")

# Cross-validated AUC
cv = StratifiedKFold(n_splits=5, shuffle=True, random_state=42)
auc_scores = cross_val_score(rf, df, y, cv=cv, scoring='roc_auc')
print(f"CV AUC: {auc_scores.mean():.3f} (+/- {auc_scores.std():.3f})")
```

```{python}
#| label: rf-importance-python
#| fig-cap: "Variable importance from the random forest, measured by mean decrease in Gini impurity."
importances = pd.Series(rf.feature_importances_, index=feature_names)
importances = importances.sort_values(ascending=True)

fig, ax = plt.subplots(figsize=(8, 5))
importances.plot(kind='barh', ax=ax, color='#2E86AB')
ax.set_xlabel("Importance (Gini impurity reduction)")
ax.set_title("Random Forest Variable Importance")
plt.tight_layout()
plt.show()
```

:::

### Variable Importance: Two Measures

1. **Impurity-based importance** (default): Total reduction in Gini impurity (or variance) from splits on that feature, summed across all trees. Fast but can be biased toward high-cardinality features.

2. **Permutation importance**: For each feature, randomly shuffle its values and measure how much the OOB (or test) error increases. More reliable but slower. This is the recommended approach for inference.

::: {.callout-warning}
## Variable Importance Is Not Causal
A feature with high importance in a random forest is *predictive* of the outcome, not necessarily *causally* related to it. A correlated proxy (e.g., number of medications as a proxy for disease severity) may appear important even though intervening on it would have no effect. Do not confuse prediction with causation.
:::

## Gradient Boosting

While bagging reduces variance by averaging independent trees, **gradient boosting** reduces bias by building trees *sequentially*, where each new tree corrects the mistakes of the previous ensemble.

### The Intuition: Learning from Mistakes

1. Fit a simple tree to the data.
2. Compute the **residuals** (errors) from this tree.
3. Fit a new tree to predict these residuals.
4. Add this new tree to the ensemble (scaled by a learning rate).
5. Repeat steps 2–4 for $B$ iterations.

Each new tree focuses on the observations the current ensemble gets wrong. Over many iterations, the ensemble becomes increasingly accurate.

### The Mathematics (Brief)

Gradient boosting minimizes a loss function $L(y, \hat{y})$ by gradient descent in function space. At iteration $m$:

$$\hat{f}_m(x) = \hat{f}_{m-1}(x) + \eta \cdot h_m(x)$$

where $h_m(x)$ is a tree fitted to the negative gradient of the loss (the "pseudo-residuals") and $\eta$ is the **learning rate** (also called the shrinkage parameter).

### Key Hyperparameters

| Parameter | What It Controls | Typical Range |
|-----------|-----------------|---------------|
| `n_estimators` / `nrounds` | Number of boosting iterations | 100–5000 |
| `learning_rate` / `eta` | Step size for each tree | 0.01–0.3 |
| `max_depth` | Depth of each tree | 3–8 (shallow trees!) |
| `subsample` | Fraction of data used per tree | 0.5–1.0 |
| `colsample_bytree` | Fraction of features per tree | 0.5–1.0 |
| `min_child_weight` | Minimum sum of instance weights in a leaf | 1–10 |
| `reg_alpha` (L1) / `reg_lambda` (L2) | Regularization | 0–10 |

::: {.callout-tip}
## The Learning Rate / Number of Trees Tradeoff
A smaller learning rate requires more trees but generally produces better results. A common strategy: set the learning rate to 0.01–0.05, then tune the number of trees via early stopping (stop when validation error stops improving).
:::

### XGBoost and LightGBM

**XGBoost** (eXtreme Gradient Boosting) by Chen and Guestrin (2016) added regularization, efficient computation, and handling of missing values to gradient boosting, making it the dominant algorithm for tabular data competitions and many clinical prediction tasks.

**LightGBM** by Microsoft further improves efficiency with gradient-based one-side sampling (GOSS) and exclusive feature bundling. It is often faster than XGBoost on large datasets.

::: {.panel-tabset}

#### R

```{r}
#| label: xgb-r
library(xgboost)

# Prepare data for xgboost (requires numeric matrix)
X_train <- readmit_data %>%
  select(age, length_of_stay, num_comorbidities, prior_admissions,
         discharge_hemoglobin, discharge_creatinine, has_diabetes, has_chf) %>%
  as.matrix()

y_train <- as.numeric(readmit_data$readmitted == "Yes")

dtrain <- xgb.DMatrix(data = X_train, label = y_train)

# Fit XGBoost with cross-validation to find optimal nrounds
set.seed(42)
xgb_cv <- xgb.cv(
  params = list(
    objective = "binary:logistic",
    eval_metric = "auc",
    eta = 0.05,
    max_depth = 4,
    subsample = 0.8,
    colsample_bytree = 0.8
  ),
  data = dtrain,
  nrounds = 1000,
  nfold = 5,
  early_stopping_rounds = 50,
  verbose = 0
)

cat("Best iteration:", xgb_cv$best_iteration, "\n")
cat("Best CV AUC:", max(xgb_cv$evaluation_log$test_auc_mean), "\n")
```

```{r}
#| label: xgb-fit-r
# Fit final model with optimal nrounds
xgb_model <- xgb.train(
  params = list(
    objective = "binary:logistic",
    eval_metric = "auc",
    eta = 0.05,
    max_depth = 4,
    subsample = 0.8,
    colsample_bytree = 0.8
  ),
  data = dtrain,
  nrounds = xgb_cv$best_iteration,
  verbose = 0
)

# Variable importance
importance_matrix <- xgb.importance(model = xgb_model)
xgb.plot.importance(importance_matrix, top_n = 8,
                     main = "XGBoost Variable Importance")
```

#### Python

```{python}
#| label: xgb-python
import xgboost as xgb
from sklearn.model_selection import cross_val_score, StratifiedKFold
import numpy as np

np.random.seed(42)

xgb_model = xgb.XGBClassifier(
    n_estimators=1000,
    learning_rate=0.05,
    max_depth=4,
    subsample=0.8,
    colsample_bytree=0.8,
    eval_metric='auc',
    early_stopping_rounds=50,
    random_state=42,
    use_label_encoder=False
)

# Cross-validation
cv = StratifiedKFold(n_splits=5, shuffle=True, random_state=42)
scores = cross_val_score(
    xgb.XGBClassifier(
        n_estimators=500, learning_rate=0.05, max_depth=4,
        subsample=0.8, colsample_bytree=0.8,
        random_state=42, use_label_encoder=False, eval_metric='logloss'
    ),
    df, y, cv=cv, scoring='roc_auc'
)
print(f"XGBoost CV AUC: {scores.mean():.3f} (+/- {scores.std():.3f})")
```

```{python}
#| label: xgb-importance-python
#| fig-cap: "XGBoost variable importance based on the number of times each feature is used in splits (weight)."
# Fit and plot importance
xgb_final = xgb.XGBClassifier(
    n_estimators=500, learning_rate=0.05, max_depth=4,
    subsample=0.8, colsample_bytree=0.8,
    random_state=42, use_label_encoder=False, eval_metric='logloss'
)
xgb_final.fit(df, y)

fig, ax = plt.subplots(figsize=(8, 5))
xgb.plot_importance(xgb_final, ax=ax, importance_type='gain',
                     title='XGBoost Feature Importance (Gain)')
plt.tight_layout()
plt.show()
```

:::

## Hyperparameter Tuning

Let us start from the very beginning, because this idea trips up almost everyone the first time.

When you fit a random forest, there are two completely different kinds of "settings" involved. The first kind, the model **learns by itself** from the data --- for a tree, that means the actual questions at each split ("prior admissions ≥ 2?"). You never choose these; the algorithm finds them. The second kind, **you must choose before training even starts** --- for example, how many trees to grow, or how deep each tree may go. These pre-set dials are called **hyperparameters**, and the model cannot learn them from the data because they govern *how the learning itself happens*.

::: {.callout-note}
## Parameters vs. hyperparameters (the one distinction to remember)
- A **parameter** is something the model estimates from the data (a regression coefficient, a tree's split rules). You do not set these.
- A **hyperparameter** is a knob *you* set beforehand that controls the learning process (number of trees, tree depth, learning rate). The model cannot figure these out on its own.

An analogy: baking bread, the *recipe quantities the dough develops into* (how the crumb sets) are like parameters --- they emerge from the process. The **oven temperature and baking time** are hyperparameters --- you dial them in beforehand, and getting them wrong ruins the loaf no matter how good the dough.
:::

**Why bother tuning them?** Every library ships with default hyperparameter values, and they are reasonable starting points --- but they are generic, not tailored to *your* dataset. The best settings depend on how much data you have, how noisy it is, and how many predictors there are. Tuning simply means **trying several combinations of settings and keeping the one that performs best on held-out data** (via the cross-validation you met earlier). Done well, tuning often buys a meaningful bump in performance for free; done carelessly (tuning on the test set), it produces a model that looks great and then disappoints.

The hyperparameters you will see in the code below, in plain language:

| Hyperparameter (R / Python) | What it controls | Rule of thumb |
|---|---|---|
| `trees` / `n_estimators` | How many trees are in the forest | More is safer (just slower); a few hundred is usually plenty |
| `mtry` / `max_features` | How many predictors each split is allowed to consider | Smaller values make trees more different from each other, which can help |
| `min_n` / `min_samples_leaf` | The smallest group of patients allowed in a leaf | Larger values = simpler, smoother trees that overfit less |
| `max_depth` | How many questions deep a tree may grow | Shallower = simpler; limits overfitting |

You do not need to memorise these --- the point is that each one trades flexibility against the risk of overfitting, and tuning finds the sweet spot for your data.

There are two common strategies for searching through the possible combinations.

### Grid Search

**Grid search** means you write down a list of candidate values for each hyperparameter, and the computer tries *every possible combination*. If you list 4 values for `mtry` and 5 values for `min_n`, grid search trains and evaluates all $4 \times 5 = 20$ combinations and reports which did best. It is thorough and easy to reason about, but the cost explodes as you add hyperparameters: 4 settings each for just five hyperparameters is already $4^5 = 1024$ models to fit --- and each one is itself fit five times under 5-fold cross-validation. This multiplicative blow-up is grid search's main drawback.

### Random Search

**Random search** instead samples a fixed number of combinations *at random* from the ranges you specify (say, 20 combinations, regardless of how many hyperparameters there are). Bergstra and Bengio (2012) showed this is usually *more* efficient than grid search, and the reason is intuitive: in most problems only one or two hyperparameters really matter, but you do not know in advance which. A grid wastes much of its budget trying many values of the unimportant dials while testing only a few values of the important one. Random search, for the same number of model fits, ends up trying *more distinct values of every dial* --- so it is more likely to stumble onto a good value of whichever one actually matters. **In practice, start with random search**; it gives you most of the benefit for a fraction of the compute.

::: {.panel-tabset}

#### R

```{r}
#| label: tune-r
library(tidymodels)

# Define the model with tunable parameters
rf_spec <- rand_forest(
  trees = 500,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# Create recipe
rf_recipe <- recipe(readmitted ~ ., data = readmit_data)

# Workflow
rf_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(rf_recipe)

# Define tuning grid
rf_grid <- grid_random(
  mtry(range = c(2, 7)),
  min_n(range = c(5, 30)),
  size = 20
)

# Cross-validation
set.seed(42)
folds <- vfold_cv(readmit_data, v = 5, strata = readmitted)

# Tune (this may take a moment)
rf_tune_results <- tune_grid(
  rf_wf,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_set(roc_auc)
)

# Best parameters
show_best(rf_tune_results, metric = "roc_auc", n = 5)
```

```{r}
#| label: tune-finalize-r
# Select best and finalize
best_params <- select_best(rf_tune_results, metric = "roc_auc")
cat("Best mtry:", best_params$mtry, "\n")
cat("Best min_n:", best_params$min_n, "\n")

final_wf <- finalize_workflow(rf_wf, best_params)
final_fit <- fit(final_wf, data = readmit_data)
```

#### Python

```{python}
#| label: tune-python
from sklearn.ensemble import RandomForestClassifier
from sklearn.model_selection import RandomizedSearchCV, StratifiedKFold
import numpy as np

np.random.seed(42)

# Define parameter distributions
param_dist = {
    'n_estimators': [100, 200, 500],
    'max_features': ['sqrt', 'log2', 3, 5],
    'min_samples_leaf': [5, 10, 15, 20, 30],
    'max_depth': [5, 10, 15, 20, None]
}

rf = RandomForestClassifier(random_state=42, n_jobs=-1)
cv = StratifiedKFold(n_splits=5, shuffle=True, random_state=42)

# Random search
search = RandomizedSearchCV(
    rf, param_dist, n_iter=20, cv=cv, scoring='roc_auc',
    random_state=42, n_jobs=-1, verbose=0
)
search.fit(df, y)

print(f"Best AUC: {search.best_score_:.3f}")
print(f"Best parameters: {search.best_params_}")
```

:::

**What the code is doing, step by step.** Both versions automate the "try many settings, keep the best" recipe. Reading the R version line by line:

1. `rand_forest(... mtry = tune(), min_n = tune())` --- we declare a random forest but, instead of fixing `mtry` and `min_n`, we mark them with `tune()`, which is a placeholder meaning *"leave this blank for now; we will search for a good value."*
2. The `recipe` and `workflow` just bundle the model with the outcome and predictors --- think of the workflow as a sealed envelope containing everything needed to fit the model.
3. `grid_random(...)` builds the list of 20 random combinations of `mtry` and `min_n` to try (these are the candidate "oven settings").
4. `vfold_cv(...)` creates the 5-fold cross-validation splits. This is the crucial part: **every candidate setting is judged on held-out patients, not on the data it was trained on.** That is what stops us from simply picking the most complex, overfit model.
5. `tune_grid(...)` does the heavy lifting --- it fits the forest for each of the 20 settings, across all 5 folds, and records the cross-validated AUC for each.
6. `show_best(...)` then prints the top settings.

**What the output shows.** `show_best` returns a small table: each row is one hyperparameter combination, with its mean cross-validated AUC and the spread across folds. You read it top-down --- the first row is the winning combination. `select_best` grabs that winner, and `finalize_workflow` + `fit` retrain the forest on *all* the data using those best settings, giving you the final model. The Python `RandomizedSearchCV` does exactly the same thing in one object, printing `Best AUC` and the `Best parameters` dictionary at the end. The specific numbers are not the lesson (the data here are simulated); the lesson is the *workflow*: define the dials to search, score each candidate honestly on held-out folds, and keep the winner.

::: {.callout-important}
## The cardinal rule of tuning
Tuning must happen **inside** cross-validation, using only the training data --- never tune by looking at your final test set. If you try hundreds of settings and pick the one that scores best on the test set, that test score is no longer an honest estimate of future performance: you have, in effect, let the test set leak into model development. Keep a truly untouched test set (or external validation data) for the *final* check, after all tuning is done.
:::

## Comparing Methods: When to Use What

| Method | Strengths | Weaknesses | Best For |
|--------|-----------|------------|----------|
| **Single tree** | Interpretable, handles non-linearity | Unstable, overfits | Exploration, simple rules |
| **Random forest** | Robust, low tuning needed, OOB error | Less interpretable than single tree | General-purpose prediction |
| **Gradient boosting** | Often highest accuracy, flexible | More tuning, can overfit, slower | Maximizing prediction accuracy |
| **Logistic regression** | Interpretable, well-understood inference | Assumes linearity (without transforms) | Inference, small samples, reporting ORs |

::: {.callout-note}
## Practical Advice for Clinical Prediction Models
Start with logistic regression as your baseline. If you need better prediction and have enough data (typically hundreds of events), try a random forest. If you want to squeeze out the last bit of performance and are willing to tune carefully, use XGBoost. Always compare against the logistic regression baseline — if the improvement is marginal, prefer the simpler model.
:::

## Clinical Example: Readmission Risk Stratification

Let us bring everything together by comparing all three methods on the hospital readmission dataset.

::: {.panel-tabset}

#### R

```{r}
#| label: comparison-r
library(tidymodels)

set.seed(42)
folds <- vfold_cv(readmit_data, v = 10, strata = readmitted)

# Logistic regression
lr_spec <- logistic_reg() %>% set_engine("glm")
lr_wf <- workflow() %>%
  add_model(lr_spec) %>%
  add_recipe(recipe(readmitted ~ ., data = readmit_data))
lr_res <- fit_resamples(lr_wf, resamples = folds,
                        metrics = metric_set(roc_auc))

# Random forest
rf_spec <- rand_forest(trees = 500, mtry = 3, min_n = 10) %>%
  set_engine("ranger") %>%
  set_mode("classification")
rf_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(recipe(readmitted ~ ., data = readmit_data))
rf_res <- fit_resamples(rf_wf, resamples = folds,
                        metrics = metric_set(roc_auc))

# Boosted trees (via xgboost engine in tidymodels)
xgb_spec <- boost_tree(trees = 500, tree_depth = 4, learn_rate = 0.05,
                        min_n = 10) %>%
  set_engine("xgboost") %>%
  set_mode("classification")
xgb_wf <- workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(recipe(readmitted ~ ., data = readmit_data))
xgb_res <- fit_resamples(xgb_wf, resamples = folds,
                         metrics = metric_set(roc_auc))

# Collect and compare results
bind_rows(
  collect_metrics(lr_res) %>% mutate(model = "Logistic Regression"),
  collect_metrics(rf_res) %>% mutate(model = "Random Forest"),
  collect_metrics(xgb_res) %>% mutate(model = "XGBoost")
) %>%
  select(model, .metric, mean, std_err) %>%
  arrange(desc(mean))
```

#### Python

```{python}
#| label: comparison-python
from sklearn.linear_model import LogisticRegression
from sklearn.ensemble import RandomForestClassifier, GradientBoostingClassifier
from sklearn.model_selection import cross_val_score, StratifiedKFold
from sklearn.preprocessing import StandardScaler
from sklearn.pipeline import make_pipeline
import numpy as np
import pandas as pd

np.random.seed(42)
cv = StratifiedKFold(n_splits=10, shuffle=True, random_state=42)

models = {
    'Logistic Regression': make_pipeline(
        StandardScaler(), LogisticRegression(max_iter=1000)
    ),
    'Random Forest': RandomForestClassifier(
        n_estimators=500, max_features='sqrt', min_samples_leaf=10,
        random_state=42, n_jobs=-1
    ),
    'Gradient Boosting': GradientBoostingClassifier(
        n_estimators=500, learning_rate=0.05, max_depth=4,
        subsample=0.8, random_state=42
    )
}

results = []
for name, model in models.items():
    scores = cross_val_score(model, df, y, cv=cv, scoring='roc_auc')
    results.append({
        'Model': name,
        'Mean AUC': f"{scores.mean():.3f}",
        'Std': f"{scores.std():.3f}"
    })
    print(f"{name:25s}: AUC = {scores.mean():.3f} (+/- {scores.std():.3f})")

print(pd.DataFrame(results).to_string(index=False))
```

:::

## Exercises

### Exercise 1: Build and Prune a Classification Tree

Using the readmission dataset (or a dataset of your choice):

1. Fit a full, unpruned classification tree. How many terminal nodes does it have?
2. Use cross-validation to find the optimal complexity parameter (cp in R) or maximum depth (in Python).
3. Prune the tree and plot it. How does it compare to the full tree?
4. What are the top 3 splitting variables? Do they make clinical sense?

::: {.panel-tabset}

#### R Starter Code

```{r}
#| label: ex1-tree-r
#| eval: false
library(rpart)      # rpart(), rpart.control()
library(rpart.plot) # rpart.plot() for visualising the tree

# Fit a full tree (set cp very low to avoid early pruning)
full_tree <- rpart(readmitted ~ ., data = readmit_data, method = "class",
                   control = rpart.control(cp = 0.001))

# Your code:
# 1. Count terminal nodes
# 2. Use printcp() and plotcp() to find optimal cp
# 3. Prune and plot
# 4. Examine variable importance: full_tree$variable.importance
```

#### Python Starter Code

```{python}
#| label: ex1-tree-python
#| eval: false
from sklearn.tree import DecisionTreeClassifier
from sklearn.model_selection import cross_val_score

# Your code:
# 1. Fit trees with different max_depth values
# 2. Cross-validate each
# 3. Plot the optimal tree
# 4. Examine feature_importances_
```

:::

### Exercise 2: Random Forest vs XGBoost Tuning Challenge

1. Split the readmission data into 80% training and 20% test sets.
2. Using the training set with 5-fold cross-validation:
   a. Tune a random forest (try different `mtry`/`max_features` and `min_n`/`min_samples_leaf` values).
   b. Tune an XGBoost model (try different `learning_rate`, `max_depth`, and `n_estimators` values).
3. Select the best hyperparameters for each model.
4. Evaluate both models on the held-out test set. Which performs better?
5. Create variable importance plots for both models. Do they agree on the most important features?

::: {.panel-tabset}

#### R Starter Code

```{r}
#| label: ex2-tune-r
#| eval: false
library(tidymodels) # initial_split(), training(), testing(), tune_grid()

# Split data
set.seed(42)
split <- initial_split(readmit_data, prop = 0.8, strata = readmitted)
train <- training(split)
test <- testing(split)

# Your code: define models with tune(), create grids, run tune_grid()
# Finalize models, predict on test set, compare
```

#### Python Starter Code

```{python}
#| label: ex2-tune-python
#| eval: false
from sklearn.model_selection import train_test_split, RandomizedSearchCV
from sklearn.ensemble import RandomForestClassifier
import xgboost as xgb

X_train, X_test, y_train, y_test = train_test_split(
    df, y, test_size=0.2, stratify=y, random_state=42
)

# Your code: define param_dist for RF and XGBoost
# Run RandomizedSearchCV for each
# Evaluate on test set with roc_auc_score
```

:::

### Exercise 3: Interpreting a Clinical Prediction Model

You have built a gradient-boosted model for 30-day hospital readmission. Your hospital administration asks you to present the results.

1. Write a non-technical summary (3–4 sentences) explaining what the model does and how well it performs, suitable for a hospital quality committee.
2. The model identifies "number of prior admissions" and "discharge creatinine" as the two most important predictors. Explain to a clinical audience what this means and *does not* mean (recall the caution about variable importance and causation).
3. How would you propose to validate this model before deploying it in clinical practice? What could go wrong if you skip external validation?

## Summary

| Method | How It Works | Key Advantage | Key Risk |
|--------|-------------|---------------|----------|
| Decision tree | Sequential yes/no splits | Highly interpretable | Overfits, unstable |
| Random forest | Average of many decorrelated trees | Robust, minimal tuning | Less interpretable |
| Gradient boosting | Sequential trees correcting errors | Often best accuracy | Can overfit, many hyperparameters |

The progression from single trees to random forests to gradient boosting follows a clear logic:

- **Single trees** are interpretable but unstable and prone to overfitting.
- **Random forests** reduce variance by averaging many decorrelated trees, with the OOB error providing a built-in validation estimate.
- **Gradient boosting** reduces both bias and variance through sequential learning, often achieving the best predictive accuracy at the cost of more tuning.

For clinical prediction models, always compare against a logistic regression baseline. Use cross-validation for honest evaluation, tune hyperparameters systematically, and remember that the most complex model is not always the best choice.

## References and Further Reading

The foundational paper on random forests is Breiman (2001), "Random Forests," published in *Machine Learning*. This paper introduced the algorithm, the concept of out-of-bag error estimation, and permutation-based variable importance. It remains one of the most cited papers in machine learning.

For XGBoost, the key reference is Chen and Guestrin (2016), "XGBoost: A Scalable Tree Boosting System," presented at KDD 2016. This paper describes the algorithmic innovations (regularized objective, approximate split finding, sparsity-aware computation) that made XGBoost the dominant method for tabular data.

Smits et al. (2026) provide an accessible treatment of tree-based methods in the context of clinical research, including practical guidance on when ensemble methods offer meaningful advantages over logistic regression and how to interpret variable importance in clinical terms.

For a thorough treatment of the CART algorithm, see Breiman, Friedman, Olshen, and Stone's *Classification and Regression Trees* (1984, Wadsworth). Although the original text is somewhat dated, the core ideas remain current.

Hastie, Tibshirani, and Friedman's *The Elements of Statistical Learning* (Springer, 2nd edition, 2009) covers bagging (Section 8.7), random forests (Chapter 15), and boosting (Chapter 10) with mathematical rigor. James et al.'s *An Introduction to Statistical Learning* (ISLR, 2nd edition, 2021) provides a gentler introduction in Chapters 8 and covers tree-based methods with R lab exercises.

Boehmke and Greenwell's *Hands-On Machine Learning with R* dedicates several chapters to tree-based methods with practical tidymodels code, and is an excellent companion for the R exercises in this chapter.

For the gradient boosting framework, Friedman (2001), "Greedy Function Approximation: A Gradient Boosting Machine," in *Annals of Statistics*, is the foundational statistical treatment. The paper introduced the general gradient boosting framework that XGBoost and LightGBM build upon.

Regarding clinical applications and best practices, the TRIPOD statement (Collins et al., 2015) provides guidelines for reporting clinical prediction models, including those based on machine learning. Any readmission model (or similar clinical prediction tool) developed with these methods should be reported following TRIPOD guidelines and externally validated before clinical deployment.
