# Evaluating Models: Beyond Accuracy {#sec-model-evaluation}

## Introduction

You have built a model. It produces predictions. But how do you know if those predictions are any good? The answer is not as simple as asking "how often is it right?" In clinical medicine, the consequences of different types of errors are rarely symmetric, the outcomes we care about are often rare, and the population we apply a model to may look nothing like the population we trained it on. This chapter introduces the core concepts and tools for evaluating the performance of classification models, with a focus on clinical prediction.

We will move beyond the seductive simplicity of "accuracy" and learn to ask better questions: Does the model separate sick from well? Are the predicted probabilities trustworthy? Would using this model actually help patients? And does it help all patients equally?

## The Accuracy Trap

Consider the following scenario. You are developing a model to predict a rare but serious condition --- say, pancreatic cancer --- in a primary care population. The prevalence is approximately 0.1%. You build a classifier and proudly report 99.9% accuracy. Your colleague, less impressed, points out that a model which simply predicts "no cancer" for every single patient would also achieve 99.9% accuracy. It would miss every true case, which is the entire point of the exercise, but its accuracy would be nearly perfect.

This is the **accuracy paradox**: when outcomes are imbalanced, accuracy becomes meaningless because it is dominated by the majority class. In clinical medicine, outcomes are almost always imbalanced. Most people do not have the disease you are screening for. Most surgical patients do not die within 30 days. Most pregnancies do not result in pre-eclampsia. A metric that ignores the structure of errors is not useful for clinical decision-making.

::: {.callout-important}
## The Golden Rule of Model Evaluation
Never report accuracy alone. Always examine how the model performs separately for those with and without the outcome.
:::

## Sensitivity and Specificity

The most fundamental decomposition of model performance separates its behaviour in two groups: those who truly have the condition and those who do not.

**Sensitivity** (also called recall or the true positive rate) answers: among everyone who truly has the disease, what proportion did the model correctly identify?

$$\text{Sensitivity} = \frac{\text{True Positives}}{\text{True Positives} + \text{False Negatives}}$$

**Specificity** (the true negative rate) answers: among everyone who is truly disease-free, what proportion did the model correctly identify as negative?

$$\text{Specificity} = \frac{\text{True Negatives}}{\text{True Negatives} + \text{False Positives}}$$

These two metrics are properties of the **test itself** and do not depend on the prevalence of the disease in the population. This makes them useful for describing a test's intrinsic discriminative ability.

### Clinical Context: Screening vs Confirmation

The relative importance of sensitivity and specificity depends on the clinical purpose:

- **Screening tests** (e.g., mammography for breast cancer, the PHQ-2 for depression) prioritise **high sensitivity**. The goal is to catch as many true cases as possible. We accept more false positives because the next step is a confirmatory test, not treatment. Missing a case (false negative) is the costly error.

- **Confirmatory tests** (e.g., biopsy for cancer, Western blot for HIV) prioritise **high specificity**. The goal is to be sure that a positive result is real, because a positive typically triggers treatment, which may carry risks. A false positive leading to unnecessary surgery or chemotherapy is the costly error.

::: {.callout-tip}
## Memory Aid
**Sn**N**out**: a test with high **S**e**n**sitivity, when **N**egative, rules **out** the disease.
**Sp**P**in**: a test with high **Sp**ecificity, when **P**ositive, rules **in** the disease.
:::

## Predictive Values: When Prevalence Matters

While sensitivity and specificity describe the test, clinicians and patients need to answer a different question: given my test result, what is the probability I actually have the disease?

**Positive Predictive Value (PPV)**: among everyone who tested positive, what proportion truly has the disease?

$$\text{PPV} = \frac{\text{True Positives}}{\text{True Positives} + \text{False Positives}}$$

**Negative Predictive Value (NPV)**: among everyone who tested negative, what proportion is truly disease-free?

$$\text{NPV} = \frac{\text{True Negatives}}{\text{True Negatives} + \text{False Negatives}}$$

Critically, PPV and NPV depend on the **prevalence** of the disease in the population being tested. A test with 95% sensitivity and 95% specificity will have very different PPVs depending on context:

| Prevalence | PPV   | NPV   |
|------------|-------|-------|
| 50%        | 95.0% | 95.0% |
| 10%        | 67.9% | 99.4% |
| 1%         | 16.1% | 99.9% |
| 0.1%       | 1.9%  | 100%  |

: PPV and NPV for a test with 95% sensitivity and 95% specificity at different prevalence levels. {#tbl-ppv-prevalence}

At 0.1% prevalence, even with an excellent test, fewer than 2% of positive results are true positives. This is why mass screening for rare diseases generates so many false alarms, and why testing should be targeted to higher-risk populations whenever possible.

### Exercise: Bayes' Theorem in Practice

::: {.panel-tabset}

#### R

```{r}
#| label: ppv-calculation-r
#| eval: false

# Calculate PPV using Bayes' theorem
# PPV = (Sensitivity * Prevalence) /
#        (Sensitivity * Prevalence + (1 - Specificity) * (1 - Prevalence))

calculate_ppv <- function(sensitivity, specificity, prevalence) {
  numerator <- sensitivity * prevalence
  denominator <- numerator + (1 - specificity) * (1 - prevalence)
  return(numerator / denominator)
}

# Example: HIV rapid test (sensitivity = 99.7%, specificity = 99.5%)
# In a general population (prevalence ~ 0.4%)
ppv_general <- calculate_ppv(0.997, 0.995, 0.004)
cat("PPV in general population:", round(ppv_general * 100, 1), "%\n")

# In an STI clinic population (prevalence ~ 5%)
ppv_clinic <- calculate_ppv(0.997, 0.995, 0.05)
cat("PPV in STI clinic:", round(ppv_clinic * 100, 1), "%\n")

# In a population with known exposure (prevalence ~ 30%)
ppv_exposed <- calculate_ppv(0.997, 0.995, 0.30)
cat("PPV with known exposure:", round(ppv_exposed * 100, 1), "%\n")

# Plot PPV across a range of prevalences
prevalences <- seq(0.001, 0.5, by = 0.001)
ppvs <- sapply(prevalences, function(p) calculate_ppv(0.997, 0.995, p))

plot(prevalences * 100, ppvs * 100,
     type = "l", lwd = 2, col = "steelblue",
     xlab = "Prevalence (%)", ylab = "Positive Predictive Value (%)",
     main = "PPV depends heavily on prevalence",
     las = 1)
abline(h = 50, lty = 2, col = "grey50")
```

#### Python

```{python}
#| label: ppv-calculation-py
#| eval: false

import numpy as np
import matplotlib.pyplot as plt

def calculate_ppv(sensitivity, specificity, prevalence):
    """Calculate PPV using Bayes' theorem."""
    numerator = sensitivity * prevalence
    denominator = numerator + (1 - specificity) * (1 - prevalence)
    return numerator / denominator

# Example: HIV rapid test (sensitivity = 99.7%, specificity = 99.5%)
# In a general population (prevalence ~ 0.4%)
ppv_general = calculate_ppv(0.997, 0.995, 0.004)
print(f"PPV in general population: {ppv_general*100:.1f}%")

# In an STI clinic population (prevalence ~ 5%)
ppv_clinic = calculate_ppv(0.997, 0.995, 0.05)
print(f"PPV in STI clinic: {ppv_clinic*100:.1f}%")

# In a population with known exposure (prevalence ~ 30%)
ppv_exposed = calculate_ppv(0.997, 0.995, 0.30)
print(f"PPV with known exposure: {ppv_exposed*100:.1f}%")

# Plot PPV across a range of prevalences
prevalences = np.linspace(0.001, 0.5, 500)
ppvs = [calculate_ppv(0.997, 0.995, p) for p in prevalences]

plt.figure(figsize=(8, 5))
plt.plot(prevalences * 100, np.array(ppvs) * 100, lw=2, color="steelblue")
plt.axhline(y=50, linestyle="--", color="grey", alpha=0.7)
plt.xlabel("Prevalence (%)")
plt.ylabel("Positive Predictive Value (%)")
plt.title("PPV depends heavily on prevalence")
plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** This puts the prevalence lesson on its feet with a real test: an HIV rapid test with near-perfect sensitivity (99.7%) and specificity (99.5%). The `calculate_ppv` function is just Bayes' theorem written out --- it asks "of everyone who tests positive, what fraction are truly positive?" The three worked examples show the same excellent test producing wildly different answers depending on *who* you test: a positive result means roughly a 44% chance of true infection in the general population (prevalence 0.4%), about 91% in an STI clinic (5%), and around 99% after known exposure (30%). The plot then sweeps prevalence from near zero to 50% and traces the PPV: a steep climb at low prevalence that flattens out, with the dashed line marking where a positive result becomes more likely true than false. The clinical message a health researcher should take away is that the same test can be trustworthy or near-useless depending entirely on the pre-test probability of the patient in front of you --- which is why we target testing to higher-risk groups.

## The Confusion Matrix

A **confusion matrix** is the fundamental bookkeeping device for classification evaluation. For a binary outcome, it is a 2x2 table that cross-classifies predictions against truth:

|                        | Condition Positive | Condition Negative |
|------------------------|-------------------:|-------------------:|
| **Predicted Positive** | True Positive (TP) | False Positive (FP)|
| **Predicted Negative** | False Negative (FN)| True Negative (TN) |

Every metric discussed in this chapter can be derived from this table. The crucial insight is that a single model can produce **different** confusion matrices depending on the **threshold** you choose for classifying a predicted probability as "positive" or "negative."

### Building and Inspecting Confusion Matrices

::: {.panel-tabset}

#### R

```{r}
#| label: confusion-matrix-r
#| eval: false

library(caret)

# Simulate predicted probabilities and true outcomes
set.seed(42)
n <- 1000
true_outcome <- rbinom(n, 1, 0.15)  # 15% prevalence
# Simulate a moderately good model
pred_prob <- plogis(rnorm(n, mean = -1 + 2 * true_outcome, sd = 1.2))

# Confusion matrix at threshold = 0.5
pred_class_50 <- ifelse(pred_prob >= 0.5, 1, 0)
cm_50 <- confusionMatrix(factor(pred_class_50), factor(true_outcome), positive = "1")
print(cm_50)

# Confusion matrix at threshold = 0.2 (more sensitive)
pred_class_20 <- ifelse(pred_prob >= 0.2, 1, 0)
cm_20 <- confusionMatrix(factor(pred_class_20), factor(true_outcome), positive = "1")
print(cm_20)

cat("\nAt threshold 0.5:\n")
cat("  Sensitivity:", round(cm_50$byClass["Sensitivity"], 3), "\n")
cat("  Specificity:", round(cm_50$byClass["Specificity"], 3), "\n")

cat("\nAt threshold 0.2:\n")
cat("  Sensitivity:", round(cm_20$byClass["Sensitivity"], 3), "\n")
cat("  Specificity:", round(cm_20$byClass["Specificity"], 3), "\n")
```

#### Python

```{python}
#| label: confusion-matrix-py
#| eval: false

import numpy as np
from sklearn.metrics import confusion_matrix, classification_report
from scipy.special import expit

np.random.seed(42)
n = 1000
true_outcome = np.random.binomial(1, 0.15, n)  # 15% prevalence
# Simulate a moderately good model
pred_prob = expit(np.random.normal(-1 + 2 * true_outcome, 1.2))

# Confusion matrix at threshold = 0.5
pred_class_50 = (pred_prob >= 0.5).astype(int)
print("=== Threshold = 0.5 ===")
print(confusion_matrix(true_outcome, pred_class_50))
print(classification_report(true_outcome, pred_class_50, target_names=["Negative", "Positive"]))

# Confusion matrix at threshold = 0.2 (more sensitive)
pred_class_20 = (pred_prob >= 0.2).astype(int)
print("=== Threshold = 0.2 ===")
print(confusion_matrix(true_outcome, pred_class_20))
print(classification_report(true_outcome, pred_class_20, target_names=["Negative", "Positive"]))
```

:::

**What the code is showing.** This demonstrates the single most important fact about a confusion matrix: it changes when you move the threshold, even though the model itself never changes. We simulate 1,000 patients at 15% prevalence and a moderately good model, then build two confusion matrices from the *same* predicted probabilities --- one classifying anyone above 0.5 as positive, the other lowering the bar to 0.2. The printed tables and the sensitivity/specificity lines tell the story: at the strict 0.5 threshold the model is cautious, so specificity is high but it misses many true cases (low sensitivity); dropping to 0.2 catches far more true cases (sensitivity rises) at the cost of more false alarms (specificity falls). Reading these outputs side by side is the intuition behind every ROC curve that follows: there is no single "right" confusion matrix, only the one that matches the clinical cost of missing a case versus raising a false alarm.

## ROC Curves

The **Receiver Operating Characteristic (ROC) curve** is perhaps the most widely used graphical tool for evaluating classification models. It plots sensitivity (true positive rate) on the y-axis against 1 minus specificity (false positive rate) on the x-axis, tracing out the trade-off as the classification threshold varies from 1 to 0.

### How to Read an ROC Curve

- The **diagonal line** from (0,0) to (1,1) represents a model with no discriminative ability --- equivalent to flipping a coin.
- A curve that bows toward the **upper-left corner** indicates good discrimination. The model achieves high sensitivity without sacrificing too much specificity.
- The **Area Under the ROC Curve (AUC or C-statistic)** summarises the curve in a single number. It equals the probability that, given one randomly chosen person with the disease and one without, the model assigns a higher predicted risk to the diseased person.

| AUC Range   | Interpretation          |
|-------------|-------------------------|
| 0.90--1.00  | Excellent discrimination |
| 0.80--0.90  | Good discrimination     |
| 0.70--0.80  | Acceptable              |
| 0.60--0.70  | Poor                    |
| 0.50--0.60  | Fail (near chance)      |

: Rough guide to AUC interpretation. These benchmarks are context-dependent. {#tbl-auc-benchmarks}

::: {.callout-warning}
## AUC Is Not Enough
A model can have a high AUC but still produce poorly calibrated probabilities. Two models with the same AUC can have very different clinical utility. AUC tells you about ranking (discrimination), not about the accuracy of the predicted probabilities themselves. Always supplement AUC with calibration assessment (see @sec-performance-validation).
:::

### Plotting ROC Curves

::: {.panel-tabset}

#### R

```{r}
#| label: roc-curve-r
#| eval: false

library(pROC)

# Using the simulated data from above
set.seed(42)
n <- 1000
true_outcome <- rbinom(n, 1, 0.15)
pred_prob <- plogis(rnorm(n, mean = -1 + 2 * true_outcome, sd = 1.2))

roc_obj <- roc(true_outcome, pred_prob)

# Plot the ROC curve
plot(roc_obj,
     main = paste("ROC Curve (AUC =", round(auc(roc_obj), 3), ")"),
     col = "steelblue", lwd = 2,
     print.auc = TRUE, print.auc.y = 0.4,
     legacy.axes = TRUE)  # Use 1-Specificity on x-axis

# Add confidence interval for AUC
ci_auc <- ci.auc(roc_obj)
cat("95% CI for AUC:", round(ci_auc[1], 3), "-", round(ci_auc[3], 3), "\n")

# Find the optimal threshold (Youden's J)
coords_best <- coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"))
cat("Optimal threshold (Youden):", round(coords_best$threshold, 3), "\n")
cat("Sensitivity at optimal:", round(coords_best$sensitivity, 3), "\n")
cat("Specificity at optimal:", round(coords_best$specificity, 3), "\n")
```

#### Python

```{python}
#| label: roc-curve-py
#| eval: false

from sklearn.metrics import roc_curve, roc_auc_score
import matplotlib.pyplot as plt
import numpy as np
from scipy.special import expit

np.random.seed(42)
n = 1000
true_outcome = np.random.binomial(1, 0.15, n)
pred_prob = expit(np.random.normal(-1 + 2 * true_outcome, 1.2))

fpr, tpr, thresholds = roc_curve(true_outcome, pred_prob)
auc_score = roc_auc_score(true_outcome, pred_prob)

plt.figure(figsize=(7, 7))
plt.plot(fpr, tpr, color="steelblue", lw=2,
         label=f"Model (AUC = {auc_score:.3f})")
plt.plot([0, 1], [0, 1], color="grey", linestyle="--", label="Chance")
plt.xlabel("False Positive Rate (1 - Specificity)")
plt.ylabel("True Positive Rate (Sensitivity)")
plt.title("ROC Curve")
plt.legend(loc="lower right")
plt.tight_layout()
plt.show()

# Youden's J statistic to find optimal threshold
j_scores = tpr - fpr
best_idx = np.argmax(j_scores)
print(f"Optimal threshold (Youden): {thresholds[best_idx]:.3f}")
print(f"Sensitivity: {tpr[best_idx]:.3f}")
print(f"Specificity: {1 - fpr[best_idx]:.3f}")
```

:::

**What the code is showing.** This draws the ROC curve and pulls two numbers off it. The plot traces sensitivity against 1 − specificity as the threshold slides from high to low; a curve hugging the top-left corner is good, the diagonal is a coin flip. The reported AUC (with its 95% confidence interval) summarises the whole curve in one figure --- recall it is the probability the model scores a randomly chosen diseased patient higher than a non-diseased one, so the confidence interval matters because a wide one means the estimate could be much better or much worse than the point value suggests. The code also finds the threshold that maximises Youden's J (sensitivity + specificity − 1), reporting the sensitivity and specificity there. Treat that "optimal" threshold as a *statistical* suggestion only: as the later section warns, it assumes a false negative and a false positive cost the same, which is rarely true in medicine.

## Precision-Recall Curves

When outcomes are highly imbalanced --- as they often are in clinical prediction --- ROC curves can paint an overly optimistic picture. Because specificity is calculated over the large number of true negatives, even a substantial number of false positives barely moves the false positive rate. The **precision-recall (PR) curve** addresses this by focusing entirely on the positive class.

- **Precision** (= PPV): of all predicted positives, how many are truly positive?
- **Recall** (= sensitivity): of all true positives, how many were detected?

The PR curve plots precision on the y-axis against recall on the x-axis. A good model maintains high precision even as recall increases. The baseline for a PR curve is a horizontal line at the prevalence level (not the diagonal, as for ROC).

The **Area Under the Precision-Recall Curve (AUPRC)** is a useful summary, particularly for rare outcomes. While a model may boast an AUC-ROC of 0.95, its AUPRC might reveal that achieving high recall requires accepting very low precision.

### Exercise: Comparing ROC and PR Curves

::: {.panel-tabset}

#### R

```{r}
#| label: pr-curve-r
#| eval: false

library(PRROC)

# Simulate an imbalanced dataset (2% prevalence)
set.seed(123)
n <- 5000
true_outcome <- rbinom(n, 1, 0.02)
pred_prob <- plogis(rnorm(n, mean = -2 + 3 * true_outcome, sd = 1.5))

# PR curve
pr <- pr.curve(scores.class0 = pred_prob[true_outcome == 1],
               scores.class1 = pred_prob[true_outcome == 0],
               curve = TRUE)
plot(pr, main = paste("Precision-Recall Curve\nAUPRC =", round(pr$auc.integral, 3)))

# For comparison, the ROC curve
roc <- roc.curve(scores.class0 = pred_prob[true_outcome == 1],
                 scores.class1 = pred_prob[true_outcome == 0],
                 curve = TRUE)
plot(roc, main = paste("ROC Curve\nAUROC =", round(roc$auc, 3)))

cat("Notice how the AUROC looks excellent but the AUPRC reveals\n")
cat("the difficulty of achieving high precision with rare outcomes.\n")
```

#### Python

```{python}
#| label: pr-curve-py
#| eval: false

from sklearn.metrics import (precision_recall_curve, average_precision_score,
                              roc_curve, roc_auc_score)
import matplotlib.pyplot as plt
import numpy as np
from scipy.special import expit

np.random.seed(123)
n = 5000
true_outcome = np.random.binomial(1, 0.02, n)  # 2% prevalence
pred_prob = expit(np.random.normal(-2 + 3 * true_outcome, 1.5))

fig, axes = plt.subplots(1, 2, figsize=(14, 6))

# ROC curve
fpr, tpr, _ = roc_curve(true_outcome, pred_prob)
auroc = roc_auc_score(true_outcome, pred_prob)
axes[0].plot(fpr, tpr, color="steelblue", lw=2)
axes[0].plot([0, 1], [0, 1], "--", color="grey")
axes[0].set_title(f"ROC Curve (AUROC = {auroc:.3f})")
axes[0].set_xlabel("False Positive Rate")
axes[0].set_ylabel("True Positive Rate")

# PR curve
precision, recall, _ = precision_recall_curve(true_outcome, pred_prob)
auprc = average_precision_score(true_outcome, pred_prob)
prevalence = true_outcome.mean()
axes[1].plot(recall, precision, color="darkorange", lw=2)
axes[1].axhline(y=prevalence, linestyle="--", color="grey", label=f"Baseline (prevalence={prevalence:.2f})")
axes[1].set_title(f"Precision-Recall Curve (AUPRC = {auprc:.3f})")
axes[1].set_xlabel("Recall (Sensitivity)")
axes[1].set_ylabel("Precision (PPV)")
axes[1].legend()

plt.tight_layout()
plt.show()

print("The AUROC looks impressive, but AUPRC reveals the real challenge.")
```

:::

**What the code is showing.** This is a deliberate side-by-side demonstration of how ROC curves can flatter a model on rare outcomes. We simulate a realistically imbalanced dataset --- only 2% of 5,000 patients have the event --- and plot the ROC and precision-recall curves for the *same* predictions. The ROC curve and its AUROC look excellent, because specificity is computed against the huge pool of true negatives, so even hundreds of false positives barely nudge the false-positive rate. The PR curve tells the uncomfortable truth: its baseline is the prevalence (the dashed line at 0.02), and the curve shows that to catch most true cases (high recall) you must wade through many false positives (precision collapses). The much lower AUPRC is the honest summary for a rare-disease screening problem. The lesson for a health researcher: when your outcome is rare, report the precision-recall curve, not just the ROC.

## Choosing the Threshold: A Clinical Decision

A prediction model typically outputs a probability. To make a binary decision (treat or don't treat, refer or don't refer), you must choose a **threshold**. This is fundamentally a **clinical** decision, not a statistical one.

Consider a model predicting 30-day mortality after surgery:

- If the intervention for high-risk patients is simply closer monitoring (low cost, minimal harm), you might choose a **low threshold** (e.g., 5%), accepting many false positives to catch as many at-risk patients as possible.
- If the intervention is a risky re-operation, you might choose a **high threshold** (e.g., 30%), requiring strong evidence before acting.

The optimal threshold depends on the **relative costs** of false positives and false negatives, which are determined by the clinical context, not by the data.

::: {.callout-note}
## Youden's Index Is Not Always Optimal
Youden's J statistic (sensitivity + specificity - 1) finds the threshold that maximises the sum of sensitivity and specificity. This implicitly assumes equal costs for false positives and false negatives --- an assumption that is rarely true in medicine. Use Youden's index as a starting point, but always adjust based on clinical reasoning.
:::

### The Threshold-Performance Trade-off

::: {.panel-tabset}

#### R

```{r}
#| label: threshold-tradeoff-r
#| eval: false

library(pROC)

set.seed(42)
n <- 1000
true_outcome <- rbinom(n, 1, 0.15)
pred_prob <- plogis(rnorm(n, mean = -1 + 2 * true_outcome, sd = 1.2))

roc_obj <- roc(true_outcome, pred_prob)

# Extract sensitivity and specificity at various thresholds
thresholds <- seq(0.05, 0.95, by = 0.05)
results <- data.frame(
  threshold = thresholds,
  sensitivity = sapply(thresholds, function(t) {
    coords(roc_obj, t, input = "threshold", ret = "sensitivity")
  }),
  specificity = sapply(thresholds, function(t) {
    coords(roc_obj, t, input = "threshold", ret = "specificity")
  })
)

# Plot the trade-off
plot(results$threshold, results$sensitivity, type = "l", lwd = 2, col = "red",
     xlab = "Classification Threshold", ylab = "Metric Value",
     main = "Sensitivity-Specificity Trade-off Across Thresholds",
     ylim = c(0, 1), las = 1)
lines(results$threshold, results$specificity, lwd = 2, col = "blue")
legend("right", legend = c("Sensitivity", "Specificity"),
       col = c("red", "blue"), lwd = 2)
```

#### Python

```{python}
#| label: threshold-tradeoff-py
#| eval: false

import numpy as np
import matplotlib.pyplot as plt
from sklearn.metrics import roc_curve
from scipy.special import expit

np.random.seed(42)
n = 1000
true_outcome = np.random.binomial(1, 0.15, n)
pred_prob = expit(np.random.normal(-1 + 2 * true_outcome, 1.2))

fpr, tpr, thresholds = roc_curve(true_outcome, pred_prob)
specificity = 1 - fpr

plt.figure(figsize=(8, 5))
plt.plot(thresholds, tpr[:-1] if len(tpr) > len(thresholds) else tpr,
         color="red", lw=2, label="Sensitivity")
plt.plot(thresholds, specificity[:-1] if len(specificity) > len(thresholds) else specificity,
         color="blue", lw=2, label="Specificity")
plt.xlabel("Classification Threshold")
plt.ylabel("Metric Value")
plt.title("Sensitivity-Specificity Trade-off Across Thresholds")
plt.legend()
plt.xlim(0, 1)
plt.ylim(0, 1)
plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** This makes the threshold trade-off visible as a single picture. We step the classification threshold across its range and, at each value, record the model's sensitivity and specificity, then plot both as lines on the same axes. The two curves move in opposite directions: as you raise the threshold (demand stronger evidence before calling a patient positive), sensitivity falls while specificity rises; lower it, and the reverse happens. Where the lines cross is the threshold that balances the two, but --- as the surrounding text stresses --- the *right* operating point is wherever the clinical costs of a missed case and a false alarm balance, which may be far from the crossing point. Reading this plot lets you pick a threshold deliberately rather than defaulting to 0.5.

## Fairness: Does the Model Work for Everyone?

A model that performs well on average may perform very differently across subgroups defined by age, sex, race, ethnicity, or socioeconomic status. This is not a theoretical concern --- it has been documented repeatedly in clinical prediction models.

A well-known example is the pulse oximeter, which systematically overestimates blood oxygen saturation in patients with darker skin pigmentation. Prediction models trained on predominantly White populations may have lower sensitivity for detecting disease in Black or Hispanic patients, potentially widening existing health disparities.

### Key Fairness Concepts

- **Equal performance**: Does the model have similar sensitivity and specificity across demographic groups?
- **Calibration fairness**: Are predicted probabilities equally well-calibrated in each group? A model might predict 20% risk for two groups, but if the actual event rate is 20% in one group and 30% in another, the model is miscalibrated for the latter.
- **Demographic parity**: Are positive predictions made at similar rates across groups? (Note: this may conflict with calibration if true prevalence differs.)

### Exercise: Evaluating Fairness

::: {.panel-tabset}

#### R

```{r}
#| label: fairness-r
#| eval: false

library(pROC)

# Simulate data with two demographic groups
set.seed(42)
n <- 2000

group <- sample(c("A", "B"), n, replace = TRUE)
# Group B has slightly different disease prevalence and predictor distribution
true_outcome <- ifelse(group == "A",
                       rbinom(n, 1, 0.10),
                       rbinom(n, 1, 0.15))
# Model performs slightly worse for Group B
pred_prob <- ifelse(group == "A",
                    plogis(rnorm(n, -1.5 + 2.5 * true_outcome, 1.0)),
                    plogis(rnorm(n, -1.5 + 1.8 * true_outcome, 1.2)))

# Calculate AUC by group
for (g in c("A", "B")) {
  idx <- group == g
  roc_g <- roc(true_outcome[idx], pred_prob[idx])
  cat(sprintf("Group %s: AUC = %.3f, Prevalence = %.1f%%\n",
              g, auc(roc_g), mean(true_outcome[idx]) * 100))
}

# Compare sensitivity at a fixed threshold
threshold <- 0.3
for (g in c("A", "B")) {
  idx <- group == g
  pred_class <- ifelse(pred_prob[idx] >= threshold, 1, 0)
  sens <- sum(pred_class == 1 & true_outcome[idx] == 1) / sum(true_outcome[idx] == 1)
  spec <- sum(pred_class == 0 & true_outcome[idx] == 0) / sum(true_outcome[idx] == 0)
  cat(sprintf("Group %s at threshold %.1f: Sensitivity = %.3f, Specificity = %.3f\n",
              g, threshold, sens, spec))
}
```

#### Python

```{python}
#| label: fairness-py
#| eval: false

import numpy as np
from sklearn.metrics import roc_auc_score
from scipy.special import expit

np.random.seed(42)
n = 2000

group = np.random.choice(["A", "B"], n)
true_outcome = np.where(group == "A",
                        np.random.binomial(1, 0.10, n),
                        np.random.binomial(1, 0.15, n))
pred_prob = np.where(group == "A",
                     expit(np.random.normal(-1.5 + 2.5 * true_outcome, 1.0)),
                     expit(np.random.normal(-1.5 + 1.8 * true_outcome, 1.2)))

# AUC by group
for g in ["A", "B"]:
    mask = group == g
    auc = roc_auc_score(true_outcome[mask], pred_prob[mask])
    prev = true_outcome[mask].mean()
    print(f"Group {g}: AUC = {auc:.3f}, Prevalence = {prev*100:.1f}%")

# Sensitivity and specificity at a fixed threshold
threshold = 0.3
for g in ["A", "B"]:
    mask = group == g
    pred_class = (pred_prob[mask] >= threshold).astype(int)
    tp = ((pred_class == 1) & (true_outcome[mask] == 1)).sum()
    fn = ((pred_class == 0) & (true_outcome[mask] == 1)).sum()
    tn = ((pred_class == 0) & (true_outcome[mask] == 0)).sum()
    fp = ((pred_class == 1) & (true_outcome[mask] == 0)).sum()
    sens = tp / (tp + fn) if (tp + fn) > 0 else 0
    spec = tn / (tn + fp) if (tn + fp) > 0 else 0
    print(f"Group {g} at threshold {threshold}: Sensitivity = {sens:.3f}, Specificity = {spec:.3f}")
```

:::

**What the code is showing.** This is a minimal but realistic fairness audit. We simulate two groups, A and B, where group B has both a different disease prevalence and a model that genuinely discriminates a little worse (we built that gap in deliberately). The first loop reports the AUC and prevalence *separately* for each group --- the single overall AUC would have hidden the disparity. The second loop fixes one threshold and reports sensitivity and specificity per group, exposing the more clinically alarming problem: at the same cut-off, the model may catch a smaller fraction of true cases in group B, meaning patients in that group are more likely to be missed. The takeaway for a health researcher is procedural: always disaggregate your performance metrics by the demographic groups you care about, because a model that looks fine on average can quietly underperform for a subgroup and widen existing health inequities.

## Putting It All Together: A Complete Evaluation

When evaluating a clinical prediction model, no single metric tells the whole story. The following checklist summarises what you should report:

1. **Sample description**: How many patients? How many events? What is the prevalence?
2. **Discrimination**: AUC-ROC with confidence interval. Consider AUPRC for rare outcomes.
3. **Calibration**: Are predicted probabilities accurate? (Covered in detail in @sec-performance-validation.)
4. **Confusion matrix**: At a clinically relevant threshold, not just the default 0.5.
5. **Sensitivity, specificity, PPV, NPV**: At the chosen threshold.
6. **Subgroup performance**: Does the model perform equitably across key demographic groups?
7. **Clinical utility**: Would using this model lead to better decisions than the alternatives? (Covered in @sec-performance-validation.)

::: {.callout-important}
## Remember
A model is not good or bad in isolation. It is good or bad **for a specific purpose, in a specific population, at a specific threshold**. Always evaluate with the intended clinical use in mind.
:::

## Summary

- **Accuracy** is misleading for imbalanced outcomes; decompose errors into sensitivity and specificity.
- **Sensitivity** matters most for screening; **specificity** matters most for confirmation.
- **PPV and NPV** depend on prevalence, so a test's clinical usefulness changes with the population.
- **ROC curves** display the sensitivity-specificity trade-off across all thresholds; **AUC** summarises discriminative ability.
- **Precision-recall curves** are more informative than ROC for rare outcomes.
- The **classification threshold** should be chosen based on clinical consequences, not statistical optimality.
- **Fairness** requires checking that model performance is consistent across demographic groups.

## References and Further Reading

The foundational concepts of sensitivity, specificity, and predictive values are covered in every epidemiology textbook, but their application to prediction models requires additional nuance. Van Calster and colleagues provide an excellent overview of performance assessment in their 2025 paper in The Lancet Digital Health, which organises evaluation into five domains: discrimination, calibration, overall performance, classification, and clinical utility. This framework is essential reading for anyone developing or evaluating clinical prediction models. The comprehensive textbook by Smits, van Kuijk, and Wynants (2026) devotes several chapters to model evaluation and provides practical guidance on choosing appropriate metrics for different clinical contexts. For the statistical foundations of ROC analysis, the classic text by Pepe (2003), "The Statistical Evaluation of Medical Tests for Classification and Prediction," remains the definitive reference. Saito and Rehmsmeier (2015) explain why precision-recall curves are preferable to ROC curves for imbalanced datasets in their paper "The Precision-Recall Plot Is More Informative than the ROC Plot When Evaluating Binary Classifiers on Imbalanced Datasets" in PLOS ONE. On the topic of fairness in clinical prediction, Obermeyer and colleagues published a landmark 2019 paper in Science demonstrating racial bias in a widely used healthcare algorithm, which is required reading for anyone working in this space. Vickers and Elkin (2006) introduced decision curve analysis and net benefit as tools for evaluating clinical utility, which we discuss further in the next chapter.
