# Performance Assessment and Model Validation {#sec-performance-validation}

## Introduction

You have developed a clinical prediction model. Its coefficients look reasonable, the predictors make clinical sense, and the apparent performance seems promising. But how good is this model, really? And will it work on patients beyond your development dataset?

This chapter addresses these questions by covering the five domains of model performance --- discrimination, calibration, overall performance, classification, and clinical utility --- and the methods for internal and external validation. The structure follows the comprehensive framework proposed by Van Calster and colleagues (2025) in The Lancet Digital Health, which we strongly recommend as a companion reading.

The core message is this: **no single metric tells the full story.** A model with excellent discrimination (high AUC) can still produce dangerously miscalibrated risk predictions. A well-calibrated model might not discriminate well enough to be useful. Clinical utility must be assessed separately from statistical performance. You need the full picture.

## The Five Performance Domains

Van Calster et al. (2025) organise model performance assessment into five complementary domains:

| Domain | Question | Key Metrics |
|--------|----------|-------------|
| **Discrimination** | Can the model rank patients by risk? | C-statistic, AUC |
| **Calibration** | Are predicted probabilities accurate? | Calibration plot, O:E ratio, calibration slope |
| **Overall performance** | How close are predictions to observed outcomes? | Brier score, Nagelkerke R-squared |
| **Classification** | How well does the model classify at a threshold? | Sensitivity, specificity, PPV, NPV |
| **Clinical utility** | Does using the model improve clinical decisions? | Net benefit, decision curve analysis |

We will examine each in detail.

## Domain 1: Discrimination

### The C-Statistic / AUC

Discrimination measures how well a model separates patients who experience the outcome from those who do not. The most common measure is the **concordance statistic (C-statistic)**, which for binary outcomes is equivalent to the area under the ROC curve (AUC) discussed in @sec-model-evaluation.

Recall the interpretation: the C-statistic is the probability that a randomly selected patient with the outcome has a higher predicted probability than a randomly selected patient without the outcome.

::: {.panel-tabset}

### R

```{r}
#| label: discrimination-r
#| message: false

library(rms)

# Simulate clinical data: predicting 30-day mortality after stroke
set.seed(2024)
n <- 1500

stroke_data <- data.frame(
  age = round(rnorm(n, 72, 12)),
  nihss = round(pmax(0, rnorm(n, 8, 6))),       # NIH Stroke Scale
  glucose = round(rnorm(n, 140, 50)),
  afib = rbinom(n, 1, 0.25),                      # Atrial fibrillation
  thrombolysis = rbinom(n, 1, 0.30)
)

# True model for 30-day mortality
lp <- -5 + 0.04 * stroke_data$age +
  0.12 * stroke_data$nihss +
  0.003 * stroke_data$glucose +
  0.3 * stroke_data$afib -
  0.5 * stroke_data$thrombolysis

stroke_data$death_30d <- rbinom(n, 1, plogis(lp))
cat("30-day mortality rate:", mean(stroke_data$death_30d), "\n")

# Fit prediction model
dd <- datadist(stroke_data)
options(datadist = "dd")

fit <- lrm(death_30d ~ age + nihss + glucose + afib + thrombolysis,
           data = stroke_data, x = TRUE, y = TRUE)

cat("C-statistic (apparent):", fit$stats["C"], "\n")
```

### Python

```{python}
#| label: discrimination-py

import numpy as np
import pandas as pd
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import roc_auc_score
from scipy.special import expit

np.random.seed(2024)
n = 1500

stroke_data = pd.DataFrame({
    "age": np.round(np.random.normal(72, 12, n)),
    "nihss": np.round(np.maximum(0, np.random.normal(8, 6, n))),
    "glucose": np.round(np.random.normal(140, 50, n)),
    "afib": np.random.binomial(1, 0.25, n),
    "thrombolysis": np.random.binomial(1, 0.30, n)
})

lp = (-5 + 0.04 * stroke_data["age"] +
      0.12 * stroke_data["nihss"] +
      0.003 * stroke_data["glucose"] +
      0.3 * stroke_data["afib"] -
      0.5 * stroke_data["thrombolysis"])

stroke_data["death_30d"] = np.random.binomial(1, expit(lp))
print(f"30-day mortality rate: {stroke_data['death_30d'].mean():.3f}")

predictors = ["age", "nihss", "glucose", "afib", "thrombolysis"]
X = stroke_data[predictors].values
y = stroke_data["death_30d"].values

model = LogisticRegression(max_iter=5000, random_state=42)
model.fit(X, y)

pred_prob = model.predict_proba(X)[:, 1]
auc = roc_auc_score(y, pred_prob)
print(f"C-statistic (apparent): {auc:.3f}")
```

:::

### Limitations of the C-Statistic

The C-statistic has important limitations that Van Calster et al. (2025) highlight:

1. **Insensitive to calibration:** A model can perfectly rank patients (C = 1.0) but assign completely wrong probabilities.
2. **Context-blind:** It weights all parts of the ROC curve equally, even clinically irrelevant regions.
3. **Hard to improve:** In many clinical domains, the achievable C-statistic is inherently limited. Adding a new biomarker might genuinely improve predictions but barely move the C-statistic.
4. **Misleading comparisons:** A C-statistic of 0.75 in one population is not comparable to 0.75 in another if the case-mix differs.

## Domain 2: Calibration

### Why Calibration Matters More Than You Think

Calibration assesses whether predicted probabilities match observed outcomes. If your model says a patient has a 20% risk, do approximately 20 out of 100 such patients actually experience the outcome?

This matters enormously in clinical practice. When a clinician tells a patient "your model-predicted 10-year cardiovascular risk is 15%," the patient and clinician rely on that number being accurate --- not just the patient's relative ranking. **Risk-based treatment decisions require calibrated predictions.**

### Calibration-in-the-Large: The O:E Ratio

The simplest calibration check is the **observed-to-expected (O:E) ratio:**

$$
\text{O:E ratio} = \frac{\text{Observed event rate}}{\text{Mean predicted probability}}
$$

An O:E ratio of 1 indicates perfect calibration-in-the-large. An O:E ratio of 1.5 means the model underestimates risk by 50%.

### Calibration Slope

The **calibration slope** is obtained by regressing the outcome on the linear predictor (log-odds of predictions). A calibration slope of 1 indicates perfect calibration. A slope less than 1 indicates overfitting (predictions are too extreme --- high risks are too high, low risks are too low). A slope greater than 1 suggests the opposite.

### Calibration Plots

The most informative way to assess calibration is visually, through a **calibration plot.** This plots predicted probabilities (x-axis) against observed proportions (y-axis). A perfectly calibrated model follows the 45-degree diagonal.

Two types are common:

- **Grouped calibration plot:** Patients are divided into groups (e.g., deciles) by predicted probability, and the observed proportion within each group is plotted.
- **Smoothed calibration plot:** A smooth curve (e.g., loess or spline) is fit to the predicted vs. observed data.

::: {.panel-tabset}

### R

```{r}
#| label: calibration-r
#| message: false
#| fig-cap: "Calibration plot showing observed outcomes versus predicted probabilities."

library(rms)

# Use the stroke model from above
pred_prob <- predict(fit, type = "fitted")

# Grouped calibration plot
cal <- val.prob(pred_prob, stroke_data$death_30d,
                m = 100, cex = 0.5,
                main = "Calibration Plot: 30-Day Stroke Mortality Model")
```

```{r}
#| label: calibration-metrics-r

# Calculate calibration metrics manually
obs_rate <- mean(stroke_data$death_30d)
mean_pred <- mean(pred_prob)

cat("Observed event rate:", round(obs_rate, 3), "\n")
cat("Mean predicted probability:", round(mean_pred, 3), "\n")
cat("O:E ratio:", round(obs_rate / mean_pred, 3), "\n")

# Calibration slope
cal_model <- glm(stroke_data$death_30d ~ qlogis(pred_prob),
                 family = binomial)
cat("Calibration slope:", round(coef(cal_model)[2], 3), "\n")
cat("Calibration intercept:", round(coef(cal_model)[1], 3), "\n")

# Brier score
brier <- mean((pred_prob - stroke_data$death_30d)^2)
cat("Brier score:", round(brier, 4), "\n")
```

### Python

```{python}
#| label: calibration-py
#| fig-cap: "Calibration plot showing observed outcomes versus predicted probabilities."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.calibration import calibration_curve
from sklearn.metrics import brier_score_loss

# Use the stroke model predictions from above
pred_prob = model.predict_proba(X)[:, 1]
y_true = stroke_data["death_30d"].values

# Grouped calibration curve
prob_true, prob_pred = calibration_curve(y_true, pred_prob,
                                         n_bins=10, strategy="quantile")

fig, axes = plt.subplots(1, 2, figsize=(14, 6))

# Left panel: calibration plot
axes[0].plot(prob_pred, prob_true, "o-", color="steelblue", lw=2,
             markersize=8, label="Model")
axes[0].plot([0, 1], [0, 1], "--", color="grey", label="Perfect calibration")
axes[0].set_xlabel("Mean Predicted Probability")
axes[0].set_ylabel("Observed Proportion")
axes[0].set_title("Calibration Plot (Decile Groups)")
axes[0].legend()
axes[0].set_xlim(0, max(prob_pred) * 1.1)
axes[0].set_ylim(0, max(prob_true) * 1.1)

# Right panel: distribution of predicted probabilities
axes[1].hist(pred_prob[y_true == 0], bins=30, alpha=0.5, color="steelblue",
             label="No event", density=True)
axes[1].hist(pred_prob[y_true == 1], bins=30, alpha=0.5, color="coral",
             label="Event", density=True)
axes[1].set_xlabel("Predicted Probability")
axes[1].set_ylabel("Density")
axes[1].set_title("Distribution of Predicted Probabilities")
axes[1].legend()

plt.tight_layout()
plt.show()

# Calibration metrics
obs_rate = y_true.mean()
mean_pred = pred_prob.mean()
oe_ratio = obs_rate / mean_pred
brier = brier_score_loss(y_true, pred_prob)

print(f"\nCalibration metrics:")
print(f"  Observed event rate: {obs_rate:.3f}")
print(f"  Mean predicted probability: {mean_pred:.3f}")
print(f"  O:E ratio: {oe_ratio:.3f}")
print(f"  Brier score: {brier:.4f}")
```

:::

### Interpreting Miscalibration

Common patterns and their clinical meaning:

- **O:E > 1 (underestimation):** The model systematically underestimates risk. Patients may not receive treatments they need.
- **O:E < 1 (overestimation):** The model systematically overestimates risk. Patients may receive unnecessary treatments.
- **Calibration slope < 1:** Predictions are too extreme. Patients at high predicted risk actually have lower risk than predicted, and patients at low predicted risk have higher risk. This is the hallmark of overfitting.
- **Calibration slope > 1:** Predictions are too moderate. This is less common and may occur after excessive shrinkage.

## Domain 3: Overall Performance

The **Brier score** combines discrimination and calibration into a single measure:

$$
\text{Brier score} = \frac{1}{N} \sum_{i=1}^{N} (p_i - y_i)^2
$$

where $p_i$ is the predicted probability and $y_i$ is the observed outcome (0 or 1). The Brier score ranges from 0 (perfect) to 1 (worst). For a reference, a model that always predicts the overall event rate achieves a Brier score equal to the prevalence times (1 minus the prevalence).

The **scaled Brier score** (or Brier skill score) expresses improvement over a non-informative model:

$$
\text{Scaled Brier} = 1 - \frac{\text{Brier score}}{\text{Brier}_{\text{max}}}
$$

where $\text{Brier}_{\text{max}} = \bar{y}(1 - \bar{y})$.

## Domain 4: Classification

Classification metrics (sensitivity, specificity, PPV, NPV) were covered in @sec-model-evaluation. The key point for model validation is that classification performance depends entirely on the chosen threshold. When reporting classification metrics, **always specify the threshold and justify its choice based on the clinical context.**

## Domain 5: Clinical Utility

### Net Benefit and Decision Curve Analysis

A model might have good discrimination and calibration, yet using it may not actually improve clinical decisions. **Decision curve analysis (DCA)**, proposed by Vickers and Elkin (2006), addresses this by quantifying the **net benefit** of using the model compared to the default strategies of treating all patients or treating no patients.

The net benefit at a given threshold probability $p_t$ is:

$$
\text{Net benefit} = \frac{TP}{N} - \frac{FP}{N} \times \frac{p_t}{1 - p_t}
$$

The term $\frac{p_t}{1 - p_t}$ represents the exchange rate between false positives and true positives, determined by the threshold probability. If the threshold is 10%, you are willing to accept up to 9 false positives for every true positive identified.

::: {.panel-tabset}

### R

```{r}
#| label: dca-r
#| message: false
#| fig-cap: "Decision curve analysis comparing the prediction model to default strategies."

library(dcurves)

# Add predicted probabilities to the dataset
stroke_data$pred_risk <- predict(fit, type = "fitted")

# Decision curve analysis
dca_result <- dca(death_30d ~ pred_risk,
                  data = stroke_data,
                  thresholds = seq(0, 0.5, by = 0.01))

plot(dca_result,
     smooth = TRUE,
     show_ggplot_code = FALSE) +
  ggplot2::ggtitle("Decision Curve Analysis:\n30-Day Stroke Mortality Model")
```

### Python

```{python}
#| label: dca-py
#| fig-cap: "Decision curve analysis comparing the prediction model to default strategies."

import numpy as np
import matplotlib.pyplot as plt

def net_benefit(y_true, y_pred, threshold):
    """Calculate net benefit at a given threshold."""
    n = len(y_true)
    pred_pos = (y_pred >= threshold).astype(int)
    tp = np.sum((pred_pos == 1) & (y_true == 1))
    fp = np.sum((pred_pos == 1) & (y_true == 0))
    nb = tp / n - fp / n * (threshold / (1 - threshold))
    return nb

thresholds = np.arange(0.01, 0.51, 0.01)
pred_prob_model = model.predict_proba(X)[:, 1]
y_true = stroke_data["death_30d"].values
n_total = len(y_true)

# Net benefit for model
nb_model = [net_benefit(y_true, pred_prob_model, t) for t in thresholds]

# Net benefit for "treat all" strategy
nb_all = [(y_true.mean() - (1 - y_true.mean()) * t / (1 - t))
          for t in thresholds]

# Net benefit for "treat none" is always 0

plt.figure(figsize=(9, 6))
plt.plot(thresholds, nb_model, color="steelblue", lw=2,
         label="Prediction Model")
plt.plot(thresholds, nb_all, color="coral", lw=2, ls="--",
         label="Treat All")
plt.axhline(y=0, color="black", lw=1, label="Treat None")
plt.xlabel("Threshold Probability")
plt.ylabel("Net Benefit")
plt.title("Decision Curve Analysis: 30-Day Stroke Mortality Model")
plt.legend()
plt.ylim(-0.05, max(max(nb_model), max(nb_all)) * 1.1)
plt.tight_layout()
plt.show()
```

:::

### Interpreting the Decision Curve

The decision curve shows net benefit (y-axis) across a range of threshold probabilities (x-axis). Three strategies are compared:

- **Treat none** (horizontal line at 0): no patients receive the intervention.
- **Treat all** (declining curve): all patients receive the intervention regardless of risk.
- **Model-guided** (the model's curve): only patients above the threshold receive the intervention.

The model is clinically useful at thresholds where its net benefit exceeds both the "treat all" and "treat none" lines. If the model curve is always below one of the default strategies, the model adds no value.

## Internal Validation

### Why Internal Validation Is Necessary

As discussed in @sec-building-prediction-models, apparent performance (measured on the training data) is optimistically biased. Internal validation estimates and corrects for this optimism. It does not replace external validation, but it provides a more honest assessment of likely performance.

### Bootstrap Optimism Correction

The recommended approach (Steyerberg 2019, Van Calster et al. 2025) is **bootstrap optimism correction:**

1. Draw a bootstrap sample (with replacement) from the original data.
2. Develop the model in the bootstrap sample (including all modelling decisions).
3. Measure performance in the bootstrap sample (apparent bootstrap performance).
4. Apply the bootstrap model to the original data (test performance).
5. Optimism = apparent bootstrap performance minus test performance.
6. Repeat steps 1--5 many times (e.g., 200+).
7. Average optimism across repetitions.
8. Corrected performance = original apparent performance minus average optimism.

::: {.panel-tabset}

### R

```{r}
#| label: bootstrap-validation-r
#| message: false

library(rms)

# Using the stroke model
fit <- lrm(death_30d ~ age + nihss + glucose + afib + thrombolysis,
           data = stroke_data, x = TRUE, y = TRUE)

# Bootstrap validation with 200 resamples
set.seed(42)
val <- validate(fit, B = 200)

cat("Bootstrap Validation Results:\n")
print(val)

# Extract optimism-corrected C-statistic
dxy_corrected <- val["Dxy", "index.corrected"]
c_corrected <- (dxy_corrected + 1) / 2

cat("\nApparent C-statistic:", fit$stats["C"], "\n")
cat("Optimism:", val["Dxy", "optimism"] / 2, "\n")
cat("Optimism-corrected C-statistic:", c_corrected, "\n")

# Calibration slope from bootstrap
cat("\nCalibration slope (optimism-corrected):",
    val["Slope", "index.corrected"], "\n")
```

### Python

```{python}
#| label: bootstrap-validation-py

import numpy as np
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import roc_auc_score, brier_score_loss

np.random.seed(42)

predictors = ["age", "nihss", "glucose", "afib", "thrombolysis"]
X = stroke_data[predictors].values
y = stroke_data["death_30d"].values
n = len(y)

# Apparent performance
full_model = LogisticRegression(max_iter=5000, random_state=42)
full_model.fit(X, y)
pred_full = full_model.predict_proba(X)[:, 1]
apparent_auc = roc_auc_score(y, pred_full)
apparent_brier = brier_score_loss(y, pred_full)

# Bootstrap optimism correction
n_boot = 200
optimism_auc = []
optimism_brier = []

for b in range(n_boot):
    boot_idx = np.random.choice(n, n, replace=True)
    X_boot, y_boot = X[boot_idx], y[boot_idx]

    boot_model = LogisticRegression(max_iter=5000, random_state=42)
    boot_model.fit(X_boot, y_boot)

    # Apparent performance on bootstrap sample
    pred_boot = boot_model.predict_proba(X_boot)[:, 1]
    auc_boot = roc_auc_score(y_boot, pred_boot)
    brier_boot = brier_score_loss(y_boot, pred_boot)

    # Test performance on original data
    pred_orig = boot_model.predict_proba(X)[:, 1]
    auc_orig = roc_auc_score(y, pred_orig)
    brier_orig = brier_score_loss(y, pred_orig)

    optimism_auc.append(auc_boot - auc_orig)
    optimism_brier.append(brier_boot - brier_orig)

mean_opt_auc = np.mean(optimism_auc)
mean_opt_brier = np.mean(optimism_brier)
corrected_auc = apparent_auc - mean_opt_auc
corrected_brier = apparent_brier - mean_opt_brier

print(f"Apparent AUC:              {apparent_auc:.3f}")
print(f"Mean optimism (AUC):       {mean_opt_auc:.4f}")
print(f"Corrected AUC:             {corrected_auc:.3f}")
print(f"\nApparent Brier:            {apparent_brier:.4f}")
print(f"Mean optimism (Brier):     {mean_opt_brier:.4f}")
print(f"Corrected Brier:           {corrected_brier:.4f}")
```

:::

### Cross-Validation

Cross-validation (typically 10-fold) provides an alternative internal validation approach. The data is split into $k$ folds; each fold serves once as the validation set while the remaining folds are used for model development. The average performance across folds estimates the model's performance on new data.

Cross-validation is computationally simpler but has limitations for prediction model development: it does not directly yield a single final model or a clear calibration slope correction. Bootstrap optimism correction is generally preferred for clinical prediction models.

## External Validation

### Types of External Validation

Internal validation tells you how optimistic your apparent performance is. **External validation** tells you whether the model works on genuinely new data --- the real test. Van Calster et al. (2025) and Smits et al. (2026) distinguish:

- **Temporal validation:** Same setting, different time period. Example: model developed on 2015--2019 data, validated on 2020--2022 data. This captures changes in clinical practice and patient populations over time.
- **Geographical validation:** Different hospital or region, same time period. This tests transportability across settings.
- **Domain validation:** Different clinical context entirely (e.g., model developed in a tertiary care centre, validated in primary care). This is the most stringent test.

### What to Report in External Validation

Following Van Calster et al. (2025), an external validation study should report at minimum:

1. **C-statistic** with 95% confidence interval
2. **Calibration plot** (both grouped and smoothed)
3. **O:E ratio** with 95% confidence interval
4. **Calibration slope**
5. **Net benefit** (decision curve) at clinically relevant thresholds
6. **Distribution of predicted probabilities** (to understand the range and spread)
7. **Performance by subgroups** (age, sex, comorbidities, etc.)

## Model Updating

### When Performance Degrades

External validation often reveals that a model does not perform as well in new settings as it did internally. This does not necessarily mean the model is useless. **Model updating** can adapt an existing model to a new population without developing a completely new model.

### Updating Strategies

From least to most complex:

1. **Recalibration-in-the-large:** Adjust only the intercept. Corrects for differences in baseline risk (e.g., higher mortality rate in the new population).

2. **Logistic recalibration:** Adjust the intercept and apply a correction to the calibration slope. This corrects for both level and spread of predictions.

3. **Model revision:** Re-estimate some or all coefficients, or add new predictors. This requires larger sample sizes in the new population.

::: {.panel-tabset}

### R

```{r}
#| label: recalibration-r
#| message: false

# Simulate external validation data (different setting)
set.seed(99)
n_ext <- 800

# New population: older, sicker
ext_data <- data.frame(
  age = round(rnorm(n_ext, 78, 10)),   # Older
  nihss = round(pmax(0, rnorm(n_ext, 11, 7))),  # Higher severity
  glucose = round(rnorm(n_ext, 155, 55)),
  afib = rbinom(n_ext, 1, 0.35),
  thrombolysis = rbinom(n_ext, 1, 0.20)  # Less thrombolysis
)

lp_ext <- -5 + 0.04 * ext_data$age +
  0.12 * ext_data$nihss +
  0.003 * ext_data$glucose +
  0.3 * ext_data$afib -
  0.5 * ext_data$thrombolysis

ext_data$death_30d <- rbinom(n_ext, 1, plogis(lp_ext))

# Apply original model to external data
ext_data$pred_original <- predict(fit, newdata = ext_data, type = "fitted")

cat("External validation (before updating):\n")
cat("  Observed mortality:", mean(ext_data$death_30d), "\n")
cat("  Mean predicted:", mean(ext_data$pred_original), "\n")
cat("  O:E ratio:", round(mean(ext_data$death_30d) / mean(ext_data$pred_original), 3), "\n")

# Calibration slope in external data
lp_orig <- qlogis(ext_data$pred_original)
cal_ext <- glm(death_30d ~ lp_orig, data = ext_data, family = binomial)
cat("  Calibration slope:", round(coef(cal_ext)[2], 3), "\n")
cat("  Calibration intercept:", round(coef(cal_ext)[1], 3), "\n")

# Recalibration: update intercept and slope
ext_data$pred_recalibrated <- predict(cal_ext, type = "response")

cat("\nAfter logistic recalibration:\n")
cat("  Mean predicted:", round(mean(ext_data$pred_recalibrated), 3), "\n")
cat("  O:E ratio:", round(mean(ext_data$death_30d) / mean(ext_data$pred_recalibrated), 3), "\n")
```

### Python

```{python}
#| label: recalibration-py

import numpy as np
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import roc_auc_score

np.random.seed(99)
n_ext = 800

ext_data = {
    "age": np.round(np.random.normal(78, 10, n_ext)),
    "nihss": np.round(np.maximum(0, np.random.normal(11, 7, n_ext))),
    "glucose": np.round(np.random.normal(155, 55, n_ext)),
    "afib": np.random.binomial(1, 0.35, n_ext),
    "thrombolysis": np.random.binomial(1, 0.20, n_ext)
}

lp_ext = (-5 + 0.04 * ext_data["age"] +
          0.12 * ext_data["nihss"] +
          0.003 * ext_data["glucose"] +
          0.3 * ext_data["afib"] -
          0.5 * ext_data["thrombolysis"])

y_ext = np.random.binomial(1, expit(lp_ext))

# Apply original model
X_ext = np.column_stack([ext_data[k] for k in predictors])
pred_original = full_model.predict_proba(X_ext)[:, 1]

obs_rate = y_ext.mean()
mean_pred = pred_original.mean()
print(f"External validation (before updating):")
print(f"  Observed mortality: {obs_rate:.3f}")
print(f"  Mean predicted: {mean_pred:.3f}")
print(f"  O:E ratio: {obs_rate / mean_pred:.3f}")
print(f"  AUC: {roc_auc_score(y_ext, pred_original):.3f}")

# Logistic recalibration: regress outcome on logit(predictions)
from scipy.special import logit
lp_original = logit(np.clip(pred_original, 1e-6, 1 - 1e-6))

recal_model = LogisticRegression(max_iter=5000)
recal_model.fit(lp_original.reshape(-1, 1), y_ext)

pred_recalibrated = recal_model.predict_proba(lp_original.reshape(-1, 1))[:, 1]

print(f"\nAfter logistic recalibration:")
print(f"  Mean predicted: {pred_recalibrated.mean():.3f}")
print(f"  O:E ratio: {y_ext.mean() / pred_recalibrated.mean():.3f}")
print(f"  Calibration slope: {recal_model.coef_[0][0]:.3f}")
print(f"  Calibration intercept: {recal_model.intercept_[0]:.3f}")
```

:::

## Recommended Reporting: The Minimum Set

Based on Van Calster et al. (2025) and Smits et al. (2026), every prediction model evaluation should include at minimum:

1. **AUROC (C-statistic)** with confidence interval --- for discrimination
2. **Calibration plot** (smoothed and/or grouped) --- for calibration
3. **Decision curve** --- for clinical utility
4. **Risk distribution plot** --- to show the spread of predicted probabilities

These four visualisations together give a comprehensive picture. A model may look excellent on one plot and problematic on another --- which is exactly why you need all four.

::: {.panel-tabset}

### R

```{r}
#| label: four-plots-r
#| message: false
#| fig-height: 10
#| fig-width: 10
#| fig-cap: "The recommended four-panel evaluation display for clinical prediction models."

library(pROC)

pred_prob <- predict(fit, type = "fitted")
y_obs <- stroke_data$death_30d

par(mfrow = c(2, 2))

# Panel 1: ROC curve
roc_obj <- roc(y_obs, pred_prob, quiet = TRUE)
plot(roc_obj, col = "steelblue", lwd = 2,
     main = "A. Discrimination (ROC Curve)",
     print.auc = TRUE, print.auc.y = 0.4)

# Panel 2: Calibration plot
val.prob(pred_prob, y_obs, m = 100, cex = 0.5,
         main = "B. Calibration Plot")

# Panel 3: Risk distribution
hist(pred_prob[y_obs == 0], breaks = 30, col = rgb(0.4, 0.6, 0.8, 0.5),
     main = "C. Risk Distribution",
     xlab = "Predicted Probability", freq = FALSE, ylim = c(0, 12))
hist(pred_prob[y_obs == 1], breaks = 30, col = rgb(0.9, 0.4, 0.4, 0.5),
     add = TRUE, freq = FALSE)
legend("topright", legend = c("No event", "Event"),
       fill = c(rgb(0.4, 0.6, 0.8, 0.5), rgb(0.9, 0.4, 0.4, 0.5)))

# Panel 4: Simplified decision curve
thresholds <- seq(0.01, 0.5, by = 0.01)
nb_model <- nb_all <- numeric(length(thresholds))
n_total <- length(y_obs)

for (i in seq_along(thresholds)) {
  t <- thresholds[i]
  pred_pos <- as.numeric(pred_prob >= t)
  tp <- sum(pred_pos == 1 & y_obs == 1)
  fp <- sum(pred_pos == 1 & y_obs == 0)
  nb_model[i] <- tp / n_total - fp / n_total * (t / (1 - t))
  nb_all[i] <- mean(y_obs) - (1 - mean(y_obs)) * (t / (1 - t))
}

plot(thresholds, nb_model, type = "l", lwd = 2, col = "steelblue",
     main = "D. Decision Curve Analysis",
     xlab = "Threshold Probability", ylab = "Net Benefit",
     ylim = c(-0.05, max(nb_model, nb_all) * 1.1))
lines(thresholds, nb_all, lwd = 2, col = "coral", lty = 2)
abline(h = 0, col = "black")
legend("topright", legend = c("Model", "Treat All", "Treat None"),
       col = c("steelblue", "coral", "black"),
       lty = c(1, 2, 1), lwd = 2, cex = 0.8)

par(mfrow = c(1, 1))
```

### Python

```{python}
#| label: four-plots-py
#| fig-height: 10
#| fig-width: 10
#| fig-cap: "The recommended four-panel evaluation display for clinical prediction models."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.metrics import roc_curve, auc
from sklearn.calibration import calibration_curve

pred_prob = full_model.predict_proba(X)[:, 1]
y_obs = stroke_data["death_30d"].values

fig, axes = plt.subplots(2, 2, figsize=(12, 10))

# Panel A: ROC curve
fpr, tpr, _ = roc_curve(y_obs, pred_prob)
roc_auc = auc(fpr, tpr)
axes[0, 0].plot(fpr, tpr, color="steelblue", lw=2,
                label=f"AUC = {roc_auc:.3f}")
axes[0, 0].plot([0, 1], [0, 1], "--", color="grey")
axes[0, 0].set_xlabel("1 - Specificity")
axes[0, 0].set_ylabel("Sensitivity")
axes[0, 0].set_title("A. Discrimination (ROC Curve)")
axes[0, 0].legend()

# Panel B: Calibration plot
prob_true, prob_pred = calibration_curve(y_obs, pred_prob, n_bins=10,
                                          strategy="quantile")
axes[0, 1].plot(prob_pred, prob_true, "o-", color="steelblue", lw=2)
axes[0, 1].plot([0, 1], [0, 1], "--", color="grey")
axes[0, 1].set_xlabel("Mean Predicted Probability")
axes[0, 1].set_ylabel("Observed Proportion")
axes[0, 1].set_title("B. Calibration Plot")

# Panel C: Risk distribution
axes[1, 0].hist(pred_prob[y_obs == 0], bins=30, alpha=0.5,
                color="steelblue", label="No event", density=True)
axes[1, 0].hist(pred_prob[y_obs == 1], bins=30, alpha=0.5,
                color="coral", label="Event", density=True)
axes[1, 0].set_xlabel("Predicted Probability")
axes[1, 0].set_ylabel("Density")
axes[1, 0].set_title("C. Risk Distribution")
axes[1, 0].legend()

# Panel D: Decision curve
thresholds = np.arange(0.01, 0.51, 0.01)
nb_model = []
nb_all = []
n_total = len(y_obs)

for t in thresholds:
    pred_pos = (pred_prob >= t).astype(int)
    tp = np.sum((pred_pos == 1) & (y_obs == 1))
    fp = np.sum((pred_pos == 1) & (y_obs == 0))
    nb_model.append(tp / n_total - fp / n_total * (t / (1 - t)))
    nb_all.append(y_obs.mean() - (1 - y_obs.mean()) * (t / (1 - t)))

axes[1, 1].plot(thresholds, nb_model, color="steelblue", lw=2,
                label="Model")
axes[1, 1].plot(thresholds, nb_all, color="coral", lw=2, ls="--",
                label="Treat All")
axes[1, 1].axhline(y=0, color="black", lw=1, label="Treat None")
axes[1, 1].set_xlabel("Threshold Probability")
axes[1, 1].set_ylabel("Net Benefit")
axes[1, 1].set_title("D. Decision Curve Analysis")
axes[1, 1].legend(fontsize=9)

plt.tight_layout()
plt.show()
```

:::

## Exercises

### Exercise 1: Calibration Assessment

Using the stroke mortality model developed in this chapter:

a. Create a calibration plot using deciles of predicted risk. Is the model well-calibrated?
b. Calculate the O:E ratio. What does it tell you?
c. Calculate the calibration slope. Is there evidence of overfitting?
d. Apply bootstrap optimism correction. How much does the C-statistic decrease? How much does the calibration slope decrease?

### Exercise 2: External Validation Simulation

::: {.panel-tabset}

### R

```{r}
#| label: ex-external-val-r
#| eval: false

# Create three external validation populations that differ from development:
# Population A: Same demographics, 3 years later (temporal)
# Population B: Different hospital, younger patients (geographical)
# Population C: Primary care setting with lower severity (domain)

# For each population:
# a. Calculate C-statistic, O:E ratio, calibration slope
# b. Create calibration plots
# c. Determine which population shows worst calibration and explain why
# d. Perform logistic recalibration and show the improvement
```

### Python

```{python}
#| label: ex-external-val-py
#| eval: false

# Same exercise as R tab - create three external populations
# and assess model performance in each
```

:::

### Exercise 3: Decision Curve Interpretation

Consider a model for predicting preeclampsia in pregnant women. The model has an AUC of 0.82 and good calibration. You plot the decision curve.

a. At what range of threshold probabilities is the model useful?
b. A colleague argues that the model should not be used because the AUC is "only" 0.82. Using the decision curve, construct a counter-argument.
c. How would the decision curve change if the follow-up action (closer monitoring) were very low-cost versus very high-cost?

### Exercise 4: Complete Evaluation

Choose a clinical prediction model from the literature (e.g., QRISK3, Wells score, APACHE II). Using published validation data:

a. Report the C-statistic with confidence interval.
b. Describe the calibration (if a calibration plot is available).
c. Is there a decision curve analysis? If not, what threshold range would be clinically relevant?
d. Has the model been externally validated? In what populations?
e. Based on your assessment, would you recommend implementing this model in your setting?

## References and Further Reading

The primary framework for this chapter is drawn from Van Calster, McLernon, van Smeden, Worster,"; Steyerberg, and colleagues (2025) in The Lancet Digital Health, whose guide to assessing the performance of prediction models is the most comprehensive and up-to-date reference available on this topic, covering all five performance domains and providing clear recommendations for what to report. The textbook by Smits, van Kuijk, and Wynants (2026) complements this with a highly accessible treatment of internal and external validation, model updating, and practical guidance for applied researchers. Decision curve analysis was introduced by Vickers and Elkin (2006) in Medical Decision Making and is now considered an essential component of model evaluation; the `dcurves` R package and accompanying tutorials make it straightforward to implement. For calibration assessment specifically, the work of Van Calster, Nieboer, Vergouwe, De Cock, Pencina, and Steyerberg (2016) in the Journal of Clinical Epidemiology provides an excellent treatment of different calibration measures and their interpretation. Bootstrap validation methodology is thoroughly described in Steyerberg's "Clinical Prediction Models" (2019, Springer) and in Harrell's "Regression Modeling Strategies" (2015, Springer). The Brier score and its decomposition were formalised by Brier (1950) in Monthly Weather Review, and its application to clinical prediction models is discussed in Steyerberg and colleagues (2010) in Epidemiology.
