# PCA, t-SNE, and UMAP: Making Sense of High-Dimensional Data {#sec-dimensionality-reduction}

```{r}
#| include: false
library(tidyverse)
library(knitr)
```

Imagine you have measured 500 gene expression values, or two dozen lab tests, on each of your patients. You cannot plot 500 axes, and your eyes can only make sense of two or three. Dimensionality reduction is the set of techniques that squeeze those many measurements down to a handful of summary axes you *can* plot --- ideally without throwing away the patterns that matter. The reward is a picture: a single scatter plot where each dot is a patient, and where patients with similar profiles sit close together, often revealing subgroups, gradients, or outliers that were invisible in the raw spreadsheet. This chapter shows three ways to make that picture and, just as importantly, how to read it without being misled.

::: {.callout-note}
These chapters are set to `eval: false`, so the rendered page shows the code but not its live output. As you read, picture the figure each block *would* produce --- the surrounding text describes what to expect so you can run the code yourself and recognise a sensible result.
:::

## Introduction

The previous chapters covered supervised learning --- methods that predict an outcome from labelled data. In this part, we turn to **unsupervised** methods that explore and summarize data without a specific outcome in mind.

A single patient encounter can generate hundreds of variables: vital signs, lab results, imaging features, medication lists, genomic markers, clinical notes. When the number of features ($p$) is large relative to the number of observations ($n$), standard statistical methods break down. Regression models become unstable, distances between data points become uninformative, and visualisation is impossible.

**Dimensionality reduction** addresses this by projecting high-dimensional data into a lower-dimensional space while preserving as much meaningful structure as possible. This chapter covers three complementary methods: **PCA** (principal component analysis), **t-SNE** (t-distributed stochastic neighbour embedding), and **UMAP** (uniform manifold approximation and projection). Each serves a different purpose, and understanding when to use which is a core skill in modern data analysis.

## The Curse of Dimensionality

The "curse of dimensionality" refers to a collection of phenomena that arise when data live in high-dimensional spaces. The most important for clinical data analysis:

**Distances become meaningless.** In high dimensions, the distance between the nearest and farthest data points converges. If all points are approximately equidistant, distance-based methods (k-nearest neighbours, clustering) lose their ability to discriminate.

**Sparsity.** The volume of a high-dimensional space grows exponentially with the number of dimensions. Data points become increasingly sparse --- you need exponentially more observations to maintain the same data density. A study with 200 patients and 500 gene expression features has data that are extraordinarily sparse.

**Overfitting.** With many features, models can find spurious patterns in the noise. A logistic regression with 200 predictors and 100 observations is almost guaranteed to overfit, even with regularisation.

**Visualisation is impossible.** Humans perceive at most three spatial dimensions. To explore the structure of a 50-dimensional dataset, we must reduce it to 2 or 3 dimensions.

## PCA: Principal Component Analysis

PCA is the oldest and most widely used dimensionality reduction technique. It finds the directions (linear combinations of the original variables) along which the data vary the most.

### The Core Idea

Given a data matrix $\mathbf{X}$ with $n$ observations and $p$ variables:

1. **Centre** the data: subtract the mean of each variable.
2. **Compute the covariance matrix** $\mathbf{C} = \frac{1}{n-1}\mathbf{X}^T\mathbf{X}$.
3. **Find the eigenvectors** of $\mathbf{C}$. These are the **principal components** --- the directions of maximum variance.
4. **The eigenvalues** tell you how much variance each component captures.

The first principal component (PC1) captures the most variance, PC2 captures the second most (orthogonal to PC1), and so on. By keeping only the first $k$ components, you reduce the data from $p$ dimensions to $k$ dimensions.

In plain terms, a **principal component** is just a recipe: a new variable made by mixing the original measurements in fixed proportions (the weights in that recipe are called the **loadings**). PC1 is the single mixture that spreads patients out as much as possible --- the direction along which they differ most. If glucose, HbA1c, and triglycerides all rise and fall together, PCA notices and bundles them into one axis, so a patient's position on PC1 might summarise their whole "metabolic" profile in a single number. "Variance explained" is then simply the share of the data's total spread that each component accounts for: PC1 might capture 40% of all the variation, meaning that one axis alone tells you a large chunk of what distinguishes your patients.

### Key Properties

- PCA is a **linear** method: each component is a weighted sum of the original variables.
- It is **deterministic**: running PCA twice on the same data gives the same result (up to sign flips).
- It preserves **global structure** (distances and variance) well, but may miss non-linear patterns.

### Choosing the Number of Components

**Scree plot.** Plot the eigenvalues (or proportion of variance explained) against the component number. Look for an "elbow" --- a point where the curve flattens, suggesting that additional components contribute little.

**Cumulative variance threshold.** Keep enough components to explain a target percentage of total variance (commonly 80--90%).

**Kaiser criterion.** Keep components with eigenvalues > 1 (only meaningful when data are standardised).

### Clinical Example: Metabolic Panel PCA

A common application: reducing a panel of correlated lab values to a few interpretable dimensions. Consider a dataset with glucose, HbA1c, triglycerides, LDL, HDL, creatinine, ALT, and albumin.

::: {.panel-tabset}

### R

```{r}
#| label: fig-pca-metabolic-r
#| fig-cap: "PCA of a simulated metabolic panel: scree plot and biplot."
#| fig-width: 12
#| fig-height: 5

set.seed(42)
n <- 300

# Simulate correlated metabolic data
# Create correlation structure: glucose/HbA1c/triglycerides cluster;
# creatinine/albumin cluster
library(MASS)
Sigma <- matrix(c(
  1.0, 0.7, 0.5, 0.3,-0.2, 0.1, 0.2,-0.1,
  0.7, 1.0, 0.4, 0.2,-0.2, 0.1, 0.1,-0.1,
  0.5, 0.4, 1.0, 0.4,-0.3, 0.1, 0.3,-0.1,
  0.3, 0.2, 0.4, 1.0,-0.5, 0.1, 0.2, 0.0,
 -0.2,-0.2,-0.3,-0.5, 1.0,-0.1,-0.1, 0.2,
  0.1, 0.1, 0.1, 0.1,-0.1, 1.0, 0.1,-0.3,
  0.2, 0.1, 0.3, 0.2,-0.1, 0.1, 1.0,-0.2,
 -0.1,-0.1,-0.1, 0.0, 0.2,-0.3,-0.2, 1.0
), nrow = 8)

z <- mvrnorm(n, mu = rep(0, 8), Sigma = Sigma)
metabolic <- tibble(
  glucose      = round(z[,1] * 30 + 100),
  hba1c        = round(z[,2] * 1.0 + 5.8, 1),
  triglycerides = round(z[,3] * 50 + 150),
  ldl          = round(z[,4] * 30 + 120),
  hdl          = round(z[,5] * 12 + 55),
  creatinine   = round(z[,6] * 0.3 + 1.0, 2),
  alt          = round(z[,7] * 15 + 30),
  albumin      = round(z[,8] * 0.4 + 4.0, 1)
)

# PCA on scaled data
pca_result <- prcomp(metabolic, scale. = TRUE)

# Scree plot and biplot side by side
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Scree plot
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
barplot(var_explained, names.arg = paste0("PC", 1:8),
        col = "steelblue", main = "Scree Plot",
        ylab = "Proportion of Variance", xlab = "Component",
        ylim = c(0, 0.4))
abline(h = 1/8, lty = 2, col = "firebrick")  # Kaiser criterion line
text(9, 1/8 + 0.015, "Kaiser threshold", col = "firebrick", cex = 0.8)

# Biplot
biplot(pca_result, scale = 0, cex = 0.6,
       col = c("grey70", "firebrick"),
       main = "PCA Biplot: Metabolic Panel")
```

### Python

```{python}
#| label: fig-pca-metabolic-py
#| fig-cap: "PCA of a simulated metabolic panel: scree plot and biplot."

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA

np.random.seed(42)
n = 300

Sigma = np.array([
    [1.0, 0.7, 0.5, 0.3,-0.2, 0.1, 0.2,-0.1],
    [0.7, 1.0, 0.4, 0.2,-0.2, 0.1, 0.1,-0.1],
    [0.5, 0.4, 1.0, 0.4,-0.3, 0.1, 0.3,-0.1],
    [0.3, 0.2, 0.4, 1.0,-0.5, 0.1, 0.2, 0.0],
    [-0.2,-0.2,-0.3,-0.5, 1.0,-0.1,-0.1, 0.2],
    [0.1, 0.1, 0.1, 0.1,-0.1, 1.0, 0.1,-0.3],
    [0.2, 0.1, 0.3, 0.2,-0.1, 0.1, 1.0,-0.2],
    [-0.1,-0.1,-0.1, 0.0, 0.2,-0.3,-0.2, 1.0]
])

z = np.random.multivariate_normal(np.zeros(8), Sigma, n)
labels = ["glucose", "hba1c", "triglycerides", "ldl",
          "hdl", "creatinine", "alt", "albumin"]
metabolic = pd.DataFrame(z, columns=labels)

# Scale and fit PCA
scaler = StandardScaler()
X_scaled = scaler.fit_transform(metabolic)
pca = PCA()
scores = pca.fit_transform(X_scaled)

fig, axes = plt.subplots(1, 2, figsize=(14, 5))

# Scree plot
axes[0].bar(range(1, 9), pca.explained_variance_ratio_,
            color="steelblue", edgecolor="white")
axes[0].axhline(y=1/8, color="firebrick", linestyle="--", label="Kaiser threshold")
axes[0].set_xlabel("Component")
axes[0].set_ylabel("Proportion of Variance")
axes[0].set_title("Scree Plot")
axes[0].set_xticks(range(1, 9))
axes[0].legend()

# Biplot
loadings = pca.components_[:2].T  # first 2 PCs
scale_factor = 3
axes[1].scatter(scores[:, 0], scores[:, 1], alpha=0.3, s=10, color="grey")
for i, lab in enumerate(labels):
    axes[1].annotate("", xy=(loadings[i, 0]*scale_factor,
                              loadings[i, 1]*scale_factor),
                     xytext=(0, 0),
                     arrowprops=dict(arrowstyle="->", color="firebrick", lw=1.5))
    axes[1].text(loadings[i, 0]*scale_factor*1.15,
                 loadings[i, 1]*scale_factor*1.15,
                 lab, fontsize=9, color="firebrick", ha="center")
axes[1].set_xlabel(f"PC1 ({pca.explained_variance_ratio_[0]:.1%} var)")
axes[1].set_ylabel(f"PC2 ({pca.explained_variance_ratio_[1]:.1%} var)")
axes[1].set_title("PCA Biplot: Metabolic Panel")
axes[1].axhline(0, color="grey", linewidth=0.5)
axes[1].axvline(0, color="grey", linewidth=0.5)

plt.tight_layout()
plt.show()
```

:::

**What the code is doing.** We simulate 300 patients with eight correlated lab values, standardise them (so glucose in the hundreds does not drown out HbA1c around 6), and run PCA. The code then draws two figures. The **scree plot** (left) is a bar chart of how much variance each of the eight components captures, tallest on the left; the dashed "Kaiser threshold" line marks the level a component would reach if it explained no more than its fair share --- bars above it are worth keeping. You would look for the bars to drop off sharply after the first two or three, telling you most of the signal lives in just a few dimensions. The **biplot** (right) places every patient on the PC1--PC2 plane and overlays arrows for the original labs; how to read those arrows is explained next.

### Interpreting the Biplot

The biplot overlays two things:

- **Points** (grey): individual patients projected onto the first two principal components.
- **Arrows** (red): the original variables. The direction and length of each arrow show how strongly and in which direction that variable contributes to each PC.

In our metabolic panel:

- **PC1** is dominated by glucose, HbA1c, and triglycerides pointing in one direction, with HDL pointing in the opposite direction. This is a "metabolic syndrome" axis: patients to the right tend to have higher glucose and triglycerides and lower HDL.
- **PC2** separates creatinine and albumin, capturing a "renal/hepatic" dimension.

This kind of interpretation is exactly what makes PCA useful in clinical research: it reveals which variables "travel together" and suggests underlying biological dimensions.

## t-SNE: Non-Linear Dimensionality Reduction

### The Core Idea

PCA preserves global, linear structure. But many datasets contain non-linear patterns --- clusters, manifolds, or curves --- that PCA cannot capture well. **t-SNE** (t-distributed stochastic neighbour embedding) was designed specifically for **visualisation** of high-dimensional data in 2 or 3 dimensions.

t-SNE works in two steps:

1. In the high-dimensional space, compute the probability that any two points are "neighbours" (using a Gaussian kernel). Nearby points have high probability; distant points have low probability.
2. In the low-dimensional (2D) embedding, try to arrange points so that the same neighbourhood probabilities are preserved, using a **t-distribution** kernel (heavier tails than a Gaussian, which helps prevent crowding).

The algorithm minimises the KL divergence between the high-dimensional and low-dimensional probability distributions using gradient descent.

The intuition is worth holding onto even if the maths is not: t-SNE asks "who are each patient's closest neighbours?" in the full high-dimensional space, then arranges the 2D plot so that those same neighbourly relationships are preserved. The payoff for a health researcher is a map where patients who are genuinely alike clump together, making subtypes pop out visually --- which is exactly why t-SNE became popular for single-cell and other omics data where distinct cell populations or patient phenotypes hide in hundreds of dimensions.

### The Perplexity Parameter

Perplexity (typically 5--50) controls how many neighbours each point considers. It roughly corresponds to the effective number of local neighbours.

- **Low perplexity** (5--10): focuses on very local structure. May produce many small, tight clusters.
- **High perplexity** (30--50): considers broader neighbourhoods. Produces smoother, more global layouts.

There is no single correct value; try several and see how the structure changes.

### Critical Pitfalls and Misinterpretations

::: {.callout-warning}
## t-SNE is for exploration, not inference
t-SNE is a visualisation tool. It is not a statistical test. Do not draw conclusions about statistical significance from a t-SNE plot.
:::

**Distances between clusters are meaningless.** Two clusters that appear far apart may not actually be more different than two clusters that appear close. t-SNE distorts global distances to preserve local neighbourhoods.

**Cluster sizes are meaningless.** A tight cluster in t-SNE space does not mean those points are more similar than a spread-out cluster.

**Different runs give different layouts.** t-SNE uses stochastic optimisation. Running it twice with different seeds gives different-looking plots (though the same clusters should emerge).

**It does not scale well.** Standard t-SNE is $O(n^2)$, making it slow for datasets with more than ~10,000 observations. Approximations (Barnes-Hut, FIt-SNE) help.

### Clinical Example: Visualising Patient Phenotypes

::: {.panel-tabset}

### R

```{r}
#| label: fig-tsne-r
#| fig-cap: "t-SNE visualisation of simulated patient data with three disease subtypes."
#| fig-width: 10
#| fig-height: 5

library(Rtsne)

set.seed(42)

# Simulate 3 patient subtypes with 20 features
n_per_group <- 100
p <- 20

group1 <- matrix(rnorm(n_per_group * p, mean = 0, sd = 1), ncol = p)
group2 <- matrix(rnorm(n_per_group * p, mean = 2, sd = 1.2), ncol = p)
group3 <- matrix(rnorm(n_per_group * p, mean = -1, sd = 0.8), ncol = p)
# Make groups differ in specific feature subsets
group2[, 1:5] <- group2[, 1:5] + 3
group3[, 10:15] <- group3[, 10:15] - 4

X <- rbind(group1, group2, group3)
labels <- factor(rep(c("Type A", "Type B", "Type C"), each = n_per_group))

# Run t-SNE with two perplexity values
tsne_low  <- Rtsne(X, perplexity = 10, dims = 2, verbose = FALSE, max_iter = 1000)
tsne_high <- Rtsne(X, perplexity = 40, dims = 2, verbose = FALSE, max_iter = 1000)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
cols <- c("steelblue", "firebrick", "forestgreen")

plot(tsne_low$Y, col = cols[labels], pch = 16, cex = 0.8,
     main = "t-SNE (perplexity = 10)", xlab = "t-SNE 1", ylab = "t-SNE 2")
legend("topright", levels(labels), col = cols, pch = 16, cex = 0.8)

plot(tsne_high$Y, col = cols[labels], pch = 16, cex = 0.8,
     main = "t-SNE (perplexity = 40)", xlab = "t-SNE 1", ylab = "t-SNE 2")
legend("topright", levels(labels), col = cols, pch = 16, cex = 0.8)
```

### Python

```{python}
#| label: fig-tsne-py
#| fig-cap: "t-SNE visualisation of simulated patient data with three disease subtypes."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.manifold import TSNE

np.random.seed(42)
n_per = 100
p = 20

g1 = np.random.normal(0, 1, (n_per, p))
g2 = np.random.normal(2, 1.2, (n_per, p))
g3 = np.random.normal(-1, 0.8, (n_per, p))
g2[:, :5] += 3
g3[:, 10:15] -= 4

X = np.vstack([g1, g2, g3])
labels = np.repeat(["Type A", "Type B", "Type C"], n_per)
colors = {"Type A": "steelblue", "Type B": "firebrick", "Type C": "forestgreen"}

fig, axes = plt.subplots(1, 2, figsize=(12, 5))
for ax, perp in zip(axes, [10, 40]):
    tsne = TSNE(n_components=2, perplexity=perp, random_state=42, max_iter=1000)
    emb = tsne.fit_transform(X)
    for lab in ["Type A", "Type B", "Type C"]:
        mask = labels == lab
        ax.scatter(emb[mask, 0], emb[mask, 1], c=colors[lab],
                   s=15, alpha=0.7, label=lab)
    ax.set_title(f"t-SNE (perplexity = {perp})")
    ax.set_xlabel("t-SNE 1")
    ax.set_ylabel("t-SNE 2")
    ax.legend(fontsize=9)

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** We build a fake dataset of 300 patients across three "disease subtypes" (each subtype differs on a different handful of the 20 features), then run t-SNE twice --- once with perplexity 10, once with 40 --- and plot the two resulting maps side by side, colouring each dot by its true subtype. Each dot is a patient; the axes ("t-SNE 1/2") have no units and no inherent meaning, so do not try to read values off them. What you look for is *separation*: three coloured blobs that do not overlap means the subtypes really do have distinct profiles.

Notice how perplexity changes the visual appearance but the three groups are consistently separated. This is reassuring --- the cluster structure is real, not an artefact of a particular perplexity setting. (The perplexity setting, recall, is roughly the number of near neighbours each point pays attention to: a small value zooms in on tight local clumps, a large value smooths things into broader groupings.)

## UMAP: Uniform Manifold Approximation and Projection

### How UMAP Differs from t-SNE

UMAP (McInnes, Healy, and Melville, 2018) is a newer method that has largely replaced t-SNE in many applications. Key advantages:

| Feature | t-SNE | UMAP |
|---|---|---|
| Speed | Slow ($O(n^2)$ or $O(n \log n)$ approximate) | Fast (typically 3--10x faster) |
| Global structure | Poorly preserved | Better preserved |
| Reproducibility | Stochastic, varies across runs | More stable (with fixed seed) |
| Scalability | Struggles above ~50k points | Handles millions of points |
| Embedding dimensions | Typically 2--3 | 2 to arbitrary $d$ |
| Use beyond visualisation | Rarely | Commonly used as a preprocessing step |

: Comparison of t-SNE and UMAP {#tbl-tsne-umap}

UMAP works by constructing a topological representation (a weighted graph) of the data in high dimensions, then optimising a low-dimensional layout that preserves the graph structure.

### Key Parameters

**n_neighbors** (analogous to perplexity in t-SNE): controls the balance between local and global structure. Small values (5--15) emphasise local clusters; large values (50--200) emphasise global connectivity.

**min_dist**: controls how tightly points are packed in the embedding. Small values (0.0--0.1) produce tight, well-separated clusters; large values (0.5--1.0) produce a more uniform spread.

### When to Use Which Method

- **PCA**: use for initial exploration, as a preprocessing step (reduce to 20--50 PCs before t-SNE/UMAP), or when you need interpretable components. Use when linear relationships are sufficient.
- **t-SNE**: use for detailed visualisation of moderate-sized datasets (< 10,000 points) when you want to emphasise local cluster structure. Avoid for quantitative downstream analysis.
- **UMAP**: use as your default non-linear method. Faster, better global structure, and can serve as a preprocessing step for clustering.

::: {.panel-tabset}

### R

```{r}
#| label: fig-umap-comparison-r
#| fig-cap: "PCA, t-SNE, and UMAP visualisations of the same simulated patient dataset."
#| fig-width: 14
#| fig-height: 4.5

library(Rtsne)
library(uwot)

set.seed(42)

pca_scores <- prcomp(X, scale. = TRUE)$x[, 1:2]
tsne_emb   <- Rtsne(X, perplexity = 30, dims = 2, verbose = FALSE)$Y
umap_emb   <- umap(X, n_neighbors = 15, min_dist = 0.1, verbose = FALSE)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
cols <- c("steelblue", "firebrick", "forestgreen")[as.numeric(labels)]

plot(pca_scores, col = cols, pch = 16, cex = 0.8,
     main = "PCA", xlab = "PC1", ylab = "PC2")

plot(tsne_emb, col = cols, pch = 16, cex = 0.8,
     main = "t-SNE (perplexity = 30)", xlab = "t-SNE 1", ylab = "t-SNE 2")

plot(umap_emb, col = cols, pch = 16, cex = 0.8,
     main = "UMAP (n_neighbors = 15)", xlab = "UMAP 1", ylab = "UMAP 2")
```

### Python

```{python}
#| label: fig-umap-comparison-py
#| fig-cap: "PCA, t-SNE, and UMAP visualisations of the same simulated patient dataset."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.decomposition import PCA
from sklearn.manifold import TSNE
from umap import UMAP

pca_2d = PCA(n_components=2).fit_transform(X)
tsne_2d = TSNE(n_components=2, perplexity=30, random_state=42).fit_transform(X)
umap_2d = UMAP(n_components=2, n_neighbors=15, min_dist=0.1,
                random_state=42).fit_transform(X)

fig, axes = plt.subplots(1, 3, figsize=(16, 5))
titles = ["PCA", "t-SNE (perplexity=30)", "UMAP (n_neighbors=15)"]
embeddings = [pca_2d, tsne_2d, umap_2d]

for ax, emb, title in zip(axes, embeddings, titles):
    for lab, col in colors.items():
        mask = labels == lab
        ax.scatter(emb[mask, 0], emb[mask, 1], c=col, s=15, alpha=0.7, label=lab)
    ax.set_title(title)
    ax.set_xlabel("Dim 1")
    ax.set_ylabel("Dim 2")
    ax.legend(fontsize=8)

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** We take the *same* three-subtype dataset and run all three methods on it --- PCA, t-SNE, and UMAP --- then plot their 2D outputs in one row so you can compare them like for like, with patients coloured by true subtype. The point of the figure is not any single panel but the contrast between them: it lets you see, on real-looking data, how a linear method (PCA) and two non-linear methods behave on the same problem.

All three methods recover the three groups, but with different visual characteristics. PCA shows the groups with some overlap (because the separation is partly non-linear). t-SNE and UMAP separate them more clearly, with UMAP preserving more of the relative distances between clusters.

## Practical Workflow: Scale, Reduce, Visualise, Interpret

A standard workflow for high-dimensional clinical data:

1. **Scale** the data. PCA, t-SNE, and UMAP are all sensitive to the scale of variables. Always standardise (mean = 0, SD = 1) before applying any dimensionality reduction.

2. **Reduce with PCA first.** If the original data have hundreds or thousands of features, first reduce to 20--50 principal components. This removes noise and speeds up t-SNE/UMAP dramatically.

3. **Visualise with UMAP (or t-SNE).** Apply the non-linear method to the PCA-reduced data. Colour points by known labels (diagnosis, treatment group) or by continuous variables (lab values, scores).

4. **Interpret cautiously.** Look for clusters, gradients, or outliers. Generate hypotheses --- but remember that these are visualisations, not tests. Validate any apparent structure with formal statistical methods.

::: {.panel-tabset}

### R

```{r}
#| label: fig-workflow-r
#| fig-cap: "The standard workflow: PCA for initial reduction, then UMAP for visualisation."
#| fig-width: 10
#| fig-height: 5

library(uwot)

set.seed(42)

# Simulate high-dimensional data: 500 patients, 100 features
n <- 500; p <- 100
group <- sample(c("Healthy", "Mild", "Severe"), n, replace = TRUE,
                prob = c(0.4, 0.35, 0.25))

X_hd <- matrix(rnorm(n * p), ncol = p)
# Add signal in first few features
X_hd[group == "Mild", 1:10]   <- X_hd[group == "Mild", 1:10] + 1.5
X_hd[group == "Severe", 1:10] <- X_hd[group == "Severe", 1:10] + 3
X_hd[group == "Severe", 11:20] <- X_hd[group == "Severe", 11:20] - 2

# Step 1: Scale
X_scaled <- scale(X_hd)

# Step 2: PCA to 20 components
pca_20 <- prcomp(X_scaled)$x[, 1:20]

# Step 3: UMAP on PCA-reduced data
umap_result <- umap(pca_20, n_neighbors = 15, min_dist = 0.1, verbose = FALSE)

# Step 4: Visualise
plot_df <- tibble(
  UMAP1 = umap_result[, 1],
  UMAP2 = umap_result[, 2],
  Group = factor(group, levels = c("Healthy", "Mild", "Severe"))
)

ggplot(plot_df, aes(x = UMAP1, y = UMAP2, colour = Group)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_colour_manual(values = c("Healthy" = "steelblue",
                                  "Mild" = "goldenrod",
                                  "Severe" = "firebrick")) +
  labs(title = "UMAP After PCA Reduction",
       subtitle = "500 patients, 100 features → 20 PCs → 2D UMAP") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Python

```{python}
#| label: fig-workflow-py
#| fig-cap: "The standard workflow: PCA for initial reduction, then UMAP for visualisation."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA
from umap import UMAP

np.random.seed(42)
n, p = 500, 100
group = np.random.choice(["Healthy", "Mild", "Severe"], n, p=[0.4, 0.35, 0.25])

X_hd = np.random.normal(0, 1, (n, p))
X_hd[group == "Mild", :10] += 1.5
X_hd[group == "Severe", :10] += 3
X_hd[group == "Severe", 10:20] -= 2

# Workflow
X_scaled = StandardScaler().fit_transform(X_hd)
pca_20 = PCA(n_components=20).fit_transform(X_scaled)
umap_2d = UMAP(n_neighbors=15, min_dist=0.1, random_state=42).fit_transform(pca_20)

fig, ax = plt.subplots(figsize=(8, 6))
color_map = {"Healthy": "steelblue", "Mild": "goldenrod", "Severe": "firebrick"}
for g in ["Healthy", "Mild", "Severe"]:
    mask = group == g
    ax.scatter(umap_2d[mask, 0], umap_2d[mask, 1],
               c=color_map[g], s=12, alpha=0.6, label=g)
ax.set_xlabel("UMAP 1")
ax.set_ylabel("UMAP 2")
ax.set_title("UMAP After PCA Reduction\n500 patients, 100 features → 20 PCs → 2D")
ax.legend()
plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** This puts the four-step recipe into practice on a larger, more realistic problem: 500 patients, 100 features, and three severity groups (Healthy, Mild, Severe) whose differences are buried in only the first 20 features. We scale, compress the 100 features down to 20 principal components, then hand those to UMAP to draw a single 2D map, colouring each patient by severity. You would expect to see the three colours form a gradient or three loose clusters --- Healthy at one end, Severe at the other --- which would confirm that the severity signal survived the squeeze from 100 dimensions to 2. Doing PCA first is not just for speed: it strips out noise so UMAP focuses on the real structure.

## Exercises

### Exercise 1: PCA on Clinical Lab Data

Using the simulated metabolic panel data from this chapter (or a real dataset if available):

a. Perform PCA on the standardised data.
b. Create a scree plot and determine how many components to retain using the cumulative variance threshold (80%) and the Kaiser criterion.
c. Examine the loadings of the first two PCs. Which variables load most strongly on each? Propose a clinical interpretation.
d. Create a biplot coloured by a simulated diabetes status variable. Do diabetic patients separate along PC1?

::: {.panel-tabset}

### R

```{r}
#| label: exercise-1-solution
#| code-fold: true
#| code-summary: "Show solution"
#| fig-width: 10
#| fig-height: 5

library(tidyverse)  # %>% pipe, select()

# Add diabetes status (correlated with glucose/HbA1c)
set.seed(99)
metabolic$diabetes <- factor(ifelse(metabolic$glucose > 110 & metabolic$hba1c > 6.2,
                                     "Diabetic", "Non-diabetic"))

pca_fit <- prcomp(metabolic %>% select(-diabetes), scale. = TRUE)

# (a) & (b) Scree plot with cumulative variance
var_prop <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
cum_var  <- cumsum(var_prop)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
barplot(var_prop, names.arg = paste0("PC", 1:8), col = "steelblue",
        main = "Variance Explained", ylab = "Proportion")
abline(h = 1/8, col = "firebrick", lty = 2)

plot(1:8, cum_var, type = "b", pch = 16, col = "steelblue",
     main = "Cumulative Variance", xlab = "Components",
     ylab = "Cumulative proportion", ylim = c(0, 1))
abline(h = 0.80, col = "firebrick", lty = 2)
text(6, 0.82, "80% threshold", col = "firebrick", cex = 0.8)

cat("Components for 80% variance:", which(cum_var >= 0.80)[1], "\n")

# (c) Loadings
cat("\nPC1 loadings:\n")
print(round(sort(pca_fit$rotation[, 1], decreasing = TRUE), 3))
cat("\nPC2 loadings:\n")
print(round(sort(pca_fit$rotation[, 2], decreasing = TRUE), 3))
```

### Python

```{python}
#| label: exercise-1-solution-py
#| code-fold: true
#| code-summary: "Show solution"

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA

# Recreate metabolic data (using earlier arrays)
diabetes = ["Diabetic" if metabolic.iloc[i]['glucose'] > 110 and
            metabolic.iloc[i]['hba1c'] > 6.2 else "Non-diabetic"
            for i in range(len(metabolic))]

X = metabolic[['glucose','hba1c','triglycerides','ldl','hdl',
               'creatinine','alt','albumin']].values
X_s = StandardScaler().fit_transform(X)
pca = PCA().fit(X_s)

cum_var = np.cumsum(pca.explained_variance_ratio_)
n_80 = np.argmax(cum_var >= 0.80) + 1
print(f"Components for 80% variance: {n_80}")

# Loadings
feat_names = ['glucose','hba1c','triglycerides','ldl','hdl',
              'creatinine','alt','albumin']
for pc in [0, 1]:
    loadings = pca.components_[pc]
    order = np.argsort(np.abs(loadings))[::-1]
    print(f"\nPC{pc+1} loadings:")
    for idx in order:
        print(f"  {feat_names[idx]:>15s}: {loadings[idx]:+.3f}")
```

:::

**What the solution shows.** The code labels patients as diabetic or not (based on glucose and HbA1c), re-runs PCA, and prints two things: how many components are needed to reach 80% of the variance, and the loadings of PC1 and PC2 sorted by size. Reading the printed loadings is the key step --- the labs with the largest positive or negative numbers are the ones that define each axis. If glucose, HbA1c, and triglycerides top the PC1 list, then PC1 is effectively a "metabolic" score, and you would expect diabetic patients to sit toward one end of that axis.

### Exercise 2: t-SNE Sensitivity to Perplexity

Using the three-group simulated dataset:

a. Run t-SNE with perplexity values of 5, 15, 30, and 50.
b. Create a 2x2 panel of the resulting embeddings.
c. At which perplexity value do the three groups first become clearly separated?
d. Are the distances between clusters consistent across perplexity values? What does this tell you about interpreting t-SNE?

### Exercise 3: UMAP Parameter Exploration

Using the same dataset:

a. Run UMAP with `n_neighbors` in {5, 15, 50, 100} while holding `min_dist = 0.1`.
b. Run UMAP with `min_dist` in {0.0, 0.1, 0.5, 1.0} while holding `n_neighbors = 15`.
c. Create a panel of plots for each parameter sweep.
d. Describe how each parameter affects the visual appearance. Which combination would you recommend for this dataset?

### Exercise 4: Full Workflow on Simulated Genomic Data

Simulate a dataset of 1,000 patients with 500 gene expression features and 4 cancer subtypes.

a. Scale the data and perform PCA. How many PCs are needed for 80% variance?
b. Apply UMAP to the first 30 PCs. Colour by cancer subtype. Are the subtypes visually separable?
c. Repeat with t-SNE. Compare the two visualisations.
d. One of the four subtypes is rare (5% of patients). Can you still identify it in the UMAP plot?

## Summary

Dimensionality reduction transforms high-dimensional clinical data into interpretable, lower-dimensional representations. PCA provides linear, interpretable components that are ideal for initial exploration and preprocessing. t-SNE and UMAP offer non-linear embeddings that reveal complex structure invisible to PCA, though their outputs must be interpreted with care. The standard workflow --- scale, reduce with PCA, embed with UMAP, interpret cautiously --- is a reliable starting point for any high-dimensional dataset in clinical research.

::: {.callout-tip}
## Key Takeaways
1. Always scale your data before dimensionality reduction.
2. PCA is linear, deterministic, and interpretable. Use it first.
3. t-SNE is for visualisation only. Distances between clusters are meaningless; cluster sizes are meaningless.
4. UMAP is faster, preserves more global structure, and can serve as a preprocessing step for downstream analysis.
5. Non-linear methods are exploratory. Validate any apparent structure with formal statistical methods or domain expertise.
:::

## References and Further Reading

The foundational reference for UMAP is McInnes, Healy, and Melville's 2018 paper "UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction," available on arXiv (1802.03426). For t-SNE, van der Maaten and Hinton's 2008 paper in the *Journal of Machine Learning Research* introduced the method, and Wattenberg, Viegas, and Johnson's 2016 interactive article "How to Use t-SNE Effectively" on Distill.pub remains the best practical guide to understanding its behaviour and pitfalls.

For PCA in a clinical context, Jolliffe and Cadima's 2016 tutorial "Principal Component Analysis: A Review and Recent Developments" in the *Philosophical Transactions of the Royal Society A* provides a modern overview. Becht et al.'s 2019 paper "Dimensionality Reduction for Visualizing Single-Cell Data Using UMAP" in *Nature Biotechnology* demonstrates the advantages of UMAP over t-SNE in a biological context that generalises well to other high-dimensional clinical data.

A comprehensive 2024 review by Xiang et al. in *BMC Bioinformatics* compares PCA, t-SNE, UMAP, and newer methods (PHATE, TriMap) across multiple benchmarks, providing guidance on method selection for different data types. For the mathematical foundations, Murphy's *Probabilistic Machine Learning: An Introduction* (2022) covers PCA in depth, while the UMAP documentation (umap-learn.readthedocs.io) includes excellent tutorials with worked examples. James, Witten, Hastie, and Tibshirani's *An Introduction to Statistical Learning* (2nd edition, 2021) provides an accessible introduction to PCA that is well suited for students with a biostatistics background.
