# Clustering: Discovering Patient Subgroups {#sec-clustering}

```{r}
#| include: false
library(tidyverse)
library(knitr)
```

Clustering answers a question every clinician has asked: "are all my patients really the same, or are there hidden subgroups in here?" Given a table of measurements and no labels, a clustering algorithm proposes a grouping --- this patient belongs with those patients --- so that you can then ask whether the groups differ in outcome, biology, or treatment response. That is how the now-famous sepsis and ARDS subtypes were discovered. This chapter walks through three popular algorithms and, crucially, how to tell a real subgroup from a mirage, because clustering will hand you groups whether or not any exist.

::: {.callout-note}
These chapters use `eval: false`, so the rendered page shows code but no live figures or printouts. Each code block is followed by a description of what its output *would* look like, so you can run it yourself and know what counts as a sensible result.
:::

## Introduction

Every classification method we have studied so far is **supervised**: you train a model on labelled data and use it to predict labels for new observations. Clustering is fundamentally different. It is **unsupervised** --- there are no labels. The goal is to discover natural groupings in the data that you did not know existed.

In clinical research, clustering is used to identify **patient phenotypes** --- subgroups of patients who share similar clinical profiles but may not correspond to any established diagnostic category. These phenotypes can reveal heterogeneity within a disease, suggest new treatment targets, or identify patients who respond differently to therapy. Sepsis, ARDS, heart failure, and asthma have all been re-examined through clustering analyses that revealed clinically meaningful subtypes hidden within traditional diagnostic labels.

This chapter covers three major clustering algorithms (K-means, hierarchical clustering, and DBSCAN), methods for choosing the number of clusters and validating results, and the integration of clustering with the dimensionality reduction methods from the previous chapter.

::: {.callout-important}
## Clusters can be artefacts
Clustering algorithms will always produce clusters, even in completely random data. The existence of clusters in your output does not mean they are real. Validation --- both statistical and clinical --- is essential.
:::

## K-Means Clustering

### The Algorithm

K-means is the simplest and most widely used clustering algorithm. Given $k$ (the number of clusters), it works as follows:

1. **Initialise**: randomly place $k$ cluster centroids in the feature space.
2. **Assign**: assign each data point to the nearest centroid (using Euclidean distance).
3. **Update**: recompute each centroid as the mean of all points assigned to it.
4. **Repeat** steps 2--3 until assignments stop changing (convergence).

K-means minimises the **within-cluster sum of squares (WCSS)**: the total squared distance from each point to its assigned centroid.

$$
\text{WCSS} = \sum_{k=1}^{K} \sum_{i \in C_k} \| x_i - \mu_k \|^2
$$

Read this formula as "for every patient, measure how far they sit from their own cluster's centre ($\mu_k$ is that centre, or **centroid**), square it, and add it all up." Tight, well-defined clusters give a small WCSS; loose, scattered ones give a large WCSS. K-means simply shuffles the group assignments until this total is as small as it can make it.

### Choosing K: Elbow Method, Silhouette Scores, Gap Statistic

The most important decision in K-means is the number of clusters $k$. There is no single correct answer, but several heuristics help.

**Elbow method.** Plot WCSS against $k$. As $k$ increases, WCSS always decreases (more clusters = points closer to centroids). Look for an "elbow" where the rate of decrease sharply changes. The elbow suggests that adding more clusters yields diminishing returns.

**Silhouette scores.** For each point, the silhouette score measures how much closer it is to its own cluster than to the nearest other cluster. Values range from --1 (misclassified) to +1 (well clustered). The average silhouette score across all points summarises clustering quality. Higher is better; choose $k$ that maximises the average silhouette score.

**Gap statistic.** Compares the WCSS of your data to the WCSS of data drawn from a reference null distribution (uniform). The gap is largest at the $k$ where your data's structure most exceeds what would be expected by chance.

### Limitations of K-Means

- **Assumes spherical, equally sized clusters.** K-means uses Euclidean distance, so it finds globular clusters. Elongated, ring-shaped, or irregularly shaped clusters will be poorly recovered.
- **Sensitive to initialisation.** Different random starting points can give different results. Modern implementations (K-means++) use smart initialisation to mitigate this, and running the algorithm multiple times with different seeds is standard practice.
- **Sensitive to outliers.** A single extreme point can pull a centroid substantially.
- **Requires scaled data.** Features on different scales will dominate the distance calculations. Always standardise before running K-means.

### Clinical Example: Patient Phenotyping from Lab Data

::: {.panel-tabset}

### R

```{r}
#| label: fig-kmeans-r
#| fig-cap: "K-means clustering of simulated patient lab data: choosing k and visualising clusters."
#| fig-width: 14
#| fig-height: 5

library(cluster)

set.seed(42)
n <- 400

# Simulate 3 patient phenotypes with distinct lab profiles
pheno1 <- tibble(  # Metabolic syndrome phenotype
  glucose = rnorm(n * 0.35, 140, 20),
  crp     = rnorm(n * 0.35, 5, 2),
  albumin = rnorm(n * 0.35, 3.5, 0.3),
  wbc     = rnorm(n * 0.35, 8, 2)
)
pheno2 <- tibble(  # Inflammatory phenotype
  glucose = rnorm(n * 0.35, 100, 15),
  crp     = rnorm(n * 0.35, 15, 4),
  albumin = rnorm(n * 0.35, 3.0, 0.4),
  wbc     = rnorm(n * 0.35, 14, 3)
)
pheno3 <- tibble(  # Healthy-ish phenotype
  glucose = rnorm(n * 0.30, 90, 10),
  crp     = rnorm(n * 0.30, 2, 1),
  albumin = rnorm(n * 0.30, 4.2, 0.2),
  wbc     = rnorm(n * 0.30, 7, 1.5)
)

lab_data <- bind_rows(pheno1, pheno2, pheno3)
lab_scaled <- scale(lab_data)

# Elbow and silhouette plots
wcss <- numeric(10)
sil_avg <- numeric(10)
for (k in 2:10) {
  km <- kmeans(lab_scaled, centers = k, nstart = 25)
  wcss[k] <- km$tot.withinss
  sil_avg[k] <- mean(silhouette(km$cluster, dist(lab_scaled))[, 3])
}

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# Elbow plot
plot(2:10, wcss[2:10], type = "b", pch = 16, col = "steelblue",
     xlab = "Number of clusters (k)", ylab = "Within-cluster SS",
     main = "Elbow Method")

# Silhouette plot
plot(2:10, sil_avg[2:10], type = "b", pch = 16, col = "firebrick",
     xlab = "Number of clusters (k)", ylab = "Average silhouette",
     main = "Silhouette Scores")

# K-means with k=3, visualise on first 2 PCs
km3 <- kmeans(lab_scaled, centers = 3, nstart = 25)
pca2 <- prcomp(lab_scaled)$x[, 1:2]
plot(pca2, col = c("steelblue", "firebrick", "forestgreen")[km3$cluster],
     pch = 16, cex = 0.7,
     main = "K-means (k=3) on PCA", xlab = "PC1", ylab = "PC2")
```

### Python

```{python}
#| label: fig-kmeans-py
#| fig-cap: "K-means clustering: elbow method, silhouette scores, and cluster visualisation."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.preprocessing import StandardScaler
from sklearn.cluster import KMeans
from sklearn.metrics import silhouette_score
from sklearn.decomposition import PCA

np.random.seed(42)

# Simulate phenotypes
g1 = np.column_stack([np.random.normal(140, 20, 140),
                       np.random.normal(5, 2, 140),
                       np.random.normal(3.5, 0.3, 140),
                       np.random.normal(8, 2, 140)])
g2 = np.column_stack([np.random.normal(100, 15, 140),
                       np.random.normal(15, 4, 140),
                       np.random.normal(3.0, 0.4, 140),
                       np.random.normal(14, 3, 140)])
g3 = np.column_stack([np.random.normal(90, 10, 120),
                       np.random.normal(2, 1, 120),
                       np.random.normal(4.2, 0.2, 120),
                       np.random.normal(7, 1.5, 120)])

X = np.vstack([g1, g2, g3])
X_scaled = StandardScaler().fit_transform(X)

ks = range(2, 11)
wcss = []
sils = []
for k in ks:
    km = KMeans(n_clusters=k, n_init=25, random_state=42)
    km.fit(X_scaled)
    wcss.append(km.inertia_)
    sils.append(silhouette_score(X_scaled, km.labels_))

fig, axes = plt.subplots(1, 3, figsize=(16, 5))

axes[0].plot(list(ks), wcss, "o-", color="steelblue")
axes[0].set_xlabel("Number of clusters (k)")
axes[0].set_ylabel("Within-cluster SS")
axes[0].set_title("Elbow Method")

axes[1].plot(list(ks), sils, "o-", color="firebrick")
axes[1].set_xlabel("Number of clusters (k)")
axes[1].set_ylabel("Average silhouette")
axes[1].set_title("Silhouette Scores")

km3 = KMeans(n_clusters=3, n_init=25, random_state=42).fit(X_scaled)
pca2 = PCA(n_components=2).fit_transform(X_scaled)
colors = ["steelblue", "firebrick", "forestgreen"]
for c in range(3):
    mask = km3.labels_ == c
    axes[2].scatter(pca2[mask, 0], pca2[mask, 1], c=colors[c],
                    s=12, alpha=0.6, label=f"Cluster {c+1}")
axes[2].set_xlabel("PC1")
axes[2].set_ylabel("PC2")
axes[2].set_title("K-means (k=3) on PCA")
axes[2].legend()

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** We simulate 400 patients drawn from three hidden lab-profile phenotypes (a metabolic, an inflammatory, and a healthier group), scale the labs, and then ask K-means to try every value of $k$ from 2 to 10, recording two diagnostics each time. The figure has three panels. The **elbow plot** (left) shows within-cluster spread falling as $k$ rises --- you hunt for the "elbow" where it stops dropping steeply, the point of diminishing returns. The **silhouette plot** (middle) shows average cluster quality; here you want the *peak*. The **PCA scatter** (right) shows the actual $k=3$ clustering, with each patient coloured by assigned cluster and placed on the first two principal components (a 2D summary, since we cannot plot four lab axes at once) --- well-separated colours mean the clusters occupy distinct regions.

Both the elbow and silhouette plots suggest $k = 3$ --- consistent with our simulation of three phenotypes. In practice, the signal is rarely this clean, and multiple values of $k$ may appear reasonable. Domain knowledge should guide the final choice.

## Hierarchical Clustering

### Agglomerative (Bottom-Up) Clustering

Hierarchical clustering does not require specifying $k$ in advance. The **agglomerative** approach starts with each observation as its own cluster and iteratively merges the closest pair of clusters until all observations belong to a single cluster.

The result is a **dendrogram** --- a tree-like diagram that shows the merging history. You choose the number of clusters by "cutting" the dendrogram at a height that produces a meaningful number of groups.

### Linkage Methods

The key decision is how to define the distance between two clusters (each containing multiple points):

**Single linkage.** Distance = minimum distance between any two points in the two clusters. Tends to produce long, chain-like clusters. Sensitive to outliers and noise.

**Complete linkage.** Distance = maximum distance between any two points. Produces compact, roughly equal-sized clusters but can be distorted by outliers.

**Average linkage.** Distance = average of all pairwise distances. A compromise between single and complete.

**Ward's method.** Merges the pair of clusters that causes the smallest increase in total WCSS. Tends to produce compact, spherical clusters of similar size. **This is usually the best default choice.**

::: {.panel-tabset}

### R

```{r}
#| label: fig-dendrogram-r
#| fig-cap: "Dendrogram from hierarchical clustering (Ward's method) with 3 clusters highlighted."
#| fig-width: 10
#| fig-height: 6

# Use a smaller subset for readable dendrogram
set.seed(42)
idx <- sample(nrow(lab_scaled), 80)
hc <- hclust(dist(lab_scaled[idx, ]), method = "ward.D2")

# Plot dendrogram
plot(hc, labels = FALSE, main = "Hierarchical Clustering (Ward's Method)",
     xlab = "", sub = "", hang = -1, cex = 0.6)
rect.hclust(hc, k = 3, border = c("steelblue", "firebrick", "forestgreen"))
```

```{r}
#| label: fig-linkage-comparison-r
#| fig-cap: "Comparison of linkage methods on the same dataset."
#| fig-width: 14
#| fig-height: 4

par(mfrow = c(1, 4), mar = c(2, 2, 3, 1))
methods <- c("single", "complete", "average", "ward.D2")
titles  <- c("Single", "Complete", "Average", "Ward's")

for (i in seq_along(methods)) {
  hc_i <- hclust(dist(lab_scaled[idx, ]), method = methods[i])
  plot(hc_i, labels = FALSE, main = titles[i], xlab = "", sub = "",
       hang = -1, cex = 0.5)
  rect.hclust(hc_i, k = 3, border = "firebrick")
}
```

### Python

```{python}
#| label: fig-dendrogram-py
#| fig-cap: "Dendrogram from hierarchical clustering (Ward's method)."

import numpy as np
import matplotlib.pyplot as plt
from scipy.cluster.hierarchy import dendrogram, linkage, fcluster

np.random.seed(42)
idx = np.random.choice(len(X_scaled), 80, replace=False)
X_sub = X_scaled[idx]

Z = linkage(X_sub, method='ward')

fig, ax = plt.subplots(figsize=(12, 5))
dendrogram(Z, ax=ax, no_labels=True, color_threshold=Z[-3, 2])
ax.set_title("Hierarchical Clustering (Ward's Method)")
ax.set_xlabel("Patients")
ax.set_ylabel("Distance")
ax.axhline(y=Z[-3, 2], color='firebrick', linestyle='--', linewidth=1)
ax.text(5, Z[-3, 2] + 0.5, "Cut for k=3", color="firebrick", fontsize=10)
plt.tight_layout()
plt.show()
```

```{python}
#| label: fig-linkage-comparison-py
#| fig-cap: "Comparison of linkage methods."

fig, axes = plt.subplots(1, 4, figsize=(18, 4))
methods = ['single', 'complete', 'average', 'ward']
titles = ['Single', 'Complete', 'Average', "Ward's"]

for ax, method, title in zip(axes, methods, titles):
    Z_i = linkage(X_sub, method=method)
    dendrogram(Z_i, ax=ax, no_labels=True, color_threshold=0)
    ax.set_title(title)
    ax.set_xlabel("")

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** The first block builds a **dendrogram** --- the upside-down tree that records, branch by branch, which patients (then which groups of patients) got merged and at what distance. The coloured rectangles show where we "cut" the tree to get three clusters. The second block redraws the dendrogram four times, once per linkage rule (single, complete, average, Ward's), so you can see how the choice of linkage reshapes the tree. The practical lesson from that comparison: single linkage often produces one lopsided "chain" with stragglers, while Ward's gives balanced, compact groups --- which is why Ward's is the recommended default for clinical data.

### Reading a Dendrogram

The dendrogram encodes the full merging history:

- The **height** of each horizontal line represents the distance at which two clusters were merged.
- **Long vertical lines** before a merge indicate well-separated clusters.
- **Short vertical lines** indicate merges of similar clusters --- less clear separation.

Cutting the dendrogram at different heights gives different numbers of clusters. A large "gap" between merging heights (a long vertical line) suggests a natural number of clusters.

## DBSCAN: Density-Based Clustering

### When K-Means Fails

K-means assumes clusters are spherical and roughly equal in size. In clinical data, clusters may be irregularly shaped, overlap, or contain noise (outlier patients who do not belong to any group). **DBSCAN** (Density-Based Spatial Clustering of Applications with Noise) addresses these limitations.

### How It Works

DBSCAN defines clusters as dense regions separated by sparse regions. Two parameters control its behaviour:

- **eps ($\varepsilon$)**: the radius of the neighbourhood around each point.
- **min_samples**: the minimum number of points required within radius $\varepsilon$ to form a dense region.

The algorithm classifies each point as:

1. **Core point**: has at least `min_samples` neighbours within radius $\varepsilon$.
2. **Border point**: within $\varepsilon$ of a core point but does not have enough neighbours to be core itself.
3. **Noise point**: neither core nor border. These are **outliers** --- DBSCAN explicitly identifies them.

Clusters are formed by connecting core points that are within $\varepsilon$ of each other.

### Advantages and Limitations

**Advantages:**

- Does not require specifying $k$ in advance.
- Can find arbitrarily shaped clusters.
- Identifies outliers explicitly.

**Limitations:**

- Sensitive to the choice of $\varepsilon$ and `min_samples`.
- Struggles when clusters have very different densities.
- Not ideal for very high-dimensional data (use after dimensionality reduction).

::: {.panel-tabset}

### R

```{r}
#| label: fig-dbscan-r
#| fig-cap: "DBSCAN identifies irregularly shaped clusters and noise points."
#| fig-width: 10
#| fig-height: 5

library(dbscan)

set.seed(42)

# Create data with two crescent-shaped clusters + noise
n_each <- 150
theta1 <- seq(0, pi, length.out = n_each)
theta2 <- seq(0, pi, length.out = n_each)

x1 <- cos(theta1) + rnorm(n_each, 0, 0.08)
y1 <- sin(theta1) + rnorm(n_each, 0, 0.08)
x2 <- 1 - cos(theta2) + rnorm(n_each, 0, 0.08)
y2 <- 1 - sin(theta2) - 0.5 + rnorm(n_each, 0, 0.08)

# Add noise
x_noise <- runif(30, -0.5, 2)
y_noise <- runif(30, -1, 1.5)

crescent_data <- rbind(
  cbind(x1, y1),
  cbind(x2, y2),
  cbind(x_noise, y_noise)
)

# Compare K-means and DBSCAN
km_crescents <- kmeans(crescent_data, centers = 2, nstart = 25)
db_crescents <- dbscan(crescent_data, eps = 0.15, minPts = 5)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
cols_km <- c("steelblue", "firebrick")[km_crescents$cluster]
plot(crescent_data, col = cols_km, pch = 16, cex = 0.7,
     main = "K-means (k=2)", xlab = "x", ylab = "y")

# DBSCAN: cluster 0 = noise (grey), others = colours
db_cols <- c("grey50", "steelblue", "firebrick", "forestgreen")
plot(crescent_data,
     col = db_cols[db_crescents$cluster + 1],
     pch = ifelse(db_crescents$cluster == 0, 4, 16),
     cex = 0.7,
     main = "DBSCAN (eps=0.15)", xlab = "x", ylab = "y")
legend("topright", c("Noise", "Cluster 1", "Cluster 2"),
       col = db_cols[1:3], pch = c(4, 16, 16), cex = 0.8)
```

### Python

```{python}
#| label: fig-dbscan-py
#| fig-cap: "DBSCAN correctly identifies crescent-shaped clusters that K-means cannot handle."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.cluster import KMeans, DBSCAN

np.random.seed(42)
n_each = 150

theta1 = np.linspace(0, np.pi, n_each)
theta2 = np.linspace(0, np.pi, n_each)

x1 = np.cos(theta1) + np.random.normal(0, 0.08, n_each)
y1 = np.sin(theta1) + np.random.normal(0, 0.08, n_each)
x2 = 1 - np.cos(theta2) + np.random.normal(0, 0.08, n_each)
y2 = 1 - np.sin(theta2) - 0.5 + np.random.normal(0, 0.08, n_each)

x_noise = np.random.uniform(-0.5, 2, 30)
y_noise = np.random.uniform(-1, 1.5, 30)

X_crescent = np.vstack([
    np.column_stack([x1, y1]),
    np.column_stack([x2, y2]),
    np.column_stack([x_noise, y_noise])
])

km = KMeans(n_clusters=2, n_init=25, random_state=42).fit(X_crescent)
db = DBSCAN(eps=0.15, min_samples=5).fit(X_crescent)

fig, axes = plt.subplots(1, 2, figsize=(12, 5))

colors_km = ["steelblue" if c == 0 else "firebrick" for c in km.labels_]
axes[0].scatter(X_crescent[:, 0], X_crescent[:, 1], c=colors_km, s=10)
axes[0].set_title("K-means (k=2)")

color_map = {-1: "grey", 0: "steelblue", 1: "firebrick"}
colors_db = [color_map.get(c, "forestgreen") for c in db.labels_]
markers = ["x" if c == -1 else "o" for c in db.labels_]
for c_val in set(db.labels_):
    mask = db.labels_ == c_val
    m = "x" if c_val == -1 else "o"
    label = "Noise" if c_val == -1 else f"Cluster {c_val+1}"
    axes[1].scatter(X_crescent[mask, 0], X_crescent[mask, 1],
                    c=color_map.get(c_val, "forestgreen"),
                    marker=m, s=15, label=label)
axes[1].set_title("DBSCAN (eps=0.15)")
axes[1].legend(fontsize=9)

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** We deliberately build a hard case: two interlocking crescent (banana) shapes plus a scatter of random "noise" patients who belong to neither. We then cluster it two ways and plot them side by side. The left panel is K-means asked for two clusters; because it can only carve out round blobs, it slices each crescent in half --- the wrong answer. The right panel is DBSCAN, which grows clusters by following dense trails of points, so it traces each crescent correctly and marks the scattered outliers as "noise" (the crosses). Here `eps = 0.15` sets how close two patients must be to count as neighbours, and `minPts = 5` sets how many neighbours a point needs to anchor a dense region.

The figure shows a classic example: K-means splits the crescent-shaped clusters incorrectly, while DBSCAN identifies them perfectly and correctly labels the noise points.

## Cluster Validation

### Internal Metrics

Internal metrics evaluate clustering quality using only the data itself (no external labels).

**Silhouette score.** For each point $i$, define $a(i)$ as the mean distance to other points in the same cluster and $b(i)$ as the mean distance to points in the nearest other cluster. The silhouette score is:

$$
s(i) = \frac{b(i) - a(i)}{\max(a(i), b(i))}
$$

Values range from --1 to +1. A high average silhouette score indicates well-separated, compact clusters. The intuition: for each patient, compare "how far am I from my own group" ($a$) with "how far am I from the nearest rival group" ($b$). If the rival is much farther away ($b \gg a$), the score is near +1 and the patient is comfortably in the right cluster; if the patient is actually closer to a rival group, the score goes negative, flagging a likely misassignment.

**Calinski-Harabasz index.** The ratio of between-cluster variance to within-cluster variance (higher is better). It favours compact, well-separated clusters.

### Stability-Based Validation

Internal metrics can be misleading --- they can be high even for meaningless clusters. **Bootstrap stability** provides a stronger test:

1. Resample the data with replacement many times.
2. Re-run the clustering on each bootstrap sample.
3. Measure how often the same points end up in the same cluster.

If clusters are stable across resamples, they likely reflect real structure. If membership changes dramatically, the clusters may be artefacts.

::: {.panel-tabset}

### R

```{r}
#| label: fig-silhouette-r
#| fig-cap: "Silhouette plot for K-means with k=3 on the lab data."
#| fig-width: 8
#| fig-height: 5

library(cluster)

km3 <- kmeans(lab_scaled, centers = 3, nstart = 25)
sil <- silhouette(km3$cluster, dist(lab_scaled))

plot(sil, col = c("steelblue", "firebrick", "forestgreen"),
     main = "Silhouette Plot (k = 3)",
     border = NA)
cat("Average silhouette width:", round(mean(sil[, 3]), 3), "\n")
```

### Python

```{python}
#| label: fig-silhouette-py
#| fig-cap: "Silhouette scores by cluster for K-means with k=3."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.metrics import silhouette_score, silhouette_samples

km3 = KMeans(n_clusters=3, n_init=25, random_state=42).fit(X_scaled)
sil_vals = silhouette_samples(X_scaled, km3.labels_)
avg_sil = silhouette_score(X_scaled, km3.labels_)

fig, ax = plt.subplots(figsize=(8, 5))
y_lower = 0
colors = ["steelblue", "firebrick", "forestgreen"]
for c in range(3):
    c_sil = np.sort(sil_vals[km3.labels_ == c])
    y_upper = y_lower + len(c_sil)
    ax.barh(range(y_lower, y_upper), c_sil, height=1.0,
            color=colors[c], edgecolor="none")
    y_lower = y_upper + 5

ax.axvline(avg_sil, color="black", linestyle="--",
           label=f"Average: {avg_sil:.3f}")
ax.set_xlabel("Silhouette Coefficient")
ax.set_ylabel("Patients (sorted within cluster)")
ax.set_title("Silhouette Plot (k=3)")
ax.legend()
plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** This draws a **silhouette plot** for the $k=3$ lab-data solution: one horizontal bar per patient, grouped and coloured by cluster, with each bar's length equal to that patient's silhouette score. A healthy cluster looks like a wide, mostly positive block of bars; bars that are short or dip below zero are patients sitting near (or on the wrong side of) a cluster boundary. The dashed line marks the overall average, the single number you would quote to summarise cluster quality. If one cluster's bars are systematically shorter than the others, that cluster is poorly defined and worth scrutinising.

## Clinical Applications

### Patient Phenotyping

The most impactful use of clustering in clinical research is identifying disease subtypes that were previously unrecognised. Notable examples:

- **ARDS phenotypes.** Calfee et al. (2014) applied latent class analysis to two ARDS clinical trial datasets and identified two phenotypes --- "hyperinflammatory" and "hypoinflammatory" --- with different mortality rates and different responses to treatment. The hyperinflammatory phenotype had higher IL-6, higher vasopressor use, and worse outcomes.

- **Sepsis endotypes.** Seymour et al. (2019) used K-means clustering on EHR data from over 20,000 sepsis patients and identified four phenotypes (alpha, beta, gamma, delta) with distinct organ dysfunction patterns and 28-day mortality rates ranging from 5% to 40%.

- **Heart failure subtypes.** Shah et al. (2015) applied hierarchical clustering to echocardiographic and clinical data from HFpEF patients and identified three phenotypes with different pathophysiology and prognosis.

### Disease Subtyping from Lab Panels

A practical workflow for phenotyping:

1. Select clinically relevant variables (labs, vitals, demographics).
2. Handle missing data (imputation or exclusion).
3. Standardise all variables.
4. Reduce dimensionality (PCA to 10--20 components if many variables).
5. Apply multiple clustering algorithms (K-means, hierarchical) and compare.
6. Validate: silhouette scores, stability analysis, and --- most importantly --- clinical interpretation.
7. Compare clusters on outcomes (mortality, length of stay, treatment response).

## Integrating Clustering with PCA/UMAP

The standard workflow for high-dimensional clinical data combines dimensionality reduction and clustering:

::: {.panel-tabset}

### R

```{r}
#| label: fig-cluster-umap-r
#| fig-cap: "The complete workflow: scale, PCA, cluster, visualise with UMAP."
#| fig-width: 10
#| fig-height: 5

library(uwot)
library(cluster)

set.seed(42)

# High-dimensional data: 500 patients, 50 features, 4 subtypes
n <- 500; p <- 50
true_subtype <- sample(1:4, n, replace = TRUE, prob = c(0.3, 0.3, 0.25, 0.15))
X_hd <- matrix(rnorm(n * p), ncol = p)

# Add subtype-specific signals
for (s in 1:4) {
  feature_start <- (s - 1) * 8 + 1
  feature_end   <- min(s * 8, p)
  X_hd[true_subtype == s, feature_start:feature_end] <-
    X_hd[true_subtype == s, feature_start:feature_end] + 2.5
}

# Step 1: Scale
X_s <- scale(X_hd)

# Step 2: PCA to 15 components
pca_15 <- prcomp(X_s)$x[, 1:15]

# Step 3: K-means on PCA scores
km4 <- kmeans(pca_15, centers = 4, nstart = 50)

# Step 4: UMAP for visualisation
umap_2d <- umap(pca_15, n_neighbors = 15, min_dist = 0.1, verbose = FALSE)

plot_df <- tibble(
  UMAP1 = umap_2d[, 1],
  UMAP2 = umap_2d[, 2],
  Cluster = factor(km4$cluster),
  True_subtype = factor(true_subtype)
)

p1 <- ggplot(plot_df, aes(x = UMAP1, y = UMAP2, colour = Cluster)) +
  geom_point(alpha = 0.6, size = 1.2) +
  scale_colour_manual(values = c("steelblue", "firebrick",
                                  "forestgreen", "goldenrod")) +
  labs(title = "K-means Clusters on UMAP") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p2 <- ggplot(plot_df, aes(x = UMAP1, y = UMAP2, colour = True_subtype)) +
  geom_point(alpha = 0.6, size = 1.2) +
  scale_colour_manual(values = c("steelblue", "firebrick",
                                  "forestgreen", "goldenrod")) +
  labs(title = "True Subtypes on UMAP", colour = "Subtype") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

library(patchwork)
p1 + p2
```

### Python

```{python}
#| label: fig-cluster-umap-py
#| fig-cap: "The complete workflow: scale, PCA, cluster, visualise with UMAP."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA
from sklearn.cluster import KMeans
from umap import UMAP

np.random.seed(42)
n, p = 500, 50
true_sub = np.random.choice(4, n, p=[0.3, 0.3, 0.25, 0.15])
X_hd = np.random.normal(0, 1, (n, p))

for s in range(4):
    start = s * 8
    end = min((s + 1) * 8, p)
    X_hd[true_sub == s, start:end] += 2.5

X_s = StandardScaler().fit_transform(X_hd)
pca_15 = PCA(n_components=15).fit_transform(X_s)
km4 = KMeans(n_clusters=4, n_init=50, random_state=42).fit(pca_15)
umap_2d = UMAP(n_neighbors=15, min_dist=0.1, random_state=42).fit_transform(pca_15)

fig, axes = plt.subplots(1, 2, figsize=(14, 5))
colors_4 = ["steelblue", "firebrick", "forestgreen", "goldenrod"]

for c in range(4):
    mask = km4.labels_ == c
    axes[0].scatter(umap_2d[mask, 0], umap_2d[mask, 1],
                    c=colors_4[c], s=10, alpha=0.6, label=f"Cluster {c+1}")
axes[0].set_title("K-means Clusters on UMAP")
axes[0].legend(fontsize=8)

for s in range(4):
    mask = true_sub == s
    axes[1].scatter(umap_2d[mask, 0], umap_2d[mask, 1],
                    c=colors_4[s], s=10, alpha=0.6, label=f"Subtype {s+1}")
axes[1].set_title("True Subtypes on UMAP")
axes[1].legend(fontsize=8)

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** This is the full pipeline end to end: 500 patients with 50 features and four built-in subtypes, which we scale, compress to 15 principal components, cluster with K-means, and finally lay out on a 2D UMAP map for viewing. The figure has two panels showing the *same* UMAP coordinates twice --- on the left, dots are coloured by the cluster K-means assigned; on the right, by the patients' true (simulated) subtype. Because the layout is identical, you read them by comparing colours position-for-position: if the two panels colour the same blobs the same way, the algorithm recovered the real structure.

Comparing the two panels lets you assess how well the clustering algorithm recovered the true subtypes. In practice you do not know the true subtypes --- the comparison would instead be against clinical outcomes, treatment response, or other external validation.

## Cautionary Notes

::: {.callout-warning}
## Clusters are hypotheses, not facts
Finding three clusters in your ICU dataset does not mean there are three types of ICU patients. It means that your data, with your chosen features, using your chosen algorithm, with your chosen parameters, can be partitioned into three groups. Whether those groups are biologically meaningful requires external validation.
:::

**Always validate clinically.** Clusters should differ on outcomes, biomarkers, or treatment response --- not just on the features used to create them. Clustering on lab values and then showing that the clusters differ on lab values is circular reasoning.

**Try multiple algorithms.** If K-means, hierarchical clustering, and DBSCAN all find the same groups, you have more confidence that the structure is real. If they disagree, the structure may be fragile.

**Report negative results.** Many clustering analyses find no meaningful structure. This is a legitimate finding --- it means the disease population is more homogeneous than expected.

**Beware of overfitting to noise.** With enough features, clustering will always find patterns. Use stability analysis and, if possible, validate clusters in an independent dataset.

## Exercises

### Exercise 1: K-Means on Simulated Patient Data

Simulate a dataset of 600 patients with 6 clinical variables (heart rate, respiratory rate, temperature, systolic BP, creatinine, lactate) drawn from 3 underlying phenotypes.

a. Scale the data and apply K-means for $k = 2, 3, 4, 5, 6$.
b. Create elbow and silhouette plots. Which $k$ appears optimal?
c. Visualise the $k = 3$ solution using PCA.
d. Profile the clusters: compute the mean of each variable within each cluster. Do the profiles make clinical sense?

::: {.panel-tabset}

### R

```{r}
#| label: exercise-1-solution
#| code-fold: true
#| code-summary: "Show solution"
#| fig-width: 10
#| fig-height: 5

library(tidyverse)  # tibble(), bind_rows(), group_by/summarise/across

set.seed(42)
n <- 600

# Phenotype 1: Septic shock (high HR, RR, lactate, low SBP)
p1 <- tibble(hr = rnorm(200, 115, 12), rr = rnorm(200, 28, 5),
             temp = rnorm(200, 38.5, 0.8), sbp = rnorm(200, 80, 12),
             creat = rnorm(200, 2.5, 0.8), lactate = rnorm(200, 5, 2))
# Phenotype 2: Stable critical (moderate vitals)
p2 <- tibble(hr = rnorm(200, 90, 10), rr = rnorm(200, 20, 3),
             temp = rnorm(200, 37.2, 0.5), sbp = rnorm(200, 110, 15),
             creat = rnorm(200, 1.2, 0.3), lactate = rnorm(200, 1.5, 0.5))
# Phenotype 3: Febrile, preserved hemodynamics
p3 <- tibble(hr = rnorm(200, 100, 10), rr = rnorm(200, 22, 4),
             temp = rnorm(200, 39.2, 0.7), sbp = rnorm(200, 120, 10),
             creat = rnorm(200, 1.0, 0.2), lactate = rnorm(200, 2.0, 0.8))

dat <- bind_rows(p1, p2, p3)
dat_scaled <- scale(dat)

# Elbow + silhouette
library(cluster)
wcss <- sil <- numeric(6)
for (k in 2:6) {
  km <- kmeans(dat_scaled, k, nstart = 25)
  wcss[k] <- km$tot.withinss
  sil[k] <- mean(silhouette(km$cluster, dist(dat_scaled))[, 3])
}

par(mfrow = c(1, 2))
plot(2:6, wcss[2:6], type = "b", pch = 16, col = "steelblue",
     xlab = "k", ylab = "WCSS", main = "Elbow Method")
plot(2:6, sil[2:6], type = "b", pch = 16, col = "firebrick",
     xlab = "k", ylab = "Avg Silhouette", main = "Silhouette")

# Profile clusters
km3 <- kmeans(dat_scaled, 3, nstart = 25)
dat$cluster <- km3$cluster
cat("\nCluster profiles (original scale):\n")
dat %>%
  group_by(cluster) %>%
  summarise(across(hr:lactate, mean), .groups = "drop") %>%
  print()
```

### Python

```{python}
#| label: exercise-1-solution-py
#| code-fold: true
#| code-summary: "Show solution"

import numpy as np
import pandas as pd
from sklearn.preprocessing import StandardScaler
from sklearn.cluster import KMeans
from sklearn.metrics import silhouette_score

np.random.seed(42)

p1 = np.column_stack([np.random.normal(115,12,200), np.random.normal(28,5,200),
                       np.random.normal(38.5,0.8,200), np.random.normal(80,12,200),
                       np.random.normal(2.5,0.8,200), np.random.normal(5,2,200)])
p2 = np.column_stack([np.random.normal(90,10,200), np.random.normal(20,3,200),
                       np.random.normal(37.2,0.5,200), np.random.normal(110,15,200),
                       np.random.normal(1.2,0.3,200), np.random.normal(1.5,0.5,200)])
p3 = np.column_stack([np.random.normal(100,10,200), np.random.normal(22,4,200),
                       np.random.normal(39.2,0.7,200), np.random.normal(120,10,200),
                       np.random.normal(1.0,0.2,200), np.random.normal(2.0,0.8,200)])

X = np.vstack([p1, p2, p3])
cols = ['hr', 'rr', 'temp', 'sbp', 'creat', 'lactate']
df = pd.DataFrame(X, columns=cols)
X_s = StandardScaler().fit_transform(X)

for k in range(2, 7):
    km = KMeans(n_clusters=k, n_init=25, random_state=42).fit(X_s)
    sil = silhouette_score(X_s, km.labels_)
    print(f"k={k}: WCSS={km.inertia_:.0f}, Silhouette={sil:.3f}")

km3 = KMeans(n_clusters=3, n_init=25, random_state=42).fit(X_s)
df['cluster'] = km3.labels_
print("\nCluster profiles:")
print(df.groupby('cluster')[cols].mean().round(2))
```

:::

**What the solution shows.** The code simulates 600 patients from three vital-sign phenotypes (a septic-shock pattern, a stable pattern, and a febrile pattern), then runs the elbow and silhouette diagnostics and prints a **cluster profile table** --- the mean of each variable within each cluster, on the original scale. That profile table is the payoff: it is where you turn anonymous "Cluster 1, 2, 3" labels into clinical stories by reading off, say, that one cluster has high heart rate, high lactate, and low blood pressure and recognising it as the septic-shock group. A clustering you cannot interpret this way is rarely useful.

### Exercise 2: Hierarchical Clustering with Different Linkage Methods

Using the same dataset from Exercise 1:

a. Apply hierarchical clustering with single, complete, average, and Ward's linkage.
b. Cut each dendrogram at $k = 3$ and compare the resulting cluster assignments.
c. Which linkage method produces clusters most similar to the K-means solution? Use the adjusted Rand index (ARI) to quantify agreement.
d. Which linkage method would you recommend for clinical data and why?

### Exercise 3: DBSCAN for Outlier Detection

Add 30 "outlier" patients to the Exercise 1 dataset: patients with extreme values across multiple variables (e.g., HR = 180, lactate = 15, creatinine = 8).

a. Run K-means with $k = 3$. Where do the outliers end up?
b. Run DBSCAN with appropriate eps and min_samples. How many outliers does it identify?
c. Compare the cluster assignments for non-outlier patients between K-means and DBSCAN.
d. In a clinical phenotyping study, which approach would you prefer for handling outliers?

### Exercise 4: Full Pipeline --- PCA + Clustering + UMAP Visualisation

Simulate a dataset of 800 patients with 30 clinical variables and 4 underlying subtypes.

a. Standardise the data and apply PCA. Use a scree plot to choose the number of components.
b. Run K-means ($k = 2, 3, 4, 5$) on the PCA scores. Use silhouette scores to choose $k$.
c. Visualise the chosen clustering on a UMAP embedding.
d. Profile the clusters: compute mean values for each variable and present in a table.
e. Discuss: how would you validate these clusters in a real clinical study?

## Summary

Clustering is a powerful tool for discovering patient subgroups, but it requires careful application and sceptical interpretation. K-means is simple and effective for spherical clusters but requires specifying $k$ and is sensitive to outliers. Hierarchical clustering provides a dendrogram that reveals the full structure of merging and does not require pre-specifying $k$. DBSCAN handles irregular cluster shapes and identifies outliers. No single method is best for all situations; using multiple methods and checking for consistency is good practice. Above all, clusters are statistical constructs --- they must be validated against clinical outcomes and replicated in independent datasets before they can inform patient care.

::: {.callout-tip}
## Key Takeaways
1. Always scale your data before clustering.
2. K-means is fast and effective but assumes spherical clusters and requires choosing $k$.
3. Hierarchical clustering (Ward's method) is a solid default that does not require pre-specifying $k$.
4. DBSCAN excels when clusters are irregularly shaped or when outlier detection matters.
5. Validate clusters with internal metrics (silhouette), stability analysis, and --- most importantly --- clinical outcomes.
6. Clustering will always find groups, even in random data. The burden of proof is on you to show the groups are meaningful.
:::

## References and Further Reading

The most clinically impactful clustering study in recent critical care literature is Seymour et al.'s 2019 paper "Derivation, Validation, and Potential Treatment Implications of Novel Clinical Phenotypes for Sepsis" in *JAMA*, which identified four sepsis phenotypes using K-means clustering on electronic health record data. Calfee et al.'s 2014 paper "Subphenotypes in Acute Respiratory Distress Syndrome: Latent Class Analysis of Data From Two Randomised Controlled Trials" in *The Lancet Respiratory Medicine* applied a related method (latent class analysis) to identify ARDS subtypes with differential treatment response. Shah et al.'s 2015 paper in *Circulation* demonstrated the use of hierarchical clustering for heart failure with preserved ejection fraction phenotyping.

For the statistical foundations, Hastie, Tibshirani, and Friedman's *The Elements of Statistical Learning* (2nd edition, 2009) provides comprehensive coverage of K-means, hierarchical clustering, and spectral methods. James, Witten, Hastie, and Tibshirani's *An Introduction to Statistical Learning* (2nd edition, 2021) offers a more accessible treatment suitable for students. The original DBSCAN paper by Ester et al. (1996) is surprisingly readable and remains the standard reference for the algorithm.

For practical guidance, a 2024 review by Rodriguez et al. in *Critical Care Medicine* synthesises the rapidly growing literature on clustering-based phenotyping in critical illness, covering methodological best practices and the gap between statistical clusters and actionable clinical subtypes. Hennig et al.'s *Handbook of Cluster Analysis* (2016) is the most comprehensive reference, covering both methodology and applications. For validation, the clValid R package and its accompanying paper by Brock et al. (2008) in the *Journal of Statistical Software* provide tools for comparing clustering algorithms using internal, stability-based, and biological metrics.
