# Clustering: Discovering Groups in Data {#sec-clustering}

```{r}
#| include: false
#| eval: true
library(tidyverse)
library(knitr)
```

Clustering answers a question every clinician has asked: "are all my patients really the same, or are there hidden subgroups in here?" Given a collection of unlabelled data, **clustering** algorithms propose a grouping of samples so that you can then ask whether the groups differ in various covariates of interest. In clinical research, clustering can be used to identify **patient phenotypes**, that is subgroups of patients with similar clinical profiles but may not correspond to any established diagnostic category. These phenotypes can reveal heterogeneity within a disease, suggest new treatment targets, or identify patients who respond differently to therapy. A well-known example of clustering in practice is the identification of sepsis and ARDS subtypes.

In this chapter, we will detail three popular algorithms ($k$-means, hierarchical clustering, and DBSCAN), their potential pitfalls, methods for choosing the number of clusters and validating results, and how to integrate clustering with the dimensionality reduction methods from the previous chapter.

## K-Means Clustering

### General Algorithm

$k$-means is the simplest and most widely used clustering algorithm. It aims to partition the data into $k$ clusters, where each sample belongs to the cluster with the nearest *centroid* (the mean of the points in that cluster). The algorithm requires you to specify $k$ (the number of clusters) and then iteratively refines the cluster assignments and centroids until convergence. The naive algorithm works as follows:

1. **Initialise**: place $k$ **centroids** in the feature space. A naive way to do so is to pick $k$ random patients as the starting centroids, but there are more sophisticated methods (like $k$-means++) for speeding up convergence and avoiding poor local minima.
2. **Assign**: assign each patient to the nearest centroid (using **Euclidean distance**, the ordinary straight-line distance between two points).
3. **Update**: recompute each centroid as the mean of all patients assigned to it.
4. **Repeat steps 2--3** until assignments converge.

The intuition: drop $k$ flags on the map, send each patient to their nearest flag, then move each flag to the middle of the patients who joined it, and repeat until nobody switches flags.

::: {.callout-note}
## Scaling before clustering
Because $k$-means uses Euclidean distance, features on different scales will dominate the distance calculations. Standardising (mean = 0, SD = 1) before running $k$-means is thus common practice.
:::

$k$-means minimises the **within-cluster sum of squares (WCSS)**, that is the total squared distance from each patient to their assigned centroid. Intuitively, this number measures how tightly packed the groups are: tight, well-defined clusters give a small WCSS; loose, scattered ones give a large WCSS.

### Choosing K: Elbow Method, Silhouette Scores, Gap Statistic

By construction, $k$-means requires two inputs: the data and the number of clusters $k$. The data is given, but how to choose $k$? Although there is no single correct answer, several heuristics can help guide the choice.

**Elbow method.** Run $k$-means for a range of $k$ values and plot the optimal WCSS for each $k$. You should observe an "elbow" shape in the plot, where the WCSS decreases rapidly at first and then levels off. The "elbow" point suggests a good choice for $k$, as adding more clusters beyond this point yields diminishing returns in terms of reducing WCSS.

**Silhouette score.** For each sample, the silhouette score measures how similar it is to its own cluster compared to other clusters. Scores run from --1 (closer to another group) through 0 (sitting on a boundary) to +1 (comfortably in the right group). From this individual score, a common heuristic is to pick the $k$ that maximises the average silhouette score across all samples.

**Gap statistic.** The gap statistic compares the tightness (WCSS) of your real data to the tightness you would get from structureless random data spread evenly across the same space (the "reference null"). The "gap" is largest at the $k$ where your data shows more grouping than random noise would, i.e. where the structure is most likely to be real.

In practice, treat these as advisers rather than oracles: run all three, see where they agree, and let clinical sense settle disagreements.

### Limitations of K-Means

- **Assumes spherical, equally sized clusters.** Because $k$-means uses the Euclidean distance to measure similarity, it finds globular clusters. Thus, irregularly shaped clusters (which are not that uncommon in clinical data) can be difficult to recover.
- **Sensitive to initialisation.** Despite approaches like $k$-means++ and repeated initialisations to mitigate initialisation sensitivity, $k$-means can converge to local minima, especially when clusters are not well-separated or when the data is noisy.
- **Sensitive to outliers.** A single extreme point can pull a centroid substantially.

### Clinical Example: Patient Phenotyping from Lab Data

::: {.panel-tabset}

### R

```{r}
#| label: fig-kmeans-r
#| eval: true
#| fig-cap: "$k$-means clustering of simulated patient lab data: choosing k and visualising clusters."
#| fig-width: 14
#| fig-height: 5

library(cluster)

set.seed(42)
n <- 400

# Simulate 3 patient phenotypes with distinct lab profiles
pheno1 <- tibble(
  # Metabolic syndrome phenotype
  glucose = rnorm(n * 0.35, 140, 20),
  crp = rnorm(n * 0.35, 5, 2),
  albumin = rnorm(n * 0.35, 3.5, 0.3),
  wbc = rnorm(n * 0.35, 8, 2)
)
pheno2 <- tibble(
  # Inflammatory phenotype
  glucose = rnorm(n * 0.35, 100, 15),
  crp = rnorm(n * 0.35, 15, 4),
  albumin = rnorm(n * 0.35, 3.0, 0.4),
  wbc = rnorm(n * 0.35, 14, 3)
)
pheno3 <- tibble(
  # Healthy-ish phenotype
  glucose = rnorm(n * 0.30, 90, 10),
  crp = rnorm(n * 0.30, 2, 1),
  albumin = rnorm(n * 0.30, 4.2, 0.2),
  wbc = rnorm(n * 0.30, 7, 1.5)
)

lab_data <- bind_rows(pheno1, pheno2, pheno3)
lab_scaled <- scale(lab_data)

# Elbow and silhouette plots
wcss <- numeric(10)
sil_avg <- numeric(10)
for (k in 2:10) {
  km <- kmeans(lab_scaled, centers = k, nstart = 25)
  wcss[k] <- km$tot.withinss
  sil_avg[k] <- mean(silhouette(km$cluster, dist(lab_scaled))[, 3])
}

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# Elbow plot
plot(
  2:10,
  wcss[2:10],
  type = "b",
  pch = 16,
  col = "steelblue",
  xlab = "Number of clusters (k)",
  ylab = "Within-cluster SS",
  main = "Elbow Method"
)

# Silhouette plot
plot(
  2:10,
  sil_avg[2:10],
  type = "b",
  pch = 16,
  col = "firebrick",
  xlab = "Number of clusters (k)",
  ylab = "Average silhouette",
  main = "Silhouette Scores"
)

# $k$-means with k=3, visualise on first 2 PCs
km3 <- kmeans(lab_scaled, centers = 3, nstart = 25)
pca2 <- prcomp(lab_scaled)$x[, 1:2]
plot(
  pca2,
  col = c("steelblue", "firebrick", "forestgreen")[km3$cluster],
  pch = 16,
  cex = 0.7,
  main = "$k$-means (k=3) on PCA",
  xlab = "PC1",
  ylab = "PC2"
)
```

```{r}
#| label: fig-silhouette-r
#| eval: true
#| fig-cap: "Silhouette plot for $k$-means with k=3 on the lab data."
#| fig-width: 8
#| fig-height: 5

sil <- silhouette(km3$cluster, dist(lab_scaled))

plot(
  sil,
  col = c("steelblue", "firebrick", "forestgreen"),
  main = "Silhouette Plot (k = 3)",
  border = NA
)
cat("Average silhouette width:", round(mean(sil[, 3]), 3), "\n")
```

### Python

```{python}
#| label: fig-kmeans-py
#| eval: true
#| fig-cap: "$k$-means clustering: elbow method, silhouette scores, and cluster visualisation."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.preprocessing import StandardScaler
from sklearn.cluster import KMeans
from sklearn.metrics import silhouette_score
from sklearn.decomposition import PCA

np.random.seed(42)

# Simulate phenotypes
g1 = np.column_stack([np.random.normal(140, 20, 140),
                       np.random.normal(5, 2, 140),
                       np.random.normal(3.5, 0.3, 140),
                       np.random.normal(8, 2, 140)])
g2 = np.column_stack([np.random.normal(100, 15, 140),
                       np.random.normal(15, 4, 140),
                       np.random.normal(3.0, 0.4, 140),
                       np.random.normal(14, 3, 140)])
g3 = np.column_stack([np.random.normal(90, 10, 120),
                       np.random.normal(2, 1, 120),
                       np.random.normal(4.2, 0.2, 120),
                       np.random.normal(7, 1.5, 120)])

X = np.vstack([g1, g2, g3])
X_scaled = StandardScaler().fit_transform(X)

ks = range(2, 11)
wcss = []
sils = []
for k in ks:
    km = KMeans(n_clusters=k, n_init=25, random_state=42)
    km.fit(X_scaled)
    wcss.append(km.inertia_)
    sils.append(silhouette_score(X_scaled, km.labels_))

fig, axes = plt.subplots(1, 3, figsize=(16, 5))

axes[0].plot(list(ks), wcss, "o-", color="steelblue")
axes[0].set_xlabel("Number of clusters (k)")
axes[0].set_ylabel("Within-cluster SS")
axes[0].set_title("Elbow Method")

axes[1].plot(list(ks), sils, "o-", color="firebrick")
axes[1].set_xlabel("Number of clusters (k)")
axes[1].set_ylabel("Average silhouette")
axes[1].set_title("Silhouette Scores")

km3 = KMeans(n_clusters=3, n_init=25, random_state=42).fit(X_scaled)
pca2 = PCA(n_components=2).fit_transform(X_scaled)
colors = ["steelblue", "firebrick", "forestgreen"]
for c in range(3):
    mask = km3.labels_ == c
    axes[2].scatter(pca2[mask, 0], pca2[mask, 1], c=colors[c],
                    s=12, alpha=0.6, label=f"Cluster {c+1}")
axes[2].set_xlabel("PC1")
axes[2].set_ylabel("PC2")
axes[2].set_title("$k$-means (k=3) on PCA")
axes[2].legend()

plt.tight_layout()
plt.show()
```

```{python}
#| label: fig-silhouette-py
#| eval: true
#| fig-cap: "Silhouette scores by cluster for $k$-means with k=3."

from sklearn.metrics import silhouette_samples

sil_vals = silhouette_samples(X_scaled, km3.labels_)
avg_sil = silhouette_score(X_scaled, km3.labels_)

fig, ax = plt.subplots(figsize=(8, 5))
y_lower = 0
colors = ["steelblue", "firebrick", "forestgreen"]
for c in range(3):
    c_sil = np.sort(sil_vals[km3.labels_ == c])
    y_upper = y_lower + len(c_sil)
    ax.barh(range(y_lower, y_upper), c_sil, height=1.0,
            color=colors[c], edgecolor="none")
    y_lower = y_upper + 5

ax.axvline(avg_sil, color="black", linestyle="--",
           label=f"Average: {avg_sil:.3f}")
ax.set_xlabel("Silhouette Coefficient")
ax.set_ylabel("Patients (sorted within cluster)")
ax.set_title("Silhouette Plot (k=3)")
ax.legend()
plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** We simulate 400 patients drawn from three hidden lab-profile phenotypes (a metabolic, an inflammatory, and a healthier group), scale the labs, and then ask $k$-means to try every value of $k$ from 2 to 10, recording two diagnostics each time. The **elbow plot** (left) shows within-cluster spread falling as $k$ rises, with an elbow around $k = 3$. The **silhouette-vs-k plot** (middle) also peaks around $k = 3$. The **PCA scatter** (right) depicts the first two principal components, with each patient coloured by assigned cluster for $k = 3$, showing well-separated colours overall. Finally, the **per-point silhouette plot** shows one horizontal bar per patient: wide, positive bars mean the patient sits comfortably in its cluster; short or negative bars flag patients near a cluster boundary. If one cluster's bars are systematically shorter, that cluster is poorly defined.

## Agglomerative Clustering

### General algorithm

**Agglomerative clustering** is a *hierarchical* clustering method, meaning that it involves building a hierarchy of clusters rather than a single flat partition. Most implementations require you to specify three parameters: $k$ cluster (like in $k$-means), a **linkage method** (how to measure distance between clusters, see below), and a **distance metric** (how to measure distance between patients). The general algorithm involves two steps:

- Start with each patient in their own cluster, and iteratively merge the closest clusters (as determined by the linkage method) until all patients are in a single cluster. The resulting hierarchy can be represented as a **dendrogram**.
- Cut the dendrogram into $k$ clusters by choosing a height at which to "cut" the tree. The height corresponds to the distance at which clusters are merged, so cutting low gives many small clusters, while cutting high gives a few large clusters.

To pick $k$, you can use silhouette scores and the gap statistic as in $k$-means. The elbow method is less natural here as we do not explicitly minimise a metric such as the WCSS (except for Ward's method, see *Linkage Methods*). The dendrogram itself can also provide insight into the natural number of clusters by looking for large gaps in the merging heights (see *Reading a Dendrogram*).

```{mermaid}
%%| echo: false
%%| eval: true
%%| label: fig-dendrogram-concept
%%| fig-cap: "A dendrogram in miniature: similar patients merge low down (early), dissimilar groups merge higher up. Cutting across the tree at a chosen height defines the clusters."
flowchart TD
    M2["Merge: all patients"] --> M1["Merge A+B"]
    M2 --> P5["Patient E"]
    M1 --> A["Patient A"]
    M1 --> M0["Merge C+D"]
    M0 --> C["Patient C"]
    M0 --> D["Patient D"]
```

### Linkage Methods

**Linkage methods** are the crux of the merging rule in agglomerative clustering, as they determine how to measure the distance between clusters. Depending on the use case, several linkage methods can be employed:

**Single linkage.** Distance = the *closest* pair of patients across the two clusters. Two groups count as near if they have even one near-neighbour, which tends to produce long, straggly, chain-like clusters.

**Complete linkage.** Distance = the *farthest* pair of patients. Groups merge only when even their most distant members are reasonably close. This tends to produce tight, roughly equal-sized clusters.

**Average linkage.** Distance = the *average* of all patient-to-patient distances between the two clusters. A compromise between single and complete.

**Ward's method.** At each step, merges the pair of clusters that increases total within-cluster spread (WCSS) the least. This tends to produce compact, roughly spherical clusters of similar size. This is a common default choice in modern package implementations.

::: {.panel-tabset}

### R

```{r}
#| label: fig-dendrogram-r
#| eval: true
#| fig-cap: "Dendrogram from hierarchical clustering (Ward's method) with 3 clusters highlighted."
#| fig-width: 10
#| fig-height: 6

# Use a smaller subset for readable dendrogram
set.seed(42)
idx <- sample(nrow(lab_scaled), 80)
hc <- hclust(dist(lab_scaled[idx, ]), method = "ward.D2")

# Plot dendrogram
plot(
  hc,
  labels = FALSE,
  main = "Hierarchical Clustering (Ward's Method)",
  xlab = "",
  sub = "",
  hang = -1,
  cex = 0.6
)
rect.hclust(hc, k = 3, border = c("steelblue", "firebrick", "forestgreen"))
```

```{r}
#| label: fig-linkage-comparison-r
#| eval: true
#| fig-cap: "Comparison of linkage methods on the same dataset."
#| fig-width: 14
#| fig-height: 4

par(mfrow = c(1, 4), mar = c(2, 2, 3, 1))
methods <- c("single", "complete", "average", "ward.D2")
titles <- c("Single", "Complete", "Average", "Ward's")

for (i in seq_along(methods)) {
  hc_i <- hclust(dist(lab_scaled[idx, ]), method = methods[i])
  plot(
    hc_i,
    labels = FALSE,
    main = titles[i],
    xlab = "",
    sub = "",
    hang = -1,
    cex = 0.5
  )
  rect.hclust(hc_i, k = 3, border = "firebrick")
}
```

### Python

```{python}
#| label: fig-dendrogram-py
#| eval: true
#| fig-cap: "Dendrogram from hierarchical clustering (Ward's method)."

import numpy as np
import matplotlib.pyplot as plt
from scipy.cluster.hierarchy import dendrogram, linkage, fcluster

np.random.seed(42)
idx = np.random.choice(len(X_scaled), 80, replace=False)
X_sub = X_scaled[idx]

Z = linkage(X_sub, method="ward")

fig, ax = plt.subplots(figsize=(12, 5))
dendrogram(Z, ax=ax, no_labels=True, color_threshold=Z[-3, 2])
ax.set_title("Hierarchical Clustering (Ward's Method)")
ax.set_xlabel("Patients")
ax.set_ylabel("Height")
ax.axhline(y=Z[-3, 2], color="firebrick", linestyle="--", linewidth=1)
ax.text(5, Z[-3, 2] + 0.5, "Cut for k=3", color="firebrick", fontsize=10)
plt.tight_layout()
plt.show()
```

```{python}
#| label: fig-linkage-comparison-py
#| eval: true
#| fig-cap: "Comparison of linkage methods."

fig, axes = plt.subplots(1, 4, figsize=(18, 4))
methods = ['single', 'complete', 'average', 'ward']
titles = ['Single', 'Complete', 'Average', "Ward's"]

for ax, method, title in zip(axes, methods, titles):
    Z_i = linkage(X_sub, method=method)
    dendrogram(Z_i, ax=ax, no_labels=True, color_threshold=0)
    ax.set_title(title)
    ax.set_xlabel("")

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** The first block builds a **dendrogram**, the upside-down tree that records, branch by branch, which patients (then which groups of patients) got merged and at what distance. The coloured rectangles show where we "cut" the tree to get three clusters. The second block redraws the dendrogram four times, once per linkage rule (single, complete, average, Ward's), so you can see how the choice of linkage reshapes the tree. The practical lesson from that comparison: single linkage often produces one lopsided "chain" with stragglers, while Ward's gives balanced, compact groups, which is why Ward's is the recommended default for clinical data.

### Reading a Dendrogram

The dendrogram encodes the full merging history:

- The **height** of each horizontal line represents the distance at which two clusters were merged.
- **Long vertical lines** before a merge indicate well-separated clusters.
- **Short vertical lines** indicate merges of similar clusters, i.e. less clear separation.

Cutting the dendrogram at different heights gives different numbers of clusters. A large "gap" between merging heights (a long vertical line) suggests a natural number of clusters.

## DBSCAN: Density-Based Clustering

As mentioned earlier, clusters in clinical data may be irregularly shaped, overlap, or contain noise (e.g., outlier patients who do not belong to any group). Compared to $k$-means and agglomerative clustering, **DBSCAN** (Density-Based Spatial Clustering of Applications with Noise) takes a different approach that defines clusters based on density rather than distance. Importantly, DBSCAN does not require specifying the number of clusters in advance, and it can identify outliers explicitly.

DBSCAN defines clusters as dense (crowded) regions of data points separated by sparse (empty) regions, similar to spotting towns on a night-time satellite photo by their clusters of lights. Two settings control its behaviour:

- **eps ($\varepsilon$)**: how close two samples must be to count as neighbours, i.e. the radius of the neighbourhood drawn around each sample.
- **min_samples**: how many neighbours a sample needs within that radius before the surrounding area counts as "crowded" (dense).

The algorithm then sorts every sample into one of three types:

1. **Core point**: sits in a crowded area, has at least `min_samples` neighbours within radius $\varepsilon$.
2. **Border point**: on the edge of a crowd, close to a core point but without enough neighbours of its own to be core.
3. **Noise point**: out on its own, neither core nor border. These are the **outliers**, and DBSCAN flags them explicitly instead of forcing them into a group.

Clusters grow by linking up core points that fall within $\varepsilon$ of each other, so a cluster can snake into any shape the dense region happens to take.

### Advantages and Limitations

**Advantages:**

- Does not require specifying $k$ in advance.
- Can find arbitrarily shaped clusters.
- Identifies outliers explicitly.

**Limitations:**

- Sensitive to the choice of $\varepsilon$ and `min_samples`.
- Struggles when clusters have very different densities.
- Not ideal for very high-dimensional data. In high dimensions, pairwise distances concentrate, which makes it hard to distinguish dense from sparse regions. Applying dimensionality reduction first (PCA) can help.

::: {.panel-tabset}

### R

```{r}
#| label: fig-dbscan-r
#| eval: true
#| fig-cap: "DBSCAN identifies irregularly shaped clusters and noise points."
#| fig-width: 10
#| fig-height: 5

library(dbscan)

set.seed(42)

# Create data with two crescent-shaped clusters + noise
n_each <- 150
theta1 <- seq(0, pi, length.out = n_each)
theta2 <- seq(0, pi, length.out = n_each)

x1 <- cos(theta1) + rnorm(n_each, 0, 0.08)
y1 <- sin(theta1) + rnorm(n_each, 0, 0.08)
x2 <- 1 - cos(theta2) + rnorm(n_each, 0, 0.08)
y2 <- 1 - sin(theta2) - 0.5 + rnorm(n_each, 0, 0.08)

# Add noise
x_noise <- runif(30, -0.5, 2)
y_noise <- runif(30, -1, 1.5)

crescent_data <- rbind(
  cbind(x1, y1),
  cbind(x2, y2),
  cbind(x_noise, y_noise)
)

# Compare $k$-means and DBSCAN
km_crescents <- kmeans(crescent_data, centers = 2, nstart = 25)
db_crescents <- dbscan(crescent_data, eps = 0.15, minPts = 5)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
cols_km <- c("steelblue", "firebrick")[km_crescents$cluster]
plot(
  crescent_data,
  col = cols_km,
  pch = 16,
  cex = 0.7,
  main = "$k$-means (k=2)",
  xlab = "x",
  ylab = "y"
)

# DBSCAN: cluster 0 = noise (grey), others = colours
db_cols <- c("grey50", "steelblue", "firebrick", "forestgreen")
plot(
  crescent_data,
  col = db_cols[db_crescents$cluster + 1],
  pch = ifelse(db_crescents$cluster == 0, 4, 16),
  cex = 0.7,
  main = "DBSCAN (eps=0.15)",
  xlab = "x",
  ylab = "y"
)
legend(
  "topright",
  c("Noise", "Cluster 1", "Cluster 2"),
  col = db_cols[1:3],
  pch = c(4, 16, 16),
  cex = 0.8
)
```

### Python

```{python}
#| label: fig-dbscan-py
#| eval: true
#| fig-cap: "DBSCAN correctly identifies crescent-shaped clusters that $k$-means cannot handle."

import numpy as np
import matplotlib.pyplot as plt
from sklearn.cluster import KMeans, DBSCAN

np.random.seed(42)
n_each = 150

theta1 = np.linspace(0, np.pi, n_each)
theta2 = np.linspace(0, np.pi, n_each)

x1 = np.cos(theta1) + np.random.normal(0, 0.08, n_each)
y1 = np.sin(theta1) + np.random.normal(0, 0.08, n_each)
x2 = 1 - np.cos(theta2) + np.random.normal(0, 0.08, n_each)
y2 = 1 - np.sin(theta2) - 0.5 + np.random.normal(0, 0.08, n_each)

x_noise = np.random.uniform(-0.5, 2, 30)
y_noise = np.random.uniform(-1, 1.5, 30)

X_crescent = np.vstack(
    [
        np.column_stack([x1, y1]),
        np.column_stack([x2, y2]),
        np.column_stack([x_noise, y_noise]),
    ]
)

km = KMeans(n_clusters=2, n_init=25, random_state=42).fit(X_crescent)
db = DBSCAN(eps=0.15, min_samples=5).fit(X_crescent)

fig, axes = plt.subplots(1, 2, figsize=(12, 5))

colors_km = ["steelblue" if c == 0 else "firebrick" for c in km.labels_]
axes[0].scatter(X_crescent[:, 0], X_crescent[:, 1], c=colors_km, s=10)
axes[0].set_title("$k$-means (k=2)")

color_map = {-1: "grey", 0: "steelblue", 1: "firebrick"}
colors_db = [color_map.get(c, "forestgreen") for c in db.labels_]
markers = ["x" if c == -1 else "o" for c in db.labels_]
for c_val in set(db.labels_):
    mask = db.labels_ == c_val
    m = "x" if c_val == -1 else "o"
    label = "Noise" if c_val == -1 else f"Cluster {c_val+1}"
    axes[1].scatter(
        X_crescent[mask, 0],
        X_crescent[mask, 1],
        c=color_map.get(c_val, "forestgreen"),
        marker=m,
        s=15,
        label=label,
    )
axes[1].set_title("DBSCAN (eps=0.15)")
axes[1].legend(fontsize=9)

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** We deliberately build a hard case: two interlocking crescent (banana) shapes plus a scatter of random "noise" patients who belong to neither. The left panel is $k$-means asked for two clusters. Because it can only carve out round blobs, it slices each crescent in half. The right panel is DBSCAN, which grows clusters by following dense trails of points, so it traces each crescent correctly and marks the scattered outliers as "noise" (the crosses).

## Cluster Validation

Beyond internal metrics described above (e.g., the silhouette score, as well as other existing ones such as the Calinski-Harabasz index), a stronger test is whether the clusters are **stable**. Internal metrics can be misleading: they can be high even for meaningless clusters.

**Bootstrap stability** provides a more robust check:

1. Resample the data with replacement many times.
2. Re-run the clustering on each bootstrap sample.
3. Measure how often the same points end up in the same cluster.

If clusters are stable across resamples, they likely reflect real structure. If membership changes dramatically, the clusters may be artefacts.

## Clinical Applications

### Patient Phenotyping

The most impactful use of clustering in clinical research is identifying disease subtypes that were previously unrecognised. Notable examples:

- **ARDS phenotypes.** @calfee2014ards applied latent class analysis to two ARDS clinical trial datasets and identified two phenotypes ("hyperinflammatory" and "hypoinflammatory") with different mortality rates and different responses to treatment. The hyperinflammatory phenotype had higher IL-6, higher vasopressor use, and worse outcomes.

- **Sepsis endotypes.** @seymour2019sepsis used $k$-means clustering on EHR data from over 20,000 sepsis patients and identified four phenotypes (alpha, beta, gamma, delta) with distinct organ dysfunction patterns and 28-day mortality rates ranging from 5% to 40%.

- **Heart failure subtypes.** @shah2015phenomapping applied hierarchical clustering to echocardiographic and clinical data from HFpEF patients and identified three phenotypes with different pathophysiology and prognosis.

### Disease Subtyping from Lab Panels

A practical workflow for phenotyping:

1. Select clinically relevant variables (labs, vitals, demographics).
2. Handle missing data (imputation or exclusion).
3. Standardise all variables.
4. Reduce dimensionality (PCA to 10--20 components if many variables).
5. Apply multiple clustering algorithms ($k$-means, hierarchical) and compare.
6. Validate: silhouette scores, stability analysis, and (most importantly) clinical interpretation.
7. Compare clusters on outcomes (mortality, length of stay, treatment response).

## Visualising Clusters

::: {.callout-warning}
## Never cluster on t-SNE or UMAP coordinates

t-SNE and UMAP distort distances and densities. They can create apparent clusters where none exist, and deform continuous structure into misleading shapes [@chari2023]. Always **cluster first on the original scaled data**, then use t-SNE/UMAP only for visualisation. PCA does not suffer from this problem because it is a linear projection that preserves distances.
:::

To see why this matters, consider the following toy datasets. First, we build a single multivariate Gaussian with no apparent clusters at all. We draw 1000 points from a 100-dimensional Gaussian where all variables are correlated ($\rho = 0.8$), then colour each point by its rank along the first principal component (blue = low, red = high).

::: {.panel-tabset}

#### R

```{r}
#| label: fig-deformed-structure-r
#| eval: true
#| fig-cap: "A single correlated Gaussian (no clusters). t-SNE and UMAP deform the continuous structure; PCA preserves it."
#| fig-width: 12
#| fig-height: 4

library(uwot)
library(Rtsne)
library(MASS)
library(patchwork)

set.seed(42)

n <- 1000
p <- 100
Sigma <- matrix(0.8, p, p)
diag(Sigma) <- 1

x <- mvrnorm(n, rep(0, p), Sigma)

ranks <- rank(rowMeans(x))
colours <- colorRampPalette(c("steelblue", "white", "firebrick"))(n)[ranks]

pca_2d <- prcomp(x)$x[, 1:2]
tsne_2d <- Rtsne(x, dims = 2, verbose = FALSE)$Y
umap_2d <- umap(x, verbose = FALSE)

plot_df <- data.frame(
  x = c(tsne_2d[, 1], umap_2d[, 1], pca_2d[, 1]),
  y = c(tsne_2d[, 2], umap_2d[, 2], pca_2d[, 2]),
  rank = rep(ranks, 3),
  method = factor(
    rep(c("t-SNE", "UMAP", "PCA"), each = n),
    levels = c("t-SNE", "UMAP", "PCA")
  )
)

ggplot(plot_df, aes(x, y, colour = rank)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_colour_gradient2(
    low = "steelblue",
    mid = "white",
    high = "firebrick",
    midpoint = median(ranks)
  ) +
  facet_wrap(~method, scales = "free") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.title = element_blank(),
    axis.text = element_blank()
  )
```

#### Python

```{python}
#| label: fig-deformed-structure-py
#| eval: true
#| fig-cap: "A single correlated Gaussian (no clusters). t-SNE and UMAP deform the continuous structure; PCA preserves it."

import numpy as np
import matplotlib.pyplot as plt
from matplotlib.colors import Normalize
import matplotlib.cm as cm
from scipy.stats import rankdata
from sklearn.decomposition import PCA
from sklearn.manifold import TSNE
from umap import UMAP

np.random.seed(42)

n, p = 1000, 100
Sigma = np.full((p, p), 0.8)
np.fill_diagonal(Sigma, 1.0)
x = np.random.multivariate_normal(np.zeros(p), Sigma, size=n)

ranks = rankdata(x.mean(axis=1))
norm = Normalize(vmin=ranks.min(), vmax=ranks.max())
colors = [cm.coolwarm(norm(r)) for r in ranks]

pca_2d  = PCA(n_components=2).fit_transform(x)
tsne_2d = TSNE(random_state=42).fit_transform(x)
umap_2d = UMAP(random_state=42).fit_transform(x)

fig, axs = plt.subplots(1, 3, figsize=(15, 5))

for ax, coords, title in zip(
    axs,
    [tsne_2d, umap_2d, pca_2d],
    ["t-SNE", "UMAP", "PCA"],
):
    ax.scatter(coords[:, 0], coords[:, 1], c=colors, s=10, alpha=0.6)
    ax.set_title(title, fontsize=16)
    ax.tick_params(labelbottom=False, labelleft=False)

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** The data is a single smooth cloud, yet t-SNE folds the gradient into a ribbon and UMAP bends it into an arc or splits it into blobs. Both create shapes that could invite false clustering. PCA shows the truth: one continuous cloud with a smooth colour gradient.

In a second example, we draw just 100 observations from a 3-dimensional correlated Gaussian (the same generating process as above, but with $p = 3$ and $n = 100$). Once again, there are no clusters in this data.

::: {.panel-tabset}

#### R

```{r}
#| label: fig-phantom-clusters-r
#| eval: true
#| fig-cap: "100 points from a single 3D Gaussian. t-SNE and UMAP create phantom clusters; PCA shows the truth."
#| fig-width: 12
#| fig-height: 4

library(uwot)
library(Rtsne)

set.seed(42)

n <- 100
p <- 3
Sigma <- matrix(0.8, p, p)
diag(Sigma) <- 1
x <- mvrnorm(n, rep(0, p), Sigma)

ranks <- rank(rowMeans(x))

pca_2d <- prcomp(x)$x[, 1:2]
tsne_2d <- Rtsne(x, dims = 2, verbose = FALSE)$Y
umap_2d <- umap(x, verbose = FALSE)

plot_df <- data.frame(
  x = c(tsne_2d[, 1], umap_2d[, 1], pca_2d[, 1]),
  y = c(tsne_2d[, 2], umap_2d[, 2], pca_2d[, 2]),
  rank = rep(ranks, 3),
  method = factor(
    rep(c("t-SNE", "UMAP", "PCA"), each = n),
    levels = c("t-SNE", "UMAP", "PCA")
  )
)

ggplot(plot_df, aes(x, y, colour = rank)) +
  geom_point(alpha = 0.8, size = 1.5) +
  scale_colour_gradient2(
    low = "steelblue",
    mid = "white",
    high = "firebrick",
    midpoint = median(ranks)
  ) +
  facet_wrap(~method, scales = "free") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.title = element_blank(),
    axis.text = element_blank()
  )
```

#### Python

```{python}
#| label: fig-phantom-clusters-py
#| eval: true
#| fig-cap: "100 points from a single 3D Gaussian. t-SNE and UMAP create phantom clusters; PCA shows the truth."

import numpy as np
import matplotlib.pyplot as plt
from matplotlib.colors import Normalize
import matplotlib.cm as cm
from scipy.stats import rankdata
from sklearn.decomposition import PCA
from sklearn.manifold import TSNE
from umap import UMAP

np.random.seed(42)

n, p = 100, 3
Sigma = np.full((p, p), 0.8)
np.fill_diagonal(Sigma, 1.0)
x = np.random.multivariate_normal(np.zeros(p), Sigma, size=n)

ranks = rankdata(x.mean(axis=1))
norm = Normalize(vmin=ranks.min(), vmax=ranks.max())
colors = [cm.coolwarm(norm(r)) for r in ranks]

pca_2d  = PCA(n_components=2).fit_transform(x)
tsne_2d = TSNE(random_state=42).fit_transform(x)
umap_2d = UMAP(random_state=42).fit_transform(x)

fig, axs = plt.subplots(1, 3, figsize=(15, 5))

for ax, coords, title in zip(
    axs,
    [tsne_2d, umap_2d, pca_2d],
    ["t-SNE", "UMAP", "PCA"],
):
    ax.scatter(coords[:, 0], coords[:, 1], c=colors, s=80, alpha=0.8)
    ax.set_title(title, fontsize=16)
    ax.tick_params(labelbottom=False, labelleft=False)

plt.tight_layout()
plt.show()
```

:::

**What the code is showing.** Here, t-SNE and UMAP fragment the smooth cloud into what look like discrete groups, while PCA correctly shows a single continuous spread. With small sample sizes, this "phantom cluster" effect can be especially pronounced.

## Cautionary Notes

::: {.callout-important}
## Clusters can be artefacts
Clustering algorithms will always produce clusters, even in completely random data. The existence of clusters in your output does not mean they are real. Validation (both statistical and clinical) is essential.
:::

::: {.callout-warning}
## Clusters are hypotheses, not facts
Finding three clusters in your ICU dataset does not mean there are three types of ICU patients. It means that your data, with your chosen features, using your chosen algorithm, with your chosen parameters, can be partitioned into three groups. Whether those groups are biologically meaningful requires external validation.
:::

**Always validate clinically.** Clusters should differ on outcomes, biomarkers, or treatment response, not just on the features used to create them. Clustering on lab values and then showing that the clusters differ on lab values is circular reasoning.

**Try multiple algorithms.** If $k$-means, hierarchical clustering, and DBSCAN all find the same groups, you have more confidence that the structure is real. If they disagree, the structure may be fragile.

**Report negative results.** Many clustering analyses find no meaningful structure. This is a legitimate finding: it means the disease population is more homogeneous than expected.

**Beware of overfitting to noise.** With enough features, clustering will always find patterns. Use stability analysis and, if possible, validate clusters in an independent dataset.

## Exercises

::: {.callout-tip title="Exercise 1: K-Means on Simulated Patient Data"}

Simulate a dataset of 600 patients with 6 clinical variables (heart rate, respiratory rate, temperature, systolic BP, creatinine, lactate) drawn from 3 underlying phenotypes.

a. Scale the data and apply $k$-means for $k = 2, 3, 4, 5, 6$.
b. Create elbow and silhouette plots. Which $k$ appears optimal?
c. Visualise the $k = 3$ solution using PCA.
d. Profile the clusters: compute the mean of each variable within each cluster. Do the profiles make clinical sense?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}

### R
```{r}
#| eval: false
#| file: ../solutions/R/ch16_ex1.R
```
### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch16_ex1.py
```

:::
:::

:::

::: {.callout-tip title="Exercise 2: Hierarchical Clustering with Different Linkage Methods"}

Using the same dataset from Exercise 1:

a. Apply hierarchical clustering with single, complete, average, and Ward's linkage.
b. Cut each dendrogram at $k = 3$ and compare the resulting cluster assignments.
c. Which linkage method produces clusters most similar to the $k$-means solution? You can perform that qualitatively or quantitatively using the adjusted Rand index ([Python](https://scikit-learn.org/stable/modules/generated/sklearn.metrics.adjusted_rand_score.html); [R](https://search.r-project.org/CRAN/refmans/mclust/html/adjustedRandIndex.html)).
d. Which linkage method would you recommend for clinical data and why?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch16_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch16_ex2.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 3: DBSCAN for Outlier Detection"}

Add 30 "outlier" patients to the Exercise 1 dataset: patients with extreme values across multiple variables (e.g., HR = 180, lactate = 15, creatinine = 8).

a. Run $k$-means with $k = 3$. Where do the outliers end up?
b. Run DBSCAN with appropriate eps and min_samples. How many outliers does it identify?
c. Compare the cluster assignments for non-outlier patients between $k$-means and DBSCAN.
d. In a clinical phenotyping study, which approach would you prefer for handling outliers?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch16_ex3.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch16_ex3.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 4: Full Pipeline (PCA + Clustering + UMAP Visualisation)"}

Simulate a dataset of 800 patients with 30 clinical variables and 4 underlying subtypes.

a. Standardise the data and apply PCA. Use a scree plot to choose the number of components.
b. Run $k$-means ($k = 2, 3, 4, 5$) on the PCA scores. Use silhouette scores to choose $k$.
c. Visualise the chosen clustering on a UMAP embedding.
d. Profile the clusters: compute mean values for each variable and present in a table.
e. Discuss: how would you validate these clusters in a real clinical study?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ch16_ex4.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ch16_ex4.py
```
:::
:::

:::

## Summary

Clustering is a powerful tool for discovering patient subgroups. $k$-means is simple and effective for spherical clusters but requires specifying $k$ and is sensitive to outliers. Hierarchical agglomerative clustering provides a dendrogram that reveals the full structure of merging. DBSCAN handles irregular cluster shapes and identifies outliers. No single method is best for all situations; using multiple methods and checking for consistency is good practice. Above all, clusters are statistical constructs: they must be validated against clinical outcomes and replicated in independent datasets before they can inform patient care.

## References and Further Reading

- For clinical phenotyping applications, see @seymour2019sepsis (A practical guide to PyMC for applied researchers), @calfee2014ards (applied latent class analysis to identify ARDS subtypes with differential treatment response), and @shah2015phenomapping (demonstrated the use of hierarchical clustering for HFpEF phenotyping).
- For clustering algorithms and methods, see @rodriguez2024clustering (a review synthesising methodological best practices and the gap between statistical clusters and actionable clinical subtypes), @ester1996dbscan (the original DBSCAN paper; surprisingly readable), and @james2021islr.

::: {.sectionrefs}
:::
