# Developing Clinical Prediction Models: A Complete Workflow {#sec-building-prediction-models}

## Introduction

A clinical prediction model estimates a patient's risk of an outcome (for example, the chance of a heart attack in the next 10 years) from information you can collect at the bedside. Building one well is not simply a matter of fitting a regression or running a machine learning algorithm. It is a structured process that begins with a clearly defined clinical question and proceeds through study design, data preparation, model specification, and evaluation. At every step, decisions must be guided by clinical knowledge, not just statistical convenience.

This chapter walks you through the complete workflow of developing a clinical prediction model, following the framework described by @smits2026prediction and the foundational principles laid out by @steyerberg2019. We will use the Framingham Heart Study data as a running example to illustrate each step, building a model to predict 10-year cardiovascular disease (CVD) risk.

@fig-workflow shows the whole journey at a glance. Each step in this chapter corresponds to one box.

```{mermaid}
%%| echo: false
%%| eval: true
%%| label: fig-workflow
%%| fig-cap: "The end-to-end workflow for developing a clinical prediction model. Each box is one step in this chapter; the dashed arrow is a reminder that the process is iterative, not strictly one-directional. Conceptual schematic based on the prediction-model development framework of @steyerberg2019 and @riley2019minimum, aligned with TRIPOD+AI reporting guidance [@collins2024tripodai]."
flowchart TD
    A[Define the clinical question] --> B[Design study & sample size]
    B --> C[Handle missing data]
    C --> D[Select predictors]
    D --> E[Specify & fit the model]
    E --> F[Assess: discrimination & calibration]
    F --> G[Validate internally]
    G --> H[Report transparently]
    F -. revise .-> E
```

## Step 1: Define the Clinical Question

Before touching any data, you must answer four questions:

1. **Who** is the target population? (e.g., adults aged 40--70 without prior CVD)
2. **What** is the outcome? (e.g., first CVD event within 10 years)
3. **When** is the prediction made? (e.g., at a routine primary care visit)
4. **Why** is the prediction needed? (e.g., to guide statin therapy decisions)

**Why a clinician should care.** The intended use determines everything that follows. A model designed to triage emergency department patients needs different predictors, a different time horizon, and different performance requirements than a model for long-term cardiovascular risk stratification. Getting the question wrong here cannot be fixed later by clever statistics.

::: {.callout-tip}
## The PICOTS Framework
You can adapt the familiar PICOTS framework from clinical research:
**P**opulation, **I**ntended use (Index model), **C**omparators (existing tools), **O**utcome, **T**iming, **S**etting.
:::

### Common Pitfalls in Problem Definition

- **Vague outcomes**: "clinical deterioration" is not a well-defined outcome. Define exactly what events count, how they are ascertained, and over what time window.
- **Leaky predictors**: including information that would not be available at the moment of prediction. If you are predicting ICU admission at the time of hospital arrival, you cannot use ICU vital signs as predictors.
- **Wrong population**: developing a model in hospitalised patients and hoping to apply it in primary care. The **case-mix** — the blend of ages, severities and comorbidities that makes up the patient group — will be entirely different, and a model tuned to one blend can behave quite differently in another.

## Step 2: Study Design and Sample Size

### Study Design

Clinical prediction models are typically developed from **cohort studies** (prospective or retrospective) or **existing databases** (electronic health records, registries). Randomised trials can also be used, though their selected populations may limit generalisability.

Key design requirements:

- **Inception cohort**: all patients should be enrolled at a similar, well-defined point in their clinical trajectory (the "moment of prediction").
- **Complete follow-up**: outcomes must be ascertained for as many patients as possible. Differential loss to follow-up can bias estimates.
- **Outcome rate**: there must be a sufficient number of events to support the number of candidate predictors.

### Sample Size: The Events Per Variable Rule and Beyond

**Why a clinician should care.** A model built on too few patients will look impressive in the paper that describes it and then fail when used on your own patients. Working out the sample size in advance is how you avoid that disappointment.

The classical rule of thumb requires at least **10 events per variable (EPV)** in the model. "Events per variable" means the number of outcome events (here, CVD events) divided by the number of predictors you are estimating. So for a logistic regression with 8 predictors, you would need at least 80 events. With 10% prevalence, that means at least 800 patients. The intuition: each predictor needs a certain amount of "evidence" (events) to be estimated reliably, and it is the rarer of the two groups (the events) that does the heavy lifting.

However, this rule is a rough guide: it asks for the same number of events whether the outcome is common or rare and whether the predictors are strong or weak. Riley and colleagues [@riley2019minimum] replaced it with a framework that requires the sample size to satisfy **three criteria at once** when the outcome is binary. Each one guards against a different failure:

1. **The coefficients should not need much shrinking.** A model fitted to too little data exaggerates its predictor effects, so honest predictions require multiplying every coefficient by a *shrinkage factor* below 1 (discussed later in this chapter). The criterion asks for the expected shrinkage factor to be at least **0.9** — the effects need scaling back by no more than about 10%. This is usually the criterion that drives the answer.
2. **Apparent performance should not be much better than honest performance.** Judge a model on the same patients it was fitted to and it always looks better than it really is; that gap is called *optimism*. The criterion puts a ceiling on it: the model's flattering self-assessment and its honest one must differ by no more than **0.05** on a measure of explained variation (the Nagelkerke R-squared, explained in the box below).
3. **The average risk should be estimated precisely.** Every individual prediction starts from the overall event rate in the population and is nudged up or down from there by the patient's predictors. If that starting point is imprecise, every prediction inherits the imprecision. The criterion asks for the overall risk to be pinned down to within about **±5 percentage points**.

(For a *continuous* outcome the framework adds a fourth criterion, about how much variation the model leaves unexplained, which is why you may see it described as having four parts.)

::: {.callout-note}
## What is a Cox-Snell R-squared, and how do I come up with one?

Criteria 1 and 2 need an advance guess at how well the model will perform, expressed as a **Cox-Snell R-squared** ($R^2_{CS}$). It is not an effect measure for a single predictor, so it is a different kind of quantity from a hazard or risk ratio: an HR says how much *one* predictor moves the outcome, while $R^2_{CS}$ summarises how much of the outcome the *whole model* accounts for.

In ordinary linear regression — where the outcome is a measurement like blood pressure — R-squared has a comfortable meaning. Patients differ from one another, and R-squared is the share of that variation the model accounts for. A yes/no outcome does not work like that: each patient either had the event or did not, so there is no range of values to account for and the familiar definition has nothing to latch onto.

The replacement is built around a different question: **how much better does the model predict what actually happened than a model that knows nothing except the overall event rate?** A good model should assign a high probability to what really did occur — high risk to the patients who went on to have the event, low risk to those who did not. $R^2_{CS}$ scores how well it manages that. Zero means it does no better than simply quoting the overall rate to everybody, and higher is better.

One awkward property matters in practice: **the maximum is not 1.** With a yes/no outcome the scale gets squashed, and how far depends on how common the outcome is. At 15% prevalence, even a hypothetical model that called every patient's outcome correctly and with total certainty would only score about 0.57. So read 0.10 against a ceiling of 0.57, not against 1 — which makes it a respectable model rather than a hopeless one. The **Nagelkerke R-squared** simply divides by that ceiling so the number does run from 0 to 1 ($0.10 / 0.57 \approx 0.18$). The sample-size formulas use Cox-Snell; Nagelkerke is the friendlier one to report.

You rarely need to guess $R^2_{CS}$ yourself. Take a **C-statistic** from a published model in a similar population (almost always reported; 0.70–0.80 is typical) and let `pmsampsize` convert it via the `cstatistic` argument, as below. Be conservative — assuming a better model than you will get makes the required sample size look smaller than it is.
:::

`pmsampsize` is an R package that does these calculations for you; the name is short for "prediction model sample size". It reports the largest sample size demanded by the three criteria.

::: {.panel-tabset}

#### R

```{r}
#| label: sample-size-r
#| eval: false

library(pmsampsize)

# Sample size for a logistic regression prediction model
# 10 candidate predictors, anticipated outcome prevalence of 15%,
# and an expected C-statistic of 0.75 taken from a published model.
# pmsampsize converts the C-statistic into a Cox-Snell R-squared for us.
ss <- pmsampsize(
  type = "b", # binary outcome
  cstatistic = 0.75, # anticipated C-statistic
  parameters = 10, # number of candidate predictors
  prevalence = 0.15
) # anticipated prevalence

print(ss) # one row per criterion, plus the final (largest) requirement
cat("\nMinimum sample size:", ss$sample_size, "\n")
cat("Minimum number of events:", ceiling(ss$events), "\n")

# If you already have a Cox-Snell R-squared in mind, supply it directly
# with csrsquared = (or nagrsquared = for a Nagelkerke R-squared).
# Note: exactly one of cstatistic, csrsquared or nagrsquared must be given;
# the older `rsquared =` argument is no longer accepted on its own.
pmsampsize(type = "b", csrsquared = 0.10, parameters = 10, prevalence = 0.15)
```

#### Python

```{python}
#| label: sample-size-py
#| eval: false

# Riley et al. sample size criteria for a binary outcome.
# There is no Python equivalent of pmsampsize, so the three criteria
# are implemented directly here. Unlike the R package, this version
# does not convert a C-statistic into a Cox-Snell R-squared for you.
import numpy as np


def pmsampsize_binary(cs_rsquared, parameters, prevalence,
                      shrinkage=0.9, delta=0.05, moe=0.05):
    """
    Minimum sample size for developing a binary-outcome prediction model.

    cs_rsquared : anticipated Cox-Snell R-squared
    parameters  : number of predictor parameters to be estimated
    prevalence  : anticipated outcome prevalence
    shrinkage   : target shrinkage factor for criterion 1 (default 0.9)
    delta       : acceptable apparent-vs-adjusted R-squared gap (criterion 2)
    moe         : margin of error for the overall risk (criterion 3)
    """
    phi = prevalence

    # The maximum attainable Cox-Snell R-squared, which depends only on
    # prevalence. Nagelkerke R-squared rescales Cox-Snell by this ceiling.
    max_r2 = 1 - np.exp(2 * (phi * np.log(phi) + (1 - phi) * np.log(1 - phi)))

    # Criterion 1: expected shrinkage factor of at least `shrinkage`
    n1 = parameters / ((shrinkage - 1) * np.log(1 - cs_rsquared / shrinkage))

    # Criterion 2: apparent and adjusted Nagelkerke R-squared differ by <= delta
    s2 = cs_rsquared / (cs_rsquared + delta * max_r2)
    n2 = parameters / ((s2 - 1) * np.log(1 - cs_rsquared / s2))

    # Criterion 3: overall risk estimated to within +/- moe
    n3 = (1.96 / moe) ** 2 * phi * (1 - phi)

    n_min = int(np.ceil(max(n1, n2, n3)))
    events = int(np.ceil(n_min * phi))

    print(f"Max possible Cox-Snell R-squared at {phi:.0%} prevalence: {max_r2:.3f}")
    print(f"Equivalent Nagelkerke R-squared: {cs_rsquared / max_r2:.3f}\n")
    print(f"Criterion 1 (shrinkage >= {shrinkage}):   n = {int(np.ceil(n1))}")
    print(f"Criterion 2 (R-squared gap <= {delta}): n = {int(np.ceil(n2))}")
    print(f"Criterion 3 (risk +/- {moe}):        n = {int(np.ceil(n3))}")
    print(f"\nMinimum sample size: {n_min}")
    print(f"Minimum events: {events}")
    print(f"Events per parameter: {events / parameters:.2f}")

    return n_min


# A C-statistic of 0.75 at 15% prevalence corresponds to a
# Cox-Snell R-squared of roughly 0.10
pmsampsize_binary(cs_rsquared=0.10, parameters=10, prevalence=0.15)
```

:::

**What the code shows.** This runs a sample-size calculation *before* any data are collected. You supply the expected outcome prevalence (15%), the number of candidate predictors (10), and an anticipated performance (a C-statistic of 0.75, which `pmsampsize` converts to a Cox-Snell R-squared of 0.1028). The output gives one row per criterion and a final row that is simply the largest of them: criterion 1 asks for 825 patients, criterion 2 for 327 and criterion 3 for 196, so **825 patients — about 124 events, or 12.4 events per parameter — is the requirement**. Criterion 1 is the one that decides the answer here, as it usually is, and that answer sits above the 100 events the old 10-EPV rule would have accepted. The Python implementation uses a rounded Cox-Snell R-squared of 0.10 rather than 0.1028 and so returns 850; the small difference is just the input, not a disagreement between the two.

The point of all this is feasibility. An underpowered study produces a model that looks impressive in development and falls apart in validation, and it is far cheaper to discover that now than after recruitment. If the required sample size is well beyond what you can realistically recruit, the levers are to reduce the number of candidate predictors, find a larger or additional data source, or reconsider the study — not to proceed and hope.

::: {.callout-note}
## What is EPV?
**Events per variable (EPV)** is the number of outcome events divided by the number of parameters you are estimating. If you have 80 events and 8 predictors, you have an EPV of 10. Low EPV is the single most common cause of overfitting in clinical models: with too few events, the model "memorises" the development sample rather than learning generalisable patterns.
:::

## Step 3: Handle Missing Data

Missing data is the norm in clinical research, not the exception. Laboratory values are missing because they were not ordered. Follow-up data is missing because patients moved. Lifestyle variables are missing because patients did not answer the questionnaire.

**Why a clinician should care.** How you deal with missing values can change which predictors look important and how trustworthy the model's risk estimates are. The wrong choice (usually the default one) can quietly bias the whole model.

### Missing Data Mechanisms

Understanding **why** data are missing is essential for choosing the right approach. The three standard patterns differ in one respect only: *what determines whether a value ends up missing?* It helps to walk through the same variable — a missing HbA1c measurement — under each.

- **MCAR (Missing Completely At Random)**: missingness has nothing to do with the patient at all — not with anything you recorded, and not with the missing value itself. Example: the laboratory analyser broke down one morning, so that day's samples were lost. Which patients have a missing HbA1c is effectively a coin flip. This is the benign case, and it is rare in practice.
- **MAR (Missing At Random)**: missingness depends on things you **did** record, but not on the missing value itself. Example: HbA1c is mostly measured in patients already known to have diabetes, so missingness depends on recorded diabetes status. Among patients with the *same* recorded diabetes status, who has a missing HbA1c is again essentially a coin flip. The name is unhelpful: "at random" here means random *once you condition on the observed data*, not haphazard.
- **MNAR (Missing Not At Random)**: missingness depends on the **unobserved** value itself, and no observed variable explains it away. Example: the patients too unwell to attend their appointment are also the patients with the worst glycaemic control — so it is precisely the highest HbA1c values that are missing, and nothing in your dataset tells you that.

The distinction matters because it decides what you can fix. Under MCAR and MAR, the observed data contain enough information to fill in the gaps sensibly, which is exactly what multiple imputation exploits. Under MNAR the information simply is not there: no imputation method can recover it from the data alone, and the honest response is a sensitivity analysis that asks how different the conclusions would be if the missing values were systematically worse (or better) than the observed ones. In practice you cannot prove which mechanism you are facing — MCAR versus MAR is partly testable, but MAR versus MNAR is not — so the mechanism is an assumption you argue for from clinical knowledge of how the data were collected, and then state explicitly.

### Why Complete Case Analysis Fails

Deleting rows with any missing data (complete case analysis) is the default in most software, and it is almost always wrong:

- It **wastes data**: if 10 variables each have 5% missing (independently), only 60% of rows are complete.
- It **biases estimates**: if missingness is related to the outcome or predictors (MAR or MNAR), the complete cases are a non-random subset.
- It **reduces power**: smaller effective sample sizes lead to wider confidence intervals and less stable models.

### Multiple Imputation

**Multiple imputation (MI)** is the recommended approach for handling missing data under the MAR assumption. "Imputation" simply means filling in a missing value with an informed guess based on the patient's other, observed values. "Multiple" means doing this several times so the uncertainty in those guesses is not hidden. The procedure has three steps:

1. **Impute**: create $m$ complete datasets, each one filling in the missing values with plausible guesses informed by the observed data. The guesses differ slightly between datasets, reflecting our genuine uncertainty.
2. **Analyse**: fit the prediction model separately in each of the $m$ filled-in datasets.
3. **Pool**: combine the $m$ sets of results using Rubin's rules, a standard recipe that widens the confidence intervals to account for the fact that the missing values were guessed, not measured.

Typically, $m = 20$ or more imputations are recommended.

```{mermaid}
%%| echo: false
%%| eval: true
%%| label: fig-multiple-imputation
%%| fig-cap: "The three steps of multiple imputation: fill in the gaps several times, analyse each completed dataset, then pool the results so the uncertainty from guessing is reflected in the final confidence intervals. The impute--analyse--pool procedure follows Rubin's rules [@wijesuriya2025mi]."
flowchart LR
    A[Dataset with<br/>missing values] --> B[Impute:<br/>m filled-in datasets]
    B --> C[Analyse:<br/>fit model in each]
    C --> D[Pool:<br/>combine with Rubin's rules]
```

::: {.panel-tabset}

#### R

```{r}
#| label: multiple-imputation-r
#| eval: false

library(mice)
library(dplyr)

# Simulate clinical data with missing values
set.seed(42)
n <- 500
data <- data.frame(
  age = rnorm(n, 55, 10),
  sbp = rnorm(n, 130, 20),
  cholesterol = rnorm(n, 220, 40),
  smoking = rbinom(n, 1, 0.25),
  bmi = rnorm(n, 27, 5)
)
data$cvd <- rbinom(
  n,
  1,
  plogis(
    -4 +
      0.03 * data$age +
      0.01 * data$sbp +
      0.005 * data$cholesterol +
      0.5 * data$smoking
  )
)

# Introduce MAR missingness
# Cholesterol more likely missing in younger patients
data$cholesterol[data$age < 50 & runif(n) < 0.3] <- NA
# BMI more likely missing in non-smokers
data$bmi[data$smoking == 0 & runif(n) < 0.2] <- NA

cat("Missing data pattern:\n")
md.pattern(data, rotate.names = TRUE)

# Perform multiple imputation
imp <- mice(data, m = 20, method = "pmm", seed = 42, printFlag = FALSE)

# Fit model in each imputed dataset and pool
fit_imp <- with(
  imp,
  glm(cvd ~ age + sbp + cholesterol + smoking + bmi, family = binomial)
)
pooled <- pool(fit_imp)
summary(pooled)
```

#### Python

```{python}
#| label: multiple-imputation-py
#| eval: false

import numpy as np
import pandas as pd
from sklearn.experimental import enable_iterative_imputer
from sklearn.impute import IterativeImputer
from sklearn.linear_model import LogisticRegression
from scipy.special import expit

np.random.seed(42)
n = 500

data = pd.DataFrame({
    'age': np.random.normal(55, 10, n),
    'sbp': np.random.normal(130, 20, n),
    'cholesterol': np.random.normal(220, 40, n),
    'smoking': np.random.binomial(1, 0.25, n),
    'bmi': np.random.normal(27, 5, n)
})
lp = -4 + 0.03*data['age'] + 0.01*data['sbp'] + 0.005*data['cholesterol'] + 0.5*data['smoking']
data['cvd'] = np.random.binomial(1, expit(lp))

# Introduce MAR missingness
mask_chol = (data['age'] < 50) & (np.random.uniform(size=n) < 0.3)
data.loc[mask_chol, 'cholesterol'] = np.nan
mask_bmi = (data['smoking'] == 0) & (np.random.uniform(size=n) < 0.2)
data.loc[mask_bmi, 'bmi'] = np.nan

print("Missing values per column:")
print(data.isnull().sum())

# Simple multiple imputation using IterativeImputer
# Note: for proper MI with pooling, consider the miceforest package
predictors = ['age', 'sbp', 'cholesterol', 'smoking', 'bmi']
X = data[predictors].values
y = data['cvd'].values

# Create multiple imputed datasets and pool coefficients
m = 20
coefs = []
for i in range(m):
    imputer = IterativeImputer(random_state=i, max_iter=10, sample_posterior=True)
    X_imp = imputer.fit_transform(X)
    model = LogisticRegression(max_iter=1000, penalty=None)
    model.fit(X_imp, y)
    coefs.append(np.concatenate([model.intercept_, model.coef_[0]]))

# Pool results (simplified Rubin's rules)
coefs = np.array(coefs)
pooled_mean = coefs.mean(axis=0)
within_var = coefs.var(axis=0)  # simplified
names = ['intercept'] + predictors
print("\nPooled coefficients:")
for name, coef in zip(names, pooled_mean):
    print(f"  {name}: {coef:.4f}")
```

:::

**What the code shows.** This demonstrates multiple imputation end-to-end on a simulated cohort where cholesterol and BMI are made deliberately missing in a way that depends on observed variables (MAR). In R, `md.pattern()` prints a table showing which combinations of variables are missing together — useful for spotting structure in the missingness. `mice()` then creates 20 filled-in datasets, the model is fitted in each, and `pool()` combines them with Rubin's rules so the reported standard errors honestly reflect the extra uncertainty from imputing. The pooled coefficients are what you would report. The key takeaway for a health researcher: you do not get a single "best guess" for each missing value — you get several plausible values, and the spread between them is folded into your confidence intervals. The Python version mimics this with `IterativeImputer`; note the comment that for fully correct pooling in Python you would reach for a dedicated MI package such as `miceforest`.

## Step 4: Variable Selection

**Variable selection** (also called predictor selection) means deciding which patient characteristics go into the model. A "candidate predictor" is any variable you are considering, such as age, blood pressure, or smoking status.

### Choosing Candidate Predictors

The best predictor selection strategy is **expert knowledge**. Decades of cardiovascular research tell us that age, sex, blood pressure, cholesterol, smoking, and diabetes are important predictors of CVD. Starting with established clinical knowledge prevents the discovery of spurious associations (chance patterns that will not hold up in new patients) and produces more generalisable models.

When expert knowledge is insufficient or the domain is novel, data-driven selection can supplement clinical reasoning. However, the approach matters enormously.

### The Dangers of Stepwise Selection

**Stepwise selection** is an automated procedure that adds or removes predictors one at a time based on statistical tests (forward = start empty and add; backward = start full and remove; both = a mix). Despite being among the most commonly used strategies, it is also one of the most harmful:

- It inflates the apparent significance of selected variables (multiple testing without correction).
- It produces unstable models: small changes in the data can lead to entirely different variable selections.
- It biases regression coefficients away from zero (selected variables appear stronger than they are).
- It yields overly optimistic performance estimates.

::: {.callout-warning}
## Avoid Univariable Screening
A particularly damaging practice is pre-selecting variables based on univariable p-values (e.g., keeping only variables with p < 0.20). This ignores confounding, suppression, and the fact that weak individual predictors can be strong in combination. If you must reduce the candidate set, use clinical knowledge, not univariable screening.
:::

### Better Approaches

- **Full model with all pre-specified predictors**: "pre-specified" means you decide the predictor list in advance, before looking at the outcome. If sample size permits, include all clinically plausible predictors and apply shrinkage (see below) to keep the model from over-reacting to noise. This is the recommended approach by Steyerberg (2019).
- **Penalised regression**: a family of methods that deliberately hold back the model's effect estimates ("penalise" large coefficients) so it does not overfit. LASSO (L1 penalty) performs automatic variable selection by shrinking some coefficients all the way to zero, effectively dropping those predictors. Elastic net combines L1 and L2 penalties. These are covered in detail in @sec-penalised-regression.
- **Background knowledge with data-driven fine-tuning**: start with a core set of established predictors and consider a small number of additional candidates.

## Step 5: Understand and Combat Overfitting

### What Is Overfitting?

Overfitting occurs when a model captures not only the true underlying patterns in the data but also the **noise** --- the random variation specific to the development sample. An overfitted model performs well on the data it was trained on but poorly on new data. Think of a student who memorises the answers to last year's exam: they ace the practice paper but stumble on this year's questions.

**Why a clinician should care.** An overfitted risk score will overstate how confident it is, telling you a patient is at 2% or 90% risk when the truth is closer to the average. Acting on those exaggerated numbers can lead to over- or under-treatment.

Clinical data is especially prone to overfitting because:

- **Sample sizes are often small** relative to the number of candidate predictors.
- **Events may be rare**, limiting the information available to estimate parameters.
- **Complex interactions** are tempting to include but expensive. (An "interaction term" lets one predictor's effect depend on another — for example, smoking mattering more in older patients. Each one adds a parameter the model must estimate, using up scarce "degrees of freedom", the budget of information your events can support.)

### Detecting Overfitting: Optimism

The **optimism** of a model is the difference between its apparent performance (how good it looks on the data it was built from) and its expected performance on new data. Every model flatters itself a little on its own data, so the apparent numbers are always a bit too rosy. You can estimate the size of that flattery using internal validation techniques such as bootstrapping — repeatedly re-fitting the model on random resamples of the data (covered in @sec-performance-validation).

If a model shows high optimism, it is overfitted. The solution is **shrinkage**.

### Shrinkage Methods

Shrinkage means deliberately pulling the model's effect estimates back toward zero so its predictions are less extreme and generalise better.

**Uniform shrinkage** multiplies all regression coefficients by a single factor $s$ (a number between 0 and 1) estimated from the data, typically via bootstrapping. Shrinking every coefficient would also drag the overall level of predicted risk downward, so one number — the model's baseline term, or *intercept* — is re-estimated afterwards to put the average predicted risk back where it belongs. That keeps the model's average honest, which is one part of what "calibration" means; the full picture is in @sec-calibration. The effect of the whole operation is to pull extreme predictions back toward the average, reducing the impact of noise.

**Penalised methods** build shrinkage directly into the model-fitting step rather than applying it afterwards:

- **Ridge regression** (L2 penalty) shrinks all coefficients toward zero but never sets any to exactly zero — every predictor stays in the model, just toned down.
- **LASSO** (L1 penalty) can shrink coefficients to exactly zero, so it also performs variable selection by dropping weak predictors.
- **Elastic net** combines both penalties.

How hard to shrink (the penalty strength) is chosen by cross-validation — repeatedly setting aside part of the data, trying different penalty strengths, and keeping the one that predicts the held-out part best.

::: {.panel-tabset}

#### R

```{r}
#| label: shrinkage-r
#| eval: false

library(glmnet)

# Simulate Framingham-like data. We deliberately make this a setting where
# overfitting is a real threat: 500 patients, and 20 pure-noise variables
# mixed in among the 6 genuine predictors.
set.seed(42)
n <- 500
age <- rnorm(n, 55, 10)
male <- rbinom(n, 1, 0.5)
sbp <- rnorm(n, 130, 20)
chol <- rnorm(n, 220, 40)
smoking <- rbinom(n, 1, 0.25)
diabetes <- rbinom(n, 1, 0.10)

# True model: only these six variables actually affect risk
lp <- -6 +
  0.05 * age +
  0.3 * male +
  0.01 * sbp +
  0.005 * chol +
  0.4 * smoking +
  0.5 * diabetes
cvd <- rbinom(n, 1, plogis(lp))

# 20 candidate predictors that are unrelated to the outcome
noise <- matrix(rnorm(n * 20), n, 20)
colnames(noise) <- paste0("noise", 1:20)

df <- data.frame(age, male, sbp, chol, smoking, diabetes, noise, cvd)
X <- model.matrix(cvd ~ ., data = df)[, -1]
y <- df$cvd

# Fit the same model three ways
fit_full <- glm(cvd ~ ., data = df, family = binomial)
fit_ridge <- cv.glmnet(X, y, family = "binomial", alpha = 0, nfolds = 10)
fit_lasso <- cv.glmnet(X, y, family = "binomial", alpha = 1, nfolds = 10)

# Compare the six real predictors against their known true values
real <- c("age", "male", "sbp", "chol", "smoking", "diabetes")
pick <- function(fit) {
  as.vector(coef(fit, s = "lambda.min"))[match(
    real,
    c("(Intercept)", colnames(X))
  )]
}
comparison <- data.frame(
  truth = c(0.05, 0.3, 0.01, 0.005, 0.4, 0.5),
  unpenalised = coef(fit_full)[real],
  ridge = pick(fit_ridge),
  lasso = pick(fit_lasso)
)
print(round(comparison, 4))

# How many of the 20 noise variables did each method keep?
kept_noise <- function(fit) {
  b <- as.vector(coef(fit, s = "lambda.min"))[-1]
  sum(b[grepl("noise", colnames(X))] != 0)
}
cat(
  "\nNoise variables kept (of 20) -- ridge:",
  kept_noise(fit_ridge),
  "| LASSO:",
  kept_noise(fit_lasso),
  "\n"
)

# Cross-validation curves
par(mfrow = c(1, 2), mar = c(4.5, 4.4, 4.6, 1.2))
for (m in list(
  list(f = fit_ridge, t = "Ridge"),
  list(f = fit_lasso, t = "LASSO")
)) {
  plot(m$f, ylab = "Prediction error (deviance)")
  title(main = paste(m$t, "- cross-validated error"), line = 3.2)
  mtext(
    "number of predictors kept",
    side = 3,
    line = 2.0,
    cex = 0.7,
    col = "grey30"
  )
  abline(v = -log(m$f$lambda.min), col = "#2166ac", lwd = 2)
  legend(
    "top",
    bty = "n",
    cex = 0.7,
    lwd = c(2, 1),
    lty = c(1, 3),
    col = c("#2166ac", "black"),
    seg.len = 1.6,
    legend = c("lambda.min (lowest error)", "lambda.1se (simpler, within 1 SE)")
  )
}
```

#### Python

```{python}
#| label: shrinkage-py
#| eval: false

import numpy as np
import pandas as pd
from sklearn.linear_model import LogisticRegression, LogisticRegressionCV
from sklearn.preprocessing import StandardScaler
from scipy.special import expit

# Same setup as the R example: 500 patients, 6 real predictors, 20 noise ones.
# The random draws differ from R's, so the numbers will not match exactly.
np.random.seed(42)
n = 500
age = np.random.normal(55, 10, n)
male = np.random.binomial(1, 0.5, n)
sbp = np.random.normal(130, 20, n)
chol = np.random.normal(220, 40, n)
smoking = np.random.binomial(1, 0.25, n)
diabetes = np.random.binomial(1, 0.10, n)

lp = -6 + 0.05*age + 0.3*male + 0.01*sbp + 0.005*chol + 0.4*smoking + 0.5*diabetes
cvd = np.random.binomial(1, expit(lp))

noise = np.random.normal(size=(n, 20))
real_names = ['age', 'male', 'sbp', 'chol', 'smoking', 'diabetes']
names = real_names + [f'noise{i}' for i in range(1, 21)]
X = np.column_stack([age, male, sbp, chol, smoking, diabetes, noise])

# Penalties act on the coefficient scale, so predictors must be standardised
# first (glmnet does this internally; scikit-learn does not). We rescale the
# coefficients afterwards so they are comparable to the true values.
scaler = StandardScaler().fit(X)
Xs = scaler.transform(X)

# Two settings matter here and are easy to get wrong:
#   scoring='neg_log_loss' -- the default is accuracy, which a "predict the
#     majority class" model maximises, so it would choose maximum shrinkage
#   Cs -- the default grid of 10 values is too coarse to locate a good penalty
grid = np.logspace(-3, 3, 60)

model_full = LogisticRegression(penalty=None, max_iter=5000).fit(X, cvd)
model_ridge = LogisticRegressionCV(Cs=grid, penalty='l2', cv=10,
                                   scoring='neg_log_loss', max_iter=5000,
                                   random_state=42).fit(Xs, cvd)
model_lasso = LogisticRegressionCV(Cs=grid, penalty='l1', solver='saga', cv=10,
                                   scoring='neg_log_loss', max_iter=20000,
                                   random_state=42).fit(Xs, cvd)

def original_scale(model):
    return model.coef_[0] / scaler.scale_

comparison = pd.DataFrame({
    'truth': [0.05, 0.3, 0.01, 0.005, 0.4, 0.5],
    'unpenalised': model_full.coef_[0][:6],
    'ridge': original_scale(model_ridge)[:6],
    'lasso': original_scale(model_lasso)[:6],
}, index=real_names)
print(comparison.round(4))

for label, model in [('ridge', model_ridge), ('lasso', model_lasso)]:
    kept = np.sum(model.coef_[0][6:] != 0)
    print(f"Noise variables kept (of 20) -- {label}: {kept}")
```

:::

**What the code shows.** Because these are simulated data we know the true coefficients, so the printed table can be read as a scorecard. The unpenalised column overshoots: diabetes comes out at about 0.88 when the truth is 0.50. That exaggeration is overfitting, and it is exactly what shrinkage exists to correct — the ridge and LASSO columns sit visibly closer to the truth. The noise count then shows the difference between the two penalties in one number: ridge keeps all 20 worthless variables (shrunk, but never removed), while LASSO sets most of them to exactly zero and drops them outright.

**How to read the cross-validation plots.** These two panels are the workhorse figure of penalised regression, and they repay a slow first read.

- **Each red dot is one fitted model.** The data are split into 10 parts; the model is fitted on 9 and its prediction error measured on the held-out one, rotating through all 10. The dot is the average of those 10 errors, so it estimates how the model would do on *patients it has not seen*.
- **The y-axis is prediction error.** The label says "binomial deviance", which is just this model family's way of scoring how far the predicted probabilities were from what actually happened. Lower is better. Do not try to interpret the number itself — 1.29 means nothing in isolation. Only the *shape* of the curve matters.
- **The x-axis is the amount of penalty**, and its direction is the part that trips people up. `glmnet` (version 5 onwards) plots $-\log(\lambda)$, the *negative* log penalty, so **left means heavy penalty and strong shrinkage, right means almost no penalty** — the far right is essentially the ordinary unpenalised model. Older textbook figures plot $\log(\lambda)$ instead, which runs the opposite way, so check the axis label before interpreting any such plot you meet.
- **The grey bars show how much that error estimate wobbles** (±1 standard error across the 10 splits) depending on which patients happened to be held out. They are a humility check: where the bars are wide, the exact position of the curve's lowest point is not to be taken too literally, because a different split of the same patients would have moved it.
- **The numbers along the top are how many predictors still have a non-zero coefficient.** For ridge this is stuck at 26 the whole way across, because ridge never eliminates anything. For LASSO it climbs from 0 on the left to 26 on the right. That contrast *is* the L1-versus-L2 difference, made visible.

The curve is U-shaped, and both arms are informative. Move right from the bottom and error rises because the model is too free and starts fitting noise — overfitting. Move left and error rises again because everything has been squashed so hard toward zero that genuine signal is lost — over-shrinking. The bottom of the U is the best available trade-off.

**The two vertical lines** mark the two conventional choices, and this is where the axis direction matters most:

- `lambda.min` (solid blue) is simply the penalty with the lowest average error — the bottom of the U.
- `lambda.1se` (dotted) is the *strongest* penalty whose error is still within one standard error of that minimum. It therefore sits to the **left** of `lambda.min`, because left means more penalty. The logic is that the minimum is itself estimated with noise, so deliberately erring toward a simpler, more heavily shrunk model is often the safer bet — and for LASSO it yields a shorter predictor list, which matters if the model has to be used at the bedside.

Either choice is defensible; what matters is deciding in advance which you will use, rather than fitting both and reporting whichever looks better.

::: {.callout-note}
## Shrinkage in plain terms
A **shrinkage factor** is a number between 0 and 1 that you multiply coefficients by to make predictions less extreme. An overfitted model produces predictions that are too confident — very high risks too high, very low risks too low. Shrinkage pulls those predictions back toward the average event rate, which is usually closer to the truth for new patients. A shrinkage factor of 0.85 means the development model was about 15% too optimistic and its effects should be scaled back accordingly.
:::

## Step 6: Build the Model --- A Framingham Example {#sec-framingham-example}

Let us now walk through a complete model development process using simulated data based on the Framingham Heart Study. We will predict 10-year CVD risk using established risk factors. This is the "derivation" (or development) cohort — the dataset in which the model is built. Later, the model should be tested in a separate "validation" cohort, ideally patients from a different time or place, to check it still works.

### Data Preparation

::: {.panel-tabset}

#### R

```{r}
#| label: framingham-build-r
#| eval: false

library(rms)
library(ggplot2)

# Simulate Framingham-like cohort
set.seed(2024)
n <- 2000

framingham <- data.frame(
  age = round(runif(n, 30, 74)),
  male = rbinom(n, 1, 0.48),
  sbp = round(rnorm(n, 130, 18)),
  total_chol = round(rnorm(n, 210, 38)),
  hdl_chol = round(rnorm(n, 52, 15)),
  smoking = rbinom(n, 1, 0.22),
  diabetes = rbinom(n, 1, 0.08),
  bp_treatment = rbinom(n, 1, 0.15)
)

# Generate outcome based on known risk factors
lp <- with(
  framingham,
  -7.5 +
    0.06 * age +
    0.4 * male +
    0.012 * sbp +
    0.005 * total_chol -
    0.02 * hdl_chol +
    0.5 * smoking +
    0.7 * diabetes +
    0.3 * bp_treatment
)
framingham$cvd_10yr <- rbinom(n, 1, plogis(lp))

cat("Event rate:", round(mean(framingham$cvd_10yr) * 100, 1), "%\n")
cat("Number of events:", sum(framingham$cvd_10yr), "\n")
cat("Events per variable:", round(sum(framingham$cvd_10yr) / 8, 1), "\n")

# Set up the rms environment for enhanced model building
dd <- datadist(framingham)
options(datadist = "dd")

# Fit the model using lrm (logistic regression model from rms)
# Allow nonlinearity for continuous variables using restricted cubic splines
fit <- lrm(
  cvd_10yr ~ rcs(age, 4) +
    male +
    rcs(sbp, 3) +
    total_chol +
    hdl_chol +
    smoking +
    diabetes +
    bp_treatment,
  data = framingham,
  x = TRUE,
  y = TRUE
)

print(fit)

# Check for nonlinearity
anova(fit)

# Visualise partial effects for age and SBP side by side.
# fun = plogis converts the model's log odds onto the risk (probability)
# scale, which is what a clinician actually wants to see.
partial <- function(var, label) {
  p <- as.data.frame(Predict(fit, name = var, fun = plogis))
  data.frame(
    x = p[[var]],
    risk = p$yhat,
    lo = p$lower,
    hi = p$upper,
    panel = label,
    row.names = NULL
  )
}
age_lab <- "Age (years)"
sbp_lab <- "Systolic blood pressure (mmHg)"
curves <- rbind(partial("age", age_lab), partial("sbp", sbp_lab))

# Where the patients actually are, drawn as a rug along the bottom
rugs <- rbind(
  data.frame(x = framingham$age, panel = age_lab),
  data.frame(x = framingham$sbp, panel = sbp_lab)
)

ggplot(curves, aes(x, risk)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey75", alpha = 0.6) +
  geom_line(colour = "#2166ac", linewidth = 0.9) +
  geom_rug(
    data = rugs,
    aes(x = x),
    inherit.aes = FALSE,
    sides = "b",
    alpha = 0.05
  ) +
  facet_wrap(~panel, scales = "free_x", strip.position = "bottom") +
  scale_y_continuous(labels = function(p) paste0(round(p * 100), "%")) +
  labs(
    title = "Predicted 10-year CVD risk across the range of each predictor",
    subtitle = paste(
      "Other predictors held fixed at: female, non-smoker,",
      "no diabetes, untreated,\ntotal cholesterol 210,",
      "HDL 52. Shaded band = 95% confidence interval."
    ),
    x = NULL,
    y = "Predicted 10-year CVD risk"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.placement = "outside",
    panel.grid.minor = element_blank(),
    plot.subtitle = element_text(size = 8.5, colour = "grey35")
  )
```

#### Python

```{python}
#| label: framingham-build-py
#| eval: false

import numpy as np
import pandas as pd
import statsmodels.api as sm
from scipy.special import expit

np.random.seed(2024)
n = 2000

framingham = pd.DataFrame({
    'age': np.random.randint(30, 75, n),
    'male': np.random.binomial(1, 0.48, n),
    'sbp': np.round(np.random.normal(130, 18, n)),
    'total_chol': np.round(np.random.normal(210, 38, n)),
    'hdl_chol': np.round(np.random.normal(52, 15, n)),
    'smoking': np.random.binomial(1, 0.22, n),
    'diabetes': np.random.binomial(1, 0.08, n),
    'bp_treatment': np.random.binomial(1, 0.15, n)
})

lp = (-7.5 + 0.06 * framingham['age'] + 0.4 * framingham['male'] +
      0.012 * framingham['sbp'] + 0.005 * framingham['total_chol'] -
      0.02 * framingham['hdl_chol'] + 0.5 * framingham['smoking'] +
      0.7 * framingham['diabetes'] + 0.3 * framingham['bp_treatment'])

framingham['cvd_10yr'] = np.random.binomial(1, expit(lp))

print(f"Event rate: {framingham['cvd_10yr'].mean()*100:.1f}%")
print(f"Number of events: {framingham['cvd_10yr'].sum()}")
print(f"Events per variable: {framingham['cvd_10yr'].sum() / 8:.1f}")

# Fit logistic regression
predictors = ['age', 'male', 'sbp', 'total_chol', 'hdl_chol',
              'smoking', 'diabetes', 'bp_treatment']
X = sm.add_constant(framingham[predictors])
y = framingham['cvd_10yr']

model = sm.Logit(y, X)
result = model.fit(disp=0)
print("\n", result.summary())

# Calculate predicted probabilities
framingham['pred_prob'] = result.predict(X)

# Summary of predicted risk distribution
print("\nPredicted risk distribution:")
print(framingham['pred_prob'].describe())
```

:::

**What the code shows.** This builds the actual prediction model on a simulated Framingham-style cohort. The first printed lines report the event rate, the number of events, and the EPV — a sanity check that we have enough events for the number of predictors before trusting anything else. In R, `lrm()` fits the logistic model and `print(fit)` reports the coefficients plus the apparent C-statistic; `rcs(age, 4)` lets age have a flexible (non-straight-line) relationship with risk using restricted cubic splines with 4 knots. The `anova(fit)` output tests whether that flexibility is actually needed (a significant nonlinear term means a straight line would misfit). The Python version fits the equivalent model with `statsmodels` and summarises the distribution of predicted risks, so you can see whether the model produces a useful spread of probabilities or bunches everyone near the average.

**How to read the partial-effect figure.** Each panel answers one question: *if I walked a single patient across the whole range of this one predictor and changed nothing else about them, how would their predicted risk move?* Three features are worth naming explicitly.

- **"Partial" means everything else is pinned.** Each curve is drawn for one reference patient — here a female non-smoker without diabetes, untreated, with typical cholesterol — and only the panel's own predictor varies. Change the reference patient and the curves shift up or down, though their *shape* stays the same. This is why such plots always carry an "adjusted to" note; it is not fine print, it tells you which patient you are looking at.
- **The y-axis is risk, not log odds.** Logistic regression reports its results on an internal scale (log odds) that means little at the bedside; `fun = plogis` converts them into plain probabilities, so the numbers are ones you could say out loud to a patient. Both panels share one y-axis, which makes the comparison the real point of the figure: across its observed range, age moves predicted risk from about 2% to about 22%, whereas systolic blood pressure moves it only from about 2% to 7%. Age is doing far more work than blood pressure in this model — a conclusion you cannot draw by glancing at coefficients, because coefficients depend on each predictor's units.
- **The shaded band is a 95% confidence interval, and the tick marks along the bottom show where the patients actually are.** Read them together. The band flares at the far left and right of each panel precisely because there are few patients out there, so the curve is extrapolating on thin evidence. Treat the ends of these curves with suspicion; that is where a smooth-looking line is least trustworthy.

One trap is worth pointing out, because this figure walks straight into it. The age curve looks bent — flat to about 45, then steepening — which invites the conclusion that age has a genuinely non-straight-line effect and that the splines have earned their keep. They have not. We simulated these data ourselves, and we built age in as a perfectly straight line; the `anova(fit)` output agrees, reporting the nonlinear term for age as nowhere near significant (p ≈ 0.45).

So where does the bend come from? Logistic regression works internally on a scale that has no ceiling, then squeezes the result into the 0–100% range that a probability has to live in. That squeezing is not uniform: near 0% there is very little room left, so a straight-line increase gets compressed, while higher up it has room to climb. Redraw *any* straight-line effect as a risk and you get a curve of exactly this shape. The lesson is practical: **a bent curve on a risk plot is not evidence of a non-straight-line effect** — the `anova` test is what tells you that. Keep both views, and use each for what it is good for: the risk scale for talking to patients, the `anova` for deciding whether the extra flexibility is real.

### Model Specification Decisions

In the Framingham example above, several important decisions were made:

1. **Pre-specified predictors**: we included the established Framingham risk factors, not variables discovered through data-dredging.
2. **Nonlinearity**: for continuous variables like age and blood pressure, we allowed the relationship with risk to bend rather than follow a straight line, using restricted cubic splines (in R's `rms` package). A spline is a smooth curve stitched together from simple pieces; it lets risk rise gently and then steeply with age, for instance, instead of forcing a single fixed slope. The relationship between age and CVD risk is genuinely curved, so a straight line would misfit.
3. **No interactions without prior evidence**: we did not fish for interactions (terms where one predictor's effect depends on another). If clinical knowledge suggests an interaction (e.g., the effect of cholesterol might differ by sex), it should be pre-specified rather than discovered by trial and error.
4. **Sample size check**: with approximately 250 events and 8 predictors (plus a few extra degrees of freedom for the splines), we have approximately 25 events per variable, which is comfortable.

## Step 7: Assess the Final Model

After fitting, we need to examine the model critically. Two questions matter most. Does it *discriminate* — rank patients who go on to have the event above those who do not? And is it *calibrated* — when it says 20% risk, do about 20% of such patients actually have the event? A model can discriminate well yet be poorly calibrated, so both must be checked.

::: {.callout-tip}
## This section is a first look; @sec-performance-validation is the full treatment

Performance assessment is a large enough topic to have the next chapter to itself, and this section deliberately does not duplicate it. What follows is the minimum you should look at before believing your own model, with just enough interpretation to read the output. If you want the definitions, the formulas and the reasoning behind any of these measures, go to @sec-performance-validation, which organises them into five domains [@vancalster2025performance] and covers:

- **Discrimination** and the C-statistic — @sec-discrimination and @sec-c-statistic, including what the number does *not* tell you (@sec-c-statistic-limits)
- **Calibration** — @sec-calibration, with the O:E ratio (@sec-calibration-large), the calibration slope (@sec-calibration-slope) and how to read a calibration plot (@sec-calibration-plots)
- **Overall performance**, including the Brier score and its no-predictor benchmark — @sec-overall-performance
- **Internal validation** and bootstrap optimism correction — @sec-internal-validation and @sec-bootstrap-optimism
- **Clinical utility** — net benefit and decision curve analysis, which ask the question none of the above do: would using this model actually improve decisions?
:::

::: {.panel-tabset}

#### R

```{r}
#| label: model-assessment-r
#| eval: false

library(rms)

# Assuming fit and framingham from above
pred_probs <- predict(fit, type = "fitted")
event_rate <- mean(framingham$cvd_10yr)
top <- ceiling(max(pred_probs) * 10) / 10 # zoom the axes to the data

cat("C-statistic (AUC):", round(fit$stats["C"], 3), "\n")

# Three complementary views of the same model, side by side
par(mfrow = c(1, 3), mar = c(4.6, 4.4, 3.6, 1.2))

# A. Does the model spread patients out, or call everyone average?
hist(
  pred_probs,
  breaks = 50,
  col = "steelblue",
  border = "white",
  main = "A. Spread of predicted risk",
  xlab = "Predicted 10-year risk",
  ylab = "Number of patients"
)
abline(v = event_rate, col = "firebrick", lwd = 2, lty = 2)
text(
  event_rate,
  par("usr")[4] * 0.92,
  paste0(" overall rate ", round(event_rate * 100), "%"),
  col = "firebrick",
  cex = 0.85,
  adj = 0
)

# B. Discrimination: do the patients who had events get higher risks?
boxplot(
  pred_probs ~ framingham$cvd_10yr,
  names = c("No CVD", "CVD"),
  main = "B. Do events get higher risks?",
  xlab = "Actual outcome",
  ylab = "Predicted 10-year risk",
  col = c("lightblue", "salmon"),
  outcex = 0.4
)

# C. Calibration: when the model says 20%, do 20% have the event?
# We draw this panel by hand rather than using val.prob()'s default plot,
# so the three lines are told apart at a glance and labelled in plain words.
y <- framingham$cvd_10yr
q95 <- quantile(pred_probs, 0.95)   # beyond here, few patients: thin evidence

# The straight-line fit: allowed only an intercept and a slope, so it can
# describe "risks uniformly too high/low" and "risks too spread out/bunched",
# and nothing more.
cal_fit <- glm(y ~ lp_hat, family = binomial,
               data = data.frame(y = y, lp_hat = qlogis(pred_probs)))
grid <- seq(min(pred_probs), max(pred_probs), length.out = 200)
straight <- plogis(predict(cal_fit, newdata = data.frame(lp_hat = qlogis(grid))))

# The flexible fit: follows the data locally, with no assumed shape at all.
flexible <- lowess(pred_probs, y, iter = 0)

plot(0, 0, type = "n", xlim = c(0, top), ylim = c(-0.06, top), yaxt = "n",
     xlab = "Predicted 10-year risk", ylab = "Observed frequency",
     main = "C. Calibration: predicted vs observed")
axis(2, at = seq(0, top, 0.2))

# Shade the region where there are hardly any patients
rect(q95, -0.06, top, top, col = rgb(0, 0, 0, 0.05), border = NA)
abline(v = q95, col = "grey45", lty = 3)
text(q95, top * 0.03, " 95% of patients\n to the left",
     cex = 0.6, col = "grey35", adj = 0)

abline(0, 1, col = "grey70", lwd = 7)                                # ideal
lines(grid, straight, col = "#2166ac", lwd = 2.4)                    # 2-number
lines(flexible$x, flexible$y, col = "#d6604d", lwd = 2.4, lty = 2)   # flexible

# Density strip: where the patients actually are
h <- hist(pred_probs, breaks = 60, plot = FALSE)
bw <- diff(h$mids)[1]
rect(h$mids - bw / 2, -0.055,
     h$mids + bw / 2, -0.055 + 0.05 * h$counts / max(h$counts),
     col = "grey55", border = NA)
mtext("where patients are", side = 1, line = -1.3,
      at = top * 0.62, cex = 0.6, col = "grey35")

legend("topleft", bty = "n", cex = 0.7, lwd = c(7, 2.4, 2.4),
       lty = c(1, 1, 2), seg.len = 1.8,
       col = c("grey70", "#2166ac", "#d6604d"),
       legend = c("Perfect calibration",
                  "Straight-line fit (intercept + slope only)",
                  "Flexible fit (no assumed shape)"))

# val.prob still gives us the statistics; pl = FALSE suppresses its own plot
val <- val.prob(pred_probs, y, pl = FALSE)

# val.prob returns 18 statistics. These are the six worth reading.
key <- c("C (ROC)", "Brier", "Intercept", "Slope", "Eavg", "Emax")
summary_table <- data.frame(
  value = round(val[key], 3),
  ideal = c("1.0", "0", "0", "1.0", "0", "0"),
  meaning = c(
    "discrimination: 0.5 = coin flip, 1.0 = perfect",
    "overall accuracy of the probabilities (lower is better)",
    "calibration-in-the-large: are risks too high or too low overall?",
    "calibration slope: are risks too extreme (<1) or too flat (>1)?",
    "average gap between predicted and observed risk",
    "worst gap between predicted and observed risk"
  )
)
print(summary_table, right = FALSE)

# Internal validation via bootstrapping (optimism-corrected)
set.seed(123)
val_boot <- validate(fit, B = 200)
print(val_boot)

cat("\nApparent C-statistic:          ", round(fit$stats["C"], 3), "\n")
cat(
  "Optimism-corrected C-statistic:",
  round(0.5 * (val_boot["Dxy", "index.corrected"] + 1), 3),
  "\n"
)
```

#### Python

```{python}
#| label: model-assessment-py
#| eval: false

import numpy as np
import matplotlib.pyplot as plt
import statsmodels.api as sm
from sklearn.metrics import roc_auc_score, brier_score_loss
from sklearn.calibration import calibration_curve

# Assuming framingham and result from above
pred_probs = framingham['pred_prob'].values
y_true = framingham['cvd_10yr'].values

# 1. Discrimination
auc = roc_auc_score(y_true, pred_probs)
print(f"C-statistic (AUC): {auc:.3f}")

print(f"Brier score:       {brier_score_loss(y_true, pred_probs):.3f}")

# Calibration intercept and slope, fitted on the linear predictor
logit_p = np.log(pred_probs / (1 - pred_probs))
cal = sm.Logit(y_true, sm.add_constant(logit_p)).fit(disp=0)
print(f"Calibration intercept: {cal.params[0]:.3f}  (ideal 0)")
print(f"Calibration slope:     {cal.params[1]:.3f}  (ideal 1)")

# Three complementary views of the same model, side by side
event_rate = y_true.mean()
top = np.ceil(pred_probs.max() * 10) / 10
fig, axes = plt.subplots(1, 3, figsize=(15, 4.6))

# A. Does the model spread patients out, or call everyone average?
axes[0].hist(pred_probs, bins=50, color="steelblue", edgecolor="white")
axes[0].axvline(event_rate, color="firebrick", ls="--", lw=2)
axes[0].annotate(f" overall rate {event_rate:.0%}", (event_rate, 0.92),
                 xycoords=("data", "axes fraction"), color="firebrick")
axes[0].set(xlabel="Predicted 10-year risk", ylabel="Number of patients",
            title="A. Spread of predicted risk")

# B. Discrimination: do the patients who had events get higher risks?
bp = axes[1].boxplot([pred_probs[y_true == 0], pred_probs[y_true == 1]],
                     tick_labels=["No CVD", "CVD"], patch_artist=True,
                     flierprops=dict(markersize=2))
for patch, colour in zip(bp['boxes'], ["lightblue", "salmon"]):
    patch.set_facecolor(colour)
axes[1].set(xlabel="Actual outcome", ylabel="Predicted 10-year risk",
            title="B. Do events get higher risks?")

# C. Calibration: when the model says 20%, do 20% have the event?
prob_true, prob_pred = calibration_curve(y_true, pred_probs, n_bins=10,
                                         strategy='quantile')
axes[2].plot([0, top], [0, top], '--', color='grey', label="Perfect calibration")
axes[2].plot(prob_pred, prob_true, 's-', color="steelblue", label="Model")
axes[2].set(xlim=(0, top), ylim=(0, top), xlabel="Predicted 10-year risk",
            ylabel="Observed frequency",
            title="C. Calibration: predicted vs observed")
axes[2].legend()

plt.tight_layout()
plt.show()
```

:::

**What the code shows.** This is a first critical look at the finished model, from three angles that answer different questions. Read them left to right.

**Panel A — does the model actually distinguish anyone?** This is the distribution of predicted risks across the 2,000 patients, with the dashed red line at the overall event rate (12%). The shape you want is a *wide* one. Predicted risks here run from under 1% to about 76%, with the middle half of patients between 4% and 16% — so the model is genuinely sorting people, not hedging. The failure mode this panel is designed to catch is a narrow spike sitting on top of the red line: a model that tells every patient "about 12%" is clinically worthless no matter how respectable its other statistics look, because no decision could ever change based on it.

**Panel B — discrimination, made concrete.** Median predicted risk is 7.5% among patients who did not have a CVD event and 18.5% among those who did. The events do get higher risks, which is what we want. But look at how much the two boxes still overlap: plenty of patients who had events were given lower risks than patients who did not. **That overlap is what a C-statistic of 0.75 looks like** — and 0.75 is a perfectly respectable clinical model. If you were expecting two cleanly separated boxes, recalibrate that expectation now; you will not see it in real risk prediction, and if you do, suspect a leaked outcome.

**Panel C — calibration.** Discrimination only asks whether the *ranking* is right. Calibration asks whether the numbers can be believed: when the model says 20%, do about 20 in 100 such patients actually have the event? Predicted risk runs along the bottom, the proportion who actually had the event up the side, and **the thick grey diagonal is the answer you want** — every predicted risk matching the observed frequency exactly.

**Why there are two lines, not one.** This is the part of the figure that confuses people, and the two lines are asking the same question with different amounts of freedom.

- **The blue straight-line fit is deliberately restricted.** It is allowed exactly two numbers — an intercept and a slope — so it can only describe two kinds of problem: predictions that are uniformly too high or too low (wrong intercept), and predictions that are too spread out or too bunched together (wrong slope). It *cannot* bend. Those two numbers are precisely the `Intercept` and `Slope` in the table below, so this line is the visual form of that two-number summary.
- **The red dashed flexible fit is allowed to go wherever the data lead.** It assumes no shape whatsoever — it simply follows the local pattern of who actually had events, which is why the technical name for it is *nonparametric* ("no assumed shape"). It can therefore reveal trouble confined to one stretch of the risk range: a model that is spot-on for low-risk patients and badly off for high-risk ones.

**Reading them together is the point.** Where the two lines agree, the tidy two-number summary is an honest description of the model's calibration. Where the flexible line wanders away from the straight one, the miscalibration has a *shape* that those two numbers are hiding — and you would never know from the table alone.

That is exactly what happens here, and it is also a lesson in not over-reading. Below the dotted vertical line — which holds 95% of the patients — the flexible fit tracks the diagonal within about 1 percentage point: the model is well calibrated where nearly all the patients are. To the right of it, in the shaded strip, the flexible fit peels upward and the gap grows to nearly 12 percentage points. But that whole region contains only about 100 of the 2,000 patients. The grey density strip along the bottom shows the same thing: patients pile up below 0.2 and thin out to almost nothing past 0.4. So the dramatic-looking divergence is built on very few people, and it is the same sparse tail that produces the alarming `Emax` of 0.117 in the table below. **Always check where the patients are before believing the ends of a calibration curve.**

::: {.callout-important}
## What the numbers mean

`val.prob()` prints 18 statistics. Six are worth your attention, and knowing which to ignore is half the skill.

| Statistic | Here | Ideal | What it means |
|---|---|---|---|
| **C (ROC)** | 0.750 | 1.0 | Pick one patient who had the event and one who did not, at random. This is the probability the model gave the *event* patient the higher risk. 0.5 is a coin flip. Full treatment in @sec-c-statistic. |
| **Brier** | 0.092 | 0 | Average squared distance between the predicted probability and what actually happened (0 or 1). Lower is better, but the scale depends on how common the outcome is — see below, and @sec-overall-performance. |
| **Intercept** | 0.000 | 0 | Calibration-in-the-large: are the predicted risks systematically too high or too low overall? Usually reported as the O:E ratio instead — @sec-calibration-large. |
| **Slope** | 1.000 | 1.0 | Calibration slope: below 1 means predictions are too extreme (high risks too high, low risks too low) — the classic signature of overfitting. See @sec-calibration-slope. |
| **Eavg** | 0.005 | 0 | Average absolute gap between predicted and observed risk, in risk units. 0.005 is half a percentage point. |
| **Emax** | 0.117 | 0 | The *worst* such gap anywhere on the curve. Almost always driven by the sparse tail, so check panel C before worrying. |

**How good is 0.75?** This is the question the raw number never answers on its own. As a rough guide: 0.5 is chance, 0.6–0.7 poor, 0.7–0.8 acceptable and where most useful clinical models live, 0.8–0.9 good, above 0.9 excellent — and, in a clinical prediction model built from routine risk factors, above 0.9 is more often a warning sign than a triumph. It usually means something in the predictor set is a consequence of the outcome rather than a cause of it. See @tbl-auc-benchmarks in @sec-model-evaluation for the fuller version, and treat all such bands as context-dependent: 0.75 for ten-year cardiovascular risk is a genuinely useful model, because whether a particular person has a heart attack is substantially a matter of chance that no set of baseline measurements can predict. There is a ceiling, and it is well below 1.

**Two traps in that table.** First, the Brier score cannot be read on its own. Predicting the overall event rate of 11.75% for every single patient — a model with no predictors at all — would score $0.1175 \times 0.8825 = 0.104$. Our model manages 0.092. That is an improvement, but a modest one, and it is a useful corrective to the more flattering C-statistic. Always compare a Brier score against this no-predictor benchmark.

Second, and more important: **the Intercept of exactly 0 and Slope of exactly 1 are not good news — they are arithmetically guaranteed.** We measured calibration on the very data the model was fitted to, and a logistic model is by construction perfectly calibrated in its own development sample. These two numbers only become informative once they are computed on different data, or corrected for optimism as below. Seeing 0 and 1 here tells you nothing except that the arithmetic worked.

Finally, the statistics *not* in the table — `Dxy`, `R2`, `D`, `U`, `Q`, `S:z`, `S:p` — mostly restate the above or serve specialist purposes. Two are worth naming: `Dxy` (Somers' D) is just the C-statistic rescaled so that 0 rather than 0.5 means no discrimination, via $D_{xy} = 2 \times (C - 0.5)$, which is why the bootstrap code below converts it back with `0.5 * (Dxy + 1)`. And `R2` is the Nagelkerke R-squared met in the sample-size section — 0.169 here.
:::

**Correcting for optimism.** `validate(fit, B = 200)` re-runs the entire modelling process on 200 bootstrap resamples to estimate how much the apparent performance is inflated by having fitted and evaluated on the same patients. For this model the apparent C-statistic of 0.750 corrects down to 0.739, and the calibration slope — pinned at exactly 1 above — corrects to 0.943, meaning the predictions are mildly too extreme and would need shrinking by about 6% to be honest. The optimism is small here because 2,000 patients for 11 parameters is comfortable; in a smaller study the gap would be far wider. The corrected figures, not the apparent ones, are what belongs in a paper. Why this procedure works, and why it is not optional, is set out in @sec-internal-validation and @sec-bootstrap-optimism.

None of the above asks the question that ultimately decides whether a model is worth deploying: would using it actually lead to better decisions than not using it? Discrimination and calibration are necessary but not sufficient for that. See the clinical utility and decision curve material in @sec-performance-validation.

::: {.callout-note}
## Optimism
**Optimism** is the gap between how well a model appears to perform on the data it was built from and how well it actually performs on new data. Because the model has partly fitted the noise in its own development sample, the apparent C-statistic is always a little too rosy. Bootstrap validation estimates this gap so you can subtract it off and report a realistic number.
:::

## The Complete Workflow: A Summary

The complete prediction model development workflow can be summarised in the following steps:

1. **Define** the clinical question (population, outcome, timing, intended use).
2. **Design** the study with adequate sample size (Riley criteria).
3. **Prepare** the data: handle missing values with multiple imputation.
4. **Select** predictors based on clinical knowledge; avoid stepwise procedures.
5. **Specify** the model: consider nonlinear terms for continuous predictors.
6. **Fit** the model, applying shrinkage to combat overfitting.
7. **Evaluate** performance: discrimination, calibration, clinical utility.
8. **Validate** internally using bootstrapping or cross-validation.
9. **Report** transparently following TRIPOD+AI guidelines (see @sec-reporting).

::: {.callout-important}
## The Modelling Pipeline Is Not Linear
In practice, you will iterate. Poor calibration might send you back to model specification. Inadequate discrimination might prompt reconsidering the candidate predictors. Internal validation might reveal severe overfitting, requiring stronger shrinkage. The important thing is that each decision is documented and justified.
:::

## Common Mistakes to Avoid

| Mistake | Why It Is Harmful | Better Approach |
|---------|-------------------|-----------------|
| Stepwise variable selection | Inflates significance, unstable | Pre-specify predictors; use penalisation |
| Univariable screening | Ignores confounding and suppression | Include all clinically plausible variables |
| Complete case analysis | Biases estimates, wastes data | Multiple imputation |
| Evaluating only on training data | Overestimates performance | Internal validation (bootstrap) |
| Dichotomising continuous predictors | Loses information, creates arbitrary groups | Model continuous variables continuously |
| Ignoring nonlinearity | Poor fit, biased predictions | Use splines or fractional polynomials |
| Reporting only accuracy or AUC | Incomplete picture | Report calibration, discrimination, clinical utility |

: Common mistakes in clinical prediction model development. {#tbl-common-mistakes}

## Exercises

These three exercises work through the same development pipeline this chapter built, mostly on the simulated Framingham cohort of @sec-framingham-example.

::: {.callout-tip title="Exercise 1: Sample size calculation"}

You are developing a model to predict pre-eclampsia (expected prevalence 4%) using 12 candidate predictors, and a published model in a comparable population reported a C-statistic of 0.72.

a) Use `pmsampsize` to determine the minimum required sample size. How many pregnancies would you need to observe?
b) Which of the criteria it evaluates is the binding one here?
c) How does the answer compare with what the 10-events-per-variable rule of thumb would have told you?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/building_prediction_models_ex1.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/building_prediction_models_ex1.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 2: Missing data simulation"}

Take the simulated Framingham dataset above and compare the regression coefficients obtained from (a) complete case analysis, (b) single mean imputation, and (c) multiple imputation.

a) Which approach yields coefficients closest to the true values used to simulate the data?
b) What happens to the standard errors under single imputation, and why is that the more serious problem?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/building_prediction_models_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/building_prediction_models_ex2.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 3: Variable selection"}

Fit the Framingham model using (a) all pre-specified predictors, (b) forward stepwise selection, and (c) LASSO.

a) Compare which variables each approach selects. Then add a dozen pure-noise candidates to the list and compare again --- selection only has work to do when some candidates do not belong.
b) Repeat the selection across 100 bootstrap samples and record how often each variable is chosen. How stable is stepwise selection compared with LASSO?
c) What does that instability imply for a paper reporting a single stepwise-selected model?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/building_prediction_models_ex3.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/building_prediction_models_ex3.py
```
:::
:::

:::

## Summary

- A prediction model starts with a **clearly defined clinical question**, not with data analysis.
- **Sample size** should be determined prospectively using the Riley et al. criteria, not just the EPV rule of thumb.
- **Missing data** should be handled with multiple imputation, not deletion.
- **Variable selection** should be driven by clinical knowledge. Stepwise methods are harmful.
- **Overfitting** is the central challenge; combat it with adequate sample size and shrinkage.
- **Penalised regression** (ridge, LASSO, elastic net) embeds shrinkage into the estimation process.
- The final model should be **internally validated** before any claims about performance.

## References and Further Reading

- For the prediction modelling workflow, see @smits2026prediction.
- For handling missing data during model development, see @vanbuuren2018flexible.
- For penalised regression and modelling strategy, see @harrell2015 and @hastie2009.

::: {.sectionrefs}
:::
