# Missing Data and Multiple Imputation {#sec-missing-data}

```{r}
#| include: false
library(tidyverse)
```

Open almost any electronic health record and you will find holes. A patient never had their cholesterol checked; a smoking-status field was left blank; a follow-up visit was missed. Missing values are not an occasional nuisance in clinical data --- they are the default state. The tempting fix is to throw away every incomplete record and analyse only the "complete" patients. This chapter is largely about why that instinct is dangerous: dropping incomplete rows wastes hard-won data and, worse, can *bias* your results in a direction you cannot see. We will build up the modern alternative --- **multiple imputation** --- gently, from what makes data go missing through to fitting and reporting a model on imputed data.

## Introduction

Suppose you are studying whether a biomarker predicts mortality, and the biomarker was measured in only 70% of patients. The quickest analysis simply drops the 30% with a missing value and runs the model on whoever remains. This is **complete-case analysis** (also called *listwise deletion*), and it is still the most common approach in clinical papers.

It has two problems. First, it is **wasteful**: a patient missing one value out of twenty is discarded entirely, even though their other nineteen values carry information. With several variables each partly missing, complete-case analysis can quietly delete the majority of your sample. Second, and more seriously, it can be **biased**: if the patients with missing values differ systematically from those without (and they usually do), the complete cases are no longer a fair representation of your population. For a short, clinically grounded overview of these trade-offs aimed at journal readers, see the *JAMA Guide to Statistics and Methods* piece by @newgard2015missing.

::: {.callout-important}
## Missingness is rarely random in clinical data
A test is ordered *because* the clinician was worried; a follow-up is missed *because* the patient was too unwell to attend, or because they recovered and stopped coming. The very reason a value is missing is often linked to the patient's condition and outcome. That is exactly the situation in which complete-case analysis goes wrong --- and exactly why we need the tools in this chapter.
:::

## Why are data missing? MCAR, MAR, MNAR

Before fixing missing data we must think about *why* it is missing. Statisticians describe three mechanisms, and the right method depends on which one you are facing. The names are unfortunate jargon, so we define each in plain terms with a clinical example.

**MCAR --- Missing Completely At Random.** The missingness has nothing to do with anything, observed or unobserved --- as if a coin decided which values to erase. *Clinical example*: a lab machine breaks for a day, so anyone sampled that day has a missing result regardless of who they are. Under MCAR, the complete cases are a genuinely random subset, so complete-case analysis is unbiased (just wasteful). MCAR is rare in practice.

**MAR --- Missing At Random.** The probability that a value is missing depends only on *other variables you have observed*, not on the missing value itself once those are accounted for. *Clinical example*: HbA1c is more often missing in younger patients --- but among patients of the same age, whether HbA1c is missing has nothing to do with the HbA1c value. Because age (which you observed) explains the missingness, you can recover the lost information by using age in the imputation. **MAR is the assumption under which multiple imputation works**, and it is the realistic target for most analyses.

**MNAR --- Missing Not At Random.** The probability of missingness depends on the *missing value itself*, even after accounting for everything you observed. *Clinical example*: patients with the highest blood pressure feel unwell and skip the clinic, so the most extreme readings are precisely the ones absent --- and no observed variable fully explains this. MNAR is the hardest case; standard imputation can still be biased, and you must reason about the mechanism directly (see the sensitivity-analysis note at the end).

For a health-researcher's walkthrough of these three mechanisms with worked observational-study examples, @pedersen2017missing in *Clinical Epidemiology* is a clear and concise companion to this section.

```{mermaid}
%%| eval: true
%%| echo: false
%%| label: fig-missing-mechanisms
%%| fig-cap: "The three missingness mechanisms. The key question is what the chance of being missing depends on."
flowchart TD
    Q["Why is the value missing?<br/>What does the chance of<br/>missingness depend on?"]
    Q --> A["MCAR: on nothing<br/>(pure chance, e.g. machine<br/>broke for a day)"]
    Q --> B["MAR: only on OBSERVED<br/>variables (e.g. missing more<br/>in younger patients)"]
    Q --> C["MNAR: on the MISSING<br/>value itself (e.g. sickest<br/>patients skip the visit)"]
```

::: {.callout-note}
## You cannot fully test which mechanism you have
Whether data are MAR or MNAR depends on the *unobserved* values, so no statistical test can prove it from the data alone. The choice rests on clinical judgement about *why* values went missing. The pragmatic strategy: make missingness as close to MAR as you can by including rich predictors in the imputation, then test robustness with an MNAR sensitivity analysis.
:::

## From single to multiple imputation

**Imputation** means filling in a missing value with a plausible estimate. The naive version is **single imputation** --- replace each gap with one number, most commonly the column's mean (*mean imputation*). This is tempting but quietly damaging: it pretends you *know* the filled-in value with certainty. It shrinks the natural variability of the data, distorts correlations, and --- crucially --- gives standard errors that are too small, because the analysis never registers that you guessed. You end up overconfident, the same trap as complete-case analysis but dressed differently.

**Multiple imputation (MI)**, developed by Rubin, fixes this by being honest about uncertainty. Instead of one filled-in value, it creates *several* complete datasets, each with the gaps filled in slightly differently to reflect the range of plausible values. You then analyse every dataset and combine the answers. The spread *between* the datasets becomes part of your final uncertainty --- so the standard errors correctly include the cost of having had to guess. The *JAMA Guide to Statistics and Methods* article by @li2015multiple is a particularly accessible one-page introduction to this idea for clinical readers.

```{mermaid}
%%| eval: true
%%| echo: false
%%| label: fig-mi-flow
%%| fig-cap: "Multiple imputation in three steps: impute several complete datasets, analyse each one separately, then pool the results with Rubin's rules."
flowchart LR
    D["Incomplete<br/>dataset"] --> I["IMPUTE:<br/>create m complete<br/>datasets"]
    I --> A["ANALYSE:<br/>fit your model on<br/>each one separately"]
    A --> P["POOL:<br/>combine estimates with<br/>Rubin's rules"]
    P --> R["One result with<br/>honest standard errors"]
```

### MICE: imputation by chained equations

The most popular engine for multiple imputation is **MICE --- Multiple Imputation by Chained Equations** (also called *fully conditional specification*). The intuition is a round-robin of small regressions: to fill in a variable's gaps, MICE predicts it from all the *other* variables; it then moves to the next incomplete variable and does the same, looping through all variables repeatedly until the filled-in values settle. Because each variable is imputed from the others, MICE naturally preserves the relationships between them --- a patient imputed to have high blood glucose will also tend to be imputed with the higher weight and HbA1c that usually accompany it. Repeating the whole loop with different random draws is what produces the several distinct completed datasets.

### Rubin's rules: pooling the results

Once you have fitted your model (say a logistic regression) on each of the $m$ imputed datasets, you have $m$ slightly different sets of estimates. **Rubin's rules** are the recipe for combining them into one:

- The **pooled estimate** is simply the *average* of the estimates across the datasets.
- The **pooled variance** has two parts added together: the average *within*-dataset variance (your ordinary uncertainty) **plus** the variance *between* datasets (the extra uncertainty from not knowing the missing values). This second part is exactly what single imputation forgets.

The result is a single coefficient, standard error, confidence interval and p-value that properly reflect the missingness. Modern software does all of this for you.

### How many imputations?

The classic advice was that 5 imputations suffice, and for modest missingness that is often fine. Current practice is more generous: a useful rule of thumb is to set the number of imputations $m$ at least equal to the *percentage* of incomplete cases (so 30% missing → roughly $m = 30$). Imputations are cheap, so when in doubt use more (20--50 is common); more imputations make your results more stable and reproducible.

::: {.callout-tip}
## Always include the outcome and helpful auxiliary variables
Two rules prevent the commonest MI mistakes. **First, include the outcome variable in the imputation model.** It feels wrong --- "using the outcome to fill in predictors?" --- but omitting it biases the very associations you are trying to estimate towards zero. **Second, include auxiliary variables**: extra variables not in your final analysis model but that help *predict* the missing values (e.g. a related lab test, an earlier measurement). They make the MAR assumption more plausible and sharpen the imputations. The imputation model should be at least as rich as your analysis model.
:::

## Worked example: complete-case vs multiple imputation

We simulate a clinical cohort, deliberately knock holes in it under a MAR mechanism, then compare three analyses: the full data (the truth we are trying to recover), complete-case analysis, and multiple imputation pooled with Rubin's rules.

::: {.panel-tabset}

### R

```{r}
#| label: missing-r-setup
#| eval: false

library(tidyverse) # data simulation and wrangling

set.seed(42)
n <- 800

# --- Simulate a complete clinical cohort ---
full <- tibble(
  age = rnorm(n, 60, 12),
  bmi = rnorm(n, 28, 5),
  sbp = 100 + 0.4 * age + 0.6 * bmi + rnorm(n, 0, 10), # systolic BP
  # Binary outcome: cardiovascular event, depends on age, bmi, sbp
  event = rbinom(n, 1, plogis(-6 + 0.04 * age + 0.03 * bmi + 0.02 * sbp))
)

# --- Induce MAR missingness in BMI ---
# Older patients are MORE likely to have missing BMI (depends on age,
# an OBSERVED variable -> Missing At Random)
p_missing <- plogis(-2 + 0.05 * (full$age - 60))
miss <- rbinom(n, 1, p_missing) == 1

missing_data <- full
missing_data$bmi[miss] <- NA

cat("Proportion of BMI missing:", round(mean(miss), 3), "\n")
cat("Complete cases retained:", sum(!miss), "of", n, "\n")
```

**What the code shows.** We create 800 patients with age, BMI, systolic blood pressure and a binary cardiovascular `event` whose risk genuinely depends on all three. We then delete BMI for some patients, making deletion more likely in *older* patients --- so missingness depends on age, an observed variable. That is textbook **MAR**. The printout reports what fraction of BMI is now missing and how many complete cases survive; notice that even moderate per-variable missingness can remove a sizeable chunk of the sample once you require complete records.

```{r}
#| label: missing-r-complete-case
#| eval: false

library(tidyverse)

set.seed(42)
n <- 800
full <- tibble(
  age = rnorm(n, 60, 12),
  bmi = rnorm(n, 28, 5),
  sbp = 100 + 0.4 * age + 0.6 * bmi + rnorm(n, 0, 10),
  event = rbinom(n, 1, plogis(-6 + 0.04 * age + 0.03 * bmi + 0.02 * sbp))
)
p_missing <- plogis(-2 + 0.05 * (full$age - 60))
missing_data <- full
missing_data$bmi[rbinom(n, 1, p_missing) == 1] <- NA

# --- Reference: model on the FULL (pre-deletion) data ---
fit_full <- glm(event ~ age + bmi + sbp, data = full, family = binomial)

# --- Complete-case analysis (glm drops rows with any NA automatically) ---
fit_cc <- glm(event ~ age + bmi + sbp, data = missing_data, family = binomial)

cat("Full-data BMI coefficient:       ", round(coef(fit_full)["bmi"], 4), "\n")
cat("Complete-case BMI coefficient:   ", round(coef(fit_cc)["bmi"], 4), "\n")
cat("Complete-case n:", length(fit_cc$y), "of", n, "\n")
```

**What the code shows.** We fit the same logistic regression twice: once on the full data (our gold-standard answer) and once on the data with holes, where `glm` silently drops every patient with a missing BMI. Compare the two `bmi` coefficients and the sample size. The complete-case model is fitted on a smaller, *older-skewed* subset (because older patients were preferentially deleted), so its estimate is noisier and can drift away from the full-data value. This is the waste-and-bias problem made concrete --- and the motivation for imputing instead of deleting.

```{r}
#| label: missing-r-mice
#| eval: false

library(tidyverse)
library(mice) # multiple imputation by chained equations

set.seed(42)
n <- 800
full <- tibble(
  age = rnorm(n, 60, 12),
  bmi = rnorm(n, 28, 5),
  sbp = 100 + 0.4 * age + 0.6 * bmi + rnorm(n, 0, 10),
  event = rbinom(n, 1, plogis(-6 + 0.04 * age + 0.03 * bmi + 0.02 * sbp))
)
p_missing <- plogis(-2 + 0.05 * (full$age - 60))
missing_data <- full
missing_data$bmi[rbinom(n, 1, p_missing) == 1] <- NA

# --- Multiple imputation with mice ---
# m = 20 imputed datasets; the imputation model uses ALL variables,
# INCLUDING the outcome 'event' (important!).
imp <- mice(missing_data, m = 20, method = "pmm", seed = 123, printFlag = FALSE)

# --- Analyse each imputed dataset, then pool with Rubin's rules ---
fits <- with(imp, glm(event ~ age + bmi + sbp, family = binomial))
pooled <- pool(fits)
summary(pooled, conf.int = TRUE)
```

**What the code shows.** `mice()` builds 20 completed datasets using **predictive mean matching** (`pmm`, a robust default that fills gaps with observed values from similar patients, so imputed BMIs stay realistic). By default `mice` uses every column --- including the outcome `event` --- as a predictor, which is exactly what we want. `with(imp, glm(...))` fits our logistic model separately on all 20 datasets, and `pool()` applies **Rubin's rules** to combine them. The `summary()` table looks like an ordinary regression output --- estimate, standard error, confidence interval, p-value --- but every number now reflects both the within-dataset uncertainty and the between-dataset spread. Compare the pooled `bmi` estimate with the full-data and complete-case values from the previous chunk: the MI estimate should track the full-data truth more closely than complete-case analysis did, while using *all* 800 patients.

::: {.callout-warning}
## Check your imputations before trusting them
After running `mice`, inspect the diagnostics. `plot(imp)` shows whether the chained-equation loops have settled (the lines for different imputations should mingle, not drift or separate). `densityplot(imp)` overlays the distribution of imputed values on the observed ones; imputed values should look plausible and broadly similar in shape --- wildly different distributions signal a problem with the imputation model.
:::

### Python

```{python}
#| label: missing-py
#| eval: false

import numpy as np
import pandas as pd
from sklearn.experimental import enable_iterative_imputer  # noqa: F401
from sklearn.linear_model import LogisticRegression
import statsmodels.api as sm

rng = np.random.default_rng(42)
n = 800

# --- Simulate a complete clinical cohort ---
age = rng.normal(60, 12, n)
bmi = rng.normal(28, 5, n)
sbp = 100 + 0.4 * age + 0.6 * bmi + rng.normal(0, 10, n)
lin = -6 + 0.04 * age + 0.03 * bmi + 0.02 * sbp
event = rng.binomial(1, 1 / (1 + np.exp(-lin)))

full = pd.DataFrame({"age": age, "bmi": bmi, "sbp": sbp, "event": event})

# --- Induce MAR missingness in BMI (depends on observed age) ---
p_missing = 1 / (1 + np.exp(-(-2 + 0.05 * (age - 60))))
miss = rng.binomial(1, p_missing) == 1
data = full.copy()
data.loc[miss, "bmi"] = np.nan
print(f"Proportion BMI missing: {miss.mean():.3f}")

# --- Complete-case logistic regression ---
cc = data.dropna()
Xcc = sm.add_constant(cc[["age", "bmi", "sbp"]])
fit_cc = sm.Logit(cc["event"], Xcc).fit(disp=0)
print("\nComplete-case BMI coefficient:", round(fit_cc.params["bmi"], 4),
      "  (n =", len(cc), ")")
```

**What the code shows.** This mirrors the R setup: a complete cohort, then MAR missingness induced in BMI as a function of age, followed by a complete-case logistic regression via `statsmodels`. The printed `bmi` coefficient and reduced sample size show the same waste-and-bias issue as in R --- the dropped patients are disproportionately older, so the surviving subset is not representative. Note we import `enable_iterative_imputer` to switch on scikit-learn's experimental multiple-imputation machinery, used next.

```{python}
#| label: missing-py-mi
#| eval: false

import numpy as np
import pandas as pd
from sklearn.experimental import enable_iterative_imputer  # noqa: F401
from sklearn.impute import IterativeImputer
import statsmodels.api as sm

rng = np.random.default_rng(42)
n = 800
age = rng.normal(60, 12, n); bmi = rng.normal(28, 5, n)
sbp = 100 + 0.4 * age + 0.6 * bmi + rng.normal(0, 10, n)
event = rng.binomial(1, 1 / (1 + np.exp(-(-6 + 0.04*age + 0.03*bmi + 0.02*sbp))))
data = pd.DataFrame({"age": age, "bmi": bmi, "sbp": sbp, "event": event})
p_missing = 1 / (1 + np.exp(-(-2 + 0.05 * (age - 60))))
data.loc[rng.binomial(1, p_missing) == 1, "bmi"] = np.nan

# --- Multiple imputation: run IterativeImputer m times with sampling on ---
# sample_posterior=True makes each run draw a DIFFERENT plausible fill-in,
# so we get genuinely multiple imputations (not one repeated answer).
# The outcome 'event' is INCLUDED among the imputation predictors.
m = 20
cols = ["age", "bmi", "sbp", "event"]
estimates, variances = [], []

for i in range(m):
    imputer = IterativeImputer(sample_posterior=True, random_state=i)
    completed = pd.DataFrame(imputer.fit_transform(data[cols]), columns=cols)
    X = sm.add_constant(completed[["age", "bmi", "sbp"]])
    fit = sm.Logit(completed["event"].round(), X).fit(disp=0)
    estimates.append(fit.params["bmi"])
    variances.append(fit.bse["bmi"] ** 2)

# --- Pool with Rubin's rules (by hand, to show the formula) ---
estimates = np.array(estimates); variances = np.array(variances)
q_bar = estimates.mean()                 # pooled estimate = average
u_bar = variances.mean()                 # within-imputation variance
b     = estimates.var(ddof=1)            # between-imputation variance
total_var = u_bar + (1 + 1 / m) * b      # Rubin's total variance
se = np.sqrt(total_var)

print(f"Pooled BMI coefficient: {q_bar:.4f}")
print(f"Pooled standard error:  {se:.4f}")
print(f"95% CI: ({q_bar - 1.96*se:.4f}, {q_bar + 1.96*se:.4f})")
```

**What the code shows.** scikit-learn's `IterativeImputer` is the Python counterpart of MICE: it imputes each variable from the others in a loop. Setting `sample_posterior=True` and a different `random_state` each pass makes the 20 runs produce *genuinely different* completed datasets --- the essence of multiple, not single, imputation. We include `event` among the imputation columns (the same "include the outcome" rule as in R), fit the logistic model on each completed dataset, and then **apply Rubin's rules by hand** so the formula is visible: the pooled estimate is the average of the coefficients, and the total variance is the within-imputation variance plus an inflation for the between-imputation spread. The reported standard error is wider than any single imputation's would be --- correctly reflecting the uncertainty from the missing data. The pooled coefficient should land closer to the true full-data value than the complete-case estimate. (For a more turnkey, MICE-faithful workflow in Python, the `miceforest` and `statsmodels` `MICE` packages are popular alternatives.)

:::

## A note on MNAR and sensitivity analysis

Everything above assumes the data are **MAR** --- that observed variables explain the missingness. If you suspect **MNAR** (the missing values are systematically different in a way no observed variable captures, e.g. the sickest patients dropping out), standard MI can still be biased, and you cannot detect this from the data alone. The responsible response is a **sensitivity analysis**: deliberately impute the missing values under pessimistic MNAR assumptions and see whether your conclusions hold.

The common technique is a **delta adjustment** (a *pattern-mixture* approach): after imputing, shift the imputed values by a fixed amount $\delta$ to represent "the missing patients were systematically worse" --- for example, assume dropouts had blood pressures 10 mmHg higher than the imputation predicted --- and refit. Repeat across a plausible range of $\delta$. If your main finding survives even under a strong, adverse shift, it is robust to MNAR; if it flips under a mild shift, you must be cautious. Reporting such a sensitivity analysis is increasingly expected in clinical trials and high-quality observational studies.

::: {.callout-note}
## Auxiliary variables that only predict missingness can backfire
A subtle 2024 finding: adding an auxiliary variable that predicts *whether* a value is missing but is *unrelated to the value itself* can, under MNAR, actually increase bias. The safe auxiliary variables are those genuinely correlated with the incomplete variable (e.g. an earlier measurement of the same quantity), not merely with the act of dropping out.
:::

## Exercises

::: {.callout-tip title="Exercise 1: Classify the mechanism (conceptual)"}

For each scenario, state whether the missingness is most plausibly MCAR, MAR, or MNAR, and justify in one sentence.

a) A weighing scale was out of service for two weeks; everyone seen in that window has missing weight.
b) Depression-score questionnaires are returned less often by male patients, but among men the chance of returning is unrelated to the score.
c) Patients with the most severe symptoms are too unwell to complete a quality-of-life survey, and severity is not otherwise recorded.


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/missing_data_ex1.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/missing_data_ex1.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 2: Quantify the cost of complete-case analysis"}

Using the simulated cohort from this chapter, additionally induce 20% MAR missingness in `sbp` (in addition to the missing BMI).

a) How many complete cases remain once both variables have gaps?
b) Fit the complete-case logistic model and compare its `age` coefficient and standard error to the full-data model.
c) In one sentence, explain why dropping rows became much more costly once a second variable was incomplete.


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/missing_data_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/missing_data_ex2.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 3: Multiple imputation end to end"}

Using the same dataset with missing BMI and SBP:

a) Run multiple imputation with `m = 30` (R `mice`, or Python `IterativeImputer`).
b) Fit the logistic model on each imputed dataset and pool with Rubin's rules.
c) Compare the pooled estimates and standard errors to both the complete-case and full-data models. Which approach recovers the truth best?
d) Inspect the imputation diagnostics. Do the imputed values look plausible?


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/missing_data_ex3.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/missing_data_ex3.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 4: MNAR sensitivity analysis (conceptual + code)"}

Suppose BMI is suspected MNAR: patients with very high BMI were less likely to be weighed.

a) Explain why standard MI (which assumes MAR) might underestimate the true association between BMI and the outcome here.
b) Describe a delta-adjustment sensitivity analysis: which way would you shift the imputed BMIs, and by how much?
c) Sketch (in words or code) how you would report the result across a range of delta values.


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/missing_data_ex4.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/missing_data_ex4.py
```
:::
:::
:::

## Summary

Missing data are ubiquitous in clinical research, and the instinctive fix --- dropping incomplete records (complete-case analysis) --- is both wasteful and potentially biased, because patients with missing values usually differ from those without. The three missingness mechanisms (MCAR, MAR, MNAR) describe *why* values are absent and determine what is safe to do; MAR, where missingness depends only on observed variables, is the realistic target and the assumption under which **multiple imputation** is valid. Single imputation (e.g. mean filling) is dangerous because it ignores the uncertainty of guessing. Multiple imputation, typically via **MICE**, instead creates several completed datasets, fits the analysis on each, and combines the results with **Rubin's rules**, so the final standard errors honestly include the cost of the missing data. Always include the outcome and helpful auxiliary variables in the imputation model, use enough imputations (roughly the percentage missing), and, where MNAR is plausible, add an MNAR sensitivity analysis.

::: {.callout-tip}
## Key Takeaways
1. Complete-case analysis wastes data and can bias results --- avoid it as a default.
2. MCAR (pure chance), MAR (depends on observed variables), MNAR (depends on the missing value itself); the mechanism dictates the method.
3. Single imputation (mean filling) is unsafe: it understates uncertainty and shrinks standard errors.
4. Multiple imputation = impute several times → analyse each → pool with Rubin's rules.
5. Include the outcome and auxiliary variables in the imputation model; the imputation model should be at least as rich as the analysis model.
6. Use enough imputations (rule of thumb: at least the percentage of incomplete cases).
7. When MNAR is plausible, run a delta-adjustment sensitivity analysis --- robust conclusions survive an adverse shift.
:::

## References and Further Reading

- For missing data theory and multiple imputation foundations, see @rubin1987multiple and @little2019statistical.
- For practical guidance on implementation, see @vanbuuren2018flexible, @vanbuuren2011mice, @sterne2009mi, @newgard2015missing, @li2015multiple, @pedersen2017missing, and @curnow2024auxiliary.

::: {.sectionrefs}
:::
