# Mixed-Effects Models for Clustered and Longitudinal Data {#sec-mixed-models}

```{r}
#| include: false
library(tidyverse)
```

Almost every clinical dataset hides a quiet problem: the rows are not independent of one another. You measure the same patient's blood pressure at five visits; you recruit patients from twelve different ICUs; a multi-site trial pools records from six hospitals. Standard regression assumes every observation is a fresh, unrelated draw --- as if each blood-pressure reading came from a different, unrelated person. When that assumption is false, the analysis quietly fools itself into thinking it has more independent information than it really does, and it hands you standard errors that are too small and p-values that are too impressive. This chapter introduces **mixed-effects models**, the standard tool for getting clustered and repeated data right, and shows how to read their output in plain clinical terms.

## Introduction

Two situations recur constantly in health research, and both break the "independent observations" assumption that ordinary regression relies on:

- **Repeated measurements on the same patient.** You record HbA1c every three months, or blood pressure at every clinic visit. Two readings from the *same* patient resemble each other far more than two readings from different patients --- a patient who runs high tends to run high every time.
- **Patients grouped (nested) within a cluster.** Patients treated in the same ICU, registered with the same GP practice, or recruited at the same trial site share an environment, staff, and case mix. Patients within a cluster resemble each other more than patients from different clusters.

In both cases we say the data are **clustered** (or *hierarchical*, or *multilevel*): observations come in groups, and observations within a group are correlated.

::: {.callout-important}
## Why ignoring clustering is dangerous
If you fit an ordinary regression that pretends every row is independent, you overstate how much information you have. The result is **standard errors that are too small**, **confidence intervals that are too narrow**, and **p-values that are too small** --- so you declare effects "significant" that are not. A study of 50 patients each measured 10 times has 500 rows, but it does *not* contain 500 patients' worth of independent information. Mixed models correct for exactly this.
:::

```{mermaid}
%%| eval: true
%%| echo: false
%%| label: fig-nesting
%%| fig-cap: "Clustered data: patients are nested within hospitals, and repeated visits are nested within patients. Observations inside the same box resemble each other."
flowchart TD
    H1["Hospital A"] --> P1["Patient 1"]
    H1 --> P2["Patient 2"]
    H2["Hospital B"] --> P3["Patient 3"]
    P1 --> V1["Visit 1, 2, 3 ..."]
    P2 --> V2["Visit 1, 2, 3 ..."]
    P3 --> V3["Visit 1, 2, 3 ..."]
```

## Fixed effects and random effects

The phrase "mixed-effects" means the model *mixes* two kinds of effect. Understanding the difference is the heart of this chapter.

A **fixed effect** is an effect you care about estimating directly and want to generalise to everyone: the average change in blood pressure per year of age, the average difference between a drug and placebo. These are the familiar regression coefficients. There is one number for the whole population, and that number is the answer to your clinical question.

A **random effect** is different. Instead of estimating a separate, free coefficient for every cluster, we assume the clusters' values are *drawn from a distribution* --- typically a normal distribution centred on the overall average. We do not care about the individual value for "Hospital B" as such; we treat the hospitals as a random sample from the larger population of hospitals, and we estimate the *spread* between them. The random effects are how the model acknowledges that clusters differ without spending a coefficient on each one.

A useful way to hear it: **fixed effects answer "what is the effect?"; random effects answer "how much do the clusters vary around that effect?"**

### Random intercepts: each cluster its own baseline

The simplest mixed model gives each cluster its own **intercept**. Imagine systolic blood pressure measured across several GP practices. Practice A's patients might average 138 mmHg, Practice B's 129, Practice C's 144 --- different *baselines*, perhaps because of differing populations or measurement habits. A **random intercept** lets every practice slide its own regression line up or down to match its baseline, while the *slope* (the effect of, say, age or treatment) stays shared across practices.

Picture a set of parallel lines, one per cluster: same tilt, different heights.

### Random slopes: each cluster its own trend

Sometimes clusters differ not just in baseline but in *how strongly* a predictor acts. Perhaps a drug lowers blood pressure more in some hospitals than others, or blood pressure climbs faster with age in one practice than another. A **random slope** lets the slope of a predictor vary by cluster too. Now the lines differ in both height *and* tilt.

You add random slopes when you believe the *effect* of a variable genuinely differs between clusters --- and you want your uncertainty to reflect that.

```{mermaid}
%%| eval: true
%%| echo: false
%%| label: fig-intercepts-slopes
%%| fig-cap: "Random intercepts give each cluster its own baseline (parallel lines); random slopes also let each cluster's trend differ (lines fan out)."
flowchart LR
    A["Random intercept<br/>= each cluster its<br/>own baseline<br/>(parallel lines)"]
    B["Random slope<br/>= each cluster its<br/>own trend<br/>(lines fan out)"]
```

### Partial pooling: borrowing strength

Here is the clever part. You could analyse each cluster completely separately (*no pooling*) --- but a practice with only three patients would give a wildly unreliable estimate. Or you could lump everyone together and ignore clusters (*complete pooling*) --- the mistake we are trying to avoid. Mixed models do something in between, called **partial pooling**: each cluster's estimate is pulled (or *shrunk*) towards the overall average, and the pull is stronger for small or noisy clusters and weaker for large, well-measured ones.

The clinical intuition: a practice that contributes only a handful of patients should not be trusted to define its own baseline from scratch; we sensibly assume it is probably not too far from the average until its own data prove otherwise. Partial pooling does this automatically, which is why mixed models give stabler, more honest cluster estimates than fitting each cluster alone.

The same structure turns up again in @sec-meta-analysis, where the clusters are entire published studies rather than clinics, and the "random-effects model" of meta-analysis is this model with the within-cluster variances already known.

### The intraclass correlation (ICC)

Once we let clusters differ, a natural question is: *how much* of the variation is between clusters versus within them? The **intraclass correlation coefficient (ICC)** answers exactly this. It is the proportion of the total variation that is *between* clusters:

$$
\text{ICC} = \frac{\sigma^2_{\text{between}}}{\sigma^2_{\text{between}} + \sigma^2_{\text{within}}}
$$

where $\sigma^2_{\text{between}}$ is the variance of the cluster-level baselines and $\sigma^2_{\text{within}}$ is the leftover (residual) variation within a cluster. Read it as a number from 0 to 1:

- **ICC near 0**: clusters are essentially interchangeable; clustering hardly matters.
- **ICC near 1**: almost all the variation is between clusters; which cluster you are in matters enormously.

An ICC of 0.20 for blood pressure across GP practices means 20% of the variation in readings is explained by *which practice* a patient attends. The ICC is also a tidy summary of "how correlated are two observations from the same cluster" --- which is the very thing ordinary regression ignores. For a conceptual, clinician-friendly walkthrough of how the ICC and multilevel models quantify this "contextual" clustering, the brief tutorial series by @merlo2005multilevel works entirely through a worked blood-pressure example.

### Continuous, binary, or count outcomes: LMM vs GLMM

When the outcome is **continuous** (blood pressure, HbA1c, weight) we fit a **linear mixed model (LMM)** --- a linear regression with random effects bolted on. When the outcome is **binary** (died/survived, readmitted/not), a **count** (number of admissions), or otherwise non-normal, we fit a **generalised linear mixed model (GLMM)** --- the mixed-model version of logistic or Poisson regression. The random-effects idea is identical; only the link between the linear predictor and the outcome changes, exactly as in ordinary generalised linear models.

### A note on GEE: marginal vs conditional models

There is a sibling approach worth knowing about: **generalised estimating equations (GEE)**. Both GEE and mixed models handle clustered data, but they answer subtly different questions.

- **Mixed models are *conditional* (subject-specific).** A coefficient describes the effect *for a given patient or cluster* --- "for this patient, treatment lowers the odds of the event by this much."
- **GEE is *marginal* (population-average).** A coefficient describes the effect *averaged over the whole population* --- "across the population, the proportion with the event shifts by this much."

For continuous outcomes with an identity link the two coincide, but for logistic models they genuinely differ, and the population-average effect is usually closer to the null. Use a **mixed model** when you care about individual or cluster-level trajectories and the variance structure itself; use **GEE** when you want a population-average effect and mainly regard the clustering as a nuisance to be corrected for. GEE is fitted with `geepack` in R. For a clinician-oriented tutorial that sets the summary-statistic, repeated-measures ANOVA, GEE and mixed-model approaches side by side --- with worked examples from published studies --- see @schober2018repeated.

## Packages

For R, `lme4` (with its `lmer` function for LMMs and `glmer` for GLMMs) is the long-standing standard, paired with `lmerTest` to obtain p-values for the fixed effects (base `lme4` deliberately omits them). `glmmTMB` is a newer, fast alternative that handles trickier cases (zero-inflation, complex variance structures) with the same formula syntax, and `geepack` provides GEE. For Python, `statsmodels` offers `MixedLM` for linear mixed models, taking data in *long format* (one row per observation, not per patient).

## Worked example: blood pressure over repeated visits across clinics

We simulate a longitudinal study: patients seen at several clinics, each with systolic blood pressure (SBP) measured at multiple visits over time. We expect (a) clinics to differ in baseline SBP, (b) SBP to drift with visit number, and (c) patients to differ from one another. We will fit a random-intercept model, then a random-slope model, and interpret the variance components.

::: {.panel-tabset}

### R

```{r}
#| label: mixed-r-data-and-fit
#| eval: false

library(tidyverse) # data wrangling and simulation
library(lme4) # lmer() for linear mixed models
library(lmerTest) # adds p-values to lmer() output

set.seed(42)

# --- Simulate a longitudinal, clustered dataset ---
n_clinic <- 8 # 8 clinics
n_patient <- 25 # 25 patients per clinic
n_visit <- 5 # 5 visits per patient

# Each clinic has its own baseline SBP (random intercept, SD = 6)
clinic_effect <- rnorm(n_clinic, 0, 6)
# Each patient has their own baseline on top of the clinic (SD = 8)
patient_effect <- rnorm(n_clinic * n_patient, 0, 8)

bp <- expand_grid(
  clinic = 1:n_clinic,
  patient = 1:n_patient,
  visit = 1:n_visit
) |>
  mutate(
    patient_id = interaction(clinic, patient), # unique patient label
    pid_index = as.integer(patient_id),
    # SBP = grand mean + clinic baseline + patient baseline
    #       + a small downward drift per visit + noise
    sbp = 135 +
      clinic_effect[clinic] +
      patient_effect[pid_index] +
      -1.5 * (visit - 1) +
      rnorm(n(), 0, 5)
  )

# --- Random-intercept model ---
# (1 | clinic/patient_id) = random intercept for clinic,
#   and for patient nested within clinic
m_intercept <- lmer(
  sbp ~ visit + (1 | clinic / patient_id),
  data = bp
)
summary(m_intercept)
```

**What the code shows.** We build a dataset of 8 clinics, 25 patients each, 5 visits each (1000 rows). Each clinic gets its own baseline shift, each patient another shift on top, and SBP drifts down about 1.5 mmHg per visit. The model `sbp ~ visit + (1 | clinic/patient_id)` has one **fixed effect** (`visit`, the shared trend we care about) and **random intercepts** at two levels: clinics, and patients nested within clinics. In the `summary()` output, look first at the **Random effects** table: it reports a variance (and standard deviation) for `clinic`, for `patient_id within clinic`, and a `Residual`. These tell you how much SBP varies between clinics, between patients, and visit-to-visit within a patient. Then the **Fixed effects** table gives the `visit` coefficient --- expect roughly $-1.5$, meaning SBP falls about 1.5 mmHg per visit on average, with a standard error and (thanks to `lmerTest`) a p-value. The clinical reading: after properly accounting for the fact that visits cluster within patients and patients within clinics, blood pressure declines modestly over follow-up.

```{r}
#| label: mixed-r-icc
#| eval: false

library(tidyverse)
library(lme4)

set.seed(42)
# (Same simulation as above, repeated so this chunk stands alone)
n_clinic <- 8
n_patient <- 25
n_visit <- 5
clinic_effect <- rnorm(n_clinic, 0, 6)
patient_effect <- rnorm(n_clinic * n_patient, 0, 8)
bp <- expand_grid(
  clinic = 1:n_clinic,
  patient = 1:n_patient,
  visit = 1:n_visit
) |>
  mutate(
    patient_id = interaction(clinic, patient),
    pid_index = as.integer(patient_id),
    sbp = 135 +
      clinic_effect[clinic] +
      patient_effect[pid_index] +
      -1.5 * (visit - 1) +
      rnorm(n(), 0, 5)
  )

m_intercept <- lmer(sbp ~ visit + (1 | clinic / patient_id), data = bp)

# --- Extract variance components and compute the ICC by hand ---
vc <- as.data.frame(VarCorr(m_intercept))
var_clinic <- vc$vcov[vc$grp == "clinic"]
var_patient <- vc$vcov[vc$grp == "patient_id:clinic"]
var_resid <- vc$vcov[vc$grp == "Residual"]

icc_clinic <- var_clinic / (var_clinic + var_patient + var_resid)
cat("ICC (clinic level):", round(icc_clinic, 3), "\n")
```

**What the code shows.** `VarCorr()` pulls out the three estimated variances. We then compute the clinic-level **ICC** --- the clinic variance divided by the total variance --- by hand, so the formula is transparent. The printed value (around 0.25--0.30 given our simulated variances: a clinic SD of 6 against a patient SD of 8 and residual SD of 5, so roughly $36/(36+64+25) \approx 0.29$) is read as: "this fraction of the total variation in blood pressure is attributable to differences *between clinics*." Even a modest ICC like this is enough to badly distort an ordinary regression's standard errors, which is precisely why we used a mixed model.

```{r}
#| label: mixed-r-slope
#| eval: false

library(tidyverse)
library(lme4)
library(lmerTest)

set.seed(42)
n_clinic <- 8
n_patient <- 25
n_visit <- 5
clinic_effect <- rnorm(n_clinic, 0, 6)
patient_effect <- rnorm(n_clinic * n_patient, 0, 8)
bp <- expand_grid(
  clinic = 1:n_clinic,
  patient = 1:n_patient,
  visit = 1:n_visit
) |>
  mutate(
    patient_id = interaction(clinic, patient),
    pid_index = as.integer(patient_id),
    sbp = 135 +
      clinic_effect[clinic] +
      patient_effect[pid_index] +
      -1.5 * (visit - 1) +
      rnorm(n(), 0, 5)
  )

# --- Random-intercept model (refit here so this chunk stands alone) ---
# (1 | clinic/patient_id) expands to (1|clinic) + (1|clinic:patient_id)
m_intercept <- lmer(sbp ~ visit + (1 | clinic / patient_id), data = bp)

# --- Random-slope model: let the visit trend vary by clinic ---
# Same nesting, but clinic now also gets its own visit slope.
m_slope <- lmer(
  sbp ~ visit + (visit | clinic) + (1 | clinic:patient_id),
  data = bp
)
summary(m_slope)

# Compare the two models with a likelihood-ratio test
anova(m_intercept, m_slope)
```

**What the code shows.** The term `(visit | clinic)` now gives each clinic both its own intercept *and* its own `visit` slope, so the blood-pressure trend is allowed to differ from clinic to clinic. In the `summary()`, the random-effects table gains a variance for the `visit` slope and a correlation between intercept and slope. The `anova()` call performs a likelihood-ratio test comparing the simpler random-intercept model with this richer random-slope model: a small p-value would say the trends genuinely differ between clinics and the slopes are worth keeping. Because our data were simulated with a *single* shared slope, expect the test to be non-significant here --- a reassuring reminder that you should add random slopes only when the data (and clinical sense) support them, since needless complexity can make models fail to converge.

### Python

```{python}
#| label: mixed-py
#| eval: false

import numpy as np
import pandas as pd
import statsmodels.formula.api as smf

rng = np.random.default_rng(42)

# --- Simulate the same longitudinal, clustered dataset (long format) ---
n_clinic, n_patient, n_visit = 8, 25, 5
clinic_effect = rng.normal(0, 6, n_clinic)

rows = []
pid = 0
for c in range(n_clinic):
    for p in range(n_patient):
        patient_effect = rng.normal(0, 8)   # each patient's own baseline
        for v in range(n_visit):
            sbp = (135 + clinic_effect[c] + patient_effect
                   - 1.5 * v + rng.normal(0, 5))
            rows.append({"clinic": c, "patient_id": pid, "visit": v, "sbp": sbp})
        pid += 1

bp = pd.DataFrame(rows)

# --- Random-intercept model: random intercept for patient ---
# statsmodels MixedLM groups by ONE variable; we group by patient.
m_intercept = smf.mixedlm("sbp ~ visit", data=bp, groups=bp["patient_id"])
res_intercept = m_intercept.fit()
print(res_intercept.summary())

# ICC at the patient level = group variance / (group variance + residual)
var_group = float(res_intercept.cov_re.iloc[0, 0])
var_resid = float(res_intercept.scale)
icc = var_group / (var_group + var_resid)
print(f"\nPatient-level ICC: {icc:.3f}")
```

**What the code shows.** We build the same data in **long format** (one row per visit) because that is what `statsmodels` expects. `mixedlm("sbp ~ visit", groups=bp["patient_id"])` fits a linear mixed model with `visit` as the fixed effect and a **random intercept per patient**. In the printed summary, the `visit` coefficient should sit near $-1.5$ (SBP falling per visit), and the `Group Var` line reports the between-patient variance. We then compute the patient-level **ICC** --- between-patient variance over total variance --- which expresses how much of the variation is due to patients being persistently different from one another. Note one practical limitation: `statsmodels` `MixedLM` groups by a *single* level at a time, so to model both clinics and patients you would typically choose the level you care most about or use the `vc_formula` argument for additional variance components; R's `lme4` handles nested levels more naturally with `(1 | clinic/patient_id)`.

```{python}
#| label: mixed-py-slope
#| eval: false

import numpy as np
import pandas as pd
import statsmodels.formula.api as smf

rng = np.random.default_rng(42)
n_clinic, n_patient, n_visit = 8, 25, 5
clinic_effect = rng.normal(0, 6, n_clinic)
rows, pid = [], 0
for c in range(n_clinic):
    for p in range(n_patient):
        patient_effect = rng.normal(0, 8)
        for v in range(n_visit):
            sbp = 135 + clinic_effect[c] + patient_effect - 1.5 * v + rng.normal(0, 5)
            rows.append({"clinic": c, "patient_id": pid, "visit": v, "sbp": sbp})
        pid += 1
bp = pd.DataFrame(rows)

# --- Random-slope model: each patient gets their own visit slope ---
m_slope = smf.mixedlm("sbp ~ visit", data=bp,
                      groups=bp["patient_id"], re_formula="~visit")
res_slope = m_slope.fit()
print(res_slope.summary())
```

**What the code shows.** Adding `re_formula="~visit"` tells `statsmodels` to give each patient their own *slope* on `visit` as well as their own intercept --- the Python equivalent of a random-slope model. The summary now reports an extra variance for the `visit` random effect (how much patients' trajectories differ in steepness) and a covariance between each patient's intercept and slope. Interpret it clinically as: not only do patients start at different blood-pressure levels, their pressures may also change at different rates over follow-up. As in R, only keep random slopes if they are justified; superfluous random slopes often cause the optimiser to warn that the model did not converge.

:::

::: {.callout-tip}
## Putting the variance components into words
The hardest part of a mixed model for newcomers is reading the random-effects table. Translate each variance into a clinical sentence: the *clinic* variance is "how much clinics differ in baseline," the *patient* variance is "how much patients differ from each other," and the *residual* is "how much a single patient's readings bounce around visit to visit." If the between-cluster variance is large relative to the residual, clustering matters a lot --- and analysing the data as if it were independent would have been seriously misleading.
:::

::: {.callout-warning}
## "Model failed to converge"
Mixed models, especially with random slopes or many variance components, sometimes report a convergence warning. This usually means the model is too complex for the data to support --- there is not enough information to estimate every variance. The remedy is usually to *simplify*: drop a random slope, or reduce the random-effects structure, rather than to ignore the warning.
:::

## Exercises

::: {.callout-tip title="Exercise 1: Spot the clustering (conceptual)"}

For each study below, state (a) what the clusters are, (b) whether the data are also longitudinal, and (c) what would go wrong if you analysed it with ordinary regression ignoring the clustering.

a) A trial of a new inhaler recruiting asthma patients from 15 GP practices.
b) A study measuring fasting glucose monthly for a year in 200 diabetics.
c) A multi-site ICU study of a sepsis bundle across 6 hospitals, one outcome per patient.


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/mixed_models_ex1.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/mixed_models_ex1.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 2: Fit and interpret a random-intercept model"}

Using the simulated blood-pressure dataset from this chapter, fit a random-intercept model with `visit` and a simulated binary `treatment` variable as fixed effects.

a) Report the `treatment` fixed-effect coefficient and interpret it clinically.
b) Compute the ICC and put it into one plain-English sentence.
c) Refit the same data with an ordinary linear regression (`lm`) that ignores clustering. Compare the standard error on `treatment`. Which is larger, and why does that matter?


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/mixed_models_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/mixed_models_ex2.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 3: Random intercept vs random slope"}

Extend the model in Exercise 2 to allow the `visit` trend to vary by clinic (a random slope).

a) Use a likelihood-ratio test (`anova` in R) to compare the random-intercept and random-slope models.
b) Does allowing clinic-specific trends improve the model? What does your answer imply clinically?


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/mixed_models_ex3.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/mixed_models_ex3.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 4: GLMM and the marginal/conditional distinction (conceptual)"}

You have a binary outcome (30-day readmission) for patients clustered within hospitals.

a) Would you fit an LMM or a GLMM? Why?
b) A colleague fits the same question with GEE and gets a smaller coefficient. Explain, in plain language, why the population-average (GEE) and subject-specific (mixed model) estimates differ for a logistic model.
c) Which would you report if a hospital manager asks "what happens to *this hospital's* readmission rate if we adopt the intervention"?


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/mixed_models_ex4.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/mixed_models_ex4.py
```
:::
:::
:::

## Summary

Clustered and longitudinal data --- repeated measurements on patients, patients nested in hospitals or practices, multi-site studies --- violate the independence assumption of ordinary regression. Ignoring the clustering produces standard errors and p-values that are too optimistic, leading to false claims of significance. Mixed-effects models solve this by combining **fixed effects** (the population-level effects you want to estimate) with **random effects** (which capture how clusters vary). Random intercepts give each cluster its own baseline; random slopes also let each cluster's trend differ. Through partial pooling, mixed models give stable, honest cluster estimates, and the intraclass correlation summarises how much variation lives between clusters. Linear mixed models handle continuous outcomes and generalised linear mixed models handle binary or count outcomes. GEE offers a population-average alternative when the clustering is a nuisance rather than the focus.

::: {.callout-tip}
## Key Takeaways
1. If observations cluster (repeated measures, or patients within sites), do not use ordinary regression --- it gives standard errors that are too small.
2. Fixed effects answer "what is the effect?"; random effects answer "how much do clusters vary around it?"
3. Random intercepts = each cluster its own baseline; random slopes = each cluster its own trend.
4. The ICC is the share of total variation that is between clusters --- a quick gauge of how much clustering matters.
5. Use a GLMM (not an LMM) for binary or count outcomes.
6. Mixed models are subject-specific (conditional); GEE is population-average (marginal). For logistic models the two differ, so know which question you are answering.
7. Keep the random-effects structure as simple as the data support --- over-complex models fail to converge.
:::

## References and Further Reading

- For mixed-effects models and longitudinal data analysis, see @twisk2013longitudinal, @gelman2007data, @schober2018repeated, and @merlo2005multilevel.
- For GEE methods, see @hubbard2010gee.
- For the R packages used in this chapter, see @bates2015lme4 (`lme4`), @kuznetsova2017lmertest (`lmerTest`), and @brooks2017glmmtmb (`glmmTMB`).

::: {.sectionrefs}
:::
