# Explainable AI & Interpretability for Tabular Clinical Data {#sec-ml-explainability}

A key challenge (and criticism) of complex machine learning models is their lack of interpretability. A deep neural network reading chest X-rays can match a radiologist's accuracy, but it is difficult to understand *why* it made a prediction. Thus, deep learning systems have often been referred to as **black boxes**: data goes in, numbers come out, and the neural layers that produced them are far too tangled to read by eye.

In tabular clinical data, the same problem arises to some extent. A random forest or XGBoost model is built from a familiar, readable component: the decision tree. So why can't we just read the trees? As shown in Chapter 8, such ensemble methods typically contain hundreds of trees, each trained on a different bootstrap sample or a different set of features, and the final prediction is their combined output. Thus, the answer to *why did the model make this prediction?* is spread across all of them, making these models rather *opaque*. However, we will show in this chapter that several statistical methods can be added to our toolkit to interrogate these models after training and get better insight into what they learned and why they made a particular prediction.

## Motivation

In the previous chapters you built tree ensembles that capture non-linear effects and interactions automatically, and you learned how to measure whether their predictions are any good. But a model that predicts well is not the same as a model you can trust. Consider why a clinician should care about *explaining* a model, not just scoring it:

- **Trust and adoption.** A clinician will not act on a risk score they cannot reconcile with their own reasoning. If the model flags a patient as high risk, the natural next question is "on what grounds?", and "the algorithm said so" is not an acceptable answer at the bedside. The case that explainability is a precondition for clinicians to accept AI tools, and the practical forms it can take, is set out for a medical audience by Reddy [@reddy2022explainability].
- **Catching the model relying on the wrong thing.** Models latch onto whatever predicts the outcome in the training data, even when it is clinically nonsensical. A well-known class of failures: a model appears to predict pneumonia severity brilliantly, but it has really learned to read which *hospital scanner* took the X-ray (sicker patients went to a particular machine), or it keys on a code that is a **proxy for access to care** rather than for the disease itself. The model is right for the wrong reason, and it will fail the moment it meets a new hospital.
- **Fairness.** Explanations help reveal whether a model leans on variables (or proxies for them) that would make it perform unequally across groups of patients.
- **Regulation and reporting.** The TRIPOD+AI reporting guideline [@collins2024tripodai], published in 2024, now expects authors of clinical prediction models to describe how the model can be interpreted. Explainability has moved from a nice-to-have to an expectation.

::: {.callout-note}
## Why a clinician should care
An accurate prediction you cannot explain is a clinical liability, not an asset. The first time a black-box model confidently makes an indefensible recommendation, and you cannot say why, trust in the whole tool collapses. Explainability is how you find the model's blindspots *before* a patient does.
:::

Model explanations can generally be divided into two categories:

- **Global explanations** answer: *"What drives the model overall, across all patients?"* These tell you which variables the model relies on most and in which direction, summarising the model's general behaviour.
- **Local explanations** answer: *"Why did the model give **this** patient **this** prediction?"* These break a single prediction down into the contribution of each of that patient's characteristics.

@fig-explain-taxonomy lays out the methods we will cover and where each one sits.

```{mermaid}
%%| eval: true
%%| echo: false
%%| label: fig-explain-taxonomy
%%| fig-cap: "A map of the explanation methods in this chapter. Global methods describe the model's overall behaviour; local methods explain one patient's prediction. Shapley values are the bridge: they explain single patients and can be aggregated to a global view."
flowchart TD
    M["Trained black-box model<br/>(random forest, XGBoost)"] --> G["GLOBAL<br/>what drives the model overall?"]
    M --> L["LOCAL<br/>why THIS patient's prediction?"]
    G --> PFI["Permutation<br/>feature importance"]
    G --> PDP["Partial dependence<br/>and ALE plots"]
    G --> SHAPG["Shapley value summary<br/>(beeswarm) plot"]
    L --> SHAPL["Shapley values for one patient<br/>(waterfall / force)"]
    L --> LIME["LIME<br/>(local linear fit)"]
    SHAPG -.same method.- SHAPL
    style M fill:#eef3fb,stroke:#33558b
    style G fill:#eef3fb,stroke:#33558b
    style L fill:#eef3fb,stroke:#33558b
    style PFI fill:#eef3fb,stroke:#33558b
    style PDP fill:#eef3fb,stroke:#33558b
    style SHAPG fill:#eef3fb,stroke:#33558b
    style SHAPL fill:#eef3fb,stroke:#33558b
    style LIME fill:#eef3fb,stroke:#33558b
```

One warning before we start, which we will return to in force at the end: every method in this chapter explains **the associations the model learned**, not the biology of disease. An explanation tells you how the model is behaving. It does not tell you that the model is *correct*, and it certainly does not establish *cause and effect*.

### Running example: readmission data and random forest

Throughout, we will use a single running scenario: a model predicting the risk of an unplanned hospital readmission within 30 days of discharge, built from routine variables (age, length of stay, number of comorbidities, prior admissions, a few discharge labs). The code below creates this simulated dataset and fits a random forest that the rest of this chapter will interrogate. The true risk is driven mainly by prior admissions and number of comorbidities, with a small age effect and near-zero contribution from discharge labs. Knowing the ground truth lets us check whether the explanation tools recover it.

::: {.panel-tabset}

#### R

```{r}
#| label: setup-data-r
#| eval: true
#| output: false
library(ranger)
library(dplyr)
library(ggplot2)
library(pROC)

set.seed(42)
n <- 1500
dat <- tibble(
  age = rnorm(n, 68, 12),
  length_of_stay = rpois(n, 5) + 1,
  num_comorbidities = rpois(n, 3),
  prior_admissions = rpois(n, 1),
  discharge_hb = rnorm(n, 11, 2),
  discharge_creat = rlnorm(n, 0.2, 0.5)
)
lin <- -3 +
  0.45 * dat$prior_admissions +
  0.20 * dat$num_comorbidities +
  0.015 * (dat$age - 68) -
  0.05 * dat$discharge_hb
dat$readmit <- factor(rbinom(n, 1, plogis(lin)), labels = c("No", "Yes"))

write.csv(dat, "../data/readmit_sim.csv", row.names = FALSE)

rf <- ranger(
  readmit ~ .,
  data = dat,
  probability = TRUE,
  num.trees = 500,
  seed = 42
)
```

#### Python

```{python}
#| label: setup-data-py
#| eval: true
#| output: false
import numpy as np
import pandas as pd
from sklearn.ensemble import RandomForestClassifier
import matplotlib.pyplot as plt

dat = pd.read_csv("../data/readmit_sim.csv")
X = dat.drop(columns="readmit").astype(float)
y = (dat["readmit"] == "Yes").astype(int)

rf = RandomForestClassifier(n_estimators=500, random_state=42, n_jobs=-1)
rf.fit(X, y)
```

:::

## Permutation Feature Importance (Global)

The simplest global question is: *which variables drive the model's predictions?* And a simple way to answer that question is to ask: *if I destroy the information in this variable, how much worse does the model get?* This is the idea behind **permutation feature importance**. Pick a variable (e.g., prior admissions in our example), shuffle (i.e., *permute*) its values randomly across patients, and re-evaluate the model's performance (e.g., AUROC in this example) on the data with the scramble feature. A performance drop indicates that the model was relying on that variable. If it barely changes, the model was not really using it. A performance increase would show that the variable was detrimental to the model's predictions, which can happen if the variable is noisy or misleading.

By repeating this process a few times for each variable, we can rank them by how much the model's performance suffers when they are permuted. This gives us a clear picture of which features are most important to the model's predictions.

Despite its apparent simplicity, this method is recommended as it works for any model and asks the right question *how much does model performance depend on this variable?*. It is more honest than the built-in "impurity" importance you met in the trees chapter, which can be biased toward variables with many distinct values.

::: {.panel-tabset}

### R

```{r}
#| label: perm-importance-r
#| eval: true
#| fig-cap: "Permutation importance: how far the model's AUC falls when each variable's values are randomly shuffled. Bigger drops mean the model relies on that variable more."

library(tidyverse)   # ggplot2 / dplyr / tibble

auc_for <- function(prob) {
  as.numeric(pROC::auc(
    dat$readmit,
    prob,
    levels = c("No", "Yes"),
    direction = "<"
  ))
}
baseline_auc <- auc_for(predict(rf, dat)$predictions[, "Yes"])

predictors <- setdiff(names(dat), "readmit")
set.seed(1)
importance <- sapply(predictors, function(v) {
  drops <- replicate(10, {
    shuffled <- dat
    shuffled[[v]] <- sample(shuffled[[v]]) # break this variable only
    baseline_auc - auc_for(predict(rf, shuffled)$predictions[, "Yes"])
  })
  mean(drops)
})

imp_df <- tibble(Variable = predictors, Importance = importance) |>
  arrange(desc(Importance))

ggplot(imp_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "#2E86AB") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Drop in AUC when shuffled",
    title = "Permutation importance: drop in AUC when each variable is shuffled"
  ) +
  theme_minimal(base_size = 13)
```

### Python

```{python}
#| label: perm-importance-py
#| eval: true
#| fig-cap: "Permutation feature importance (scikit-learn): the mean drop in AUC when each column is shuffled."

from sklearn.inspection import permutation_importance

result = permutation_importance(rf, X, y, scoring="roc_auc",
                                n_repeats=10, random_state=1)
order = result.importances_mean.argsort()

fig, ax = plt.subplots(figsize=(6, 4))
ax.barh(X.columns[order], result.importances_mean[order], color="#2E86AB")
ax.set_xlabel("Drop in AUC when shuffled")
ax.set_title("Permutation importance: drop in AUC when each variable is shuffled")
ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

:::

**What the output shows.** Each bar is the drop in AUC when that variable is shuffled: longer bars mean the model relies on that variable more. As expected, *prior admissions* dominates, with age and comorbidities contributing modestly and discharge labs contributing almost nothing. If the top bar had instead been something clinically irrelevant (a patient ID, an admission timestamp), that would be a red flag that the model learned a data-collection artefact rather than the disease process. Note: the R and Python tabs use the same data but different random forest implementations (`ranger` vs `scikit-learn`), so the exact bar lengths and middle-of-the-pack ordering may differ slightly. The overall story should be the same.

## Partial Dependence Plots (Global)

Knowing that *age* matters does not tell you *how* it matters: does risk climb steadily with age, jump only after 75, or even fall again in the very old? A way to answer this is to look at the model's predictions as you sweep one variable across its range, holding all other variables fixed. For example, take your whole dataset and ask the model to predict risk for every patient, but set *everybody's* age to 50. Then set everybody's age to 51, and so on. The average predicted risk at each age gives you a curve that shows how the model's predictions depend on age. The resulting plot is called a **partial dependence plot (PDP)**. Typical patterns include:

- Flat: the model's predictions do not change with this variable.
- Steadily rising or falling: the model's predictions increase or decrease monotonically with this variable.
- Curved: the model's predictions accelerate or decelerate past a certain threshold.

::: {.callout-warning}
## When predictors are correlated, prefer accumulated local effects (ALE)
When predictors are correlated (as clinical variables usually are), forcing one to an extreme while holding the others fixed creates *impossible patients*. For example, setting everybody's age to 90 while keeping their original bloodwork produces 90-year-olds with the lab values of a 40-year-old, combinations that never occur in practice. The model is asked to predict for these unrealistic profiles, which can distort the curve.

**Accumulated Local Effects (ALE)** plots fix this. Instead of forcing unrealistic values on everyone, ALE looks at how the prediction *changes* over small, realistic windows of the variable and accumulates those local changes. The result reads almost exactly like a PDP (prediction on the y-axis, variable on the x-axis) but it stays more honest in the presence of correlation. When in doubt with correlated clinical predictors, plot the ALE. In R, `iml` and `DALEX` both compute ALE; in Python, the `PyALE` package does.
:::

::: {.panel-tabset}

### R

```{r}
#| label: pdp-r
#| eval: true
#| fig-cap: "Partial dependence of predicted readmission risk on prior admissions: the average risk the model assigns as prior admissions is swept across its range."

library(tidyverse)   # ggplot2 / dplyr / tibble
library(pdp)

pred_fun <- function(object, newdata) {
  predict(object, newdata)$predictions[, "Yes"]
}

pd <- partial(
  rf,
  pred.var = "prior_admissions",
  pred.fun = pred_fun,
  train = dat
)

pd_avg <- pd |>
  group_by(prior_admissions) |>
  summarise(yhat = mean(yhat), .groups = "drop")

ggplot(pd_avg, aes(x = prior_admissions, y = yhat)) +
  geom_line(linewidth = 1.2, colour = "#2E86AB") +
  geom_point(size = 2, colour = "#2E86AB") +
  labs(
    x = "Prior admissions",
    y = "Average predicted risk",
    title = "Partial dependence: prior admissions"
  ) +
  theme_minimal(base_size = 13)
```

### Python

```{python}
#| label: pdp-py
#| eval: true
#| fig-cap: "Partial dependence of predicted readmission risk on prior admissions."

from sklearn.inspection import partial_dependence

pd_result = partial_dependence(rf, X, features=["prior_admissions"], kind="average")
grid_values = pd_result["grid_values"][0]
avg_pred = pd_result["average"][0]

fig, ax = plt.subplots(figsize=(6, 4))
ax.plot(grid_values, avg_pred, linewidth=1.5, color="#2E86AB", marker="o", markersize=4)
ax.set_xlabel("Prior admissions")
ax.set_ylabel("Average predicted risk")
ax.set_title("Partial dependence: prior admissions")
ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

:::

**What the output shows.** The x-axis is prior admissions; the y-axis is the model's average predicted risk at each value. The line climbs from left to right: more prior admissions means higher predicted risk, exactly as a clinician would expect. When reading any PDP, ask: does the direction make clinical sense, and is the shape linear or does it bend at a threshold? A curve that slopes the "wrong" way (e.g., predicted risk *falling* as comorbidities rise) is a red flag worth investigating.

## Shapley Values (Local and Global)

In 1953, the mathematician Lloyd Shapley posed a deceptively simple question: if a team of players jointly earns a payout, how should the credit be fairly divided among them [@shapley1953]? His answer, the **Shapley value**, considers every possible order in which players could join the team and averages each player's marginal contribution.

**From game theory to ML.** **SHAP** (**SH**apley **A**dditive ex**P**lanations) [@lundberg2017shap] applies this idea to model predictions. The "players" are a patient's variables (age, prior admissions, and so on), and the "payout" is the model's predicted risk for that patient. Each variable's Shapley value tells you how much *this patient's* value of that variable pushed the prediction up or down, compared with the average patient. SHAP has become the modern standard for explaining tabular models, and we will spend the most time on it.

The payoff of all that theory is a property that makes SHAP uniquely useful: the SHAP values for one patient **add up**. Start from the model's average prediction across everyone (the "baseline"), add each variable's SHAP value for this patient, and you arrive exactly at this patient's prediction. So SHAP gives you a complete, additive accounting of every individual prediction, and nothing is left unexplained. Because each explanation is exact and additive, you can also average the *magnitudes* of SHAP values across many patients to get a trustworthy **global** importance ranking. This is why SHAP bridges the local and global worlds (recall @fig-explain-taxonomy).

For tree ensembles there is a fast, exact algorithm called **TreeSHAP**, so computing SHAP for a random forest or XGBoost model on a clinical dataset is quick.

Every SHAP figure below is drawn from the *same* fitted model and one set of SHAP values, so we set that up once and then read the plots one at a time. We reuse the readmission data from the running example above but fit an XGBoost model (instead of the random forest) so that we can use the fast, exact TreeSHAP algorithm. The resulting SHAP values are stored in `sv` (R) / `shap_values` (Python); each figure that follows simply reuses that object.

::: {.panel-tabset}

### R

```{r}
#| label: shap-setup-r
#| eval: true

library(xgboost)
library(shapviz)

X <- dat %>% select(-readmit) %>% as.matrix()
y <- as.integer(dat$readmit == "Yes")

xgb <- xgb.train(
  params = list(
    objective = "binary:logistic",
    max_depth = 4,
    learning_rate = 0.1
  ),
  data = xgb.DMatrix(X, label = y),
  nrounds = 100,
  verbose = 0
)

sv <- shapviz(xgb, X_pred = X, X = as.data.frame(X))

# Human-readable variable labels, reused to tidy up the figures below
nice_labels <- c(
  age = "Age (years)",
  length_of_stay = "Length of stay (days)",
  num_comorbidities = "Number of comorbidities",
  prior_admissions = "Prior admissions (count)",
  discharge_hb = "Discharge haemoglobin (g/dL)",
  discharge_creat = "Discharge creatinine (mg/dL)"
)
```

### Python

```{python}
#| label: shap-setup-py
#| eval: true
#| output: false

import xgboost as xgb
import shap
import matplotlib.pyplot as plt

model = xgb.XGBClassifier(
    n_estimators=100, max_depth=4, learning_rate=0.1, eval_metric="logloss"
)
model.fit(X, y)

explainer = shap.TreeExplainer(model, feature_perturbation="tree_path_dependent")
shap_values = explainer(X)
```

:::

### The SHAP summary (beeswarm) plot: global

The single most informative SHAP figure is the **beeswarm summary plot**, and it repays careful reading. Picture this: each variable gets one horizontal row, and the rows are stacked with the most important variable at the top. Within a row, there is **one dot per patient**. The dot's horizontal position is that patient's SHAP value for that variable: to the right means this variable *pushed that patient's predicted risk up*, to the left means it *pushed it down*, and the further from the centre, the bigger the push. The dot's **colour** encodes the variable's value for that patient: typically red/high to blue/low.

So a single beeswarm row tells you both *how important* a variable is (how spread out the dots are) and *which direction* it acts (do high values, in red, sit on the right or the left?). For *prior admissions*, you would expect to see red dots (many prior admissions) clustered on the right (pushing risk up) and blue dots (few prior admissions) on the left (pushing risk down), a clean, clinically sensible gradient. If a variable's colours were scrambled with no clear left-right pattern, the model is using it in a complicated, non-monotonic way worth investigating.

::: {.panel-tabset}

### R

```{r}
#| label: shap-beeswarm-r
#| eval: true
#| fig-cap: "Beeswarm summary. One dot per patient per variable; horizontal position is the SHAP contribution to that patient's predicted risk, and colour is the variable's value (red = high, blue = low)."
# GLOBAL: beeswarm summary, importance AND direction in one figure
sv_importance(sv, kind = "beeswarm") +
  scale_y_discrete(labels = nice_labels) +
  labs(
    title = "Global SHAP summary: which variables matter, and how",
    x = "SHAP value  (left = lowers predicted risk, right = raises it)"
  )
```

### Python

```{python}
#| label: shap-beeswarm-py
#| eval: true
# GLOBAL: beeswarm summary plot
shap.plots.beeswarm(shap_values, show=False)
plt.tight_layout()
plt.show()
```

:::

**What the output shows.** Variables are ranked top-to-bottom by importance. Each dot is one patient: position right of centre means the variable pushed that patient's risk up, left means down. Dot colour encodes the variable's value (red = high, blue = low). As expected, prior admissions and number of comorbidities dominate, with red dots on the right confirming a risk-increasing effect. If a row's colours show no clear left-right gradient, the model is using that variable non-monotonically.

### SHAP dependence (scatter) plots: global

A **SHAP dependence plot** zooms into one variable. It plots, for every patient, the variable's value on the x-axis against its SHAP value on the y-axis. This is like a PDP but built from the exact per-patient contributions, and it reveals the *shape* of the variable's effect plus the spread caused by interactions with other variables (vertical scatter at a given x-value means the variable's effect depends on something else).

::: {.panel-tabset}

### R

```{r}
#| label: shap-dependence-r
#| eval: true
#| fig-cap: "Dependence plot for prior admissions. Each dot is a patient; an upward trend confirms a risk-increasing effect, and vertical spread signals interactions with other variables."
# GLOBAL: dependence plot, the shape of one variable's effect
sv_dependence(sv, v = "prior_admissions") +
  labs(
    title = "SHAP dependence: how prior admissions shape predicted risk",
    x = "Prior admissions (count)",
    y = "SHAP value for prior admissions"
  )
```

### Python

```{python}
#| label: shap-dependence-py
#| eval: true
# GLOBAL: dependence (scatter) plot for prior admissions
shap.plots.scatter(shap_values[:, "prior_admissions"], show=False)
plt.tight_layout()
plt.show()
```

:::

*What to look for.* An upward trend confirms the risk-increasing relationship; any vertical spread at a given x-value hints that the effect is modulated by other variables (an interaction).

### Explaining one patient: local

For a single patient, the **waterfall plot** (or the equivalent **force plot**) is the clinician's friend. It starts at the baseline (the average predicted risk) at the bottom and stacks one bar per variable, each pushing the running total up (red, risk-increasing) or down (blue, risk-decreasing), until it lands on this patient's final predicted risk at the top. Read top to bottom, it is a plain sentence: *"This patient's risk is high mainly because they have four prior admissions (+12%) and a low discharge haemoglobin (+4%), partly offset by their relatively young age (−3%)."* That is exactly the kind of justification you can put in a note or discuss with a patient.

::: {.panel-tabset}

### R

```{r}
#| label: shap-waterfall-r
#| eval: true
#| fig-cap: "Waterfall plot for one patient. Bars begin at the cohort-average predicted risk; each variable then pushes the running total up (risk-increasing) or down (risk-decreasing) until it reaches this patient's final prediction."
# LOCAL: explain a single patient (row 1); sv was computed in the setup chunk
sv_waterfall(sv, row_id = 1) +
  labs(title = "Why this patient? Building one prediction from the baseline up")
```

### Python

```{python}
#| label: shap-waterfall-py
#| eval: true
# LOCAL: explain a single patient (row 0); shap_values came from the setup chunk
shap.plots.waterfall(shap_values[0], show=False)
plt.tight_layout()
plt.show()
```

:::

*What to look for.* The baseline (cohort-average) risk sits at one end and this patient's final predicted risk at the other, with the coloured bars in between showing exactly which of *their* characteristics raised or lowered the prediction, and by how much. This is the figure to show when someone asks "why is *this* patient flagged?"

Taken together, SHAP lets you audit the model globally and defend any single prediction locally, using one coherent, additive accounting. That combination is why it has become the default tool. For a concrete clinical example, Lundberg and colleagues used SHAP to explain a model that predicts intraoperative hypoxaemia in real time, and showed that surfacing the per-patient risk factors to anaesthetists improved their ability to anticipate events, a demonstration that local explanations can change behaviour at the bedside, not just on paper [@lundberg2018hypoxaemia]. A practical note: TreeSHAP is exact and fast for the tree models above; for other model types SHAP falls back to slower approximations (`KernelExplainer` in Python, `kernelshap` in R), so on large datasets you may explain a representative sample of patients rather than all of them.

## LIME (Local)

Although less popular than Shapley Values for tabular clinical data, Local Interpretable Model-agnostic Explanations (**LIME**) [@ribeiro2016lime] can be useful for local explanations. To explain one patient's prediction, LIME creates a cloud of slightly altered versions of that patient (nudging variables up and down), asks the model to score all of them, then fits a *simple, transparent model* (usually a small linear regression) to that local cloud. Because the simple model only has to be accurate *near this one patient* (vs. the entire dataset), its coefficients give a readable, local approximation, e.g., "in the neighbourhood of this patient, higher creatinine and more prior admissions are what drive the risk up."

Compared to SHAP which rests on firm game theory (the fairness guarantees of Shapley values) and its local explanations add up exactly to the prediction, LIME depends on somewhat arbitrary choices (how big is the local neighbourhood? how is "local" defined?) that can make explanations unstable. Reach for SHAP first; understand LIME so you can read older literature and appreciate the shared "explain locally with a simple surrogate" idea.

## A Health-Focused Warning: What Explanations Do and Do Not Tell You

Whilst the above tools are powerful, and precisely because their output looks so authoritative, they are easy to over-trust. Ghassemi, Oakden-Rayner, and Beam make this case forcefully in *The Lancet Digital Health* [@ghassemi2021explainable], arguing that current explainability methods offer a "false hope" for patient-level decision support: an explanation can describe how the model behaves without ever assuring you that it reached its decision for a clinically valid reason, and they urge that rigorous internal and external validation, not explanation, is the more direct route to a trustworthy model. We recommend reading that paper alongside this section.

::: {.callout-important}
## Explanations show association, not causation... and not correctness
- **They describe the model, not the disease.** A SHAP plot says "the model uses high creatinine to raise predicted risk." It does *not* say "high creatinine causes readmission," and it certainly does not license "lowering creatinine will prevent readmission." The variable may be a proxy for sicker patients. **Never read an importance ranking or a SHAP plot as a causal effect.** Intervening on a predictor is a different question entirely, requiring the causal-inference methods covered elsewhere on this website.
- **A plausible explanation is not proof the model is right.** If a model has secretly learned to read the scanner artefact, SHAP will faithfully and confidently report *the scanner-related variable* as important. The explanation is "correct" about the model and yet exposes that the model is wrong. Explanations are a tool for *catching* such failures, but only if you scrutinise them against clinical knowledge rather than rationalising whatever they show.
- **Explanations can be unstable.** Re-run LIME, or compute SHAP on a different background sample, and the numbers can shift. Treat explanations as estimates with their own uncertainty, not as exact readouts.
- **Correlated features split the credit.** When two variables carry overlapping information (say, weight and BMI), the model can use either, and the importance or SHAP credit gets divided between them in ways that can look arbitrary. A variable showing "low importance" may simply have had its thunder stolen by a correlated twin, not be genuinely irrelevant.
:::

The honest stance is this: explanations are an indispensable tool for building trust, satisfying reporting standards, and (most valuably) *catching a model that is right for the wrong reasons*. They are not a substitute for external validation, for clinical judgement, or for the causal questions that ultimately matter to patients.

## Exercises

::: {.callout-tip title="Exercise 1: Permutation importance and the wrong-reason model"}

Take the readmission dataset from the permutation-importance code and add a new variable that is essentially a leak (for example, `discharge_disposition` constructed to be strongly associated with the outcome but clinically a *consequence* of risk rather than a cause). Re-fit the random forest and recompute permutation importance.

1. Where does the leaky variable rank?
2. Explain to a colleague why a high-ranking variable here is a warning sign, not a discovery.
3. What real-world variables in routine hospital data might behave like this leak?


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ml_explainability_ex1.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ml_explainability_ex1.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 2: Reading a SHAP beeswarm"}

Using the SHAP code, produce the beeswarm summary plot and answer:

1. Which two variables are globally most important?
2. For the top variable, do high values (red dots) push the prediction up or down? Is this clinically sensible?
3. Pick a variable whose dots show no clear left-right colour pattern. What might that tell you about how the model uses it?


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ml_explainability_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ml_explainability_ex2.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 3: Explaining one patient to a patient"}

Produce a SHAP waterfall plot for a single high-risk patient and a single low-risk patient.

1. Write one or two sentences, in plain language a patient could understand, explaining each prediction from its waterfall.
2. For the high-risk patient, which characteristic contributed most? Would intervening on that characteristic necessarily reduce the patient's risk? Explain, drawing on the caveats section.


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ml_explainability_ex3.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ml_explainability_ex3.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 4: PDP versus ALE with correlated predictors"}

Modify the simulation so that `discharge_creat` and `age` are strongly correlated (e.g. make creatinine rise with age). Plot the partial dependence of predicted risk on creatinine, then an ALE plot for the same variable (use `iml` or `DALEX` in R, or `PyALE` in Python).

1. Do the two curves agree?
2. If they differ, explain why the PDP may be misleading here.


::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ml_explainability_ex4.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ml_explainability_ex4.py
```
:::
:::
:::

## Summary

- Accurate tree-ensemble models (random forests, XGBoost) are **black boxes**; explainability tools open them after training so we can see what they learned.
- **Global** explanations describe the model overall; **local** explanations justify a single patient's prediction. You need both.
- **Permutation importance** ranks variables by how much shuffling each one hurts model performance, the cheapest check that the model relies on sensible variables.
- **Partial dependence plots** show the average shape of one variable's effect; switch to **ALE** when predictors are correlated.
- **SHAP** is the modern standard: it fairly allocates each prediction among the variables, gives an exact additive local explanation (waterfall/force plots), and aggregates to a global view (beeswarm summary). For tree models it is fast and exact (TreeSHAP).
- **LIME** is the older local method (a simple surrogate fitted near one patient), now largely superseded by SHAP but still seen in the literature.
- Explanations reveal **association learned by the model, not causation**, and a plausible explanation is **not proof the model is correct**. Their greatest value is catching a model that predicts well for the wrong reason.

::: {.callout-tip}
## Key Takeaways
- A prediction you cannot explain is a clinical liability. Explainability supports trust, fairness, and the TRIPOD+AI reporting expectation.
- Distinguish **global** ("what drives the model?") from **local** ("why this patient?").
- Use **permutation importance** for a quick global sanity check, **PDP/ALE** for the shape of effects, and **SHAP** as your primary tool for both global summaries (beeswarm) and per-patient explanations (waterfall).
- Always sanity-check explanations against clinical knowledge; an unexpected top variable may signal a data leak or a proxy, not a finding.
- Explanations are **not causal**. Never read a SHAP value or importance score as an intervention effect.
:::

## References and Further Reading

- For comprehensive treatments of interpretable machine learning, see @molnar2022interpretable and @biecek2021ema.

::: {.sectionrefs}
:::
