# What Is Machine Learning? Core Concepts for Clinical Researchers {#sec-ml-foundations}

```{r}
#| label: setup-ml
#| include: false
#| eval: true
library(tidyverse)
library(tidymodels)
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## What Machine Learning Is (and Is Not)

If you have worked through the regression and Bayesian chapters, you already understand the core of machine learning --- you just did not call it that.

By definition, machine learning is the study of algorithms that learn from data. Whether your algorithm is a logistic regression model that predicts 30-day readmission from 5 clinical variables, or an advanced AI model that reads chest X-rays for pneumonia detection, the goal is the same: **to discover patterns** in data to make predictions or discover structure, without being explicitly programmed with rules.

Thus, compared to traditional statistics, the difference is mainly one of emphasis:

- **Traditional statistics** focuses more on understanding relationships (*what is the effect of smoking on lung cancer risk, adjusted for age?*).
- **Machine learning** focuses more on prediction accuracy (*given these 50 variables, what is the probability this patient will be readmitted within 30 days?*).

This part of the course introduces the vocabulary, workflow, and evaluation frameworks that the ML community uses.

## Brief Overview

Machine learning involves four key components:

1. The **machine learning model** is a mathematical function that maps inputs to predicted outputs. It is the object that *learns*.
   *Examples: linear regression, random forests, and neural networks.*
2. **Data** is the information the model learns from. It can be structured (e.g., lab values, demographics) or unstructured (e.g., images, audio, text).
3. The **task** or **objective** is the prediction or structure discovery problem the model is trying to solve.
   *Examples: predicting levels of a biomarker, classifying images, clustering patients into subgroups.*
4. The **learning algorithm** is the procedure for selecting a specific model given the data.
   *Example: in linear regression, ordinary least squares can be used to find coefficients that minimise the sum of squared errors. OLS is the "learning engine" of the linear regression model.*

<!-- TODO: Add a schematic/diagram showing how the 4 components (model, data, task, optimiser) interact — e.g. a cycle: data feeds into model, model produces predictions, loss compares predictions to truth, optimiser updates model. -->

Whilst each component requires careful thought, modelling itself is only one stage of the data science pipeline. In practice, a considerable amount of work lies upstream (cleaning, linking, and transforming raw data) and downstream (evaluating, interpreting, reporting, and deploying the model).

## Learning Paradigms

There are three main paradigms of machine learning: supervised, unsupervised, and reinforcement learning. Each paradigm has a different goal and requires different types of data. This part of the course will focus on **supervised learning** --- the prediction-oriented methods that build on the regression foundations you already know.

### Supervised Learning

In **supervised learning**, the model learns from labelled examples (input-output pairs). In this framework, an input is fed to the model, which in turn makes a prediction. The model learns based on how well its predictions match the known outputs (the labels). There are two main types of supervised learning: classification and regression.

**Classification** predicts a category; **regression** predicts a continuous value.

| Task | Question | Input | Output |
|------|----------|-------|--------|
| **Classification** | Is this skin lesion malignant or benign? | Dermoscopy image | "malignant" or "benign" |
| | Will this patient be readmitted within 30 days? | Clinical variables | "yes" or "no" |
| **Regression** | What will this patient's HbA1c be in 6 months? | Clinical variables | Predicted HbA1c level |
| | What is the expected length of stay? | Clinical variables | Predicted days |

Most clinical prediction models — logistic regression, random forests, neural networks — are trained using supervised learning.

### Unsupervised Learning

**Unsupervised learning** deals with unlabelled data. Rather than predicting a known outcome, the model discovers structure in the data. Common tasks include:

- **Clustering**: Find an optimal grouping of data based on similarity. *Are there distinct phenotypes of sepsis patients?*
- **Dimensionality reduction**: Compress high-dimensional data into fewer dimensions for visualisation or preprocessing. *Can we summarise 50 lab values into a few meaningful components?*
- **Anomaly detection**: Identify unusual observations. *Which patients have lab value patterns that are outliers?*

We cover dimensionality reduction and clustering in detail later in the course.

### Reinforcement Learning

**Reinforcement learning (RL)** concerns how models (called *agents*) learn to take actions in an environment to optimise a reward signal. In RL, agents learn by trial and error by interacting with the environment and receiving feedback in the form of rewards or penalties. The goal is to learn a policy that maximises cumulative reward over time. While RL has exciting potential in medicine (e.g., optimising treatment policies for sepsis management or dynamic treatment regimes), it requires large amounts of interaction data and is rarely used in standard clinical research. We mention it for completeness but will not cover it further in this course.

::: {.callout-note}
## Where Does Traditional Statistics Fit?
Linear regression, logistic regression, and Cox regression are all forms of supervised learning. The distinction between "statistics" and "machine learning" is largely cultural and historical. Statistics emphasises inference (understanding relationships, testing hypotheses). ML emphasises prediction (making accurate forecasts on new data). The methods overlap substantially.
:::

## Supervised Learning

A core tenet of supervised learning is that the model is trained on a dataset that constitutes a sample of a (much) larger population. The model learns patterns from the training data, but its ultimate goal is to generalise to new, unseen data. This is where the concepts of **bias** and **variance** come into play.

### The Bias-Variance Tradeoff

Every supervised model navigates a tension between two sources of error:

- **Bias**: Error from oversimplifying the model. A model with high bias misses important patterns. It *underfits* the data.
- **Variance**: Error from making the model too flexible. A model with high variance captures noise as if it were signal. It *overfits* the data.

The total prediction error can be decomposed as:

$$\text{Expected Error} = \text{Bias}^2 + \text{Variance} + \text{Irreducible Noise}$$

See [this link](https://en.wikipedia.org/wiki/Bias%E2%80%93variance_tradeoff#Derivation) for a derivation of this decomposition.

We can draw a clinical analogy to make this more concrete. Think of a diagnostic criterion:

- **Too specific** (high bias, low variance): A criterion that requires 5 out of 5 symptoms to diagnose a disease. It misses many true cases (underfitting — the rule is too rigid) but rarely triggers a false alarm.
- **Too sensitive** (low bias, high variance): A criterion that diagnoses the disease if *any* 1 of 5 symptoms is present. It catches nearly every true case but also flags many healthy individuals (overfitting to noise).

The optimal diagnostic criterion balances sensitivity and specificity. Similarly, the optimal ML model balances bias and variance.

#### Visualising the Tradeoff

```{r}
#| label: bias-variance-plot
#| fig-cap: "The bias-variance tradeoff. As model complexity increases, bias decreases but variance increases. The test error (what we care about) has a U-shape."
#| echo: false
#| eval: true

library(tidyverse)   # ggplot2 / dplyr / tibble

complexity <- seq(0.1, 10, length.out = 100)
bias2 <- 5 * exp(-0.5 * complexity)
variance <- 0.3 * complexity^1.5
total <- bias2 + variance + 1 # irreducible noise = 1

df_bv <- data.frame(
  Complexity = rep(complexity, 3),
  Error = c(bias2, variance, total),
  Component = rep(c("Bias²", "Variance", "Total Error"), each = 100)
)

ggplot(
  df_bv,
  aes(x = Complexity, y = Error, color = Component, linetype = Component)
) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(
    values = c(
      "Bias²" = "#E69F00",
      "Variance" = "#56B4E9",
      "Total Error" = "#D55E00"
    )
  ) +
  geom_hline(yintercept = 1, linetype = "dotted", color = "gray50") +
  annotate(
    "text",
    x = 9,
    y = 1.3,
    label = "Irreducible noise",
    color = "gray50"
  ) +
  labs(
    x = "Model Complexity",
    y = "Error",
    title = "The Bias-Variance Tradeoff"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

::: {.callout-tip title="You have already seen bias and variance"}
In Chapter 4, you modelled non-linear relationships. A straight line through a U-shaped relationship has **high bias** — it systematically misses the true pattern. A spline with 20 knots that follows every wiggle in the training data has **high variance** — it captures noise and will perform poorly on new data. The bias-variance tradeoff is the formal name for this tension you have already experienced.
:::

In practice, we cannot directly observe bias and variance separately. We detect the tradeoff by comparing **training error** (how well the model fits the data it learned from) and **test error** (how well it performs on new data). When training error is low but test error is high, the model is overfitting.

### Training, Validation, and Test Sets

On what data should a model learn from? Since we want to strike the right balance between bias and variance, we need to evaluate the model on data it has not seen during learning. To that end, a recommended practice is to split the dataset into three parts: **training**, **validation**, and **test** sets.

Each set serves a distinct purpose.

| Set | Purpose | Typical Size |
|-----|---------|-------------|
| **Training set** | Fit the model (learn parameters) | 60–70% |
| **Validation set** | Tune hyperparameters, select among models | 15–20% |
| **Test set** | Final, unbiased performance estimate | 15–20% |

The most important methodological consideration in machine learning is that the **test set must never be used for any decision** — not for feature selection, not for hyperparameter tuning, not for choosing between models. It is opened exactly once, at the very end, to report final performance. If you use the test set to make modelling decisions, it becomes a validation set, and your reported performance will be optimistically biased.

::: {.callout-warning}
## The Cardinal Sin of Machine Learning
Using the test set during model development — and then reporting performance on that same test set — is the ML equivalent of p-hacking. It produces overoptimistic results that will not replicate. In clinical ML, this can mean deploying a model that performs worse in practice than expected, with potential patient harm.
:::

Splitting data into training, validation, and test sets may be performed at random or in a stratified manner, but the key is that the splits are **mutually exclusive**. Potential pitfalls (also called *data leakage*) include:

* Repeated measurements from the same patient appearing in several splits.
* Site- or instrument-specific patterns in multi-centre data that correlate with the outcome, so the model learns institutional identity rather than biology.
* Features derived from information collected *after* the prediction time point (*temporal leakage*).
* Variables that exist as a consequence of the outcome rather than as independent predictors (e.g., a chemotherapy flag leaking a cancer diagnosis).

In clinical research, we often have hundreds (not millions) of patients. A single 70/15/15 split wastes data and produces unstable estimates. The solution is **cross-validation**.

### Cross-Validation

**Cross-validation (CV)** is a resampling strategy that uses all the data for both training and validation, by rotating which portion is held out.

The most common approach is **k-fold cross-validation**:

1. Randomly split the data into $k$ equally sized **folds** (typically $k = 5$ or $k = 10$).
2. For each fold $i = 1, \ldots, k$:
   - Train the model on all folds except fold $i$.
   - Evaluate on fold $i$.
3. Average the $k$ performance estimates.

This gives a more stable and less biased estimate of performance than a single train/test split. It is **more stable** because the estimate is an average over $k$ different held-out sets rather than one arbitrary split, so a single unlucky partition can no longer dominate the result. It is **less biased** because every observation is used for validation exactly once and for training $k-1$ times, so the model is trained on almost the full dataset each time --- avoiding the pessimistic bias you get when a large chunk of data is permanently held out and never learned from.

#### Variants of Cross-Validation

Plain k-fold works well when the outcome is balanced and every row is an independent patient. Clinical data often violate those assumptions, so several variants exist. Each one fixes a specific problem you are likely to meet:

- **Stratified k-fold**: Ensures each fold has the same proportion of events/classes as the full dataset. *Why it matters:* with a rare outcome (say a 5% readmission rate and 10 folds), an ordinary random split could easily produce a fold with almost no events --- giving a meaningless performance estimate for that fold. Stratification guarantees every fold contains a fair share of events. **Use it by default for any classification problem.**
- **Repeated k-fold**: Repeat the entire k-fold procedure $m$ times with different random splits, then average. *Why it matters:* a single 10-fold split is itself somewhat random --- shuffle the data differently and you get a slightly different performance estimate. Repeating (e.g. 10-fold repeated 5 times = 50 estimates) averages out that luck-of-the-draw and gives a more stable estimate.
- **Leave-one-out (LOO)**: the extreme case where $k = n$, so each fold is a *single* observation --- the model is refit $n$ times, each time leaving out one patient and predicting that patient. *Why it behaves the way it does:* because each training set contains almost all the data ($n-1$ patients), the performance estimate is **nearly unbiased**. But the $n$ held-out predictions are highly correlated (the training sets barely differ from one another), so averaging them removes little noise and the estimate has **high variance** --- and refitting $n$ times is computationally expensive. For these reasons LOO is worth it only for very small datasets, where you cannot afford to hold out 10% at a time.
- **Grouped/clustered CV**: When observations are not independent (e.g., multiple visits per patient, or patients clustered within hospitals), entire groups must be kept together --- all of a patient's rows go into the *same* fold. *Why it matters:* if the same patient appears in both the training and validation folds, the model can "recognise" them at validation time and post an inflated score that will not hold up on genuinely new patients. This is one of the most common and most damaging mistakes in clinical ML. **Never split a patient's data across training and validation sets.**

::: {.panel-tabset}

#### R

```{r}
#| label: cv-r
#| eval: true
library(tidyverse)   # ggplot2 / dplyr / tibble
library(tidymodels)

# Simulate clinical data
set.seed(42)
n <- 500
clin_data <- tibble(
  age = rnorm(n, 60, 12),
  bmi = rnorm(n, 28, 5),
  sbp = rnorm(n, 135, 20),
  glucose = rnorm(n, 110, 30),
  smoking = rbinom(n, 1, 0.25),
  readmit = factor(
    rbinom(n, 1, 0.15),
    levels = c("0", "1"),
    labels = c("No", "Yes")
  )
)

# Create 10-fold CV with stratification
folds <- vfold_cv(clin_data, v = 10, strata = readmit, repeats = 5)
cat("Number of resamples:", nrow(folds), "\n")
cat("Training size per fold:", nrow(training(folds$splits[[1]])), "\n")
cat("Validation size per fold:", nrow(testing(folds$splits[[1]])), "\n")

# Fit logistic regression with cross-validation using tidymodels
log_spec <- logistic_reg() %>%
  set_engine("glm")

log_recipe <- recipe(readmit ~ ., data = clin_data)

log_wf <- workflow() %>%
  add_model(log_spec) %>%
  add_recipe(log_recipe)

cv_results <- fit_resamples(
  log_wf,
  resamples = folds,
  metrics = metric_set(roc_auc, accuracy)
)

collect_metrics(cv_results) |>
  knitr::kable(
    digits = 3,
    caption = "Cross-validated performance (mean +/- standard error across folds)."
  )
```

#### Python

```{python}
#| label: cv-python
#| output: false
#| eval: true
import numpy as np
import pandas as pd
from sklearn.model_selection import (StratifiedKFold, RepeatedStratifiedKFold,
                                      cross_val_score)
from sklearn.linear_model import LogisticRegression
from sklearn.preprocessing import StandardScaler
from sklearn.pipeline import make_pipeline

np.random.seed(42)
n = 500

X = pd.DataFrame({
    'age': np.random.normal(60, 12, n),
    'bmi': np.random.normal(28, 5, n),
    'sbp': np.random.normal(135, 20, n),
    'glucose': np.random.normal(110, 30, n),
    'smoking': np.random.binomial(1, 0.25, n)
})
y = np.random.binomial(1, 0.15, n)

# 10-fold stratified CV, repeated 5 times
cv = RepeatedStratifiedKFold(n_splits=10, n_repeats=5, random_state=42)

# Pipeline: scale features, then logistic regression
pipe = make_pipeline(StandardScaler(), LogisticRegression(max_iter=1000))

scores = cross_val_score(pipe, X, y, cv=cv, scoring='roc_auc')
print(f"Mean AUC: {scores.mean():.3f} (+/- {scores.std():.3f})")
print(f"Number of CV iterations: {len(scores)}")
```

:::

**What the code is showing.** This is a worked example of stratified, repeated cross-validation on simulated patient data. Walking through the output:

- It builds 10 folds and repeats the whole process 5 times, so it reports **50 train/validate cycles** (`Number of resamples: 50`). Each cycle trains the model on 90% of patients and checks it on the held-out 10%, with stratification keeping the event rate steady across folds.
- The headline result is the **mean AUC together with its spread** (mean ± standard deviation). The mean is your best single estimate of how well the model discriminates; the spread tells you how much that estimate wobbles depending on which patients land in which fold. A tight spread means the estimate is trustworthy; a wide spread is a warning that performance is fragile, often a sign of too little data.

The point of the exercise is not the specific number (the data are random here, so the AUC will sit near 0.5) but the *workflow*: this is the template you reuse for real data, simply swapping in your own dataset and model.

### Feature Engineering in Clinical Data

**Feature engineering** is the process of transforming raw variables into features that are more useful for modelling. In clinical ML, this often means incorporating domain knowledge.

#### Examples of Clinical Feature Engineering

| Raw Data | Engineered Feature | Rationale |
|----------|-------------------|-----------|
| Date of birth + visit date | Age at visit | More meaningful than raw dates |
| Systolic BP, Diastolic BP | Mean arterial pressure, pulse pressure | Physiologically meaningful composites |
| Serum creatinine, age, sex, race | eGFR (CKD-EPI equation) | Standard clinical measure of kidney function |
| Multiple lab values over time | Rate of change (slope), variability (SD) | Trajectory matters more than single values |
| ICD-10 codes | Charlson comorbidity index | Summarises comorbidity burden |
| Free-text clinical notes | Extracted symptoms via natural language processing (NLP) | Unlocks unstructured data |
| Medication list | Binary flags for drug classes | Captures treatment intent |

::: {.callout-tip}
#### Domain Knowledge Is Your Superpower
Clinical researchers have a massive advantage in ML: you understand the data. A computer scientist might feed raw creatinine into a model; you know to compute eGFR. A data scientist might treat blood pressure as two independent numbers; you know that pulse pressure has specific physiological meaning. Feature engineering is where clinical expertise directly improves model performance.
:::

### Feature Selection

With electronic health record data, you might have hundreds or thousands of candidate features. Including all of them risks overfitting and reduces interpretability. **Feature selection** identifies the most informative subset. Three common approaches are:

**1. Filter methods**: Rank features by some statistical criterion (correlation with outcome, mutual information, chi-squared test) *before* fitting any model. Fast but ignores feature interactions.

**2. Wrapper methods**: Iteratively fit models with different feature subsets and select the best-performing set. Examples: forward selection, backward elimination, recursive feature elimination (RFE). More accurate but computationally expensive.

**3. Embedded methods**: Feature selection happens as part of model fitting. Examples: LASSO regression (L1 penalty shrinks some coefficients to zero), random forest variable importance, elastic net. Often the best practical choice.

::: {.panel-tabset}

#### R

```{r}
#| label: feature-selection-r
#| eval: true
library(tidyverse)   # ggplot2 / dplyr / tibble
library(tidymodels)

# LASSO for feature selection
set.seed(42)
n <- 400
df_feat <- tibble(
  age = rnorm(n, 60, 12),
  bmi = rnorm(n, 28, 5),
  sbp = rnorm(n, 135, 20),
  glucose = rnorm(n, 110, 30),
  smoking = rbinom(n, 1, 0.25),
  noise1 = rnorm(n), # irrelevant
  noise2 = rnorm(n), # irrelevant
  noise3 = rnorm(n), # irrelevant
  outcome = factor(
    rbinom(n, 1, plogis(-3 + 0.02 * age + 0.05 * bmi)),
    labels = c("No", "Yes")
  )
)

# LASSO logistic regression
lasso_spec <- logistic_reg(penalty = 0.01, mixture = 1) %>%
  set_engine("glmnet")

lasso_recipe <- recipe(outcome ~ ., data = df_feat) %>%
  step_normalize(all_numeric_predictors())

lasso_wf <- workflow() %>%
  add_model(lasso_spec) %>%
  add_recipe(lasso_recipe)

lasso_fit <- fit(lasso_wf, data = df_feat)

# Extract coefficients
tidy(lasso_fit) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(abs(estimate))) %>%
  knitr::kable(
    digits = 3,
    caption = "LASSO coefficients, ordered by absolute size (variables shrunk to zero are dropped)."
  )
```

#### Python

```{python}
#| label: feature-selection-python
#| eval: true
from sklearn.linear_model import LogisticRegression
from sklearn.feature_selection import RFE
from sklearn.preprocessing import StandardScaler
import pandas as pd
import numpy as np

np.random.seed(42)
n = 400
X = pd.DataFrame({
    'age': np.random.normal(60, 12, n),
    'bmi': np.random.normal(28, 5, n),
    'sbp': np.random.normal(135, 20, n),
    'glucose': np.random.normal(110, 30, n),
    'smoking': np.random.binomial(1, 0.25, n),
    'noise1': np.random.normal(0, 1, n),
    'noise2': np.random.normal(0, 1, n),
    'noise3': np.random.normal(0, 1, n)
})
y = np.random.binomial(1, 1 / (1 + np.exp(-(-3 + 0.02 * X['age'] + 0.05 * X['bmi']))))

# LASSO feature selection
scaler = StandardScaler()
X_scaled = scaler.fit_transform(X)

lasso = LogisticRegression(penalty='l1', solver='saga', C=1.0, max_iter=5000)
lasso.fit(X_scaled, y)

coef_df = pd.DataFrame({
    'Feature': X.columns,
    'Coefficient': lasso.coef_[0]
}).sort_values('Coefficient', key=abs, ascending=False)

print("LASSO Coefficients (features with 0 coefficient are excluded):")
print(coef_df.to_string(index=False))

# Recursive Feature Elimination
rfe = RFE(LogisticRegression(max_iter=5000), n_features_to_select=4)
rfe.fit(X_scaled, y)
selected = X.columns[rfe.support_]
print(f"\nRFE selected features: {list(selected)}")
```

:::

**What the code is telling you.** Both examples are a small experiment: we built a dataset where `age` and `bmi` genuinely relate to the outcome, while `noise1`, `noise2`, and `noise3` are pure random noise that we added on purpose. A good feature-selection method should *keep the real predictors and discard the noise*. That is exactly what you should see in the output:

- The **LASSO** table lists each predictor with its coefficient, sorted by size. The L1 penalty shrinks unhelpful coefficients all the way to **exactly zero** --- so the noise variables drop out (coefficient = 0) while age and BMI survive with non-zero coefficients. Reading the table top-to-bottom tells you which variables the model considers most important.
- The **RFE** line prints the handful of features it chose to keep; the noise variables should not be among them.

The clinical point is reassurance: when you face a spreadsheet with dozens of candidate variables, these methods let the data tell you which ones carry signal, rather than you guessing or keeping everything (which invites overfitting). If a variable you *expected* to matter gets dropped, that is itself informative and worth investigating.

## ML vs Traditional Statistics: What Is Actually Different?

This is a question clinical researchers often ask, and the honest answer is: less than you think.

| Aspect | Traditional Statistics | Machine Learning |
|--------|----------------------|-----------------|
| **Primary goal** | Inference (understand relationships) | Prediction (forecast accurately) |
| **Model choice** | Guided by theory and assumptions | Guided by cross-validated performance |
| **Feature selection** | Guided by domain knowledge, parsimony | Data-driven, automated |
| **Evaluation** | p-values, confidence intervals | AUC, accuracy, calibration, cross-validation |
| **Interpretability** | Usually high (coefficients) | Varies (logistic regression = high; deep learning = low) |
| **Sample size** | Can work with small samples | Often needs more data for complex models |
| **Assumptions** | Explicit (linearity, normality, etc.) | Implicit (representative data, stationarity) |

The biggest practical difference is in **workflow**. Statistical modelling typically follows a hypothesis-driven process: specify a model based on theory, fit it, check assumptions, interpret coefficients. ML follows a more empirical process: try many models, tune them, evaluate on held-out data, pick the best performer.

Neither approach is inherently superior. For a clinical trial with 200 patients and 5 pre-specified predictors, logistic regression is perfectly appropriate and more interpretable than a random forest. For a hospital system with 500,000 patient records, 1,000 candidate features, and the goal of predicting ICU transfer, a gradient-boosted tree with cross-validated tuning may outperform logistic regression.

## Exercises

::: {.callout-tip title="Exercise 1: Feature Engineering and Cross-Validation"}

Using the simulated clinical dataset below, compare two logistic regression models using 10-fold stratified cross-validation:

- **Model A (raw features)**: Fit logistic regression on the raw variables as given.
- **Model B (engineered features)**: Add clinically motivated features — e.g., eGFR estimated from creatinine, age, and sex; a hemoglobin-to-platelet ratio; log-transformed WBC — and fit logistic regression on the expanded feature set.

Report AUC for both models. Does feature engineering improve performance? Why or why not?

::: {.panel-tabset}

#### R Starter Code

```{r}
#| label: ex1-ml-r
#| eval: false
library(tidyverse) # tibble()
library(tidymodels) # vfold_cv(), logistic_reg(), recipe(), etc.

set.seed(123)
n <- 600
ex_data <- tibble(
  age = rnorm(n, 65, 10),
  sex = rbinom(n, 1, 0.5),
  creatinine = rlnorm(n, 0, 0.5),
  hemoglobin = rnorm(n, 12, 2),
  platelets = rnorm(n, 250, 70),
  wbc = rlnorm(n, 2, 0.4),
  icu = factor(
    rbinom(
      n,
      1,
      plogis(-4 + 0.03 * age + 0.5 * creatinine)
    ),
    labels = c("No", "Yes")
  )
)

# Your code:
# 1. Create two recipes: one with raw features, one adding engineered features
#    (e.g., step_mutate to add eGFR, hb/platelet ratio, log(wbc))
# 2. Set up logistic_reg() workflows for both
# 3. Use vfold_cv with strata = icu
# 4. Compare using roc_auc
```

#### Python Starter Code

```{python}
#| label: ex1-ml-python
#| eval: false
import numpy as np
import pandas as pd
from sklearn.linear_model import LogisticRegression
from sklearn.model_selection import cross_val_score, StratifiedKFold
from sklearn.preprocessing import StandardScaler
from sklearn.pipeline import make_pipeline

np.random.seed(123)
n = 600

# Your code:
# 1. Create X with raw features (age, sex, creatinine, hemoglobin, platelets, wbc)
# 2. Create X_eng with additional engineered features
#    (e.g., eGFR from creatinine/age/sex, hb/platelet ratio, log(wbc))
# 3. Generate y from age and creatinine
# 4. Compare LogisticRegression on X vs X_eng using cross_val_score
#    with scoring='roc_auc'
```

:::

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ml_foundations_ex1.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ml_foundations_ex1.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 2: Feature Engineering Challenge"}

You have the following raw variables for a diabetes prediction model:

- `height_cm`, `weight_kg`
- `sbp`, `dbp` (systolic and diastolic blood pressure)
- `fasting_glucose`, `hba1c`
- `age`, `sex`
- `waist_circumference`, `hip_circumference`
- `total_cholesterol`, `hdl`, `ldl`, `triglycerides`

1. List at least 5 clinically meaningful engineered features you could create from these variables. For each, explain *why* it is clinically meaningful.
2. Implement the feature engineering in R (using `recipes`) or Python (using pandas/sklearn).
3. Discuss which original features might become redundant after engineering.

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ml_foundations_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ml_foundations_ex2.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 3: The Bias-Variance Tradeoff in Practice"}

Using polynomial regression on a simulated dataset:

1. Fit polynomials of degree 1, 3, 5, 10, and 20 to a training set.
2. Plot the fitted curves overlaid on the data.
3. Compute training error and test error for each degree.
4. Identify the degree that minimises test error. Explain this in terms of the bias-variance tradeoff.

::: {.panel-tabset}

#### R Starter Code

```{r}
#| label: ex3-ml-r
#| eval: false
set.seed(42)
x_train <- sort(runif(50, 0, 10))
y_train <- sin(x_train) + rnorm(50, 0, 0.3)
x_test <- sort(runif(200, 0, 10))
y_test <- sin(x_test) + rnorm(200, 0, 0.3)

# Your code: fit poly(x, degree) for degree in c(1, 3, 5, 10, 20)
# Compute RMSE on train and test for each
# Plot the results
```

#### Python Starter Code

```{python}
#| label: ex3-ml-python
#| eval: false
import numpy as np
from sklearn.preprocessing import PolynomialFeatures
from sklearn.linear_model import LinearRegression
from sklearn.metrics import mean_squared_error

np.random.seed(42)
x_train = np.sort(np.random.uniform(0, 10, 50))
y_train = np.sin(x_train) + np.random.normal(0, 0.3, 50)
x_test = np.sort(np.random.uniform(0, 10, 200))
y_test = np.sin(x_test) + np.random.normal(0, 0.3, 200)

# Your code: loop over degrees [1, 3, 5, 10, 20]
# Fit PolynomialFeatures + LinearRegression
# Compute train and test RMSE
```

:::

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ml_foundations_ex3.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ml_foundations_ex3.py
```
:::
:::

:::

::: {.callout-tip title="Exercise 4: Spot the Data Leakage"}

A colleague has built a model to predict ICU admission from routine clinical variables and is excited to report a cross-validated AUC of 0.99. The code below reproduces their analysis. Run it, confirm the suspiciously high AUC, and then:

1. Identify **which feature(s)** cause the data leakage and explain *why* they leak.
2. Fix the dataset by removing or correcting the problematic feature(s).
3. Re-run the cross-validation and report the new AUC. How does it compare?

::: {.panel-tabset}

#### R Starter Code

```{r}
#| label: ex4-ml-r
#| eval: false
library(tidyverse)
library(tidymodels)

set.seed(42)
n <- 800
ex_data <- tibble(
  age = rnorm(n, 65, 10),
  creatinine = rlnorm(n, 0, 0.5),
  hemoglobin = rnorm(n, 12, 2),
  wbc = rlnorm(n, 2, 0.4),
  icu = factor(
    rbinom(n, 1, plogis(-4 + 0.03 * age + 0.5 * creatinine)),
    labels = c("No", "Yes")
  )
)

# --- Features added by the colleague ---
# Ventilator use (only ICU patients receive mechanical ventilation)
ex_data <- ex_data %>%
  mutate(ventilator = ifelse(icu == "Yes", rbinom(n(), 1, 0.85), 0))

# ICU-specific sedation score (0 for non-ICU patients, 1-10 for ICU patients)
ex_data <- ex_data %>%
  mutate(
    sedation_score = ifelse(icu == "Yes", sample(1:10, n(), replace = TRUE), 0)
  )

# Fit and evaluate
log_wf <- workflow() %>%
  add_model(logistic_reg() %>% set_engine("glm")) %>%
  add_recipe(recipe(icu ~ ., data = ex_data))

folds <- vfold_cv(ex_data, v = 10, strata = icu)
cv_results <- fit_resamples(
  log_wf,
  resamples = folds,
  metrics = metric_set(roc_auc)
)
collect_metrics(cv_results)
# AUC ~ 0.99 — is this real?
```

#### Python Starter Code

```{python}
#| label: ex4-ml-python
#| eval: false
import numpy as np
import pandas as pd
from sklearn.linear_model import LogisticRegression
from sklearn.model_selection import cross_val_score, StratifiedKFold
from sklearn.preprocessing import StandardScaler
from sklearn.pipeline import make_pipeline

np.random.seed(42)
n = 800

X = pd.DataFrame({
    'age': np.random.normal(65, 10, n),
    'creatinine': np.random.lognormal(0, 0.5, n),
    'hemoglobin': np.random.normal(12, 2, n),
    'wbc': np.random.lognormal(2, 0.4, n),
})
y = np.random.binomial(1, 1 / (1 + np.exp(-(-4 + 0.03 * X['age'] + 0.5 * X['creatinine']))))

# --- Features added by the colleague ---
# Ventilator use (only ICU patients receive mechanical ventilation)
X['ventilator'] = np.where(y == 1, np.random.binomial(1, 0.85, n), 0)

# ICU-specific sedation score (0 for non-ICU, 1-10 for ICU)
X['sedation_score'] = np.where(y == 1, np.random.randint(1, 11, n), 0)

# Fit and evaluate
cv = StratifiedKFold(n_splits=10, shuffle=True, random_state=42)
pipe = make_pipeline(StandardScaler(), LogisticRegression(max_iter=1000))
scores = cross_val_score(pipe, X, y, cv=cv, scoring='roc_auc')
print(f"AUC: {scores.mean():.3f} (+/- {scores.std():.3f})")
# AUC ~ 0.99 — is this real?
```

:::

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/ml_foundations_ex4.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/ml_foundations_ex4.py
```
:::
:::

:::

## Summary

This chapter introduced the foundational concepts of machine learning that every clinical researcher should understand before applying ML methods:

- **ML is pattern recognition**, not magic. It requires careful validation and domain knowledge.
- **Supervised learning** (prediction from labelled data) is the most common ML paradigm in clinical research.
- The **bias-variance tradeoff** governs all model selection: too simple = underfitting, too complex = overfitting.
- **Training/validation/test splits** and **cross-validation** are essential for honest performance evaluation.
- **Feature engineering** — where clinical expertise meets data science — is often more important than model choice.
- **Feature selection** (filter, wrapper, embedded) prevents overfitting in high-dimensional settings.
- **ML and statistics are complementary**, not competing. The best clinical researchers use both.

### Common Misconceptions

We list here several misconceptions about machine learning that are particularly relevant to clinical researchers:

1. **"Complex ML models are always better than regression."** False. For many clinical prediction tasks with modest sample sizes and well-understood relationships, logistic or Cox regression can perform just as well as or better than complex ML models — and is far more interpretable. A 2019 systematic review by Christodoulou et al. in the *Journal of Clinical Epidemiology* found that ML models did not consistently outperform logistic regression for clinical prediction once methodological bias was accounted for. As we will see during the course, machine learning practitioners have to balance a trade-off between model complexity, interpretability, accuracy and flexibility to unseen data.

2. **"ML finds truth in data automatically."** False. ML models find patterns — including spurious ones. Without careful validation, feature engineering, and domain knowledge, ML models can learn noise, artefacts, and biases in the data. As the computer science adage goes: "garbage in, garbage out" (GIGO).

3. **"More data always helps."** Mostly true, but not always. Whilst research in large language models (LLMs) has shown that simply scaling up data and model size can go surprisingly far, data quality matters at least as much as data quantity in clinical settings — more biased data produces a more confidently wrong model.

## References and Further Reading

- For the relationship between statistics and machine learning, see @breiman2001cultures.
- For practical machine learning, see @boehmke2019homl and @gatto2024introml.
- For clinical prediction context, see @smits2026prediction and @christodoulou2019systematic.

::: {.sectionrefs}
:::
