# Reporting Standards, TRIPOD+AI, and Clinical Impact {#sec-reporting}

## Introduction

Thousands of clinical prediction models are published every year. Yet only a small fraction are ever validated externally, and fewer still make it into clinical practice. One of the major reasons is poor reporting. Many papers omit critical methodological details, overstate performance, or fail to describe the model in enough detail for anyone to reproduce or implement it. In 2024, the TRIPOD+AI statement was published to address this problem head-on. A reporting guideline is simply an agreed checklist of what a study must describe, so that readers can judge whether the results are trustworthy and whether the model could be used elsewhere.

This chapter walks you through the current reporting standards for clinical prediction models, explains why they matter, and gives you the tools to produce work that meets the expectations of top medical journals.

The lifecycle below frames the whole chapter. A prediction model is not finished when the paper is published. It must be reported clearly, implemented in real workflows, and monitored over time.

```{mermaid}
%%| echo: false
%%| eval: true
%%| label: fig-lifecycle
%%| fig-cap: "The lifecycle of a clinical prediction model, from development to ongoing monitoring. Conceptual schematic aligned with the TRIPOD [@moons2015tripod] and TRIPOD+AI [@collins2024tripodai] reporting frameworks."
flowchart LR
    A[Development] --> B[Validation]
    B --> C[Transparent reporting<br/>TRIPOD+AI]
    C --> D[Implementation<br/>at the bedside]
    D --> E[Monitoring<br/>and updating]
    E -.-> A
```

## Why reporting standards matter

Consider trying to use a published prediction model in your own clinical setting. You would need to know:

- What predictors are included, and exactly how they were measured
- What outcome was predicted, and at what time horizon
- How missing data were handled
- How the model was validated, and on what population
- The model's calibration, not just its discrimination
- Whether the model would be feasible to use in your context

Remarkably, many published prediction model studies omit several of these details. That makes the model impossible to trust or to reuse. A systematic review by Collins et al. found that reporting quality was poor across most published prediction model studies, which motivated the development of formal reporting guidelines.

Why should a clinician care? Because a poorly reported model is one you cannot safely apply to your own patients. If you do not know how risk was defined, in whom it was measured, or whether the predicted percentages are honest, you cannot tell whether acting on the model will help or harm someone in front of you.

## The TRIPOD+AI statement

The **Transparent Reporting of a multivariable prediction model for Individual Prognosis Or Diagnosis** (TRIPOD) guideline was first published in 2015. In 2024, it was updated as **TRIPOD+AI** to address machine learning and AI-based models.

In plain terms, TRIPOD+AI is a checklist of things every prediction-model paper should report. It lets a reader judge whether the model is trustworthy and could actually be used. Think of it the same way you think of CONSORT for randomised trials or STROBE for observational studies. These are not hurdles, but shared expectations that make your work credible and reproducible. Journals increasingly require it, and reviewers will look for it.

A related tool you may encounter is PROBAST (Prediction model Risk Of Bias ASsessment Tool). Where TRIPOD+AI tells authors what to report, PROBAST helps a reader formally assess whether a published model is at risk of bias, for example because of how patients were selected or how the model was validated. Reporting guidelines and risk-of-bias tools work hand in hand.

TRIPOD+AI applies to studies that develop, validate, or update a prediction model, whether based on regression, machine learning, or deep learning. The key additions in the +AI version include:

- **Fairness evaluation**: checking whether the model performs equally well across demographic groups (for example, whether it predicts as accurately for women as for men, or across ethnic groups)
- **Open science**: sharing code, data (where possible), and the model itself, so others can check and reuse it
- **Handling of AI-specific issues**: hyperparameter tuning (the settings that control how a machine learning model learns), computational requirements, and software dependencies

### Key TRIPOD+AI items

The full checklist contains items across title, abstract, introduction, methods, results, and discussion. Here are some of the most commonly missed items:

**Study design and participants:**

- Describe the study design (e.g., cohort, case-control, registry)
- Specify the eligibility criteria clearly
- Report dates of recruitment and follow-up

**Predictors and outcome:**

- List all candidate predictors, how they were defined, and when they were measured
- Define the outcome precisely, including the time horizon for prognostic models

**Missing data:**

- Report the amount of missing data for each variable
- Describe the method used to handle missing data (complete case analysis, imputation)
- If multiple imputation was used, report the number of imputations and imputation model

**Model development:**

- Describe the modelling approach (logistic regression, random forest, etc.)
- Report how continuous predictors were handled (linear terms, splines, categorised)
- For ML models: report hyperparameter tuning strategy, software, and random seeds

**Model performance:**

- Report discrimination (e.g., C-statistic) with confidence intervals
- Report calibration (calibration plot, calibration slope and intercept)
- Report clinical utility (decision curve analysis or net benefit) where relevant

**Validation:**

- Describe the validation approach (internal, external, temporal)
- Report performance in the validation data, not just the development data

If some of these terms are still fuzzy, here is the quick clinician's translation. *Discrimination* (the C-statistic) is whether the model ranks a patient who has the outcome above one who does not --- like a diagnostic test's ability to separate cases from non-cases. *Calibration* is whether the predicted risks are numerically honest: when the model says "20% risk", do roughly 20 in 100 such patients actually have the event? A model can discriminate well yet be badly calibrated (consistently over- or under-stating risk), which is dangerous if clinicians act on the actual numbers. *Clinical utility* (decision curve analysis / net benefit) goes one step further and asks whether using the model to guide decisions does more good than harm at a realistic risk threshold. A complete paper reports all three, in the validation data, not just the development data where performance is always flattering.

::: {.callout-tip title="Exercise 1: TRIPOD+AI audit"}
Find a recently published clinical prediction model paper (try searching PubMed for "prediction model" in your area of interest). Using the TRIPOD+AI checklist, evaluate how many items the paper reports adequately.

Focus on:

1. Is the outcome clearly defined with a time horizon?
2. Is calibration reported, or only discrimination?
3. How was missing data handled?
4. Were continuous predictors appropriately modelled (splines) or just categorised?
5. Is the model available for others to use (coefficients, code, web calculator)?

Write a brief (one paragraph) critical appraisal.

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/reporting_and_implementation_ex1.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/reporting_and_implementation_ex1.py
```
:::
:::
:::

## From model to bedside: the implementation gap

*Implementation* (also called *deployment*) means actually putting a model to work in routine care, for example building it into the electronic health record so it flags high-risk patients automatically. This is the step where most models fail. Even well-developed, well-validated models face an uphill battle to reach the bedside. Smits, van Kuijk & Wynants (2026) dedicate five chapters of their book to this topic, covering the journey from model selection through innovation development, impact evaluation, and implementation.

This matters to clinicians because a model that never reaches the bedside, or that reaches it in an unusable form, helps no patient. The gap between "the model works on paper" and "the model improves care" is where good ideas are usually lost.

### Key barriers to implementation

1. **The model doesn't address a real clinical need.** Models built for academic interest rather than to solve a genuine decision problem rarely get implemented.

2. **The model isn't embedded in clinical workflow.** A model that requires a clinician to manually enter 15 variables into a spreadsheet will not be used, regardless of how accurate it is. Equally, a model that fires too many warnings causes *alert fatigue*: clinicians become so used to seeing pop-up alerts that they start ignoring them, including the important ones.

3. **Performance hasn't been demonstrated in the target population.** External validation in the specific setting where the model will be used is essential.

4. **Clinicians don't trust the model.** Transparency, explainability, and evidence of clinical utility (not just statistical performance) build trust.

5. **There is no plan for maintenance.** Models can degrade over time as populations and practices change. This slow loss of accuracy is often called *calibration drift*: the model's predicted risks gradually stop matching the real event rates (for example, a sepsis model that becomes too pessimistic after treatment protocols change). A plan for ongoing monitoring and updating is essential.

### The prediction model-based innovation (PMBI) framework

@smits2026prediction introduce the concept of a PMBI. The idea is that the prediction model itself is only the engine. What clinicians actually use is the complete tool wrapped around it, including:

- The user interface (how predictions are displayed)
- Decision support (what actions are recommended at different risk levels)
- Integration with electronic health records
- Training materials for end users
- A monitoring and updating plan

A growing convention is to publish a *model card*: a short, standardised summary sheet describing what the model predicts, who it was built for, how it performed, and its known limitations. Think of it as a "package insert" for a prediction model, written so that a busy clinician can quickly judge whether it is fit for their setting.

The final step is an *impact study*: research that tests whether using the model in practice actually changes outcomes for patients (for example, fewer missed diagnoses or shorter hospital stays), rather than only measuring statistical accuracy on paper.

::: {.callout-note}
**Key insight:** Building a good prediction model is necessary but not sufficient for clinical impact. The model must be embedded in a workable clinical tool, validated in the target setting, and demonstrated to improve patient outcomes.
:::

## Risk communication

A prediction is only useful if the person receiving it understands it. Presenting predicted probabilities to patients and clinicians is not straightforward, and how a number is phrased can change the decision it leads to. Research in health literacy shows that:

- **Frequencies are easier to understand than probabilities.** "3 out of 100 patients like you" is clearer than "3% probability."
- **Visual aids help.** Icon arrays (showing 100 faces with 3 highlighted) are effective.
- **Framing matters.** "97% chance of survival" feels different from "3% chance of death" — both are true.
- **Uncertainty should be communicated.** Presenting a point estimate without a confidence interval overstates precision.

### Presenting results to different audiences

| Audience | Recommended format |
|----------|-------------------|
| Patients | Frequencies, icon arrays, plain language |
| Clinicians | Risk categories with action thresholds, decision curves |
| Researchers | Full performance metrics, calibration plots, code |
| Policymakers | Population-level impact, cost-effectiveness |

## Ethical considerations

### Algorithmic bias

Prediction models can perpetuate or amplify existing health disparities if:

- Training data underrepresents certain populations
- Predictors serve as proxies for protected characteristics (e.g., postcode as proxy for race/ethnicity)
- Performance is not evaluated across subgroups

TRIPOD+AI specifically requires fairness evaluation: reporting model performance stratified by relevant demographic groups.

### Informed consent and transparency

Patients have a right to know when a prediction model is being used in their care, what data it uses, and how its output influences clinical decisions. This is particularly important for models that inform treatment decisions or resource allocation.

### The EU AI Act and regulatory considerations

The European Union's AI Act (2024) classifies medical AI applications as "high risk" and imposes requirements for transparency, human oversight, and documentation. Clinical prediction models that qualify as medical devices may need regulatory approval (e.g., CE marking in the EU, FDA clearance in the US).

## Writing a statistical methods section

A well-written methods section for a prediction model study should include:

1. **Study design and setting** (one paragraph)
2. **Participants** — eligibility criteria, dates, sample size (one paragraph)
3. **Predictors** — list, definitions, measurement timing (one paragraph)
4. **Outcome** — definition, ascertainment, time horizon (one paragraph)
5. **Missing data** — amount, mechanism assumed, handling method (one paragraph)
6. **Model development** — method, how continuous variables were handled, variable selection approach (one paragraph)
7. **Model performance** — discrimination, calibration, clinical utility measures (one paragraph)
8. **Validation** — internal and/or external validation approach (one paragraph)
9. **Software** — R/Python version, key packages, random seed (one sentence)

::: {.callout-tip title="Exercise 2: Write a methods section"}
Using the prediction model you developed in earlier chapters (or a hypothetical one), write a complete statistical methods section following the structure above. Aim for approximately 500 words.

Check your methods section against the TRIPOD+AI checklist. Are all key items covered?

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/reporting_and_implementation_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/reporting_and_implementation_ex2.py
```
:::
:::
:::

## The future of clinical prediction

Several trends are shaping the next decade of clinical prediction modelling:

- **Dynamic prediction models** that update as new patient data become available (e.g., during a hospital stay)
- **Federated learning** that trains models across institutions without sharing patient data
- **Foundation models** adapted for clinical tasks (large language models, multimodal models)
- **Continuous model monitoring** with automated detection of performance degradation
- **Patient-facing prediction tools** integrated into health apps and patient portals

Regardless of the technology, the fundamentals covered in this course — proper validation, calibration assessment, clinical utility evaluation, and transparent reporting — will remain essential.

::: {.callout-tip title="Exercise 3: Critical appraisal"}
Find a prediction model paper from 2024 or 2025 in a journal relevant to your field. Answer the following:

1. What clinical question does the model address?
2. Was the model developed with regression, ML, or both?
3. Was calibration reported? If so, was it assessed using a calibration plot?
4. Was clinical utility (decision curve analysis) reported?
5. Could you implement this model in your own setting with the information provided?
6. Does the paper comply with TRIPOD+AI?

Discuss your findings with a colleague or in the course discussion forum.
:::

## References and Further Reading

- For prediction model reporting and development, see @smits2026prediction, @vancalster2025performance, and @steyerberg2019.
- For decision curve analysis, see @vickers2019dca.
- For regulatory context, see @euaiact2024 (official text available at eur-lex.europa.eu).
- For communicating risk to patients, see @gigerenzer2003risks.

::: {.sectionrefs}
:::
