# Setting Up Your Computing Environment {#sec-setup}

## Step 0: Download the Course Materials

The easiest way is to download the course materials as a ZIP file:

- Go to https://github.com/mlhealthcourse/mlhealthcourse.github.io
- Click the green "Code" button, then click "Download ZIP."

If you are familiar with **git**, you can also open a terminal and run the following command to clone the course repository:

```{bash}
git clone https://github.com/mlhealthcourse/mlhealthcourse.github.io.git
```

We won't use git for the rest of the course, but if programming is going to be a part of your career, we strongly recommend
the PhD course [Git & GitHub](https://heads.ku.dk/course/git_github/) at the University of Copenhagen.

## Step 1: Setting Up Your Computing Environment

### Why Two Languages?

Throughout this course, you will have the opportunity to work with both **R** and **Python**. This is not an accident or an attempt to double your workload. In modern biostatistics, epidemiology, and health data science, both languages appear regularly in published research, collaborative projects, and industry applications. R has deep roots in classical statistics and has an extraordinary ecosystem of packages for survival analysis, Bayesian modelling, and clinical reporting. Python dominates in machine learning, deep learning, and large-scale data engineering. By learning both, you will be able to read and contribute to a wider range of projects, collaborate with more colleagues, and choose the best tool for each task.

You do not need to be an expert programmer to succeed in this course. We will introduce code gradually and explain every line. Think of R and Python as lab instruments: you will learn to use them by doing, and we will always prioritise understanding the statistical ideas over memorising syntax.

::: {.callout-tip title="Which language should I pick?"}
If you have little or no programming experience, or prefer sticking to R, use **Setup 1: R in RStudio**. RStudio is designed for data analysis and is standard in many biostatistics departments. You can always add Python development to your toolkit later on.

If you'd prefer to solely use Python, or are already familiar with it, use **Setup 2: Python in VS Code**. VS Code is a general-purpose code editor that works well for both R and Python, and it is widely used in industry.

You may combine both setups if you would like to explore both languages.

If you are curious and/or have a little experience with the terminal, **use Setup 3: R and Python in VS Code**. We will show you how to set up a single development environment that can handle both languages, but it is more advanced and requires comfort with the command line.
:::

### Step 1.1: R and RStudio

#### What Are R and RStudio?

**R** is a programming language designed for statistical computing and graphics. It runs in a terminal or console, but working with raw R in a terminal is not the most pleasant experience. **RStudio** is an integrated development environment (IDE) that wraps around R, providing a code editor, a console, a file browser, a plot viewer, and much more in a single window. Think of R as the engine and RStudio as the dashboard and steering wheel.

### Step 1.1.1: Download and Install R

1. Open your web browser and navigate to the Comprehensive R Archive Network (CRAN): <https://cran.r-project.org/>
2. You will see links for your operating system near the top of the page:
   - **Windows**: Click "Download R for Windows," then click "base," then click the link that says something like "Download R-4.5.3 for Windows." Run the downloaded `.exe` installer and accept all default settings.
   - **macOS**: Click "Download R for macOS." Choose the appropriate version for your Mac. If you have an Apple Silicon Mac (M1, M2, M3, or M4 chip), download the `arm64` version. If you have an older Intel Mac, download the `x86_64` version. If you are unsure, click the Apple icon in the top-left corner of your screen, choose "About This Mac," and look at the "Chip" or "Processor" field. Open the downloaded `.pkg` file and follow the installation prompts.
   - **Linux**: Click "Download R for Linux," select your distribution (Ubuntu, Fedora, etc.), and follow the instructions. On Ubuntu, you can also install R from the terminal:

```{bash}
#| eval: false
# Ubuntu/Debian
sudo apt update
sudo apt install r-base r-base-dev
```

3. Verify the installation by opening a terminal (or Command Prompt on Windows) and typing:

```{bash}
#| eval: false
R --version
```

You should see output that includes the R version number (4.5.3 or later is recommended).

### Step 1.1.2: Download and Install RStudio

1. Navigate to the Posit (formerly RStudio) download page: <https://posit.co/download/rstudio-desktop/>
2. The page will detect your operating system and suggest the correct installer. Click the download button.
3. Run the installer:
   - **Windows**: Run the `.exe` file and follow the prompts.
   - **macOS**: Open the `.dmg` file and drag RStudio to your Applications folder.
   - **Linux**: Install the `.deb` or `.rpm` package using your package manager.
4. Open RStudio. You should see four panels: the Source editor (top left), the Console (bottom left), the Environment/History pane (top right), and the Files/Plots/Packages/Help pane (bottom right). If R is installed correctly, you will see a welcome message in the Console that includes the R version number.

### Step 1.1.3: A Quick Tour of RStudio

- **Console** (bottom left): Type R commands here and press Enter to run them immediately. Good for quick exploration.
- **Source Editor** (top left): Write and save R scripts (`.R` files) or Quarto documents (`.qmd` files). You can run lines or selections by pressing Ctrl+Enter (Cmd+Enter on Mac).
- **Environment** (top right): Shows all variables and data objects currently in memory.
- **Files/Plots/Packages/Help** (bottom right): Browse files on your computer, view plots, manage installed packages, and read documentation.

### Step 1.1.4: Install key course packages for R

R packages extend the language with specialised tools. Think of them as apps you install on your phone --- the base R system is the phone itself, and packages add new capabilities.

Open RStudio and run the following command in the Console. This will take several minutes the first time because it downloads and compiles many packages:

```{r}
source("scripts/install_packages.R")
```

To verify that the installation worked, you can re-run the same command. If all packages are already installed, R will skip them and print: ```All packages are already installed```.

If you get an error like `there is no package called 'xyz'`, try reinstalling that specific package using: ```install.packages('xyz')```. If compilation errors occur (common on Linux), you may need to install system-level dependencies. RStudio will usually tell you what is missing.

#### What Each Package Does (Brief Overview)

| Package | Purpose |
|---------|---------|
| `tidyverse` | A collection of packages for data wrangling and visualisation. Includes `ggplot2` (plotting), `dplyr` (data manipulation), `tidyr` (reshaping), and more. |
| `rms` | Frank Harrell's Regression Modeling Strategies package. Provides tools for fitting and validating regression models with a focus on best practices. |
| `glmnet` | Fits penalised (regularised) generalised linear models, including lasso and ridge regression. |
| `survival` | The foundational package for survival (time-to-event) analysis in R. |
| `brms` | An interface to Stan for fitting Bayesian generalised linear mixed models using familiar R formula syntax. |
| `rstanarm` | Similar to `brms` but with pre-compiled Stan models for faster startup. Great for common Bayesian regression tasks. |
| `ranger` | A fast implementation of random forests, useful for both classification and regression. |
| `xgboost` | Extreme gradient boosting --- one of the most powerful and popular machine learning algorithms. |
| `mice` | Handles missing data through multiple imputation, a principled approach to dealing with incomplete datasets. |
| `dcurves` | Implements decision curve analysis for evaluating clinical prediction models. |
| `pROC` | Computes and displays ROC curves and calculates the area under the curve (AUC). |
| `uwot` | Implements UMAP (Uniform Manifold Approximation and Projection) for dimensionality reduction. |
| `cluster` | Provides methods for cluster analysis including k-medoids (PAM), hierarchical clustering, and more. |
| `gtsummary` | Creates publication-quality summary and regression tables. |
| `MatchIt` | Implements propensity score matching and other matching methods for causal inference. |
| `tidymodels` | A unified framework for building, tuning, and evaluating machine learning models in R. |

: Key R packages for this course {.striped}

### Step 1.2: Python and VS Code

#### What Are Python and VS Code?

**Python** is a general-purpose programming language widely used in data science and machine learning. Unlike R, Python is not designed exclusively for statistics, but its ecosystem of scientific libraries makes it extremely powerful for data analysis.

There are many tools available for coding in Python. For this course, we recommend **VS Code**, general-purpose code editor from Microsoft. It supports extensions which make development in Python (and R) easy.

VS Code also allows you to write code in *notebooks*, an interactive document format that mixes code and narrative text, directly inside VS Code which is convenient for data science purposes.

Other alternatives include Positron, JupyterLab and Spyder. If you are already comfortable with one of these, feel free to use it instead of VS Code.

### Step 1.2.1: Install Python and course dependencies using pixi (Recommended)

We recommend installing Python and course dependencies through **pixi**,
a fast and lightweight tool that relies on the `conda` package manager.
We have created a so-called `lock` file that contains all the dependencies for this course, which ensures that everyone in the course is using the same versions of Python and packages, and reduces the chances of encountering "this doesn't work on my machine" issues.

1. Install pixi by following the instructions at <https://pixi.prefix.dev/latest/installation/>.
2. Open a new terminal window (Command Prompt on Windows, Terminal on macOS/Linux) and go to the directory where you downloaded the course materials, then install the dependencies with:

```{bash}
cd /path/to/course-materials
pixi install
```

### Step 1.2.2: Install VS Code

1. Download VS Code from <https://code.visualstudio.com/>
2. Install it using the standard process for your operating system.
3. Open VS Code, click the Extensions icon in the left sidebar (it looks like four small squares), and install the following extensions:
   - **Python** (by Microsoft) --- provides Python language support, debugging, and Jupyter notebook integration.
   - **Jupyter** (by Microsoft) --- adds full Jupyter notebook support inside VS Code.
   - **Quarto** (by Quarto) --- optional, but useful if you want to edit `.qmd` files in VS Code.
   - **R** (by REditorSupport) --- optional, provides R language support in VS Code.

### Step 1.2.3: Verify Python Installation

```{bash}
# Activate the environment
pixi shell

# Run the verification script
python scripts/check_packages.py
```

## Key Python Packages

| Package | Purpose |
|---------|---------|
| `pandas` | The primary data manipulation library. DataFrames (tables) are the central data structure. |
| `numpy` | Numerical computing --- arrays, linear algebra, random number generation. |
| `scikit-learn` | The standard machine learning library. Classification, regression, clustering, preprocessing, model evaluation. |
| `statsmodels` | Classical statistical models: linear regression, logistic regression, time series, and more. Provides p-values and confidence intervals that scikit-learn does not. |
| `lifelines` | Survival analysis in Python: Kaplan-Meier curves, Cox proportional hazards models, and more. |
| `xgboost` | Gradient boosted trees (same algorithm as the R version). |
| `pymc` | Bayesian statistical modelling and probabilistic programming. |
| `bambi` | BAyesian Model-Building Interface --- a high-level interface to PyMC using R-style formulas. |
| `arviz` | Visualisation and diagnostics for Bayesian models. |
| `umap-learn` | UMAP dimensionality reduction for Python. |
| `matplotlib` | The foundational plotting library for Python. |
| `seaborn` | Statistical data visualisation built on top of matplotlib. Easier to use for common statistical plots. |
| `miceforest` | Multiple imputation using random forests in Python. |

: Key Python packages for this course {.striped}

### Verifying Python Package Installation

```{python}
#| eval: false
import pandas as pd
import numpy as np
import sklearn
import statsmodels
import matplotlib.pyplot as plt

print(f"pandas:       {pd.__version__}")
print(f"numpy:        {np.__version__}")
print(f"scikit-learn: {sklearn.__version__}")
print("All key packages loaded successfully!")
```

## Hello World: Your First Analysis

Let us make sure everything works by loading a built-in dataset and creating a simple plot in both languages. We will use the classic `iris` dataset --- measurements of petal and sepal dimensions for three species of iris flowers. While not a clinical dataset, it is available in both R and Python without any downloads, making it perfect for a quick test.

::: {.panel-tabset}

### R

```{r}
#| eval: false
# Load the tidyverse (includes ggplot2 and dplyr)
library(tidyverse)

# The iris dataset is built into R
data(iris)

# Take a quick look at the first few rows
head(iris)

# Summary statistics
summary(iris)

# Create a scatter plot of Sepal Length vs. Petal Length, colored by Species
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(
    title = "Iris Dataset: Sepal Length vs. Petal Length",
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)"
  ) +
  theme_minimal()
```

### Python

```{python}
#| eval: false
import pandas as pd
import seaborn as sns
import matplotlib.pyplot as plt
from sklearn.datasets import load_iris

# Load the iris dataset
iris_data = load_iris()
iris = pd.DataFrame(
    iris_data.data,
    columns=iris_data.feature_names
)
iris["species"] = pd.Categorical.from_codes(
    iris_data.target, iris_data.target_names
)

# Take a quick look
print(iris.head())
print(iris.describe())

# Create a scatter plot
plt.figure(figsize=(8, 5))
sns.scatterplot(
    data=iris,
    x="sepal length (cm)",
    y="petal length (cm)",
    hue="species",
    alpha=0.7,
    s=60
)
plt.title("Iris Dataset: Sepal Length vs. Petal Length")
plt.tight_layout()
plt.show()
```

:::

If you see a colourful scatter plot with three clusters of points, congratulations --- your setup is working.

## How to Use the Course Materials

### Navigating This Website

This course is delivered as a **website** --- a collection of chapters that you can read in your web browser. Each chapter builds on previous ones, so we recommend reading them in order the first time through. However, you can always jump to a specific chapter using the table of contents in the left sidebar.

### Code Blocks

Throughout this website, you will encounter code blocks like this:

```{r}
#| eval: false
# This is an R code block
x <- c(1, 2, 3, 4, 5)
mean(x)
```

Many chapters present code in **tabbed panels** labelled "R" and "Python." Click the tab for your preferred language. We encourage you to try both, but if you are short on time, pick the one you are more comfortable with and come back to the other later.

### Exercises

Each chapter ends with exercises. They follow this pattern:

1. A description of the task inside a coloured callout box.
2. **Starter code** that sets up the problem and leaves blanks for you to fill in.
3. A **Solution** hidden inside a collapsible box --- try the exercise yourself before peeking!

### Downloading Notebooks

For hands-on practice, you can download the exercises as standalone notebooks:

- **R users**: Look for `.qmd` or `.Rmd` files in the course repository that you can open in RStudio.
- **Python users**: Look for `.ipynb` (Jupyter notebook) files in the course repository that you can open in JupyterLab or VS Code.

These live in the course repository you downloaded in Step 0, at
<https://github.com/mlhealthcourse/mlhealthcourse.github.io>.

### Rendering Quarto Documents

If you want to render `.qmd` files yourself (to produce HTML or PDF output), you need to install Quarto:

1. Download Quarto from <https://quarto.org/docs/get-started/>
2. Install it following the instructions for your operating system.
3. In RStudio, you can render a `.qmd` file by clicking the "Render" button. In VS Code, use the Quarto extension's render command. From the terminal:

```{bash}
#| eval: false
quarto render my_document.qmd
```

## Troubleshooting Common Setup Issues

### R and RStudio Issues

**Problem**: RStudio cannot find R.
**Solution**: Make sure you installed R *before* installing RStudio. If you installed them in the wrong order, try reinstalling RStudio. On Windows, RStudio looks for R in standard installation locations. If you installed R to a non-standard path, go to Tools > Global Options > General and set the R version manually.

**Problem**: Package installation fails with a compilation error.
**Solution**: Some R packages need to compile C++ or Fortran code. On Windows, install [Rtools](https://cran.r-project.org/bin/windows/Rtools/). On macOS, install the Xcode Command Line Tools by running `xcode-select --install` in Terminal. On Linux, install `build-essential` and `r-base-dev`.

**Problem**: `brms` or `rstanarm` fails to install.
**Solution**: These packages depend on Stan, a probabilistic programming language that requires a C++ compiler. Follow the instructions above for installing compilation tools. On Windows, make sure Rtools is on your PATH. Installation can take 10--15 minutes --- be patient.

**Problem**: Package loads but you get warnings about versions.
**Solution**: Warnings (yellow text) are usually harmless --- they often say things like "package was built under R version X.Y.Z." Errors (red text) are the ones that prevent code from running. If you get errors, try updating the package with `install.packages("package_name")`.

**Problem**: Installing `rmarkdown` fails with `object 'attr' is not exported by 'namespace:xfun'`, or installing packages fails with `undefined symbol: SET_BODY` from `yaml.so`.
**Solution**: These errors occur when system-installed packages (e.g., `xfun` or `yaml` in `/usr/local/lib/R/site-library/`) were compiled against an older version of R and are incompatible with R 4.6+. Reinstall the offending packages from source: `install.packages(c("xfun", "yaml"), type = "source")`. If the system library is read-only, use `sudo R -e 'install.packages("yaml", lib="/usr/local/lib/R/site-library", type="source")'`. If `rmarkdown` still fails after updating `xfun`, the CRAN version may lag behind; install the development version with `remotes::install_github("rstudio/rmarkdown")`.

### Python Issues

**Problem**: `python` command not found.
**Solution**: On some systems, Python 3 is accessed via `python3` instead of `python`. Try `python3 --version`. If using pixi, make sure you are inside the project directory and run commands with `pixi run` or activate the shell with `pixi shell`.

**Problem**: `ModuleNotFoundError: No module named 'xyz'`.
**Solution**: The package is not installed in your current environment. Make sure you have activated the pixi environment with `pixi shell` or are running commands via `pixi run`.

**Problem**: `pip install pymc` fails with obscure errors.
**Solution**: PyMC can be tricky to install because it depends on compiled numerical libraries. If you are using pixi, the lock file should handle this automatically --- run `pixi install` and try again.


### Common Error Messages

**Error**: `Error: object 'x' not found`
**Solution**: You haven't run the earlier code blocks that create `x`. Go back and run all preceding code chunks in order. In RStudio, use "Run All Chunks Above" from the Run menu. In Jupyter, use "Run All Above" from the Cell menu.

**Error**: `Error in library(xxx): there is no package called 'xxx'`
**Solution**: Install it first with `install.packages("xxx")` in R, or `pip install xxx` in Python. Then try loading it again.

### General Tips

- **Keep your software updated.** At the start of each semester, update R, RStudio, Python, and your packages.
- **Use separate environments.** Conda environments (Python) and `renv` (R) prevent package version conflicts between projects.
- **Read error messages carefully.** They usually tell you exactly what went wrong, even if the language is technical. Copy the last line of an error message into a search engine --- someone else has almost certainly had the same problem.
- **Ask for help.** Post on the course discussion board with the full error message, your operating system, and what you were trying to do. Screenshots are helpful.

::: {.callout-tip title="Exercise 1: Verify Your Setup"}
Confirm that both R and Python are working by completing the following tasks:

1. In R, load the `tidyverse` package and use `ggplot2` to create a histogram of the `Sepal.Width` column from the built-in `iris` dataset.
2. In Python, use `seaborn` to create a histogram of the `sepal width (cm)` column from scikit-learn's iris dataset.

::: {.panel-tabset}
## R
```{r}
#| eval: false
# Load tidyverse
library(tidyverse)

# Create a histogram of Sepal.Width from the iris dataset
# YOUR CODE HERE
```
## Python
```{python}
#| eval: false
import pandas as pd
import seaborn as sns
import matplotlib.pyplot as plt
from sklearn.datasets import load_iris

# Load iris and create a DataFrame
iris_data = load_iris()
iris = pd.DataFrame(iris_data.data, columns=iris_data.feature_names)

# Create a histogram of sepal width
# YOUR CODE HERE
```
:::

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/setup_ex1.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/setup_ex1.py
```
:::
:::
:::

::: {.callout-tip title="Exercise 2: Explore a Clinical Dataset"}
Now let us work with something more relevant to health sciences. Both R and Python include the `mtcars` dataset (or we can simulate clinical-like data). In this exercise, create a scatter plot relating two variables and add a trend line.

::: {.panel-tabset}
## R
```{r}
#| eval: false
library(tidyverse)

# Let's simulate a small clinical dataset
set.seed(42)
n <- 200
clinical <- tibble(
  age = round(rnorm(n, mean = 55, sd = 12)),
  systolic_bp = round(100 + 0.8 * age + rnorm(n, sd = 10)),
  bmi = round(rnorm(n, mean = 27, sd = 5), 1)
)

# Create a scatter plot of age vs. systolic blood pressure with a trend line
# Hint: use geom_point() and geom_smooth(method = "lm")
# YOUR CODE HERE
```
## Python
```{python}
#| eval: false
import pandas as pd
import numpy as np
import seaborn as sns
import matplotlib.pyplot as plt

# Simulate a small clinical dataset
np.random.seed(42)
n = 200
clinical = pd.DataFrame({
    "age": np.round(np.random.normal(55, 12, n)).astype(int),
})
clinical["systolic_bp"] = np.round(100 + 0.8 * clinical["age"] + np.random.normal(0, 10, n))
clinical["bmi"] = np.round(np.random.normal(27, 5, n), 1)

# Create a scatter plot of age vs. systolic blood pressure with a trend line
# Hint: use sns.regplot() or sns.lmplot()
# YOUR CODE HERE
```
:::

::: {.callout-note collapse="true" title="Solution"}
::: {.panel-tabset}
#### R
```{r}
#| eval: false
#| file: ../solutions/R/setup_ex2.R
```
#### Python
```{python}
#| eval: false
#| file: ../solutions/python/setup_ex2.py
```
:::
:::
:::

## References and Further Reading

- For R, see @wickham2017r4ds, the definitive introduction to the tidyverse ecosystem.
- For Python:
  - @vanderplas2016python covers NumPy, pandas, matplotlib, and scikit-learn in depth.
  - @johnson2021numpy offers a concise, practical introduction to NumPy.
  - @mckinney2022python is the authoritative guide to pandas, written by the library's creator.
- For Quarto, see the official guide [@posit2024quarto].

::: {.sectionrefs}
:::
