---
title: "Exercise Solutions"
---

# Exercise Solutions {#sec-solutions .unnumbered}

This appendix collects solutions for all exercises in the course. Each solution is provided in both R and Python. Use them to check your work after attempting the exercises yourself.

::: {.callout-warning}
## Try First
These solutions are most valuable after you have made a genuine attempt at each exercise. Resist the temptation to read the solution before trying.
:::

The solution files are also available in the course repository under `solutions/R/` and `solutions/python/`.


## Chapter 01: Setup

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch00-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch00_ex1.R
```

#### Python

```{python}
#| label: sol-ch00-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch00_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch00-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch00_ex2.R
```

#### Python

```{python}
#| label: sol-ch00-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch00_ex2.py
```

:::

## Chapter 02: Probability and Distributions

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch01_ex1.R
```

#### Python

```{python}
#| label: sol-ch01-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch01_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch01_ex2.R
```

#### Python

```{python}
#| label: sol-ch01-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch01_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch01_ex3.R
```

#### Python

```{python}
#| label: sol-ch01-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch01_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch01_ex4.R
```

#### Python

```{python}
#| label: sol-ch01-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch01_ex4.py
```

:::

### Exercise 5

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex5-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch01_ex5.R
```

#### Python

```{python}
#| label: sol-ch01-ex5-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch01_ex5.py
```

:::

## Chapter 04: Continuous Variables and Splines

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch04-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch04_ex1.R
```

#### Python

```{python}
#| label: sol-ch04-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch04_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch04-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch04_ex2.R
```

#### Python

```{python}
#| label: sol-ch04-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch04_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch04-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch04_ex3.R
```

#### Python

```{python}
#| label: sol-ch04-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch04_ex3.py
```

:::

## Chapter 05: Penalised Regression

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch05-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch05_ex1.R
```

#### Python

```{python}
#| label: sol-ch05-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch05_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch05-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch05_ex2.R
```

#### Python

```{python}
#| label: sol-ch05-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch05_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch05-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch05_ex3.R
```

#### Python

```{python}
#| label: sol-ch05-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch05_ex3.py
```

:::

## Chapter 06: Survival Analysis

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06_ex1.R
```

#### Python

```{python}
#| label: sol-ch06-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06_ex2.R
```

#### Python

```{python}
#| label: sol-ch06-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06_ex3.R
```

#### Python

```{python}
#| label: sol-ch06-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06_ex3.py
```

:::

## Chapter 06b: Mixed-Effects Models

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06b-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06b_ex1.R
```

#### Python

```{python}
#| label: sol-ch06b-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06b_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06b-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06b_ex2.R
```

#### Python

```{python}
#| label: sol-ch06b-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06b_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06b-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06b_ex3.R
```

#### Python

```{python}
#| label: sol-ch06b-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06b_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06b-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06b_ex4.R
```

#### Python

```{python}
#| label: sol-ch06b-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06b_ex4.py
```

:::

## Chapter 06c: Missing Data

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06c-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06c_ex1.R
```

#### Python

```{python}
#| label: sol-ch06c-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06c_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06c-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06c_ex2.R
```

#### Python

```{python}
#| label: sol-ch06c-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06c_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06c-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06c_ex3.R
```

#### Python

```{python}
#| label: sol-ch06c-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06c_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06c-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch06c_ex4.R
```

#### Python

```{python}
#| label: sol-ch06c-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch06c_ex4.py
```

:::

## Chapter 07: ML Foundations

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch07-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch07_ex1.R
```

#### Python

```{python}
#| label: sol-ch07-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch07_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch07-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch07_ex2.R
```

#### Python

```{python}
#| label: sol-ch07-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch07_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch07-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch07_ex3.R
```

#### Python

```{python}
#| label: sol-ch07-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch07_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch07-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch07_ex4.R
```

#### Python

```{python}
#| label: sol-ch07-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch07_ex4.py
```

:::

## Chapter 08: Trees and Ensembles

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch08_ex1.R
```

#### Python

```{python}
#| label: sol-ch08-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch08_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch08_ex2.R
```

#### Python

```{python}
#| label: sol-ch08-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch08_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch08_ex3.R
```

#### Python

```{python}
#| label: sol-ch08-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch08_ex3.py
```

:::

## Chapter 08b: Introduction to Neural Networks

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08b-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch08b_ex1.R
```

#### Python

```{python}
#| label: sol-ch08b-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch08b_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08b-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch08b_ex2.R
```

#### Python

```{python}
#| label: sol-ch08b-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch08b_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08b-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch08b_ex3.R
```

#### Python

```{python}
#| label: sol-ch08b-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch08b_ex3.py
```

:::

## Chapter 09: Model Evaluation

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch09_ex1.R
```

#### Python

```{python}
#| label: sol-ch09-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch09_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch09_ex2.R
```

#### Python

```{python}
#| label: sol-ch09-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch09_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch09_ex3.R
```

#### Python

```{python}
#| label: sol-ch09-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch09_ex3.py
```

:::

## Chapter 09b: ML Explainability

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09b-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch09b_ex1.R
```

#### Python

```{python}
#| label: sol-ch09b-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch09b_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09b-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch09b_ex2.R
```

#### Python

```{python}
#| label: sol-ch09b-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch09b_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09b-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch09b_ex3.R
```

#### Python

```{python}
#| label: sol-ch09b-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch09b_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09b-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch09b_ex4.R
```

#### Python

```{python}
#| label: sol-ch09b-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch09b_ex4.py
```

:::

## Chapter 11: Performance and Validation

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch11-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch11_ex1.R
```

#### Python

```{python}
#| label: sol-ch11-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch11_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch11-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch11_ex2.R
```

#### Python

```{python}
#| label: sol-ch11-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch11_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch11-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch11_ex3.R
```

#### Python

```{python}
#| label: sol-ch11-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch11_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch11-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch11_ex4.R
```

#### Python

```{python}
#| label: sol-ch11-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch11_ex4.py
```

:::

## Chapter 13: Bayesian Inference

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch13-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch13_ex1.R
```

#### Python

```{python}
#| label: sol-ch13-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch13_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch13-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch13_ex2.R
```

#### Python

```{python}
#| label: sol-ch13-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch13_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch13-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch13_ex3.R
```

#### Python

```{python}
#| label: sol-ch13-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch13_ex3.py
```

:::

## Chapter 14: Applied Bayesian Modelling

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch14-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch14_ex1.R
```

#### Python

```{python}
#| label: sol-ch14-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch14_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch14-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch14_ex2.R
```

#### Python

```{python}
#| label: sol-ch14-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch14_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch14-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch14_ex3.R
```

#### Python

```{python}
#| label: sol-ch14-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch14_ex3.py
```

:::

## Chapter 15: Dimensionality Reduction

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch15-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch15_ex1.R
```

#### Python

```{python}
#| label: sol-ch15-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch15_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch15-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch15_ex2.R
```

#### Python

```{python}
#| label: sol-ch15-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch15_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch15-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch15_ex3.R
```

#### Python

```{python}
#| label: sol-ch15-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch15_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch15-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch15_ex4.R
```

#### Python

```{python}
#| label: sol-ch15-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch15_ex4.py
```

:::

## Chapter 16: Clustering

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch16-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch16_ex1.R
```

#### Python

```{python}
#| label: sol-ch16-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch16_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch16-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch16_ex2.R
```

#### Python

```{python}
#| label: sol-ch16-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch16_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch16-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch16_ex3.R
```

#### Python

```{python}
#| label: sol-ch16-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch16_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch16-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch16_ex4.R
```

#### Python

```{python}
#| label: sol-ch16-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch16_ex4.py
```

:::

## Chapter 17: Causal Inference and Treatment Effects

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17_ex1.R
```

#### Python

```{python}
#| label: sol-ch17-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17_ex2.R
```

#### Python

```{python}
#| label: sol-ch17-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17_ex3.R
```

#### Python

```{python}
#| label: sol-ch17-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17_ex4.R
```

#### Python

```{python}
#| label: sol-ch17-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17_ex4.py
```

:::

### Exercise 5

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex5-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17_ex5.R
```

#### Python

```{python}
#| label: sol-ch17-ex5-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17_ex5.py
```

:::

### Exercise 6

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex6-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17_ex6.R
```

#### Python

```{python}
#| label: sol-ch17-ex6-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17_ex6.py
```

:::

## Chapter 17c: Mediation Analysis

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17c-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17c_ex1.R
```

#### Python

```{python}
#| label: sol-ch17c-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17c_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17c-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17c_ex2.R
```

#### Python

```{python}
#| label: sol-ch17c-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17c_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17c-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17c_ex3.R
```

#### Python

```{python}
#| label: sol-ch17c-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17c_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17c-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch17c_ex4.R
```

#### Python

```{python}
#| label: sol-ch17c-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch17c_ex4.py
```

:::

## Chapter 18: Meta-Analysis

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch18-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch18_ex1.R
```

#### Python

```{python}
#| label: sol-ch18-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch18_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch18-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch18_ex2.R
```

#### Python

```{python}
#| label: sol-ch18-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch18_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch18-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch18_ex3.R
```

#### Python

```{python}
#| label: sol-ch18-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch18_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch18-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch18_ex4.R
```

#### Python

```{python}
#| label: sol-ch18-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch18_ex4.py
```

:::

## Chapter 19: Reporting and Implementation

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch19-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch19_ex1.R
```

#### Python

```{python}
#| label: sol-ch19-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch19_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch19-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ch19_ex2.R
```

#### Python

```{python}
#| label: sol-ch19-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ch19_ex2.py
```

:::
