---
title: "Exercise Solutions"
---

# Exercise Solutions {#sec-solutions .unnumbered}

This appendix collects solutions for all exercises in the course. Each solution is provided in both R and Python. Use them to check your work after attempting the exercises yourself.

::: {.callout-warning}
## Try First
These solutions are most valuable after you have made a genuine attempt at each exercise. Resist the temptation to read the solution before trying.
:::

The solution files are also available in the course repository under `solutions/R/` and `solutions/python/`.

Chapters 3 and 4 are not listed below: their exercises are conceptual, and the answers are written out in the chapters themselves rather than as code files.

## Chapter 1: Setting Up Your Computing Environment

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-setup-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/setup_ex1.R
```

#### Python

```{python}
#| label: sol-setup-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/setup_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-setup-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/setup_ex2.R
```

#### Python

```{python}
#| label: sol-setup-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/setup_ex2.py
```

:::

## Chapter 2: Probability, Distributions, and the Language of Uncertainty

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-probability-and-distributions-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex1.R
```

#### Python

```{python}
#| label: sol-probability-and-distributions-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-probability-and-distributions-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex2.R
```

#### Python

```{python}
#| label: sol-probability-and-distributions-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-probability-and-distributions-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex3.R
```

#### Python

```{python}
#| label: sol-probability-and-distributions-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-probability-and-distributions-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex4.R
```

#### Python

```{python}
#| label: sol-probability-and-distributions-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex4.py
```

:::

### Exercise 5

::: {.panel-tabset}

#### R

```{r}
#| label: sol-probability-and-distributions-ex5-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex5.R
```

#### Python

```{python}
#| label: sol-probability-and-distributions-ex5-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex5.py
```

:::

## Chapter 5: Modelling Non-Linear Relationships

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-continuous-variables-and-splines-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/continuous_variables_and_splines_ex1.R
```

#### Python

```{python}
#| label: sol-continuous-variables-and-splines-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/continuous_variables_and_splines_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-continuous-variables-and-splines-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/continuous_variables_and_splines_ex2.R
```

#### Python

```{python}
#| label: sol-continuous-variables-and-splines-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/continuous_variables_and_splines_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-continuous-variables-and-splines-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/continuous_variables_and_splines_ex3.R
```

#### Python

```{python}
#| label: sol-continuous-variables-and-splines-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/continuous_variables_and_splines_ex3.py
```

:::

## Chapter 6: Penalised Regression

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-penalized-regression-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/penalized_regression_ex1.R
```

#### Python

```{python}
#| label: sol-penalized-regression-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/penalized_regression_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-penalized-regression-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/penalized_regression_ex2.R
```

#### Python

```{python}
#| label: sol-penalized-regression-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/penalized_regression_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-penalized-regression-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/penalized_regression_ex3.R
```

#### Python

```{python}
#| label: sol-penalized-regression-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/penalized_regression_ex3.py
```

:::

## Chapter 7: Survival Analysis

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-survival-analysis-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/survival_analysis_ex1.R
```

#### Python

```{python}
#| label: sol-survival-analysis-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/survival_analysis_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-survival-analysis-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/survival_analysis_ex2.R
```

#### Python

```{python}
#| label: sol-survival-analysis-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/survival_analysis_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-survival-analysis-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/survival_analysis_ex3.R
```

#### Python

```{python}
#| label: sol-survival-analysis-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/survival_analysis_ex3.py
```

:::

## Chapter 8: Mixed-Effects Models for Clustered and Longitudinal Data

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-mixed-models-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mixed_models_ex1.R
```

#### Python

```{python}
#| label: sol-mixed-models-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mixed_models_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-mixed-models-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mixed_models_ex2.R
```

#### Python

```{python}
#| label: sol-mixed-models-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mixed_models_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-mixed-models-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mixed_models_ex3.R
```

#### Python

```{python}
#| label: sol-mixed-models-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mixed_models_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-mixed-models-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mixed_models_ex4.R
```

#### Python

```{python}
#| label: sol-mixed-models-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mixed_models_ex4.py
```

:::

## Chapter 9: Missing Data and Multiple Imputation

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-missing-data-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/missing_data_ex1.R
```

#### Python

```{python}
#| label: sol-missing-data-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/missing_data_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-missing-data-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/missing_data_ex2.R
```

#### Python

```{python}
#| label: sol-missing-data-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/missing_data_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-missing-data-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/missing_data_ex3.R
```

#### Python

```{python}
#| label: sol-missing-data-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/missing_data_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-missing-data-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/missing_data_ex4.R
```

#### Python

```{python}
#| label: sol-missing-data-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/missing_data_ex4.py
```

:::

## Chapter 10: Bayesian Inference

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-bayesian-inference-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/bayesian_inference_ex1.R
```

#### Python

```{python}
#| label: sol-bayesian-inference-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/bayesian_inference_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-bayesian-inference-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/bayesian_inference_ex2.R
```

#### Python

```{python}
#| label: sol-bayesian-inference-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/bayesian_inference_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-bayesian-inference-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/bayesian_inference_ex3.R
```

#### Python

```{python}
#| label: sol-bayesian-inference-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/bayesian_inference_ex3.py
```

:::

## Chapter 11: Applied Bayesian Modelling

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-applied-bayesian-modeling-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/applied_bayesian_modeling_ex1.R
```

#### Python

```{python}
#| label: sol-applied-bayesian-modeling-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/applied_bayesian_modeling_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-applied-bayesian-modeling-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/applied_bayesian_modeling_ex2.R
```

#### Python

```{python}
#| label: sol-applied-bayesian-modeling-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/applied_bayesian_modeling_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-applied-bayesian-modeling-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/applied_bayesian_modeling_ex3.R
```

#### Python

```{python}
#| label: sol-applied-bayesian-modeling-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/applied_bayesian_modeling_ex3.py
```

:::

## Chapter 12: What Is Machine Learning?

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ml-foundations-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_foundations_ex1.R
```

#### Python

```{python}
#| label: sol-ml-foundations-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_foundations_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ml-foundations-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_foundations_ex2.R
```

#### Python

```{python}
#| label: sol-ml-foundations-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_foundations_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ml-foundations-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_foundations_ex3.R
```

#### Python

```{python}
#| label: sol-ml-foundations-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_foundations_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ml-foundations-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_foundations_ex4.R
```

#### Python

```{python}
#| label: sol-ml-foundations-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_foundations_ex4.py
```

:::

## Chapter 13: Decision Trees, Random Forests, and Gradient Boosting

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-trees-and-ensembles-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/trees_and_ensembles_ex1.R
```

#### Python

```{python}
#| label: sol-trees-and-ensembles-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/trees_and_ensembles_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-trees-and-ensembles-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/trees_and_ensembles_ex2.R
```

#### Python

```{python}
#| label: sol-trees-and-ensembles-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/trees_and_ensembles_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-trees-and-ensembles-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/trees_and_ensembles_ex3.R
```

#### Python

```{python}
#| label: sol-trees-and-ensembles-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/trees_and_ensembles_ex3.py
```

:::

## Chapter 14: Introduction to Neural Networks

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-introduction-to-deep-learning-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/introduction_to_deep_learning_ex1.R
```

#### Python

```{python}
#| label: sol-introduction-to-deep-learning-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/introduction_to_deep_learning_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-introduction-to-deep-learning-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/introduction_to_deep_learning_ex2.R
```

#### Python

```{python}
#| label: sol-introduction-to-deep-learning-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/introduction_to_deep_learning_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-introduction-to-deep-learning-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/introduction_to_deep_learning_ex3.R
```

#### Python

```{python}
#| label: sol-introduction-to-deep-learning-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/introduction_to_deep_learning_ex3.py
```

:::

## Chapter 15: Evaluating Models

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-model-evaluation-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/model_evaluation_ex1.R
```

#### Python

```{python}
#| label: sol-model-evaluation-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/model_evaluation_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-model-evaluation-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/model_evaluation_ex2.R
```

#### Python

```{python}
#| label: sol-model-evaluation-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/model_evaluation_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-model-evaluation-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/model_evaluation_ex3.R
```

#### Python

```{python}
#| label: sol-model-evaluation-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/model_evaluation_ex3.py
```

:::

## Chapter 16: Explainable AI & Interpretability for Tabular Clinical Data

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ml-explainability-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_explainability_ex1.R
```

#### Python

```{python}
#| label: sol-ml-explainability-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_explainability_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ml-explainability-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_explainability_ex2.R
```

#### Python

```{python}
#| label: sol-ml-explainability-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_explainability_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ml-explainability-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_explainability_ex3.R
```

#### Python

```{python}
#| label: sol-ml-explainability-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_explainability_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ml-explainability-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_explainability_ex4.R
```

#### Python

```{python}
#| label: sol-ml-explainability-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_explainability_ex4.py
```

:::

## Chapter 17: Developing Clinical Prediction Models

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-building-prediction-models-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/building_prediction_models_ex1.R
```

#### Python

```{python}
#| label: sol-building-prediction-models-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/building_prediction_models_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-building-prediction-models-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/building_prediction_models_ex2.R
```

#### Python

```{python}
#| label: sol-building-prediction-models-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/building_prediction_models_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-building-prediction-models-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/building_prediction_models_ex3.R
```

#### Python

```{python}
#| label: sol-building-prediction-models-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/building_prediction_models_ex3.py
```

:::

## Chapter 18: Performance Assessment and Model Validation

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-performance-and-validation-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/performance_and_validation_ex1.R
```

#### Python

```{python}
#| label: sol-performance-and-validation-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/performance_and_validation_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-performance-and-validation-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/performance_and_validation_ex2.R
```

#### Python

```{python}
#| label: sol-performance-and-validation-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/performance_and_validation_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-performance-and-validation-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/performance_and_validation_ex3.R
```

#### Python

```{python}
#| label: sol-performance-and-validation-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/performance_and_validation_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-performance-and-validation-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/performance_and_validation_ex4.R
```

#### Python

```{python}
#| label: sol-performance-and-validation-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/performance_and_validation_ex4.py
```

:::

## Chapter 19: Reporting Standards, TRIPOD+AI, and Clinical Impact

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-reporting-and-implementation-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/reporting_and_implementation_ex1.R
```

#### Python

```{python}
#| label: sol-reporting-and-implementation-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/reporting_and_implementation_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-reporting-and-implementation-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/reporting_and_implementation_ex2.R
```

#### Python

```{python}
#| label: sol-reporting-and-implementation-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/reporting_and_implementation_ex2.py
```

:::

## Chapter 20: Making Sense of High-Dimensional Data

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-dimensionality-reduction-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/dimensionality_reduction_ex1.R
```

#### Python

```{python}
#| label: sol-dimensionality-reduction-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/dimensionality_reduction_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-dimensionality-reduction-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/dimensionality_reduction_ex2.R
```

#### Python

```{python}
#| label: sol-dimensionality-reduction-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/dimensionality_reduction_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-dimensionality-reduction-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/dimensionality_reduction_ex3.R
```

#### Python

```{python}
#| label: sol-dimensionality-reduction-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/dimensionality_reduction_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-dimensionality-reduction-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/dimensionality_reduction_ex4.R
```

#### Python

```{python}
#| label: sol-dimensionality-reduction-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/dimensionality_reduction_ex4.py
```

:::

## Chapter 21: Clustering: Discovering Groups in Data

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-clustering-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/clustering_ex1.R
```

#### Python

```{python}
#| label: sol-clustering-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/clustering_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-clustering-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/clustering_ex2.R
```

#### Python

```{python}
#| label: sol-clustering-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/clustering_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-clustering-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/clustering_ex3.R
```

#### Python

```{python}
#| label: sol-clustering-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/clustering_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-clustering-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/clustering_ex4.R
```

#### Python

```{python}
#| label: sol-clustering-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/clustering_ex4.py
```

:::

## Chapter 22: Causal Inference and Treatment Effects

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-causal-inference-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex1.R
```

#### Python

```{python}
#| label: sol-causal-inference-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-causal-inference-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex2.R
```

#### Python

```{python}
#| label: sol-causal-inference-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-causal-inference-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex3.R
```

#### Python

```{python}
#| label: sol-causal-inference-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-causal-inference-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex4.R
```

#### Python

```{python}
#| label: sol-causal-inference-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex4.py
```

:::

### Exercise 5

::: {.panel-tabset}

#### R

```{r}
#| label: sol-causal-inference-ex5-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex5.R
```

#### Python

```{python}
#| label: sol-causal-inference-ex5-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex5.py
```

:::

### Exercise 6

::: {.panel-tabset}

#### R

```{r}
#| label: sol-causal-inference-ex6-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex6.R
```

#### Python

```{python}
#| label: sol-causal-inference-ex6-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex6.py
```

:::

## Chapter 23: Mediation Analysis

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-mediation-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mediation_ex1.R
```

#### Python

```{python}
#| label: sol-mediation-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mediation_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-mediation-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mediation_ex2.R
```

#### Python

```{python}
#| label: sol-mediation-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mediation_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-mediation-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mediation_ex3.R
```

#### Python

```{python}
#| label: sol-mediation-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mediation_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-mediation-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mediation_ex4.R
```

#### Python

```{python}
#| label: sol-mediation-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mediation_ex4.py
```

:::

## Chapter 24: Meta-Analysis for Evidence Synthesis

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-meta-analysis-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/meta_analysis_ex1.R
```

#### Python

```{python}
#| label: sol-meta-analysis-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/meta_analysis_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-meta-analysis-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/meta_analysis_ex2.R
```

#### Python

```{python}
#| label: sol-meta-analysis-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/meta_analysis_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-meta-analysis-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/meta_analysis_ex3.R
```

#### Python

```{python}
#| label: sol-meta-analysis-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/meta_analysis_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-meta-analysis-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/meta_analysis_ex4.R
```

#### Python

```{python}
#| label: sol-meta-analysis-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/meta_analysis_ex4.py
```

:::
