---
title: "Exercise Solutions"
---

# Exercise Solutions {#sec-solutions .unnumbered}

This appendix collects solutions for all exercises in the course. Each solution is provided in both R and Python. Use them to check your work after attempting the exercises yourself.

::: {.callout-warning}
## Try First
These solutions are most valuable after you have made a genuine attempt at each exercise. Resist the temptation to read the solution before trying.
:::

The solution files are also available in the course repository under `solutions/R/` and `solutions/python/`.

Three chapters are not listed below. The exercises in Chapters 3 and 4 are conceptual, and their answers are written out in the chapters themselves rather than as code files. The three exercises in Chapter 17 are extended practice tasks on the workflow that chapter builds, and are left for you to work through.

## Chapter 1: Setting Up Your Computing Environment

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch00-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/setup_ex1.R
```

#### Python

```{python}
#| label: sol-ch00-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/setup_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch00-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/setup_ex2.R
```

#### Python

```{python}
#| label: sol-ch00-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/setup_ex2.py
```

:::

## Chapter 2: Probability, Distributions, and the Language of Uncertainty

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex1.R
```

#### Python

```{python}
#| label: sol-ch01-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex2.R
```

#### Python

```{python}
#| label: sol-ch01-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex3.R
```

#### Python

```{python}
#| label: sol-ch01-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex4.R
```

#### Python

```{python}
#| label: sol-ch01-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex4.py
```

:::

### Exercise 5

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch01-ex5-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/probability_and_distributions_ex5.R
```

#### Python

```{python}
#| label: sol-ch01-ex5-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/probability_and_distributions_ex5.py
```

:::

## Chapter 5: Modelling Non-Linear Relationships

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch04-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/continuous_variables_and_splines_ex1.R
```

#### Python

```{python}
#| label: sol-ch04-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/continuous_variables_and_splines_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch04-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/continuous_variables_and_splines_ex2.R
```

#### Python

```{python}
#| label: sol-ch04-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/continuous_variables_and_splines_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch04-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/continuous_variables_and_splines_ex3.R
```

#### Python

```{python}
#| label: sol-ch04-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/continuous_variables_and_splines_ex3.py
```

:::

## Chapter 6: Penalised Regression

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch05-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/penalized_regression_ex1.R
```

#### Python

```{python}
#| label: sol-ch05-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/penalized_regression_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch05-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/penalized_regression_ex2.R
```

#### Python

```{python}
#| label: sol-ch05-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/penalized_regression_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch05-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/penalized_regression_ex3.R
```

#### Python

```{python}
#| label: sol-ch05-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/penalized_regression_ex3.py
```

:::

## Chapter 7: Survival Analysis

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/survival_analysis_ex1.R
```

#### Python

```{python}
#| label: sol-ch06-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/survival_analysis_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/survival_analysis_ex2.R
```

#### Python

```{python}
#| label: sol-ch06-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/survival_analysis_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/survival_analysis_ex3.R
```

#### Python

```{python}
#| label: sol-ch06-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/survival_analysis_ex3.py
```

:::

## Chapter 8: Mixed-Effects Models for Clustered and Longitudinal Data

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06b-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mixed_models_ex1.R
```

#### Python

```{python}
#| label: sol-ch06b-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mixed_models_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06b-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mixed_models_ex2.R
```

#### Python

```{python}
#| label: sol-ch06b-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mixed_models_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06b-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mixed_models_ex3.R
```

#### Python

```{python}
#| label: sol-ch06b-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mixed_models_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06b-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mixed_models_ex4.R
```

#### Python

```{python}
#| label: sol-ch06b-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mixed_models_ex4.py
```

:::

## Chapter 9: Missing Data and Multiple Imputation

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06c-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/missing_data_ex1.R
```

#### Python

```{python}
#| label: sol-ch06c-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/missing_data_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06c-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/missing_data_ex2.R
```

#### Python

```{python}
#| label: sol-ch06c-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/missing_data_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06c-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/missing_data_ex3.R
```

#### Python

```{python}
#| label: sol-ch06c-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/missing_data_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch06c-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/missing_data_ex4.R
```

#### Python

```{python}
#| label: sol-ch06c-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/missing_data_ex4.py
```

:::

## Chapter 10: Bayesian Inference

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch13-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/bayesian_inference_ex1.R
```

#### Python

```{python}
#| label: sol-ch13-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/bayesian_inference_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch13-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/bayesian_inference_ex2.R
```

#### Python

```{python}
#| label: sol-ch13-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/bayesian_inference_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch13-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/bayesian_inference_ex3.R
```

#### Python

```{python}
#| label: sol-ch13-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/bayesian_inference_ex3.py
```

:::

## Chapter 11: Applied Bayesian Modelling

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch14-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/applied_bayesian_modeling_ex1.R
```

#### Python

```{python}
#| label: sol-ch14-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/applied_bayesian_modeling_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch14-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/applied_bayesian_modeling_ex2.R
```

#### Python

```{python}
#| label: sol-ch14-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/applied_bayesian_modeling_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch14-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/applied_bayesian_modeling_ex3.R
```

#### Python

```{python}
#| label: sol-ch14-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/applied_bayesian_modeling_ex3.py
```

:::

## Chapter 12: What Is Machine Learning?

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch07-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_foundations_ex1.R
```

#### Python

```{python}
#| label: sol-ch07-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_foundations_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch07-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_foundations_ex2.R
```

#### Python

```{python}
#| label: sol-ch07-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_foundations_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch07-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_foundations_ex3.R
```

#### Python

```{python}
#| label: sol-ch07-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_foundations_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch07-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_foundations_ex4.R
```

#### Python

```{python}
#| label: sol-ch07-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_foundations_ex4.py
```

:::

## Chapter 13: Decision Trees, Random Forests, and Gradient Boosting

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/trees_and_ensembles_ex1.R
```

#### Python

```{python}
#| label: sol-ch08-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/trees_and_ensembles_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/trees_and_ensembles_ex2.R
```

#### Python

```{python}
#| label: sol-ch08-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/trees_and_ensembles_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/trees_and_ensembles_ex3.R
```

#### Python

```{python}
#| label: sol-ch08-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/trees_and_ensembles_ex3.py
```

:::

## Chapter 14: Introduction to Neural Networks

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08b-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/introduction_to_deep_learning_ex1.R
```

#### Python

```{python}
#| label: sol-ch08b-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/introduction_to_deep_learning_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08b-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/introduction_to_deep_learning_ex2.R
```

#### Python

```{python}
#| label: sol-ch08b-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/introduction_to_deep_learning_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch08b-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/introduction_to_deep_learning_ex3.R
```

#### Python

```{python}
#| label: sol-ch08b-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/introduction_to_deep_learning_ex3.py
```

:::

## Chapter 15: Evaluating Models

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/model_evaluation_ex1.R
```

#### Python

```{python}
#| label: sol-ch09-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/model_evaluation_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/model_evaluation_ex2.R
```

#### Python

```{python}
#| label: sol-ch09-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/model_evaluation_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/model_evaluation_ex3.R
```

#### Python

```{python}
#| label: sol-ch09-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/model_evaluation_ex3.py
```

:::

## Chapter 16: Explainable AI & Interpretability for Tabular Clinical Data

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09b-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_explainability_ex1.R
```

#### Python

```{python}
#| label: sol-ch09b-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_explainability_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09b-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_explainability_ex2.R
```

#### Python

```{python}
#| label: sol-ch09b-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_explainability_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09b-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_explainability_ex3.R
```

#### Python

```{python}
#| label: sol-ch09b-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_explainability_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch09b-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/ml_explainability_ex4.R
```

#### Python

```{python}
#| label: sol-ch09b-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/ml_explainability_ex4.py
```

:::

## Chapter 18: Performance Assessment and Model Validation

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch11-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/performance_and_validation_ex1.R
```

#### Python

```{python}
#| label: sol-ch11-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/performance_and_validation_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch11-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/performance_and_validation_ex2.R
```

#### Python

```{python}
#| label: sol-ch11-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/performance_and_validation_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch11-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/performance_and_validation_ex3.R
```

#### Python

```{python}
#| label: sol-ch11-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/performance_and_validation_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch11-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/performance_and_validation_ex4.R
```

#### Python

```{python}
#| label: sol-ch11-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/performance_and_validation_ex4.py
```

:::

## Chapter 19: Reporting Standards, TRIPOD+AI, and Clinical Impact

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch19-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/reporting_and_implementation_ex1.R
```

#### Python

```{python}
#| label: sol-ch19-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/reporting_and_implementation_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch19-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/reporting_and_implementation_ex2.R
```

#### Python

```{python}
#| label: sol-ch19-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/reporting_and_implementation_ex2.py
```

:::

## Chapter 20: Making Sense of High-Dimensional Data

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch15-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/dimensionality_reduction_ex1.R
```

#### Python

```{python}
#| label: sol-ch15-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/dimensionality_reduction_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch15-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/dimensionality_reduction_ex2.R
```

#### Python

```{python}
#| label: sol-ch15-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/dimensionality_reduction_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch15-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/dimensionality_reduction_ex3.R
```

#### Python

```{python}
#| label: sol-ch15-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/dimensionality_reduction_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch15-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/dimensionality_reduction_ex4.R
```

#### Python

```{python}
#| label: sol-ch15-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/dimensionality_reduction_ex4.py
```

:::

## Chapter 21: Clustering: Discovering Groups in Data

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch16-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/clustering_ex1.R
```

#### Python

```{python}
#| label: sol-ch16-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/clustering_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch16-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/clustering_ex2.R
```

#### Python

```{python}
#| label: sol-ch16-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/clustering_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch16-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/clustering_ex3.R
```

#### Python

```{python}
#| label: sol-ch16-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/clustering_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch16-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/clustering_ex4.R
```

#### Python

```{python}
#| label: sol-ch16-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/clustering_ex4.py
```

:::

## Chapter 22: Causal Inference and Treatment Effects

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex1.R
```

#### Python

```{python}
#| label: sol-ch17-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex2.R
```

#### Python

```{python}
#| label: sol-ch17-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex3.R
```

#### Python

```{python}
#| label: sol-ch17-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex4.R
```

#### Python

```{python}
#| label: sol-ch17-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex4.py
```

:::

### Exercise 5

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex5-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex5.R
```

#### Python

```{python}
#| label: sol-ch17-ex5-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex5.py
```

:::

### Exercise 6

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17-ex6-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/causal_inference_ex6.R
```

#### Python

```{python}
#| label: sol-ch17-ex6-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/causal_inference_ex6.py
```

:::

## Chapter 23: Mediation Analysis

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17c-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mediation_ex1.R
```

#### Python

```{python}
#| label: sol-ch17c-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mediation_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17c-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mediation_ex2.R
```

#### Python

```{python}
#| label: sol-ch17c-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mediation_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17c-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mediation_ex3.R
```

#### Python

```{python}
#| label: sol-ch17c-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mediation_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch17c-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/mediation_ex4.R
```

#### Python

```{python}
#| label: sol-ch17c-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/mediation_ex4.py
```

:::

## Chapter 24: Meta-Analysis for Evidence Synthesis

### Exercise 1

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch18-ex1-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/meta_analysis_ex1.R
```

#### Python

```{python}
#| label: sol-ch18-ex1-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/meta_analysis_ex1.py
```

:::

### Exercise 2

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch18-ex2-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/meta_analysis_ex2.R
```

#### Python

```{python}
#| label: sol-ch18-ex2-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/meta_analysis_ex2.py
```

:::

### Exercise 3

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch18-ex3-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/meta_analysis_ex3.R
```

#### Python

```{python}
#| label: sol-ch18-ex3-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/meta_analysis_ex3.py
```

:::

### Exercise 4

::: {.panel-tabset}

#### R

```{r}
#| label: sol-ch18-ex4-r
#| eval: false
#| code-fold: true
#| code-summary: "Show R solution"
#| file: ../solutions/R/meta_analysis_ex4.R
```

#### Python

```{python}
#| label: sol-ch18-ex4-py
#| eval: false
#| code-fold: true
#| code-summary: "Show Python solution"
#| file: ../solutions/python/meta_analysis_ex4.py
```

:::
